### 2022-04-07 analysis/analysis.Rmd
first result analysis

---
title: "Analysis"
author: "Martin Hinz"
date: '2022-04-04'
output:
  html_document:
    keep_md: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preface

Running through the analysis can take a long time, especially the Bayesian model. We point out the long runtimes again at the appropriate place.

## Loading the necessary libraries

For this analysis we need the following libraries:

```{r}
library(tidyverse)
library(ggdist)
library(reshape2)
library(here)
library(RcppRoll)
library(sf)
library(mapview)
library(vegan)
library(neotoma)
library(analogue)
library(rcarbon)
library(doParallel)
library(nimble)
library(coda)
library(MCMCvis)
library(fmcmc)
```

If you have not yet installed these libraries (and receive a corresponding error message), please install them!

For the aorist analysis we also use the package aoristAAR, which is not available on CRAN, but would have to be installed directly from GitHub:

```{r}
if(!require('devtools')) install.packages('devtools')
library(devtools)
install_github('ISAAKiel/aoristAAR')
require(aoristAAR)
```

## Data collection

The data used in this analysis are the following proxies obtained from the following sources:

-   An openness index generated by a PCA of five pollen profiles in the Swiss Plateau from the neotoma database [@williams2018].

-   ^14^C sum calibration, data sets from the XRONOS database as well as from the unpublished PhD thesis of Julian Laabs [\@laabs2019] and from the data collection of [\@martínez-grau2021].

-   Sites from the database of the Cantonal Archaeology <!--# which! --> as the basis of an aoristic analysis

-   The number of dated fell phases at lakeshore settlements in the Three Lakes Region from the doctoral thesis by Julian Laabs [\@laabs2019].

### Openness index

We start with the openness index, as this will set the spatial framework for our analyses. The five pollen profiles are the following:

```{r}
pollen_locations <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "pollen_locations.csv")))
pollen_locations
```

We can also map these. To do this, we first transform the coordinate system into the Swiss national coordinate system, and then display them:

```{r}
pollen_locations.sf <- st_as_sf(pollen_locations,
                                coords = c("lng", "lat"), # coordinate columns
                                crs = 4326)  # coordinate references system in epsg))
pollen_locations.sf.lv95 <- st_transform(pollen_locations.sf, crs = 2056)

mapview(pollen_locations.sf)

```

We download the data from neotoma:

```{r}
pollen_data <- get_download(pollen_locations$neotoma_id)
```

Then we load a few help scripts which can be found under "code/neotoma_helpers.R".

```{r}
source(file = normalizePath(file.path(here(), "code", "neotoma_helpers.R")))
```

Now we filter the data sets so that we only retain the pollen data that record "NISP".

```{r}
pollen_data_filtered <- filter_by_unit(filter_by_element(pollen_data, "pollen"), "NISP")
class(pollen_data_filtered) <- append(class(pollen_data_filtered), "download_list")
```

Then we convert the NISP into percentages:

```{r}
pollen_data_comp <- transform_to_percent(pollen_data_filtered)
```

Furthermore, we need a (handwritten) concordance list for the taxa names:

```{r}
common_names <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "species.csv")), row.names = 1)
common_names
```

With this concordance table we combine the species with respect to their genus:

```{r}
pollen_data_comp <- aggregate_by_joint_name(pollen_data_comp, common_names = common_names)
```

Now we can combine the individual pollen profiles into an overall table, where each row represents a sample:

```{r}
all_output <- suppressWarnings(neotoma::compile_downloads(pollen_data_comp))

```

We restrict the data set to the Holocene so that we no longer have the pre-Holocene developments in the principal component analysis:

```{r}
timeframe <- c(-7000,-1000)
all_output <- subset(all_output, 1950-age > timeframe[1] & 1950-age < timeframe[2])
```

Then we extract the actual data:

```{r}
all_data <- all_output[,11:ncol(all_output)]
all_data[is.na(all_data)] <- 0
```

We then filter the data. Only those taxa are considered which are present in at least 1/3 of all data series:

```{r}
all_data <- all_data[ , colSums(all_data) != 0]
all_data <- all_data[,colSums(all_data>0)>(nrow(all_data)/3)]
```

In addition, only those with an average frequency greater than 0.1% are included:

```{r}
all_data <- all_data[,colMeans(all_data)>0.1]
```

However, we add the *cerealia* back in, regardless of their frequency:

```{r}
if(!("Cerealia" %in% colnames(all_data)))
{
  all_data <- data.frame(all_data, Cerealia = all_output[,"Cerealia"])  
}
```

These data can now be evaluated by means of a principal component analysis:

```{r}
res.pca <- rda(all_data, scale = TRUE)
biplot(res.pca, type="text")
```

This calculation already shows the familiar structure that we have the known open land indicators at one end of the spectrum and the forest species (tree pollen) at the other. However, the result is also influenced by a gradient which has nothing to do with the effect we are interested in: The immigration of beech. To eliminate this gradient, we use a partial RDA in which only the residuals for Fagus are evaluated:

```{r}
res.pca <- rda(all_data ~ Condition(Fagus), data=all_data, scale = TRUE)
biplot(res.pca, type="text")
```

From this principal component analysis, we now extract the first dimension as the openness index, and supplement it with the datings of the individual samples. In doing so, we ensure that open land indicators end up on the positive side of the result vector:

```{r}
orientation <- sign(scores(res.pca)$species["Cerealia",1])
res.open <- data.frame(scores(res.pca, choices = 1, display = "si")*orientation)
colnames(res.open) <- "score"
res.open$age <- all_output$age
res.open$site <- all_output$site.name
ggplot(res.open) + geom_line(aes(x = 1950-age, y = score, color = site))
```

The individual profiles show a very similar development over time. We now summarise the common trend and smooth it with a 50-year moving window. For this, however, we first have to interpolate shared points in time, which we do via a linear approximation:

```{r}

sw <- 50

full_range <- full_seq(res.open$age, 1)

res.open.interpolated <- by(res.open, res.open$site, function(x) approx(x$age,x$score, full_range)$y)

res.open.interpolated <- data.frame(do.call(cbind, res.open.interpolated))

res.open.interpolated$age <- full_range

res.open.interpolated <- melt(res.open.interpolated, id.vars=c("age"))

res.open.interpolated$site <- res.open.interpolated$variable

res.open.interpolated$variable <- NULL

res.open_smoothed <- res.open.interpolated %>% 
  group_by(site) %>%
  # RcppRoll::roll_mean() is written in C++ for speed 
  mutate(moving_mean = roll_mean(value, sw, fill = NA, na.rm = TRUE))

res.open_smoothed$value<-NULL
res.open_smoothed <- na.omit(res.open_smoothed)
```

Then we pick every 50th date:

```{r}
res.open_smoothed <- res.open_smoothed[res.open_smoothed$age %in%
seq(from=1000,
    to=10000, by=50),]
ggplot(res.open_smoothed,aes(x=1950-age,y=moving_mean)) + geom_point(aes(color=site), alpha =.5)  + geom_smooth(span = 0.05, color = "black")+ scale_x_continuous(breaks=seq(timeframe[1],timeframe[2],200), limits = timeframe) + labs(y = "Openness score [PCA]",x = "Date",colour = "Sites")  + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Finally, we calculate the average of all profiles as the final supra-regional openness index.

```{r}
res.open_final <- res.open_smoothed %>%
  group_by(age) %>%
  mutate(mean=mean(moving_mean),
         sd = sd(moving_mean)) %>% 
  select(age,mean,sd) %>%
  unique()  %>% arrange(desc(age))

ggplot(res.open_final,aes(x=1950-age)) + geom_line(aes(y=mean)) + geom_segment(aes(y=mean-sd, yend =mean+sd, xend=1950-age), alpha = .25)
```

### ^14^C Summenkalibration

The data set for the 14C sum calibration is composed of the above-mentioned sources.

```{r}
c14 <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "14c_swiss_plateau.csv")))
```

We can show this on the map:

```{r}
mapview(st_as_sf(c14,
                 coords = c("lng", "lat"), 
                 crs = 4326),layer.name = "14C Data"
        )
```

Next, we calibrate the 14C data using rcarbon: <!--# uncal vs cal? -->

```{r}
c14.calibrated <- calibrate(x = c14$bp, errors = c14$std)
```

Subsequently, we perform the sum calibration, whereby we perform a binning at the level of the sites:

```{r}
sumcal <- spd(c14.calibrated, timeRange = 1950-timeframe, bins = c14$site, runm = sw)
```

Below is the visualisation of the sum calibration:

```{r}
plot(1950-sumcal$grid$calBP, sumcal$grid$PrDens,type="l", main = paste0("n=", length(c14.calibrated)))

```

Finally, we extract the values every 50 years:

```{r}
res.sumcal <- sumcal$grid[sumcal$grid$calBP %in%
seq(from=1000,
    to=10000, by=50),]
```

### Aoristic Site Count

The third proxy is the aoristic sum of the number of sites from the cantonal archaeology database. For these, we have obscured the coordinates to avoid the well-known problems that can arise from clear site coordinates.

```{r}
sites <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "site_data_fuzzy.csv")))
```

We can depict this on the map:

```{r}
mapview(st_as_sf(sites,
                 coords = c("LV03_Rechtswert_fuzzy", "LV03_Hochwert_fuzzy"),
                 crs = 21781),
        layer.name = "Sites"
)
```

The dates come in two categories, rough and very rough We use the rough datings where they are available. Where they are not, we fall back on the very rough dates. But first we load the concordance table:

```{r}

chrono_rough <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "chrono_rough.csv")), row.names = 1)
chrono_vrough <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "chrono_very_rough.csv")), row.names = 1)
```

Next, we prepare the beginning and end as a column:

```{r}
sites$begin <- NA
sites$end <- NA
```

And then we first fill these columns with the data from the very rough dating:

```{r}
for(i in 1:nrow(chrono_vrough)){
  this_phase <- chrono_vrough[i,]
  sites$begin[sites$chrono_very_rough == this_phase[,1]] <- this_phase$dat_begin
  sites$end[sites$chrono_very_rough == this_phase[,1]] <- this_phase$dat_end
}
```

Where available, we overwrite the very rough dating with the rough dating:

```{r}
for(i in 1:nrow(chrono_rough)){
  this_phase <- chrono_rough[i,]
  sites$begin[sites$chrono_rough == this_phase[,1]] <- this_phase$dat_begin
  sites$end[sites$chrono_rough == this_phase[,1]] <- this_phase$dat_end
}
```

However, this does not allow for a very precise dating:

```{r}
table(sites$begin)
```

Nevertheless, we next use the aorist sum as a mapping of the expected value for simultaneously existing settlements:

```{r}
res.aorist <- aorist(sites, from = "begin", to = "end", method = "weight")

res.aorist$age <- 1950-res.aorist$date
res.aorist <- subset(res.aorist, date >= timeframe[1] & date <= timeframe[2])
```

Again, the curve of the proxy is plotted here also:

```{r}
plot(res.aorist$date, res.aorist$sum,type="l", main = paste0("n=", nrow(sites)))
```

Once more, we extract the values every 50 years:

```{r}
res.aorist <- res.aorist[res.aorist$age %in%
seq(from=1000,
    to=10000, by=50),]
```

### Dendro Dated Settlements

This year-by-year dataset maps the number of lakeshore settlements in the Three Lakes Region. Details can be found in the publication by [@laabs2019]. To match the resolution of the other datasets, this dataset is also sampled at 50-year intervals, and smoothed accordingly beforehand using a moving average.

First we load the data:

```{r}
lss <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "dendro_years.csv")))
```

Then we complete the data set so that it covers the entire period under consideration:

```{r}
lss <- rbind(data.frame(date=-7050:min(lss$date)-1,n_settlements=0),
             lss,
             data.frame(date=max(lss$date)+1:-1,n_settlements=0))
```

Finally, we run a moving average over it with the same window as for the other data sets:

```{r}
res.lss <- lss %>% 
  mutate(moving_mean = roll_mean(n_settlements, sw, fill = NA, na.rm = TRUE))
```

Here is yet another plot of the resulting curve:

```{r}
plot(res.lss$date, res.lss$moving_mean,type="l")
```

Picking the data points every 50 years again:

```{r}
res.lss[res.lss$date %in%
seq(from=timeframe[1],
    to=timeframe[2], by=50),]
res.lss$age <- 1950-res.lss$date
```

### Combining the proxies

Thus we have all proxies available, and can display them (scaled) all together:

```{r}
ggplot() +
  geom_line(data = res.open_final, aes(x = 1950-age, y = scale(mean)), col="darkgreen", alpha = 0.75) +
  geom_line(data = res.sumcal, aes(x = 1950-calBP, y = scale(PrDens)), col="red", alpha = 0.75) +
  geom_line(data = res.aorist, aes(x = date, y = scale(sum)), col="gray", alpha = 0.75) + 
  geom_line(data = res.lss, aes(x = date, y = scale(moving_mean)), col="brown", alpha = 0.75) + scale_colour_manual(values = c("darkgreen", "red", "gray", "brown"), labels = c("Openness Indicator", "Sum Calibration", "Aoristic Sum", "Dendro Dated Settlements"))
```

We merge the dataset and restrict it to our observation window:

```{r}
all_proxies <- merge(res.sumcal, res.open_final, by.x = "calBP", by.y = "age") %>%
  merge(res.aorist, by.x = "calBP", by.y = "age") %>%
  merge(res.lss, by.x = "calBP", by.y = "age") %>%
  select(calBP, PrDens, mean, sum, moving_mean) %>%
  rename(age = calBP, sumcal = PrDens, openness = mean, aoristic_sum = sum, dendro = moving_mean) %>% filter(age < 8000 & age > 3000)

all_proxies
```

We might like to store the result for the actual analysis:

```{r}
write.csv(all_proxies, file = normalizePath(file.path(here(), "data","preprocessed_data", "all_proxies.csv"), mustWork = F))
```

## Transforming the Data

If the data has already been created in a previous run, we can also reload it directly at this point:

```{r}
all_proxies <- read.csv(file = normalizePath(file.path(here(), "data","preprocessed_data", "all_proxies.csv")), row.names = 1)
```

First we reverse the order of the data so that it is sorted from oldest to youngest:

```{r}
all_proxies <- all_proxies %>% arrange(desc(age))
```

Since the data influence the model in their respective change from the state t -\> t-1, we first normalise them by a z-transformation, and then calculate the individual difference to the predecessor. The first row is set to 0.

```{r}
all_proxies[,2:5] <- all_proxies[,2:5] %>% scale() %>% diff() %>% rbind(0,.)
```

Furthermore, it has proven practical, at least in the development phase of the model, to be able to restrict the data set in order to quickly try out responses of the model. Each data point is included in the model as a separate parameter, and it therefore does not scale linearly with the number of data points. It can also be helpful for reproducing the analysis. To do this, we define a reduction factor that determines what proportion of the data is passed to the model and reduce the data set accordingly.

```{r}
redux_factor <- 1
redux_nrow <- round(nrow(all_proxies) * redux_factor)
redux_seq <- round(seq(from=1,to=nrow(all_proxies), length.out = redux_nrow))

model_data <- all_proxies[redux_seq,]
```

Furthermore, the model contains two constraints for the maximum growth rate. These help to speed up convergence. They are given as data and must be set to 1 for the constraint to be processed.

```{r}
model_data$constraint_lambda_lower <- 1
model_data$constraint_lambda_upper <- 1
```

## The Model

The actual model consists of a nimble code block that describes the relationships and sets the priors for the parameters:

```{r}
model_code <- nimbleCode( {
  # ---- Process Model ----

  # Estimate the initial state vector of population abundances
  nEnd ~ dnorm(nSites[nYears] * MeanSiteSize / AreaSwissPlateau, sd=0.5)

  # Autoregressive process for remaining years
  for(t in 2:(nYears)) {
    # The actual number of sites per year
    nSites[t] ~ dpois(lambda[t])

    # limiting the change to a maximum value, estimated in the model
    constraint_lambda_lower[t] ~ dconstraint(
      nSites[t]/nSites[t-1] < (max_growth_rate + 1)
      )
    constraint_lambda_upper[t] ~ dconstraint(
      nSites[t-1]/nSites[t] < (max_growth_rate + 1)
      )
  }

  # ---- Observational Model ----

  # For all but the first year
  for(t in 2:(nYears)) {
    # lambda depends on the number of sites at the previous year, plus
    # changes in relation to the proxies
    log(lambda[t]) <- log(nSites[t-1]) + (
      p[1] * sumcal[t]  +
        p[2] * openness[t] +
        p[3] * aoristic_sum[t] + 
        p[4] * dendro[t]
      )
  }
  # ---- Priors ----
  # Relevance of the proxies is estimated as Dirichlet distribution
  p[1:4] ~ ddirch(alpha[1:4])

  # The parameters for the Dirichlet distribution have a weakly informative prior
  for(j in 1:4) {
    alpha[j] ~ dlnorm(mu_alpha[j],sdlog=a_alpha[j])
    a_alpha[j] ~ dexp(1)
    mu_alpha[j] ~ dlnorm(1,sdlog=0.1)
  }

  # The maximum growth rate has a prior gamma distributed between 0 and 1
  # by adding 1 in the process model, this becomes 1-2[
  max_growth_rate ~ dgamma(shape = 5, scale=0.05)

  # The mean site size
  MeanSiteSize ~ dpois(50)

  # ---- transformed data ----

  # Population density and total population as function of site number
  PopDens[1:(nYears)] <- PopTotal[1:(nYears)] / AreaSwissPlateau
  PopTotal[1:nYears] <- nSites[1:nYears] * MeanSiteSize
})
```

## Run Preparation

A few things still need to be specified for the model: constants and initial values. Let's start with the constants:

```{r}
model_constants <- list(
  # an estimation for the population density at the end of the modeled period
  nEnd = 5, 
  # the number of time slices
  nYears = redux_nrow,
  # the area of the swiss plateau for the transformation
  # of population density to Total Population
  AreaSwissPlateau = 12649
  )
```

Finally, it helps for convergence to give meaningful starting values for the chains:

```{r}

# Calculation of an exponential increase of the factor 10
r <- log(1 - (10^(1/model_constants$nYears) - 1))
model_inits <- list(
  lambda = rep(r,redux_nrow),
  PopDens = rep(model_constants$nEnd,redux_nrow),
  nSites = rep(50,redux_nrow)
)
```

Furthermore, we set the parameters for the runs until convergence:

```{r}
batches <- 100000 # run in batches of this length
thinning <- 10 # save only each ... of the chains
```

Finally, we have to define which parameters (apart from the top-level stochastic nodes of the model) are to be observed and stored. We also include the population density as a core parameter and the relative proportion of proxies p:

```{r}
model_monitors <- c("PopDens", "p")
```

## Parallelisation

The model runs significantly faster if we use multiple cores in parallel. Nimble itself does not have an option for parallelisation, so we use functions provided by the package `do_parallel` for this. The computer used for the analysis has 4 cores, you may need to adjust these values for your computer.

```{r}

# Spare 4 cores from the calculation
ncore <- detectCores() - 4
ncore <- max(ncore, 1)

# Register the cluster
cl <- makeCluster(ncore)
registerDoParallel(cl)

# Export common values for the cluster
clusterExport(cl, c("model_code", "model_inits", "model_data", "model_constants", "model_monitors", "batches", "thinning"))

# Set random seeds
for (j in seq_along(cl)) {
  set.seed(j)
}
```

## First batch

In the first run, the clusters are set up and the models are instantiated and started within the clusters.

```{r}
start_time <- Sys.time()

mcmcSamples <- clusterEvalQ(cl, {

  # Load necessary libraries
  library(nimble)
  library(coda)

  # initiate the model with the parameters and dates
  model <- nimbleModel(code=model_code,
                       data=model_data,
                       constants = model_constants,
                       inits = model_inits)

  # Compile the model
  Cmodel <- compileNimble(model)

  # Configure the model
  modelConf <- configureMCMC(model, thin = thinning)

  # Add the monitor(s)
  modelConf$addMonitors(model_monitors)

  # Build the mcmc
  modelMCMC <- buildMCMC(modelConf)

  # Compile the final model
  CmodelMCMC <- compileNimble(modelMCMC, project = model)

  # Run the model for the number of iterations specified in batches
  CmodelMCMC$run(batches, reset = FALSE)
  return(as.mcmc(as.matrix(CmodelMCMC$mvSamples)))
})

end_time <- Sys.time()
end_time - start_time
```

At the end of the 1st run, the model is instantiated in the individual cluster partitions, possibly already converged. But we now check the convergence in continued runs.

```{r}
# Initialize with non-convergence
converged <- F

# minimum psrf until the model is considered converged
min_convergence <- 1.1

start_time <- Sys.time()

# As long as not converged
while (!converged) {

  # run the model for another batch, with resetting the values (burn-in)
  mcmcSamples <- clusterEvalQ(cl, {
    CmodelMCMC$run(batches, reset = FALSE, resetMV = TRUE)
    return(as.mcmc(as.matrix(CmodelMCMC$mvSamples)))
  })

  # convert the resulting list to an mcmc.list object
  mcmcSamples <- as.mcmc.list(mcmcSamples)

  # Check the psrf values from the Gelman and Rubin's convergence diagnostic
  gelman <- gelman.diag(mcmcSamples, multivariate = F)
  psrf <- gelman$psrf[,1]

  # The model is considered converged when
  # all psrf are lower than the minimum convergence criterion
  converged <- all(psrf<min_convergence, na.rm=T)

  # optional: create an mcmc traceplot for visual inspection
  #MCMCtrace(window(mcmcSamples, start = (batches/thinning)-min((batches/thinning),5000)), params = "PopDens", pdf = F)
}

end_time <- Sys.time()
end_time - start_time
```

Once the model has converged, we can look at the result of estimating the population density based on the number of settlements. For this we extract the mean and 95% highest posterior density interval:

```{r}
result <- MCMCsummary(mcmcSamples, HPD = T, params = "PopDens")

plotmax <- max(result$`95%_HPDU`)

plot(1950-model_data$age, result$mean, type="l", ylim = c(0,plotmax))
lines(1950-model_data$age, result$`95%_HPDL`, lty=2)
lines(1950-model_data$age, result$`95%_HPDU`, lty=2)
```

This converged run is sufficient to check some initial parameter values. For example, we can see how the proportional importance of the proxies is distributed:

```{r}
MCMCchains(mcmcSamples, params = "p") %>% as.data.frame %>% rename(sumcal = 'p[1]', openness = 'p[2]', aorist = 'p[3]', dendro = 'p[4]') %>% pivot_longer(everything()) %>% ggplot(aes(x = value, y = name)) + stat_halfeye()
```

The model trusts mainly the openness index, the aoristic sum and the dendro data play a subordinate role. The sum calibration has a higher weight. However, the shape of the distributions also indicates that we may have too few effective samples to make a reliable statement about the highest posterior density intervals. Let's take a look at the effective samples:

```{r fig.width=5,fig.height=12}
MCMCsummary(mcmcSamples) %>% as.data.frame() %>% rownames_to_column() %>% ggplot() + geom_bar(aes(y=n.eff, x=rowname), stat = "identity") + coord_flip()
```

Very few parameters have already reached the effective sample size of 10000 suggested for a reliable estimate. Therefore, we extend the run until this number is reached.

## Finale Run after Convergence

First we set our new meta-parameters:

```{r}
# Starting with not enough samples
enough_samples <- F

# We want at least 10000 effective samples
min_eff_samples <- 10000

# We prolongue the batches for more time efficient sampling
sample_batches <- 1000000
```

Then we export the new batch length to the clusters:

```{r}
clusterExport(cl, c("sample_batches"))
```

In order to work memory-efficiently (the model can need a lot of memory quickly!), we only allow the current samples to be used within the cluster instances and collect them outside the clusters in a variable:

```{r}
start_time <- Sys.time()
while (!enough_samples) {

  gc() # Garbage collector for more RAM

  # Start sampling  
  mcmcSamples <- clusterEvalQ(cl, {
    CmodelMCMC$run(sample_batches, reset = FALSE, resetMV = TRUE)
    return(as.mcmc(as.matrix(CmodelMCMC$mvSamples)))
  })

  mcmcSamples <- as.mcmc.list(mcmcSamples)

  # create or append values to a collector variable
  if(!exists("collector")){
    collector <- mcmcSamples
  } else {
    collector <- append_chains(collector, mcmcSamples)
  }

  # calculate effective sample size iteratively over the chains
  # and then sum the individual values
  # more RAM efficient than the `MCMCsummary` implementation
  mcmcsummary <- lapply(collector, effectiveSize)
  ness <- Reduce(`+`, mcmcsummary)

  # we have enough samples, when all parameters have more than the minimum ESS
  enough_samples <- all(ness >  min_eff_samples)

  # just in case: store the result in a RDS file for later inspection  
  saveRDS(collector, file = "mcmc_collector.RDS")
}

end_time <- Sys.time()
end_time - start_time
```

When the model has finished running, we should stop the clusters:

```{r}
stopCluster(cl)
```

For comparison with the priors, we simulate from these, first for p:

```{r}
n_prior <- 100000
a_alpha <- mu_alpha <- alpha <- list()

for(j in 1:4) {
a_alpha[[j]] <- rexp(n_prior, 1)
mu_alpha[[j]] <- rlnorm(n_prior, 1,sdlog=0.1)
alpha[[j]] <- rlnorm(n_prior, mu_alpha[[j]],sdlog=a_alpha[[j]])
}

p_mat <- matrix(ncol = 4, nrow = n_prior)
for(i in 1:n_prior){
 p_mat[i,] <- rdirch(1, c(alpha[[1]][i], alpha[[2]][i], alpha[[3]][i], alpha[[4]][i]))
}
MCMCtrace(collector, params = "p", priors = p_mat, pdf = F, ind = T, Rhat = T)
```

And then for lambda and nSites (respectively derived for PopDens):

```{r}
nSites <- matrix(ncol = model_constants$nYears, nrow = n_prior)
MeanSiteSize <- rpois(n_prior, 50)

nEnd <- rnorm(n_prior, model_constants$nEnd, sd=0.5)

nSites[,model_constants$nYears] <- rpois(n_prior,nEnd / MeanSiteSize * model_constants$AreaSwissPlateau)

for(i in (model_constants$nYears-1):1) {
  nSites[,i] <- rpois(n_prior, nSites[,i+1])
}

PopDens <- nSites * MeanSiteSize / model_constants$AreaSwissPlateau

MCMCtrace(collector, params = "PopDens", priors = PopDens, pdf = F, ind = T, Rhat = T)
gc()
```

Finally, we can plot the result in comparison to the input data, superimposing them (scaled) on the estimation result in their original form rather than as difference data:

```{r}

# MCMCvis methods are rather memory intensive. We implement the summary from scratch
all_chains <- as.matrix(collector)

rm(collector)

gc()

params <- colnames(all_chains)

this_mcmc_summary <- data.frame()

for(this_column in colnames(all_chains)) {
   this_mcmc_summary[this_column, "mean"] <- mean(all_chains[,this_column])
   hdpi_i <- coda::HPDinterval(as.mcmc(all_chains[,this_column]))
   this_mcmc_summary[this_column, "hpdi_l"] <- hdpi_i[1]
   this_mcmc_summary[this_column, "hpdi_u"] <- hdpi_i[2]
}

gc()

PopDens_summary <- this_mcmc_summary[grep("^PopDens", params),]

all_proxies_orig <- read.csv(file = normalizePath(file.path(here(), "data","preprocessed_data", "all_proxies.csv")), row.names = 1)

ggplot(all_proxies_orig) +
  geom_line(aes(x = 1950-age, y = scale(openness), col="Openness Indicator"), alpha = 0.5) +
  geom_line(aes(x = 1950-age, y = scale(sumcal), col="Sum Calibration"), alpha = 0.5) +
  geom_line(aes(x = 1950-age, y = scale(dendro), col="Dendro Dated Settlements"), alpha = 0.5) +
  geom_line(aes(x = 1950-age, y = scale(aoristic_sum), col="Aoristic Sum"), alpha = 0.5) +
  geom_line(data = PopDens_summary, aes(x = rev(1950-all_proxies_orig$age), y=mean, col = "Estimation Result")) + 
  geom_line(data = PopDens_summary, aes(x = rev(1950-all_proxies_orig$age), y=hpdi_l, col = "Estimation Result"), lty=2) + 
  geom_line(data = PopDens_summary, aes(x = rev(1950-all_proxies_orig$age), y=hpdi_u, col = "Estimation Result"), lty=2) + theme_minimal() + 
  scale_colour_manual(values = c("Openness" = "darkgreen", "Sum Calibration" = "red", "Aoristic Sum" = "gray", "Dendro Dated Settlements" = "brown", "Estimation" = "black"), name = "") + theme(legend.position="bottom") + ylab("Scaled values for Proxies,\n P/km² for Estimation") + xlab("cal BCE")
```

And finally, the final estimate of the influence of the different proxies on the result:

```{r}
all_chains[,grep("^p\\[", params)] %>% as.data.frame %>% rename(sumcal = 'p[1]', openness = 'p[2]', aorist = 'p[3]', dendro = 'p[4]') %>% pivot_longer(everything()) %>% ggplot(aes(x = value, y = name)) + stat_halfeye()
```

### 2022-04-07 manuscript/paper.Rmd
first result analysis

---
title: "Bayesian Hierarchical Model for estimating prehistoric population development on the Swiss Plateau"
author:
  - Martin Hinz:
      email: martin.hinz@iaw.unibe.ch
      institute: [IAW, OCCR]
      correspondence: true
  - Joe Roe:
      email: joe@joeroe.io
      institute: [IAW, OCCR]
      correspondence: false
  - Julian Laabs:
      email: julian.laabs@ufg.uni-kiel.de
      institute: [SFB1266]
      correspondence: false
  - Caroline Heitz:
      email: caroline.heitz@arch.ox.ac.uk
      institute: [OX, IAW, OCCR]
      correspondence: false
  - Jan Kolář:
      email: jan.kolar@ibot.cas.cz
      institute: [IBOT, IAMFAMU]
      correspondence: false
institute:
  - IAW: Institute of Archaeological Sciences, University of Bern
  - OCCR: Oeschger Centre for Climate Change Research, University of Bern
  - OX: Associated Researcher, School of Archaeology, University of Oxford
  - IBOT: Department of Vegetation Ecology, Institute of Botany of the Czech Academy of Sciences
  - IAMFAMU: Institute of Archaeology and Museology, Faculty of Arts, Masaryk University
  - SFB1266: CRC 1266 - Scales of Transformation, University of Kiel
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::word_document2:
    fig_caption: yes
    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
    pandoc_args:
    - --lua-filter=../templates/scholarly-metadata.lua
    - --lua-filter=../templates/author-info-blocks.lua
    - --lua-filter=../templates/pagebreak.lua
  bookdown::pdf_book: default
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  In order to understand human-environment relationships in the past and the dynamics of socio-ecological systems, it is essential to have a robust estimate of population sizes. Size and density of population are crucial factors for the reconstruction of group sizes, the character and range of political institutions for social organization, the constraints and possibilities of economic practice, the number of people available for collective activities, for economic and social exchange systems, and the creation and maintenance of collective identities. And, of course, the human impact on the environment. Most estimates of population sizes currently work on limited spatial scales or use proxies that hardly allow for absolute numbers (e.g. sum calibration). The figures in the archaeological discourses on population density in general currently come from vague sources and are often decades old.
  In demography in general, the last decade has triggered an upswing in the application of Bayesian methods, so that a Bayesian demography has been announced. Bayesian demography is currently at the forefront of methodological developments in this area, but has reached such maturity that the United Nations has been using it since 2015 for population forecasts and projections. The Bayesian approach offers the possibility of combining heterogeneous data and at the same time qualifying them in terms of uncertainty and credibility. This is precisely where it becomes very interesting for archaeological data, since they are mostly inaccurate, biased, sparse, simplified and often based on simplistic assumptions.
  In this contribution, issues and possibilities of the application of Bayesian demography in the field of archaeological (prehistoric) population size and density estimation will be discussed and the results of a pilot study will be presented, which should lead to a more comprehensive project for the reconstruction of demographic developments in prehistory on a large scale.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

The study of population size and population density as important demographic factors has a long history in archaeology. Although questions of population extent and distribution have been dominant fields since the beginning of archaeological research, the lack of access to this information, data on it, or methodological tools for the use of proxies has long made the statements made on this topic for prehistoric times vague and superficial [@hassan_demographic_1981], and they were often based on uncritical transfer of ethnographic parallels.

Gordon Childe [-@childe_man_1936] was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution [@hassan_demographic_1981].

Attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack [-@hack_changing_1942], Colton [-@colton_prehistoric_1949], Frankfort [-@frankfort_town_1950] and Cook [-@cook_reconsideration_1946]. Cook and Heizer (1965, 1968) and Naroll (1962) provided rules derived from ethnographic contexts for estimating the size of prehistoric populations. Cook and Heizer (1966) made the first attempt to estimate the population growth rate during the Neolithic [@hassan_demographic_1981].

At the end of the 1960s, however, with the emergence of the processual archaeology, such investigations became increasingly popular. It was the rise of the ecological paradigm that gave even greater weight to population studies, especially those dealing with population ecology [@hassan_demographic_1981].

After research interest turned increasingly to other aspects with the emergence of post-process trends, a strong boom in the study of population sizes can be observed especially since the 2010s (or even earlier, [@shennan_population_2000]). This is probably due, on the one hand, to the new shift towards the study of human-environment relationships, for which an assessment of the size of human populations is indispensable. This is true especially for an evaluation of the human impact on the natural environment. Another reason is certainly the emergence of the 'dates as data' approach [@rick_dates_1987], which only really took off with the publications around the London UCL group. Moreover, Kintigh et al. [-@kintigh_grand_2014] listed human influence, dominance and population size and growth as one of the key elements where methodological improvements are necessary in the article on the Grand Challenges for archaeology, in the area of 'human-environment interactions'.

Müller and Diachenko [@muller_tracing_2019] has compiled a list of currently used approaches and proxies for estimating prehistoric population sizes. These can be roughly divided into ethnographic parallels, ecologic-economic deductive estimates, and the counting and spatial interpolation of archaeological features (settlements, houses, individual find types). From this short literature review three basic problems are common to all these approaches and the studies based on them:

1.  Proxy is based on a single line of evidence. Although multiproxy approaches exist, in such cases the individual proxies only serve to support each other or the proxy used in the main case. There is no combination of results to increase the accuracy of each of the inaccurate and biased proxies.
2.  The uncertainty of evidence is not adequately reflected. In most cases, individual curves are presented as estimates, but the size of the error bars for these estimates is almost never apparent.
3.  Nature of the archaeological data is taken into account only insufficiently. What is meant here is that they take little account of the inherent uncertainties of the proxies themselves, for lack of an evaluative framework, and that transfer functions from the proxy to the actual size to be observed, i.e. population size or density, are hardly ever evaluated more intensively. In the best case, a critical appraisal of the informative value is provided, but a quantification of the relationship between, for example, a doubling of a proxy value and the change in population density does not take place and cannot take place for lack of a suitable framework or external data for calibration.

\[Beispiele bringen? Bring some examples here...\]

Archaeological data used to estimate population trends have the following characteristics:

-   Limited: We have only incomplete data that can be used for these purposes, and they are usually very low in information value.
-   Unevenly distributed: Although there is a good data on settlement frequencies for some regions, and these are sometimes in very high temporal resolution, such favorable situations are very unevenly distributed over time and space.
-   Noisy: Often other influences, not related to population development, affect the characteristics of individual proxies. Preservation conditions, depositional behavior or social aspects should be mentioned here as an example.
-   Unreliable: Comprehensive information and derived data are influenced by research strategies. Systematic distortions are possible or rather the rule.
-   Highly heterogeneous: The number of artefacts at a site, the number of houses, the demographic conditions in cemeteries, the estimates of human impact based on pollen analyses: All these are data that may serve our purposes, but which are available in completely different spatio-temporal scales, granularities and information values, and simply represent technically very different data formats and scales.
-   Indirect: It is actually always proxy data that are collected as substitute values for the information that is actually needed (i.e. population density or size). The transfer functions that link the collected data with the data to be obtained are unknown and are at best established by observing the (also ethnological) presence. However, since living conditions, circumstances and economic parameters differ greatly between the present time and prehistory, these transfer functions can only be evaluated as a first approximation.
-   Contradicting: Particularly when considering several proxies, differences in transfer functions and data quality and noisiness inevitably lead to contrasts. However, this usually only leads to the rejection of a proxy for a certain time period, without mutually supporting or contradictory information leading to a quantitative evaluation of the proxy and its uncertainties.

In recent years, a new method has been established in the field of demographic estimations. Bayesian demographic forecasting has been designed to solve all these problems, which also exist for demographic estimations using recent data. In their book on the subject, Bryant and Zhang \[BrynatandZhang\] state that Bayesian data modeling can be used to deal with exactly those problems that also affect archaeological data. Especially for limited, unreliable and noisy data, Bayesian approaches are optimally suited. Various even contradictory data can be brought into a common framework and support each other. Similarly, unlikelihood and uncertainty of a model approach can be quantified. This is well-accepted in archaeological applications through the Bayesian calibration and modelling of stratigraphic conditions and radiometric data. Furthermore, spatially and temporally missing data can be taken into account: Where these data are missing, the uncertainty automatically increases, but this does not prevent general modelling and estimation. Finally, hierarchically structured model suites, in which sub-models are created for the utilisation of individual proxies, can also be used to estimate transfer functions between individual proxies and the value to be modelled, thanks to the interaction of a large number of data sources and evidence.

Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on 14C data.Most notably are the recently published papers by Crema et al. However, the approach taken by Crema et al. is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of 14C data only. This is a clear step in the direction of a model-based, and thus scientific, analysis.However, on the one hand, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool. Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series.Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented. Therefore, we would like to better exploit the capabilities of Bayesian hierarchical models through a combination of inductive data analysis and model-passed data integration of different proxies.

This contribution now represents an attempt to make these techniques usable for archaeological reconstructions. In addition to a presentation of the basics and possible procedures, we want to show in the following, in a reproducible and practical form using a case study, how Bayesian methods can also make a decisive contribution to a better assessment of population development. These assessments are crucial for the reconstruction of the human past, even in for periods for which we only have very patchy, noisy and unreliable data.

# Background

At the heart of all Bayesian statistics is the concept of updating a given prior assumption with new data and expressing this in probabilities (refs.). Our assumptions about the demographic development of the past must naturally be very conservative. In the logic of the basic Bayesian equation, these assumptions represent the prior (probability). In conjunction with the data that are included as the likelihood of the prior, a posterior (probability) results, which represents the Bayesian learning from data. It is also in the nature of the approach that in real applications there is no point prediction, but in most cases a probability distribution for the prediction. Thus we simultaneously obtained a result and an estimate of the confidence intervals, or better, the credibility interval given the data.

This Bayesian learning is iterative and sequential, so that the result of one Bayesian inference can form the prior of another, i.e. it is an additive process. Moreover, at the conceptual level, this allows different sources of information to be combined. This enables that they can be mapped to the same set and the same real-world domain, e.g. probabilities. This fact has long been exploited by archaeology in using stratigraphic information to make radiometric dating more accurate. ^14^C dates and stratigraphy are something completely different, but both can be mapped to the probability of older and younger and combined in this way. The same is also conceivable (and feasible) when it comes to the probability of population sizes or population densities or their derived dynamics (increase, decrease).

A further characteristic of Bayesian modeling is that, due to the fact that results of a Bayesian inference can be regarded as a priority for the next one, a hierarchical formulation of problem domains is possible. Parameters that are necessary for an estimation, such as the relationship of population density to the clearing signal in pollen data, need not be specified explicitly, but can be given by probability distributions and then estimated in the model itself. The more data available, the more degrees of freedom can be estimated with a reasonable loss of confidence or a reasonable width of credibility intervals. For the estimation of these parameters, in turn, submodels have to be created which describe the relationship of the data to the characteristics of the parameter. This can be carried out over several levels, depending on necessity.

This modeling technique can thus be used to combine different lines of evidence horizontally and vertically and in this way combine their results into a common conclusion and estimate, which at the same time includes an estimation of their reliability: if the data contradict each other, the overall reliability will be lower. If they support each other, the confidence interval will be smaller. And if there is no systematic bias that affects all data sources to the same extent, it should be possible to arrive at the most reliable estimate possible through the most heterogeneous set of data sources.

A special class of Bayesian hierarchical models are so-called State Space Models (also known as Hidden Markov Models). These are specifically designed for time series and follow two basic principles. First, a hidden or latent process is assumed, which represents the state of the variable of interest $x_t$ at all times. For the course of the states of this variable over time, it is assumed that every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$. At the same time, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$. This means that there is a relationship between the variable $x$ and the state of variable y over time. This implies that a relationship between the individual states of variable $y$ is generated over time via the hidden variable $x$, which itself is not observable.

This basic structure of the model makes it particularly useful and suitable for the purpose of demographic reconstruction using archaeological, but also other data, which depend on population density in the past. This population density itself is not accessible or measurable by our means. All we have at our disposal are observations derived by unknown transfer functions. These can be of very different natures, such as number of archaeologically observable settlements, or other effects that can be observed through time series, and which are influenced by population in a given area. In our example, these are the openness indicators from pollen data, which we can interpret primarily in terms of human influence and its intensity. On a more abstract level, we can also include expert estimates, as these also happen on (often unspecified) bases that are at least indirectly influenced by past population density. The specifics of the individual data we used in this modelling are discussed in the section \@ref(dataproxies).

# Methods

## Process Model

The overall model for the estimation of demographic developments is broken down into several hierarchically interconnected individual elements in accordance with the basic structure of a state space model. The basis is a process model that represents the demographic development itself in terms of a structure model, without this model already being explicitly configured with data.

In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods. In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time. This results in the following very simple formula:

$$
N_t = N_{t-1} * \lambda_t
$$

A distribution that is particularly suitable for modelling frequencies is the Poisson distribution. It is a univariate discrete probability distribution that can be used to model the number of events that occur independently of each other at a constant mean rate in a fixed time interval or spatial area. It is determined by a real parameter $\lambda$ >0, which describes the expected value and simultaneously the variance of the distribution. Thus, the relationship shown above can also be rearranged as follows:

$$
N_t \sim dpois(\lambda_t) \\
\lambda_t = N_t
$$
If we now have information about the change in population development, we can use this to enter it into the model via a change in $\lambda$. This is done in the form of a regression: If $x \in R^n$ is a vector of independent variables, then the model takes the form

$$
log({E} (Y\mid x))=\alpha + \beta' x
$$
In the case of population trends, the expected value would be equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables:

$$
log(\lambda_t) = log(N_t) + \sum_{i=1}^n \beta_i x_{t,i}
$$

The values for population size $N_t$ as well as for population change $\lambda_t$ are time-dependent. At each individual point in time in the time series, these variables can or will also take on different values. However, we can narrow down the structure of population change even further. We can assume that, considered overall over time, the population change the population change will not exceed certain limits over the entire time span, without it being possible here to specify this already.

Thus, we can define the limits, the $max\_change\_rate$ as an time independent variable, again without already specifying them with fixed values at this stage. The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively. Overall, this represents a hierarchical model that can be noted as follows:

$$
max\_growth\_rate \sim dgamma(shape = 5, scale=0.05) \\
N_t/N_{t-1} < (max\_growth\_rate + 1) \\
N_{t-1}/N_t < (max\_growth\_rate + 1)
$$

The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$ (See also Figure X).

```{r}
plot(density(rgamma(100000,shape=5,scale = 0.05)+1), xlim=c(0,5), main ="gamma(shape = 5, scale=0.05) + 1" )
```

This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model. The interaction of these parameters results in the following prior probability distribution for $\lambda$ and thus the growth (or change) rate of the population:

\[Prior Probability Simulation einbauen\]

## Data/Proxies

In principle, a large number of different data sources can be integrated into the overall model as observations, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created. A non-exhaustive list can be found in the following table:

| Proxies                                        |
|------------------------------------------------|
| **Expert estimates**                           |
| Ethnographic Analogies                         |
| Carrying Capacity                              |
| Economic modelling                             |
| Extrapolation of buried individuals            |
| Burial anthropology                            |
| Settlement data, number of houses              |
| Settlement data, settlement size               |
| **Aoristic analysis**                          |
| **Dendro dates**                               |
| Amount of archaeological objects               |
| **Radiocarbon sum calibration**                |
| Estimates based on specific object types       |
| **Human impact from pollen or colluvial data** |
| aDNA based estimates                           |
| ...                                            |

: (#tab:table_proxies) A incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.

Our working region consists of the Swiss Plateau, which on the one hand offers excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking. If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also causes a research problem with regard to the ^14^C data often used as a proxy. Since many periods of the Swiss prehistoric dry land settlements are unknown, and archaeology is based primarily on the wet land settlements, the dendro data are used primarily here, since they are much more precise. As a result, ^14^C data could be largely absent in this time period. If the amount of ^14^C data were now evaluated directly and naively over time, it would show that in the phases of particularly active settlement of the lakeshore, the ^14^C sum calibration shows minima.

```{r}
library(ggplot2)
joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))

ggplot(joined_data) + geom_line(aes(x=1950-age, y=scale(sumcal, center = F), color = "red"), alpha = 0.75)  + geom_line(aes(x=1950-age, y=scale(dendro, center = F), color = "blue"), alpha = 0.75) + theme_minimal()  +
    labs(x = "years BCE",
         y = "scaled values",
         color = "Legend") + ggtitle("Number of dendro and 14C dated sites on the Swiss Plateau (scaled)") +
    scale_color_manual(values = c("blue", "red"), labels = c("dendro dated", "14C dated"))
```

In fact, there is a not uninteresting fit between the two data series. However, it must be assumed that the two dating methods, even if they would contradict each other, actually complement each other, and thus allow a better overall unified picture of the actual settlement density than each of the individual proxies would allow on their own.

We have integrated the ^14^C data as one of the proxies in the model, in the form of a sum calibration. The dataset consists of 1135 ^14^C dates from 246 sites, and dates between 10730 and 235 BP uncal. We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites. For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [!citation!] were used in the standard settings.

\[Histogramm 14C Datierungen\]

Zugleich stand uns ein Datensatz über die Anzahl der Dendrodatierten Feuchtbodensiedlungen in der Drei-Seen-Region zur Verfügung, welcher durch Julian Laabs für seine PhD-These gesammelt worden ist und deren Hintergrunddaten am genannten Ort veröffentlicht werden. Die hier verwendete Zeitreihe läuft von X bis Y, und enthält die Anzahl von zeitliche verzeichneten Schlagphasen an einzelnen Siedlungen. Somit ergibt sich hier eine Zeitreihe, die die Besiedlung der Seeufer in der Zeit des Neolithikums und der Bronzezeit (?) wiederspiegelt.

In order to add another indicator of archaeological evidence of occupation, we have included the data of the Cantonal Archaeology, and thus the Heritage Management, which are primarily derived from scattered surface finds, and which often have a low depth of information and thus dating accuracy. This information is incorporated into our model as a typologically dated aorist time series. Although the dating accuracy is very low, the advantage here lies in the fact that we are not bound to the conditions and problems of radiocarbon dates, and thus involved in the issue of sum calibration. Much more, these data provide an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by the same transmission filters and archaeological conditions as the evaluation of ^14^C data. Data from x sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy in archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.

\[Abbildung Aoristische Kurve\]

Die durch die vielen sehen Natur räumlich gegebenen Gegebenheiten ermöglichen neben einer hoch präzisen Datierung von archäologischen Fundstellen gleichzeitig aber auch ein sehr dichtes Netz an pollenanalytischen Untersuchungen. Diesen Fakt machen wir uns zu Nutze, in dem wir aus den Pollendaten einem überregionalen Offenheitsanzeiger für den Bewuchs generieren. Genaueres dazu wird im Methoden Kapitel ausgeführt. Dieser Indikator hat spezifisch den Vorteil, dass er nicht an Überlieferungsbedingungen gebunden ist, wie ist die archäologischen Indikatoren sind. Damit wird er besonders wertvoll, um systematische Verzierungen anzuzeigen oder auszugleichen, die sich durch zeitlich spezifische Siedlungsweisen und Überlieferungsbedingungen ergeben.

<!--# Ab hier muss noch umformuliert werden, ist cp aus dem Cambridge Artikel -->

Es ist eine allgemein akzeptierte Annahme, dass der menschliche Einfluss umso größer ist, je höher die Bevölkerungsdichte in einem Gebiet ist (Lechterbeck et al. 2014), und dass die Bevölkerungsdichte eines Gebiets eng mit der landwirtschaftlichen Fläche zusammenhängt (Zimmermann et al. 2004). Hinweise auf Landrodung in Pollendiagrammen können daher zusätzliche Anhaltspunkte für die Bevölkerungsdynamik liefern. Da ein einzelnes Pollenprofil eine Kombination von (außer-)lokalen und regionalen Signalen darstellt, haben wir für überregionale Analysen verschiedene Pollendiagramme von 5 kleinen Seen im Hinterland der großen Alpenseen kombiniert (Abb. 7): Le Loclat (Hadorn 1992, 1994), Moossee (Rey et al. 2019a, 2019b), Burgäschisee (Rey et al. 2019a, 2019b), Soppensee (Hajdas und Michczynski 2010; Lotter 1999), Rotsee (Lotter 1988a, 1988b, 1990). Die Daten wurden aus der Neotoma-Datenbank extrahiert (Williams et al. 2018; dort Datensatznummern 26632, 40454, 46280, 40955, 44723, 4382). Die prozentualen Pollendaten, die auf einer Pollensumme aller terrestrischen Taxa der einzelnen Standorte basieren, wurden zu einem Datensatz zusammengefasst (Abb. 8a). Nur terrestrische Pollentaxa mit einer Häufigkeit von mehr als 1/3 und, falls vorhanden, mit einer durchschnittlichen Häufigkeit von mindestens mehr als 0,1 % wurden ausgewählt, um die potenzielle Störung durch seltene Arten zu reduzieren. Getreidepollen wurden ausdrücklich als wichtiger anthropogener Indikator beibehalten.

Wir wendeten eine Hauptkomponentenanalyse (PCA) nach der von Feeser et al. (2019) vorgeschlagenen Methode an und skalierten die relativen Pollenanteile der einzelnen Taxone durch z-Scores-Standardisierung. In der Paläoökologie können multivariate Ordinationsverfahren wie die PCA verwendet werden, um zugrundeliegende Gradienten innerhalb der Daten aufzudecken (Feeser et al. 2019; Hinz et al. 2012). Aufgrund der Verbreitung der Buche (Fagus) gibt es einen starken erklärbaren Gradienten im Datensatz, der das gewünschte Ergebnis verschleiern würde. Um dies zu eliminieren, entschieden wir uns, eine Partial Constrained PCA durchzuführen. Das Ergebnis der partiellen PCA, die die Pollendaten zwischen 7000 und 1 cal BCE einschließt, zeigt, dass 20,5 % der Variabilität im Datensatz durch die erste Achse erklärt wird (die zweite Achse macht 12,6 % aus, Abb. 8b). Zu den Taxa mit hohen Werten auf der ersten Achse gehören neben Wildgrasarten auch klassische anthropogene Indikatoren (z. B. Cerealia, Urtica, Artemisa und Plantago); negative Werte werden in Übereinstimmung mit anderen Studien im Allgemeinen arborealen Taxa zugeschrieben (vgl. Behre 1981). Da dunkle mesophile Buchen-Tannenwälder als natürliche Vegetation unseres Untersuchungsgebiets angenommen werden können (vgl. Rey et al. 2017), wird eine erhöhte Offenheit der Landschaft als Ausdruck des menschlichen Einflusses auf die natürliche Vegetation interpretiert. die natürliche Vegetation widerspiegelt. Da jede Probe absolut datiert ist, können die Daten auf der x-Achse gegen den Offenheitswert auf der y-Achse aufgetragen werden, um eine Zeitreihe Zeitreihe für die Landrodung zu erhalten.

<!--# Ende cp aus dem Cambridge Artikel -->

\[Abbildung Offenheitsanzeiger\]

Leider fallen im Bereich des Schweizer Mittelland es Daten aus Bestattungen weg, die sich auch nutzbringend in ein solches Modell einbauen lassen würden. Hier sind es auch wieder schlicht Überlieferungsbedingungen, die eine Nutzung in dieser Untersuchung verwähren. Dennoch sehen wir für andere Regionen in der Integration von demographischen Indikatoren Ausgestaltungsdaten ein sehr hohes Potenzial, um den Methodenkanon und die Proxydichte zu erweitern.

Letzt endlich haben wir uns auch dazu entschlossen, Expertenschätzungen in das Modell zu integrieren. These expert estimates are certainly very valuable, as they represent a quantification of - in the strict sense - hardly quantifiable holistic assessments and the intuitive combination of different variables. However, as they are very subjectively generated, they must be considered as relatively uncertain estimates and must at least be contrasted with each other and with independent data in order to achieve a minimum degree of verifiability and intersubjectively comprehensible methodology in data evaluation. Für unser Arbeitsgebiet gibt es eine begrenzte Anzahl von Expertenschätzungen, die verschiedene Zeit Epochen umfassen, und die sich deutlich in ihren Aussagen widersprechen.

\[Tabelle Expertenschätzungen\]

## Observational Models

Für die einzelnen Datensätze existieren unabhängige Beobachtungsmodelle, die die unbeobachtbare latente Variable der Bevölkerungsdichte mit ihrem Effekt auf die beobachteten Proxies selbst bzw. spezifika darstellen. Dies geschieht einerseits in Hinsicht darauf, wie sich die hinter den Proxies liegenden Eigenschaften der Landschaft mit Bevölkerungsdichte und ihren Auswirkungen verbinden lassen, anderseits aber auch in Hinsicht darauf, wie die Beobachteten Werte selbst zu diesen Prozessen in Beziehung stehen (z.B. Überlieferungs- und Auffindungswahrscheinlichkeiten). Die implementierung der einzelnen Modelle ist teilweise sehr unterschiedlich, in Hinsicht auf die Zahl bekannter Archäologischer Fundstellen, welche hinter verschiedenen Proxies stehen, wiederum strukturell sehr ähnlich.

### Expert Estimations

Das einfachste Modell ergibt sich hinsichtlich der Expertenschätzungen. Bei diesen gehen wir davon aus, das zeitabhängig die Expertenschätzungen normalverteilt um den realen Wert der Bevölkerungsdichte schwanken. Die eigentlichen Werte der Schätzungen sind im gleichen Mass wie die zu schätzende versteckte latente Bevölkerungsdichte, die es zu schätzen gilt. Daher ist eine explizite Modellierung einer Transferfunktion nicht nötig, bzw. es ist kein eigentlicher "Messfehler" vorhanden.

Da die Schätzungen im Allgemeinen einen konstanten Wert darstellen, welcher für eine bestimmte Zeitspanne gegeben ist, und der reale Wert der Bevölkerungsdichte in jedem Zeitschritt ein anderer sein dürfte, kann hier ein partielles Pooling nicht durchgeführt werden. Vielmehr ist der Fehler der Schätzung in jedem Zeitabschnitt ein anderer, und zwischen den einzelnen Experten kann auch keine Übertragung der Information stattfinden, da davon auszugehen ist, dass jeder Experte in unterschiedlicher Weise den realen Wert über- oder unterschätzt. Daraus ergibt sich eine experten- und Zeitschrittabhängige Formulierung des Verhältnisses von der Expertenschätzung zu den im Modell zu schätzenden Bevölkerungdichten.

$$
Z_{i, t} \sim N(N_t, \sigma_{i, t}) \\
\sigma_{i, t} \sim N(N_t, 1/N_t^2)
$$

$N_t$ stellt dabei die zum Zeitpunkt t geschätze Bevölkerungsgrösse dar. Dabei ist davon auszugehen, dass die Standardabweichung der Schätzung ($\sigma_{i,t}$) abhängig ist von der absolute Grösse des Schätzwertes sowie unabhängig je Experte. Diese wird dementsprechend aufgrund des im Modell vorhandenen Schätzwertes errechnet.

### Openness Indicator

Für den Offenheitsindikator haben wir die kombinierten überregionalen Werte, die sich aus den genutzten Pollendiagrammen bzw. deren zeitspezifischen Werten auf der ersten Dimension der PCA. Für diesen Wert nehmen wir einen Messfehler an, das Verhältnis zwischen den Beobachten Werten und der tatsächlichen Landöffnung der Umwelt zeitabhängig darstellt. Der Beobachtete Wert der Landöffung $V_t$ schwankt dabei normalverteilt um den tatsächlichen Wert $I_t$, wobei wir annehmen, dass hier eine zeitunabhängige Schätzungenauigkeit $\sigma_v$ besteht, über die ein partielles Pooling stattfinden kann. Insgesamt ergibt sich für diesen ersten Teil folgende Formel:

$$
V_t \sim N(I_t, \sigma_v) \\
\sigma_v \sim \Gamma(0.5, 0.5)
$$

Somit ist $\sigma_v$ auf die gleiche Weise spezifiziert wie $\sigma_\lambda$ des Prozessmodells: eine Abweichung muss positiv sein, werte grösser als 1 sind eher unwahrscheinlich.

In weiteren Hierarchieschritten wird nun die Beziehung der Landöffnung zur Menschlichen Besiedlung modelliert. Hierbei gehen wir davon aus, dass die Landöffnung linear zur Grösse der Bevölkerung innerhalb eines Zeitschrittes skaliert, das aber zwischen einzelnen Zeitschritten (durch technische Innovationen oder Veränderungen in den naturräumlichen Gegebenheiten oder kulturellen Präferenzen) eine unterschiedliche Intensität menschlicher Aktivitäten auf die Landöffnung möglich sind ($B_t$), wobei wir insgesamt gepooled davon ausgehen, das dieser Wert normalverteilt um einen Mittelwert schankt. Hinzu kommt eine Grundsätzliche Landöffnung, die wir zeitunabhängig annehmen, und die die natürlichen Landöffnungsprozesse (Feuer durch Blitzschlag u.ä., $BaseOpen$) wiederspiegelt.

Insgesamt ergibt sich daraus folgendes Modell:

$$
I_t = N_t * B_t + BaseOpen \\
BaseOpen \sim N(0,1000) \\
B_t \sim N(\mu_b, \sigma_b) \\
\mu_b \sim N(0.5,1) \\
\sigma_b \sim \Gamma(0.5,0.5)
$$

Für $\sigma_b$ ist wiederum eine Gamma-Verteilung mit Fokus auf positive Werte kleiner als 1 angenommen, für den Mittelwert $\mu_b$ nehmen wir eine breite Normalverteilung um 0.5 an, was bedeutet, das (in Zusammenspiel mit $BaseOpen$) darauf hinausläuft, dass wir im Mittel einen positiven Einfluss von menschlicher Bevölkerung auf Landöffnung annehmen, aber viel Spielraum für das Modell zur spezifikation des Wertes lassen (weakly informative prior).

### Aoristic Analysis

# Results

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

### 2022-04-07 paper/paper.tex
first result analysis

--- 2022-04-07 manuscript/paper.Rmd
@@
----
-title: "Bayesian Hierarchical Model for estimating prehistoric population development on the Swiss Plateau"
-author:
-  - Martin Hinz:
-      email: martin.hinz@iaw.unibe.ch
-      institute: [IAW, OCCR]
-      correspondence: true
-  - Joe Roe:
-      email: joe@joeroe.io
-      institute: [IAW, OCCR]
-      correspondence: false
-  - Julian Laabs:
-      email: julian.laabs@ufg.uni-kiel.de
-      institute: [SFB1266]
-      correspondence: false
-  - Caroline Heitz:
-      email: caroline.heitz@arch.ox.ac.uk
-      institute: [OX, IAW, OCCR]
-      correspondence: false
-  - Jan Kolář:
-      email: jan.kolar@ibot.cas.cz
-      institute: [IBOT, IAMFAMU]
-      correspondence: false
-institute:
-  - IAW: Institute of Archaeological Sciences, University of Bern
-  - OCCR: Oeschger Centre for Climate Change Research, University of Bern
-  - OX: Associated Researcher, School of Archaeology, University of Oxford
-  - IBOT: Department of Vegetation Ecology, Institute of Botany of the Czech Academy of Sciences
-  - IAMFAMU: Institute of Archaeology and Museology, Faculty of Arts, Masaryk University
-  - SFB1266: CRC 1266 - Scales of Transformation, University of Kiel
-date: "`r format(Sys.time(), '%d %B, %Y')`"
-output:
-  bookdown::word_document2:
-    fig_caption: yes
-    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-    pandoc_args:
-    - --lua-filter=../templates/scholarly-metadata.lua
-    - --lua-filter=../templates/author-info-blocks.lua
-    - --lua-filter=../templates/pagebreak.lua
-  bookdown::pdf_book: default
-bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
-abstract: |
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
+\usepackage{amsmath,amssymb}
+\usepackage{lmodern}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  pdftitle={Bayesian Hierarchical Model for estimating prehistoric population development on the Swiss Plateau},
+  pdfkeywords={keyword 1; keyword 2; keyword 3},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{5}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+
+\title{Bayesian Hierarchical Model for estimating prehistoric population development on the Swiss Plateau}
+\author{true \and true \and true \and true \and true}
+\date{07 April, 2022}
+
+\begin{document}
+\maketitle
+\begin{abstract}
 In order to understand human-environment relationships in the past and the dynamics of socio-ecological systems, it is essential to have a robust estimate of population sizes.
@@
 And, of course, the human impact on the environment.
-Most estimates of population sizes currently work on limited spatial scales or use proxies that hardly allow for absolute numbers (e.g.
-sum calibration).
+Most estimates of population sizes currently work on limited spatial scales or use proxies that hardly allow for absolute numbers (e.g.~sum calibration).
 The figures in the archaeological discourses on population density in general currently come from vague sources and are often decades old.
@@
 In this contribution, issues and possibilities of the application of Bayesian demography in the field of archaeological (prehistoric) population size and density estimation will be discussed and the results of a pilot study will be presented, which should lead to a more comprehensive project for the reconstruction of demographic developments in prehistory on a large scale.
-keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
----
+\end{abstract}
 
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
+{
+\setcounter{tocdepth}{2}
+\tableofcontents
+}
+Keywords: keyword 1; keyword 2; keyword 3
 
-<!-- With the following code you can access and display values from the yml header above. -->
+Highlights: These are the highlights.
 
-Keywords: `r rmarkdown::metadata$keywords`
-
-Highlights: `r rmarkdown::metadata$highlights`
-
-<!-- The following code chunk defines some general settings how code chunks should behave. -->
-
-```{r setup, echo = FALSE}
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  warning = FALSE,
-  message = FALSE,
-  echo = FALSE,
-  comment = "#>",
-  fig.path = "../figures/",
-  dpi = 300
-)
-```
-
-<!-- The actual document text starts here: -->
-
-# Introduction
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
 The study of population size and population density as important demographic factors has a long history in archaeology.
-Although questions of population extent and distribution have been dominant fields since the beginning of archaeological research, the lack of access to this information, data on it, or methodological tools for the use of proxies has long made the statements made on this topic for prehistoric times vague and superficial [@hassan_demographic_1981], and they were often based on uncritical transfer of ethnographic parallels.
+Although questions of population extent and distribution have been dominant fields since the beginning of archaeological research, the lack of access to this information, data on it, or methodological tools for the use of proxies has long made the statements made on this topic for prehistoric times vague and superficial (Hassan, 1981), and they were often based on uncritical transfer of ethnographic parallels.
 
-Gordon Childe [-@childe_man_1936] was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution [@hassan_demographic_1981].
+Gordon Childe (1936) was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution (Hassan, 1981).
 
-Attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack [-@hack_changing_1942], Colton [-@colton_prehistoric_1949], Frankfort [-@frankfort_town_1950] and Cook [-@cook_reconsideration_1946].
+Attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack (1942), Colton (1949), Frankfort (1950) and Cook (1946).
 Cook and Heizer (1965, 1968) and Naroll (1962) provided rules derived from ethnographic contexts for estimating the size of prehistoric populations.
-Cook and Heizer (1966) made the first attempt to estimate the population growth rate during the Neolithic [@hassan_demographic_1981].
+Cook and Heizer (1966) made the first attempt to estimate the population growth rate during the Neolithic (Hassan, 1981).
 
 At the end of the 1960s, however, with the emergence of the processual archaeology, such investigations became increasingly popular.
-It was the rise of the ecological paradigm that gave even greater weight to population studies, especially those dealing with population ecology [@hassan_demographic_1981].
+It was the rise of the ecological paradigm that gave even greater weight to population studies, especially those dealing with population ecology (Hassan, 1981).
 
-After research interest turned increasingly to other aspects with the emergence of post-process trends, a strong boom in the study of population sizes can be observed especially since the 2010s (or even earlier, [@shennan_population_2000]).
+After research interest turned increasingly to other aspects with the emergence of post-process trends, a strong boom in the study of population sizes can be observed especially since the 2010s (or even earlier, (Shennan, 2000)).
 This is probably due, on the one hand, to the new shift towards the study of human-environment relationships, for which an assessment of the size of human populations is indispensable.
 This is true especially for an evaluation of the human impact on the natural environment.
-Another reason is certainly the emergence of the 'dates as data' approach [@rick_dates_1987], which only really took off with the publications around the London UCL group.
+Another reason is certainly the emergence of the `dates as data' approach (Rick, 1987), which only really took off with the publications around the London UCL group.
 Moreover, Kintigh et al.
-[-@kintigh_grand_2014] listed human influence, dominance and population size and growth as one of the key elements where methodological improvements are necessary in the article on the Grand Challenges for archaeology, in the area of 'human-environment interactions'.
+(2014) listed human influence, dominance and population size and growth as one of the key elements where methodological improvements are necessary in the article on the Grand Challenges for archaeology, in the area of `human-environment interactions'.
 
-Müller and Diachenko [@muller_tracing_2019] has compiled a list of currently used approaches and proxies for estimating prehistoric population sizes.
+Müller and Diachenko (Müller and Diachenko, 2019) has compiled a list of currently used approaches and proxies for estimating prehistoric population sizes.
 These can be roughly divided into ethnographic parallels, ecologic-economic deductive estimates, and the counting and spatial interpolation of archaeological features (settlements, houses, individual find types).
@@
 
-1.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
   Proxy is based on a single line of evidence.
@@
 There is no combination of results to increase the accuracy of each of the inaccurate and biased proxies.
-2.  The uncertainty of evidence is not adequately reflected. In most cases, individual curves are presented as estimates, but the size of the error bars for these estimates is almost never apparent.
-3.
+\item
+  The uncertainty of evidence is not adequately reflected. In most cases, individual curves are presented as estimates, but the size of the error bars for these estimates is almost never apparent.
+\item
   Nature of the archaeological data is taken into account only insufficiently.
-What is meant here is that they take little account of the inherent uncertainties of the proxies themselves, for lack of an evaluative framework, and that transfer functions from the proxy to the actual size to be observed, i.e.
-population size or density, are hardly ever evaluated more intensively.
+What is meant here is that they take little account of the inherent uncertainties of the proxies themselves, for lack of an evaluative framework, and that transfer functions from the proxy to the actual size to be observed, i.e.~population size or density, are hardly ever evaluated more intensively.
 In the best case, a critical appraisal of the informative value is provided, but a quantification of the relationship between, for example, a doubling of a proxy value and the change in population density does not take place and cannot take place for lack of a suitable framework or external data for calibration.
+\end{enumerate}
 
@@
 
--   Limited: We have only incomplete data that can be used for these purposes, and they are usually very low in information value.
--   Unevenly distributed: Although there is a good data on settlement frequencies for some regions, and these are sometimes in very high temporal resolution, such favorable situations are very unevenly distributed over time and space.
--   Noisy: Often other influences, not related to population development, affect the characteristics of individual proxies.
+\begin{itemize}
+\tightlist
+\item
+  Limited: We have only incomplete data that can be used for these purposes, and they are usually very low in information value.
+\item
+  Unevenly distributed: Although there is a good data on settlement frequencies for some regions, and these are sometimes in very high temporal resolution, such favorable situations are very unevenly distributed over time and space.
+\item
+  Noisy: Often other influences, not related to population development, affect the characteristics of individual proxies.
 Preservation conditions, depositional behavior or social aspects should be mentioned here as an example.
--   Unreliable: Comprehensive information and derived data are influenced by research strategies. Systematic distortions are possible or rather the rule.
--   Highly heterogeneous: The number of artefacts at a site, the number of houses, the demographic conditions in cemeteries, the estimates of human impact based on pollen analyses: All these are data that may serve our purposes, but which are available in completely different spatio-temporal scales, granularities and information values, and simply represent technically very different data formats and scales.
--   Indirect: It is actually always proxy data that are collected as substitute values for the information that is actually needed (i.e.
-population density or size).
+\item
+  Unreliable: Comprehensive information and derived data are influenced by research strategies. Systematic distortions are possible or rather the rule.
+\item
+  Highly heterogeneous: The number of artefacts at a site, the number of houses, the demographic conditions in cemeteries, the estimates of human impact based on pollen analyses: All these are data that may serve our purposes, but which are available in completely different spatio-temporal scales, granularities and information values, and simply represent technically very different data formats and scales.
+\item
+  Indirect: It is actually always proxy data that are collected as substitute values for the information that is actually needed (i.e.~population density or size).
 The transfer functions that link the collected data with the data to be obtained are unknown and are at best established by observing the (also ethnological) presence.
 However, since living conditions, circumstances and economic parameters differ greatly between the present time and prehistory, these transfer functions can only be evaluated as a first approximation.
--   Contradicting: Particularly when considering several proxies, differences in transfer functions and data quality and noisiness inevitably lead to contrasts.
+\item
+  Contradicting: Particularly when considering several proxies, differences in transfer functions and data quality and noisiness inevitably lead to contrasts.
 However, this usually only leads to the rejection of a proxy for a certain time period, without mutually supporting or contradictory information leading to a quantitative evaluation of the proxy and its uncertainties.
+\end{itemize}
 
@@
 
-Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on 14C data.Most notably are the recently published papers by Crema et al.
-However, the approach taken by Crema et al.
-is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of 14C data only.
+Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on 14C data.Most notably are the recently published papers by Crema et al.~However, the approach taken by Crema et al.~is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of 14C data only.
 This is a clear step in the direction of a model-based, and thus scientific, analysis.However, on the one hand, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool.
@@
 
-# Background
+\hypertarget{background}{%
+\section{Background}\label{background}}
 
@@
 
-This Bayesian learning is iterative and sequential, so that the result of one Bayesian inference can form the prior of another, i.e.
-it is an additive process.
+This Bayesian learning is iterative and sequential, so that the result of one Bayesian inference can form the prior of another, i.e.~it is an additive process.
 Moreover, at the conceptual level, this allows different sources of information to be combined.
-This enables that they can be mapped to the same set and the same real-world domain, e.g.
-probabilities.
+This enables that they can be mapped to the same set and the same real-world domain, e.g.~probabilities.
 This fact has long been exploited by archaeology in using stratigraphic information to make radiometric dating more accurate.
-^14^C dates and stratigraphy are something completely different, but both can be mapped to the probability of older and younger and combined in this way.
+\textsuperscript{14}C dates and stratigraphy are something completely different, but both can be mapped to the probability of older and younger and combined in this way.
 The same is also conceivable (and feasible) when it comes to the probability of population sizes or population densities or their derived dynamics (increase, decrease).
@@
 These are specifically designed for time series and follow two basic principles.
-First, a hidden or latent process is assumed, which represents the state of the variable of interest $x_t$ at all times.
-For the course of the states of this variable over time, it is assumed that every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$.
-At the same time, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$.
-This means that there is a relationship between the variable $x$ and the state of variable y over time.
-This implies that a relationship between the individual states of variable $y$ is generated over time via the hidden variable $x$, which itself is not observable.
+First, a hidden or latent process is assumed, which represents the state of the variable of interest \(x_t\) at all times.
+For the course of the states of this variable over time, it is assumed that every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable \(x\) at time \(t\).
+At the same time, it is assumed that certain observations, represented in variable y, are dependent on the state of variable \(x\) at time \(t\).
+This means that there is a relationship between the variable \(x\) and the state of variable y over time.
+This implies that a relationship between the individual states of variable \(y\) is generated over time via the hidden variable \(x\), which itself is not observable.
 
@@
 On a more abstract level, we can also include expert estimates, as these also happen on (often unspecified) bases that are at least indirectly influenced by past population density.
-The specifics of the individual data we used in this modelling are discussed in the section \@ref(dataproxies).
+The specifics of the individual data we used in this modelling are discussed in the section \ref{dataproxies}.
 
-# Methods
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
-## Process Model
+\hypertarget{process-model}{%
+\subsection{Process Model}\label{process-model}}
 
@@
 In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods.
-In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.
+In principle, one can represent a population development in such a way that the population at time \(t\) results from the population at time \(t-1\) times a parameter \(\lambda\), which represents the population change at this time.
 This results in the following very simple formula:
 
-$$
+\[
 N_t = N_{t-1} * \lambda_t
-$$
+\]
 
@@
 It is a univariate discrete probability distribution that can be used to model the number of events that occur independently of each other at a constant mean rate in a fixed time interval or spatial area.
-It is determined by a real parameter $\lambda$ >0, which describes the expected value and simultaneously the variance of the distribution.
+It is determined by a real parameter \(\lambda\) \textgreater0, which describes the expected value and simultaneously the variance of the distribution.
 Thus, the relationship shown above can also be rearranged as follows:
 
-$$
+\[
 N_t \sim dpois(\lambda_t) \\
 \lambda_t = N_t
-$$
-If we now have information about the change in population development, we can use this to enter it into the model via a change in $\lambda$.
-This is done in the form of a regression: If $x \in R^n$ is a vector of independent variables, then the model takes the form
+\]
+If we now have information about the change in population development, we can use this to enter it into the model via a change in \(\lambda\).
+This is done in the form of a regression: If \(x \in R^n\) is a vector of independent variables, then the model takes the form
 
-$$
+\[
 log({E} (Y\mid x))=\alpha + \beta' x
-$$
+\]
 In the case of population trends, the expected value would be equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables:
 
-$$
+\[
 log(\lambda_t) = log(N_t) + \sum_{i=1}^n \beta_i x_{t,i}
-$$
+\]
 
-The values for population size $N_t$ as well as for population change $\lambda_t$ are time-dependent.
+The values for population size \(N_t\) as well as for population change \(\lambda_t\) are time-dependent.
 At each individual point in time in the time series, these variables can or will also take on different values.
@@
 
-Thus, we can define the limits, the $max\_change\_rate$ as an time independent variable, again without already specifying them with fixed values at this stage.
+Thus, we can define the limits, the \(max\_change\_rate\) as an time independent variable, again without already specifying them with fixed values at this stage.
 The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively.
@@
 
-$$
-max\_growth\_rate \sim dgamma(shape = 5, scale=0.05) \\
-N_t/N_{t-1} < (max\_growth\_rate + 1) \\
-N_{t-1}/N_t < (max\_growth\_rate + 1)
-$$
+\$\$
+max\_growth\_rate \sim dgamma(shape = 5, scale=0.05) \textbackslash{}
 
-The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$ (See also Figure X).
+N\_t/N\_\{t-1\} \textless{} (max\_growth\_rate + 1) \textbackslash{}
+N\_\{t-1\}/N\_t \textless{} (max\_growth\_rate + 1)
+\$\$
 
-```{r}
-plot(density(rgamma(100000,shape=5,scale = 0.05)+1), xlim=c(0,5), main ="gamma(shape = 5, scale=0.05) + 1" )
-```
+The gamma distribution used centres the probability in a range \([0,1[\), adding 1 makes this range \([1-2[\) (See also Figure X).
+
+\includegraphics{../figures/unnamed-chunk-1-1.pdf}
 
 This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model.
-The interaction of these parameters results in the following prior probability distribution for $\lambda$ and thus the growth (or change) rate of the population:
+The interaction of these parameters results in the following prior probability distribution for \(\lambda\) and thus the growth (or change) rate of the population:
 
@@
 
-## Data/Proxies
+\hypertarget{dataproxies}{%
+\subsection{Data/Proxies}\label{dataproxies}}
 
@@
 
-| Proxies                                        |
-|------------------------------------------------|
-| **Expert estimates**                           |
-| Ethnographic Analogies                         |
-| Carrying Capacity                              |
-| Economic modelling                             |
-| Extrapolation of buried individuals            |
-| Burial anthropology                            |
-| Settlement data, number of houses              |
-| Settlement data, settlement size               |
-| **Aoristic analysis**                          |
-| **Dendro dates**                               |
-| Amount of archaeological objects               |
-| **Radiocarbon sum calibration**                |
-| Estimates based on specific object types       |
-| **Human impact from pollen or colluvial data** |
-| aDNA based estimates                           |
-| ...                                            |
-
-: (#tab:table_proxies) A incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.
+\begin{longtable}[]{@{}l@{}}
+\caption{(\#tab:table\_proxies) A incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.}\tabularnewline
+\toprule
+Proxies \\
+\midrule
+\endfirsthead
+\toprule
+Proxies \\
+\midrule
+\endhead
+\textbf{Expert estimates} \\
+Ethnographic Analogies \\
+Carrying Capacity \\
+Economic modelling \\
+Extrapolation of buried individuals \\
+Burial anthropology \\
+Settlement data, number of houses \\
+Settlement data, settlement size \\
+\textbf{Aoristic analysis} \\
+\textbf{Dendro dates} \\
+Amount of archaeological objects \\
+\textbf{Radiocarbon sum calibration} \\
+Estimates based on specific object types \\
+\textbf{Human impact from pollen or colluvial data} \\
+aDNA based estimates \\
+\ldots{} \\
+\bottomrule
+\end{longtable}
 
 Our working region consists of the Swiss Plateau, which on the one hand offers excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking.
-If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also causes a research problem with regard to the ^14^C data often used as a proxy.
+If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also causes a research problem with regard to the \textsuperscript{14}C data often used as a proxy.
 Since many periods of the Swiss prehistoric dry land settlements are unknown, and archaeology is based primarily on the wet land settlements, the dendro data are used primarily here, since they are much more precise.
-As a result, ^14^C data could be largely absent in this time period.
-If the amount of ^14^C data were now evaluated directly and naively over time, it would show that in the phases of particularly active settlement of the lakeshore, the ^14^C sum calibration shows minima.
+As a result, \textsuperscript{14}C data could be largely absent in this time period.
+If the amount of \textsuperscript{14}C data were now evaluated directly and naively over time, it would show that in the phases of particularly active settlement of the lakeshore, the \textsuperscript{14}C sum calibration shows minima.
 
-```{r}
-library(ggplot2)
-joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))
-
-ggplot(joined_data) + geom_line(aes(x=1950-age, y=scale(sumcal, center = F), color = "red"), alpha = 0.75)  + geom_line(aes(x=1950-age, y=scale(dendro, center = F), color = "blue"), alpha = 0.75) + theme_minimal()  +
-    labs(x = "years BCE",
-         y = "scaled values",
-         color = "Legend") + ggtitle("Number of dendro and 14C dated sites on the Swiss Plateau (scaled)") +
-    scale_color_manual(values = c("blue", "red"), labels = c("dendro dated", "14C dated"))
-```
+\includegraphics{../figures/unnamed-chunk-2-1.pdf}
 
@@
 
-We have integrated the ^14^C data as one of the proxies in the model, in the form of a sum calibration.
-The dataset consists of 1135 ^14^C dates from 246 sites, and dates between 10730 and 235 BP uncal.
-We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites.
-For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [!citation!] were used in the standard settings.
+We have integrated the \textsuperscript{14}C data as one of the proxies in the model, in the form of a sum calibration.
+The dataset consists of 1135 \textsuperscript{14}C dates from 246 sites, and dates between 10730 and 235 BP uncal.
+We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous \textsuperscript{14}C dated sites.
+For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon {[}!citation!{]} were used in the standard settings.
 
@@
 Although the dating accuracy is very low, the advantage here lies in the fact that we are not bound to the conditions and problems of radiocarbon dates, and thus involved in the issue of sum calibration.
-Much more, these data provide an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by the same transmission filters and archaeological conditions as the evaluation of ^14^C data.
+Much more, these data provide an independent indicator with regard to the methodology of the \textsuperscript{14}C data, even if they are influenced by the same transmission filters and archaeological conditions as the evaluation of \textsuperscript{14}C data.
 Data from x sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy in archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.
@@
 
-<!--# Ab hier muss noch umformuliert werden, ist cp aus dem Cambridge Artikel -->
-
-Es ist eine allgemein akzeptierte Annahme, dass der menschliche Einfluss umso größer ist, je höher die Bevölkerungsdichte in einem Gebiet ist (Lechterbeck et al.
-2014), und dass die Bevölkerungsdichte eines Gebiets eng mit der landwirtschaftlichen Fläche zusammenhängt (Zimmermann et al.
-2004).
+Es ist eine allgemein akzeptierte Annahme, dass der menschliche Einfluss umso größer ist, je höher die Bevölkerungsdichte in einem Gebiet ist (Lechterbeck et al.~2014), und dass die Bevölkerungsdichte eines Gebiets eng mit der landwirtschaftlichen Fläche zusammenhängt (Zimmermann et al.~2004).
 Hinweise auf Landrodung in Pollendiagrammen können daher zusätzliche Anhaltspunkte für die Bevölkerungsdynamik liefern.
 Da ein einzelnes Pollenprofil eine Kombination von (außer-)lokalen und regionalen Signalen darstellt, haben wir für überregionale Analysen verschiedene Pollendiagramme von 5 kleinen Seen im Hinterland der großen Alpenseen kombiniert (Abb.
-7): Le Loclat (Hadorn 1992, 1994), Moossee (Rey et al.
-2019a, 2019b), Burgäschisee (Rey et al.
-2019a, 2019b), Soppensee (Hajdas und Michczynski 2010; Lotter 1999), Rotsee (Lotter 1988a, 1988b, 1990).
-Die Daten wurden aus der Neotoma-Datenbank extrahiert (Williams et al.
-2018; dort Datensatznummern 26632, 40454, 46280, 40955, 44723, 4382).
+7): Le Loclat (Hadorn 1992, 1994), Moossee (Rey et al.~2019a, 2019b), Burgäschisee (Rey et al.~2019a, 2019b), Soppensee (Hajdas und Michczynski 2010; Lotter 1999), Rotsee (Lotter 1988a, 1988b, 1990).
+Die Daten wurden aus der Neotoma-Datenbank extrahiert (Williams et al.~2018; dort Datensatznummern 26632, 40454, 46280, 40955, 44723, 4382).
 Die prozentualen Pollendaten, die auf einer Pollensumme aller terrestrischen Taxa der einzelnen Standorte basieren, wurden zu einem Datensatz zusammengefasst (Abb.
 8a).
-Nur terrestrische Pollentaxa mit einer Häufigkeit von mehr als 1/3 und, falls vorhanden, mit einer durchschnittlichen Häufigkeit von mindestens mehr als 0,1 % wurden ausgewählt, um die potenzielle Störung durch seltene Arten zu reduzieren.
+Nur terrestrische Pollentaxa mit einer Häufigkeit von mehr als 1/3 und, falls vorhanden, mit einer durchschnittlichen Häufigkeit von mindestens mehr als 0,1 \% wurden ausgewählt, um die potenzielle Störung durch seltene Arten zu reduzieren.
 Getreidepollen wurden ausdrücklich als wichtiger anthropogener Indikator beibehalten.
 
-Wir wendeten eine Hauptkomponentenanalyse (PCA) nach der von Feeser et al.
-(2019) vorgeschlagenen Methode an und skalierten die relativen Pollenanteile der einzelnen Taxone durch z-Scores-Standardisierung.
-In der Paläoökologie können multivariate Ordinationsverfahren wie die PCA verwendet werden, um zugrundeliegende Gradienten innerhalb der Daten aufzudecken (Feeser et al.
-2019; Hinz et al.
-2012).
+Wir wendeten eine Hauptkomponentenanalyse (PCA) nach der von Feeser et al.~(2019) vorgeschlagenen Methode an und skalierten die relativen Pollenanteile der einzelnen Taxone durch z-Scores-Standardisierung.
+In der Paläoökologie können multivariate Ordinationsverfahren wie die PCA verwendet werden, um zugrundeliegende Gradienten innerhalb der Daten aufzudecken (Feeser et al.~2019; Hinz et al.~2012).
 Aufgrund der Verbreitung der Buche (Fagus) gibt es einen starken erklärbaren Gradienten im Datensatz, der das gewünschte Ergebnis verschleiern würde.
 Um dies zu eliminieren, entschieden wir uns, eine Partial Constrained PCA durchzuführen.
-Das Ergebnis der partiellen PCA, die die Pollendaten zwischen 7000 und 1 cal BCE einschließt, zeigt, dass 20,5 % der Variabilität im Datensatz durch die erste Achse erklärt wird (die zweite Achse macht 12,6 % aus, Abb.
+Das Ergebnis der partiellen PCA, die die Pollendaten zwischen 7000 und 1 cal BCE einschließt, zeigt, dass 20,5 \% der Variabilität im Datensatz durch die erste Achse erklärt wird (die zweite Achse macht 12,6 \% aus, Abb.
 8b).
@@
 Da dunkle mesophile Buchen-Tannenwälder als natürliche Vegetation unseres Untersuchungsgebiets angenommen werden können (vgl.
-Rey et al.
-2017), wird eine erhöhte Offenheit der Landschaft als Ausdruck des menschlichen Einflusses auf die natürliche Vegetation interpretiert.
+Rey et al.~2017), wird eine erhöhte Offenheit der Landschaft als Ausdruck des menschlichen Einflusses auf die natürliche Vegetation interpretiert.
 die natürliche Vegetation widerspiegelt.
 Da jede Probe absolut datiert ist, können die Daten auf der x-Achse gegen den Offenheitswert auf der y-Achse aufgetragen werden, um eine Zeitreihe Zeitreihe für die Landrodung zu erhalten.
-
-<!--# Ende cp aus dem Cambridge Artikel -->
 
@@
 
-## Observational Models
+\hypertarget{observational-models}{%
+\subsection{Observational Models}\label{observational-models}}
 
@@
 
-### Expert Estimations
+\hypertarget{expert-estimations}{%
+\subsubsection{Expert Estimations}\label{expert-estimations}}
 
@@
 Daher ist eine explizite Modellierung einer Transferfunktion nicht nötig, bzw.
-es ist kein eigentlicher "Messfehler" vorhanden.
+es ist kein eigentlicher ``Messfehler'' vorhanden.
 
@@
 
-$$
+\[
 Z_{i, t} \sim N(N_t, \sigma_{i, t}) \\
 \sigma_{i, t} \sim N(N_t, 1/N_t^2)
-$$
+\]
 
-$N_t$ stellt dabei die zum Zeitpunkt t geschätze Bevölkerungsgrösse dar.
-Dabei ist davon auszugehen, dass die Standardabweichung der Schätzung ($\sigma_{i,t}$) abhängig ist von der absolute Grösse des Schätzwertes sowie unabhängig je Experte.
+\(N_t\) stellt dabei die zum Zeitpunkt t geschätze Bevölkerungsgrösse dar.
+Dabei ist davon auszugehen, dass die Standardabweichung der Schätzung (\(\sigma_{i,t}\)) abhängig ist von der absolute Grösse des Schätzwertes sowie unabhängig je Experte.
 Diese wird dementsprechend aufgrund des im Modell vorhandenen Schätzwertes errechnet.
 
-### Openness Indicator
+\hypertarget{openness-indicator}{%
+\subsubsection{Openness Indicator}\label{openness-indicator}}
 
@@
 Für diesen Wert nehmen wir einen Messfehler an, das Verhältnis zwischen den Beobachten Werten und der tatsächlichen Landöffnung der Umwelt zeitabhängig darstellt.
-Der Beobachtete Wert der Landöffung $V_t$ schwankt dabei normalverteilt um den tatsächlichen Wert $I_t$, wobei wir annehmen, dass hier eine zeitunabhängige Schätzungenauigkeit $\sigma_v$ besteht, über die ein partielles Pooling stattfinden kann.
+Der Beobachtete Wert der Landöffung \(V_t\) schwankt dabei normalverteilt um den tatsächlichen Wert \(I_t\), wobei wir annehmen, dass hier eine zeitunabhängige Schätzungenauigkeit \(\sigma_v\) besteht, über die ein partielles Pooling stattfinden kann.
 Insgesamt ergibt sich für diesen ersten Teil folgende Formel:
 
-$$
+\[
 V_t \sim N(I_t, \sigma_v) \\
 \sigma_v \sim \Gamma(0.5, 0.5)
-$$
+\]
 
-Somit ist $\sigma_v$ auf die gleiche Weise spezifiziert wie $\sigma_\lambda$ des Prozessmodells: eine Abweichung muss positiv sein, werte grösser als 1 sind eher unwahrscheinlich.
+Somit ist \(\sigma_v\) auf die gleiche Weise spezifiziert wie \(\sigma_\lambda\) des Prozessmodells: eine Abweichung muss positiv sein, werte grösser als 1 sind eher unwahrscheinlich.
 
 In weiteren Hierarchieschritten wird nun die Beziehung der Landöffnung zur Menschlichen Besiedlung modelliert.
-Hierbei gehen wir davon aus, dass die Landöffnung linear zur Grösse der Bevölkerung innerhalb eines Zeitschrittes skaliert, das aber zwischen einzelnen Zeitschritten (durch technische Innovationen oder Veränderungen in den naturräumlichen Gegebenheiten oder kulturellen Präferenzen) eine unterschiedliche Intensität menschlicher Aktivitäten auf die Landöffnung möglich sind ($B_t$), wobei wir insgesamt gepooled davon ausgehen, das dieser Wert normalverteilt um einen Mittelwert schankt.
-Hinzu kommt eine Grundsätzliche Landöffnung, die wir zeitunabhängig annehmen, und die die natürlichen Landöffnungsprozesse (Feuer durch Blitzschlag u.ä., $BaseOpen$) wiederspiegelt.
+Hierbei gehen wir davon aus, dass die Landöffnung linear zur Grösse der Bevölkerung innerhalb eines Zeitschrittes skaliert, das aber zwischen einzelnen Zeitschritten (durch technische Innovationen oder Veränderungen in den naturräumlichen Gegebenheiten oder kulturellen Präferenzen) eine unterschiedliche Intensität menschlicher Aktivitäten auf die Landöffnung möglich sind (\(B_t\)), wobei wir insgesamt gepooled davon ausgehen, das dieser Wert normalverteilt um einen Mittelwert schankt.
+Hinzu kommt eine Grundsätzliche Landöffnung, die wir zeitunabhängig annehmen, und die die natürlichen Landöffnungsprozesse (Feuer durch Blitzschlag u.ä., \(BaseOpen\)) wiederspiegelt.
 
@@
 
-$$
+\[
 I_t = N_t * B_t + BaseOpen \\
@@
 \sigma_b \sim \Gamma(0.5,0.5)
-$$
+\]
 
-Für $\sigma_b$ ist wiederum eine Gamma-Verteilung mit Fokus auf positive Werte kleiner als 1 angenommen, für den Mittelwert $\mu_b$ nehmen wir eine breite Normalverteilung um 0.5 an, was bedeutet, das (in Zusammenspiel mit $BaseOpen$) darauf hinausläuft, dass wir im Mittel einen positiven Einfluss von menschlicher Bevölkerung auf Landöffnung annehmen, aber viel Spielraum für das Modell zur spezifikation des Wertes lassen (weakly informative prior).
+Für \(\sigma_b\) ist wiederum eine Gamma-Verteilung mit Fokus auf positive Werte kleiner als 1 angenommen, für den Mittelwert \(\mu_b\) nehmen wir eine breite Normalverteilung um 0.5 an, was bedeutet, das (in Zusammenspiel mit \(BaseOpen\)) darauf hinausläuft, dass wir im Mittel einen positiven Einfluss von menschlicher Bevölkerung auf Landöffnung annehmen, aber viel Spielraum für das Modell zur spezifikation des Wertes lassen (weakly informative prior).
 
-### Aoristic Analysis
+\hypertarget{aoristic-analysis}{%
+\subsubsection{Aoristic Analysis}\label{aoristic-analysis}}
 
-# Results
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-# Discussion
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
-# Conclusion
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
-# Acknowledgements
-
-<!-- The following line inserts a page break  -->
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
 
@@
 
-# References
+\hypertarget{references}{%
+\section{References}\label{references}}
 
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-childe_man_1936}{}}%
+Childe, V.G. (Ed.), 1936. Man {Makes} {Himself}. Watts \& Co., London.
 
-::: {#refs}
-:::
+\leavevmode\vadjust pre{\hypertarget{ref-colton_prehistoric_1949}{}}%
+Colton, H.S., 1949. The prehistoric population of the {Flagstaff} area. Plateau 22, 21--25.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cook_reconsideration_1946}{}}%
+Cook, S.F., 1946. A reconsideration of shell mounds with respect to population and nutrition. American Antiquity 12, 51--53.
+
+\leavevmode\vadjust pre{\hypertarget{ref-frankfort_town_1950}{}}%
+Frankfort, H., 1950. Town planning in ancient {Mesopotamia}. Town Planning Review 21, 98--115.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hack_changing_1942}{}}%
+Hack, J.F., 1942. The changing physical environment of the {Hopi} {Indians} of {Arizona}, Papers of the {Peabody} {Mu}- seum of {American} {Archaeology} and {Ethnolog}. Harvard University, Cambridge.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hassan_demographic_1981}{}}%
+Hassan, F.A., 1981. Demographic archaeology, Studies in archaeology. Academic Press, New York.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kintigh_grand_2014}{}}%
+Kintigh, K.W., Altschul, J.H., Beaudry, M.C., Drennan, R.D., Kinzig, A.P., Kohler, T.A., Limp, W.F., Maschner, H.D.G., Michener, W.K., Pauketat, T.R., Peregrine, P., Sabloff, J.A., Wilkinson, T.J., Wright, H.T., Zeder, M.A., 2014.
+Grand challenges for archaeology.
+Proceedings of the National Academy of Sciences 111, 879--880.
+\url{https://doi.org/10.1073/pnas.1324000111}
+
+\leavevmode\vadjust pre{\hypertarget{ref-muller_tracing_2019}{}}%
+Müller, J., Diachenko, A., 2019. Tracing long-term demographic changes: {The} issue of spatial scales. PLOS ONE 14, e0208739. \url{https://doi.org/10.1371/journal.pone.0208739}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rick_dates_1987}{}}%
+Rick, J.W., 1987. Dates as {Data}: {An} {Examination} of the {Peruvian} {Preceramic} {Radiocarbon} {Record}. American Antiquity 52, 55--73. \url{https://doi.org/10.2307/281060}
+
+\leavevmode\vadjust pre{\hypertarget{ref-shennan_population_2000}{}}%
+Shennan, S., 2000. Population, {Culture} {History}, and the {Dynamics} of {Culture} {Change}. Current Anthropology 41, 811--835.
+
+\end{CSLReferences}
 
@@
 
-### Colophon
+\hypertarget{colophon}{%
+\subsubsection{Colophon}\label{colophon}}
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+This report was generated on 2022-04-07 08:33:33 using the following computational environment and dependencies:
 
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
-```
+\begin{verbatim}
+#> - Session info ---------------------------------------------------------------
+#>  setting  value                       
+#>  version  R version 4.1.3 (2022-03-10)
+#>  os       Manjaro Linux               
+#>  system   x86_64, linux-gnu           
+#>  ui       X11                         
+#>  language (EN)                        
+#>  collate  C                           
+#>  ctype    de_DE.UTF-8                 
+#>  tz       Europe/Zurich               
+#>  date     2022-04-07                  
+#> 
+#> - Packages -------------------------------------------------------------------
+#>  package     * version date       lib source        
+#>  assertthat    0.2.1   2019-03-21 [1] CRAN (R 4.1.0)
+#>  bookdown      0.22    2021-04-22 [1] CRAN (R 4.1.0)
+#>  cachem        1.0.5   2021-05-15 [1] CRAN (R 4.1.0)
+#>  callr         3.7.0   2021-04-20 [1] CRAN (R 4.1.0)
+#>  cli           3.1.0   2021-10-27 [1] CRAN (R 4.1.1)
+#>  colorspace    2.0-1   2021-05-04 [1] CRAN (R 4.1.0)
+#>  crayon        1.4.1   2021-02-08 [1] CRAN (R 4.1.0)
+#>  DBI           1.1.1   2021-01-15 [1] CRAN (R 4.1.0)
+#>  desc          1.4.0   2021-09-28 [1] CRAN (R 4.1.1)
+#>  devtools      2.4.2   2021-06-07 [1] CRAN (R 4.1.2)
+#>  digest        0.6.27  2020-10-24 [1] CRAN (R 4.1.0)
+#>  dplyr         1.0.7   2021-06-18 [1] CRAN (R 4.1.0)
+#>  ellipsis      0.3.2   2021-04-29 [1] CRAN (R 4.1.0)
+#>  evaluate      0.14    2019-05-28 [1] CRAN (R 4.1.0)
+#>  fansi         0.5.0   2021-05-25 [1] CRAN (R 4.1.0)
+#>  farver        2.1.0   2021-02-28 [1] CRAN (R 4.1.0)
+#>  fastmap       1.1.0   2021-01-25 [1] CRAN (R 4.1.0)
+#>  fs            1.5.0   2020-07-31 [1] CRAN (R 4.1.0)
+#>  generics      0.1.0   2020-10-31 [1] CRAN (R 4.1.0)
+#>  ggplot2     * 3.3.5   2021-06-25 [1] CRAN (R 4.1.2)
+#>  glue          1.4.2   2020-08-27 [1] CRAN (R 4.1.0)
+#>  gtable        0.3.0   2019-03-25 [1] CRAN (R 4.1.0)
+#>  highr         0.9     2021-04-16 [1] CRAN (R 4.1.0)
+#>  htmltools     0.5.2   2021-08-25 [1] CRAN (R 4.1.2)
+#>  knitr         1.33    2021-04-24 [1] CRAN (R 4.1.0)
+#>  labeling      0.4.2   2020-10-20 [1] CRAN (R 4.1.0)
+#>  lifecycle     1.0.0   2021-02-15 [1] CRAN (R 4.1.0)
+#>  magrittr      2.0.1   2020-11-17 [1] CRAN (R 4.1.0)
+#>  memoise       2.0.0   2021-01-26 [1] CRAN (R 4.1.0)
+#>  munsell       0.5.0   2018-06-12 [1] CRAN (R 4.1.0)
+#>  pillar        1.6.1   2021-05-16 [1] CRAN (R 4.1.0)
+#>  pkgbuild      1.2.0   2020-12-15 [1] CRAN (R 4.1.0)
+#>  pkgconfig     2.0.3   2019-09-22 [1] CRAN (R 4.1.0)
+#>  pkgload       1.2.1   2021-04-06 [1] CRAN (R 4.1.0)
+#>  prettyunits   1.1.1   2020-01-24 [1] CRAN (R 4.1.0)
+#>  processx      3.5.2   2021-04-30 [1] CRAN (R 4.1.0)
+#>  ps            1.6.0   2021-02-28 [1] CRAN (R 4.1.0)
+#>  purrr         0.3.4   2020-04-17 [1] CRAN (R 4.1.0)
+#>  R6            2.5.0   2020-10-28 [1] CRAN (R 4.1.0)
+#>  remotes       2.4.0   2021-06-02 [1] CRAN (R 4.1.0)
+#>  rlang         0.4.11  2021-04-30 [1] CRAN (R 4.1.0)
+#>  rmarkdown     2.11    2021-09-14 [1] CRAN (R 4.1.1)
+#>  rprojroot     2.0.2   2020-11-15 [1] CRAN (R 4.1.0)
+#>  rstudioapi    0.13    2020-11-12 [1] CRAN (R 4.1.0)
+#>  scales        1.1.1   2020-05-11 [1] CRAN (R 4.1.0)
+#>  sessioninfo   1.1.1   2018-11-05 [1] CRAN (R 4.1.0)
+#>  stringi       1.7.6   2021-11-29 [1] CRAN (R 4.1.2)
+#>  stringr       1.4.0   2019-02-10 [1] CRAN (R 4.1.0)
+#>  testthat      3.0.2   2021-02-14 [1] CRAN (R 4.1.0)
+#>  tibble        3.1.2   2021-05-16 [1] CRAN (R 4.1.0)
+#>  tidyselect    1.1.1   2021-04-30 [1] CRAN (R 4.1.0)
+#>  usethis       2.1.3   2021-10-27 [1] CRAN (R 4.1.1)
+#>  utf8          1.2.1   2021-03-12 [1] CRAN (R 4.1.0)
+#>  vctrs         0.3.8   2021-04-29 [1] CRAN (R 4.1.0)
+#>  withr         2.4.2   2021-04-18 [1] CRAN (R 4.1.0)
+#>  xfun          0.23    2021-05-15 [1] CRAN (R 4.1.0)
+#>  yaml          2.2.1   2020-02-01 [1] CRAN (R 4.1.0)
+#> 
+#> [1] /home/martin/R/x86_64-pc-linux-gnu-library/4.1
+#> [2] /usr/lib/R/library
+\end{verbatim}
+
+\end{document}

### 2022-04-07 code/neotoma_helpers.R

@@ -1,0 +1,55 @@
+filter_by_element <- function(x, element) {
+  x <- lapply(x, function(y) {
+    select <- y$taxon.list$variable.element == element
+    y$taxon.list <- y$taxon.list[select,]
+    y$counts <- y$counts[,colnames(y$counts) %in% y$taxon.list$taxon.name]
+    return(y)
+  })
+  class(x) <- append(class(x), "download_list")
+  return(x)
+}
+
+filter_by_unit <- function(x, unit) {
+  x <- lapply(x, function(y) {
+    select <- y$taxon.list$variable.unit == unit
+    y$taxon.list <- y$taxon.list[select,]
+    y$counts <- y$counts[,colnames(y$counts) %in% y$taxon.list$taxon.name]
+    return(y)
+  })
+  class(x) <- append(class(x), "download_list")
+  return(x)
+}
+
+transform_to_percent <- function(x)
+{
+  x <- lapply(x, function(y) {
+    y$counts <- analogue::tran(x = y$counts, method ="percent")
+    return(y)
+  })
+  class(x) <- append(class(x), "download_list")
+  return(x)
+}
+
+aggregate_by_joint_name <- function(x, common_names) {
+
+  out <- pollen_data_comp %>% lapply(function(y) {
+    y$taxon.list <- merge(y$taxon.list, common_names, by.x = "taxon.name", by.y = "name_orig")
+    counts.new <- data.frame(t(y$counts))
+    counts.new$name_joint <- NA
+    counts.new[y$taxon.list$taxon.name,]$name_joint <- y$taxon.list$name_joint
+    counts.collected <- counts.new  %>%
+      as_tibble() %>%
+      group_by(name_joint) %>%
+      summarise(across(where(is.numeric), sum))
+    counts.new <- data.frame(t(counts.collected[,2:ncol(counts.collected)]))
+
+    colnames(counts.new) <- counts.collected$name_joint
+    rownames(counts.new) <- rownames(y$counts)
+
+    y$counts <- counts.new
+    return(y)
+  })
+
+  class(out) <- append(class(out), "download_list")
+  return(out)
+}

### 2022-04-07 analysis/analysis.Rmd
updated materialized analysis rmd

@@
 
-mapview(pollen_locations.sf)
+pollenmap <- tempfile(fileext = ".png")
+mapshot(
+  mapview(pollen_locations.sf),
+  file = pollenmap
+  )
+knitr::include_graphics(pollenmap)
 
@@
 ```{r}
+c14map <- tempfile(fileext = ".png")
+mapshot(
   mapview(st_as_sf(c14,
@@
                    crs = 4326),layer.name = "14C Data"
+          ),
+  file = c14map
   )
+knitr::include_graphics(c14map)
 ```
@@
 ```{r}
-c14.calibrated <- calibrate(x = c14$bp, errors = c14$std)
+c14.calibrated <- calibrate(x = c14$bp, errors = c14$std, verbose = F)
 ```
@@
 ```{r}
-sumcal <- spd(c14.calibrated, timeRange = 1950-timeframe, bins = c14$site, runm = sw)
+sumcal <- spd(c14.calibrated, timeRange = 1950-timeframe, bins = c14$site, runm = sw, verbose = F)
 ```
@@
 ```{r}
+sitesmap <- tempfile(fileext = ".png")
+mapshot(
   mapview(st_as_sf(sites,
@@
           layer.name = "Sites"
+  ),
+  file = sitesmap
 )
+knitr::include_graphics(sitesmap)
 ```
@@
 ```{r}
-res.lss[res.lss$date %in%
-seq(from=timeframe[1],
-    to=timeframe[2], by=50),]
+res.lss <- res.lss[res.lss$date %in% seq(from=timeframe[1],
+                                         to=timeframe[2], by=50),
+        ]
 res.lss$age <- 1950-res.lss$date
@@
 
+The final run after convergence to get more than the min 10000 effective samples for each parameter takes quite some time.
+On my machine, it took nearly 5 hours.
+Therefore, I add a switch here for not doing this analysis, but feel free to proceed yourself, if you like:
+
+```{r}
+do_final_run = F
+```
+
 First we set our new meta-parameters:
@@
 start_time <- Sys.time()
-while (!enough_samples) {
+while (!enough_samples & do_final_run) {
 
@@
 stopCluster(cl)
+```
+
+If the final run above was skipped, but you have already the results from a previous run, you can load this at this stage:
+
+```{r}
+collector <- readRDS("mcmc_collector.RDS")
 ```

### 2022-04-07 README.Rmd
added an RMD readme and knitted

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Estimation of Population Density for the prehistoric Swiss Plateau using a Hierarchical Bayesian Poisson Regression Model"
```

# bayesian.demographic.reconstruction.2022

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/MartinHinz/bayesian.demographic.reconstruction.2021/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **repository** consists of:

  - [:file\_folder: paper](/manuscript): R Markdown source document
    for manuscript.
  - [:file\_folder: paper](/paper): A rendered version, of the manuscript
    `paper.docx` and `paper.pdf`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: paper](/analysis): R Markdown source document
    for the actual analysis. Includes code to reproduce the figures and tables
    generated by the analysis.
  - [:file\_folder: code](/code): Functions used in the analysis.
  - [:file\_folder: data](/data): Data used in the analysis.
  - [:file\_folder: figures](/figures): Plots and other illustrations

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2022-04-07 README.Rmd
minor corrections

@@
 
-  - [:file\_folder: paper](/manuscript): R Markdown source document
+  - [:file\_folder: manuscript](/manuscript): R Markdown source document
     for manuscript.
@@
     and tables in this file)
-  - [:file\_folder: paper](/analysis): R Markdown source document
+  - [:file\_folder: analysis](/analysis): R Markdown source document
     for the actual analysis. Includes code to reproduce the figures and tables

### 2022-04-07 analysis/analysis.Rmd
removed visual clutter from the library calls and made the maps work again

@@
 knitr::opts_chunk$set(echo = TRUE)
+if (!webshot::is_phantomjs_installed()) webshot::install_phantomjs()
+library(magick)
 ```
@@
 
-```{r}
+```{r message=FALSE}
 library(tidyverse)
@@
 
-```{r}
+```{r message=FALSE}
 if(!require('devtools')) install.packages('devtools')
@@
   )
-knitr::include_graphics(pollenmap)
 
+img <- magick::image_read(pollenmap)
+plot(img)
 ```
@@
   )
-knitr::include_graphics(c14map)
+
+img <- magick::image_read(c14map)
+plot(img)
 ```
@@
 )
-knitr::include_graphics(sitesmap)
+img <- magick::image_read(sitesmap)
+plot(img)
 ```

### 2022-04-07 analysis/analysis.Rmd
more optical decluttering and improvement

@@
 
-```{r}
+```{r message=FALSE}
 pollen_data <- get_download(pollen_locations$neotoma_id)
@@
     to=10000, by=50),]
-ggplot(res.open_smoothed,aes(x=1950-age,y=moving_mean)) + geom_point(aes(color=site), alpha =.5)  + geom_smooth(span = 0.05, color = "black")+ scale_x_continuous(breaks=seq(timeframe[1],timeframe[2],200), limits = timeframe) + labs(y = "Openness score [PCA]",x = "Date",colour = "Sites")  + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
+ggplot(res.open_smoothed,aes(x=1950-age,y=moving_mean)) +
+  geom_point(aes(color=site), alpha =.5) +
+  geom_smooth(span = 0.05, color = "black") +
+  scale_x_continuous(breaks=seq(timeframe[1],timeframe[2],200), limits = timeframe) +
+  labs(y = "Openness score [PCA]",x = "Date",colour = "Sites") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 ```
@@
 
-ggplot(res.open_final,aes(x=1950-age)) + geom_line(aes(y=mean)) + geom_segment(aes(y=mean-sd, yend =mean+sd, xend=1950-age), alpha = .25)
+ggplot(res.open_final,aes(x=1950-age)) +
+  geom_line(aes(y=mean)) +
+  geom_segment(aes(y=mean-sd, yend =mean+sd, xend=1950-age), alpha = .25)
 ```
@@
 
-chrono_rough <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "chrono_rough.csv")), row.names = 1)
-chrono_vrough <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "chrono_very_rough.csv")), row.names = 1)
+chrono_rough <- read.csv(file = normalizePath(
+  file.path(here(), "data","raw_data", "chrono_rough.csv")
+  ),
+  row.names = 1)
+
+chrono_vrough <- read.csv(file = normalizePath(
+  file.path(here(), "data","raw_data", "chrono_very_rough.csv")
+  ),
+  row.names = 1)
 ```
@@
   geom_line(data = res.aorist, aes(x = date, y = scale(sum)), col="gray", alpha = 0.75) + 
-  geom_line(data = res.lss, aes(x = date, y = scale(moving_mean)), col="brown", alpha = 0.75) + scale_colour_manual(values = c("darkgreen", "red", "gray", "brown"), labels = c("Openness Indicator", "Sum Calibration", "Aoristic Sum", "Dendro Dated Settlements"))
+  geom_line(data = res.lss, aes(x = date, y = scale(moving_mean)), col="brown", alpha = 0.75) +
+  scale_colour_manual(values = c("darkgreen", "red", "gray", "brown"),
+                      labels = c("Openness Indicator", "Sum Calibration", "Aoristic Sum", "Dendro Dated Settlements")
+                      )
 ```
@@
   select(calBP, PrDens, mean, sum, moving_mean) %>%
-  rename(age = calBP, sumcal = PrDens, openness = mean, aoristic_sum = sum, dendro = moving_mean) %>% filter(age < 8000 & age > 3000)
+  rename(age = calBP, sumcal = PrDens, openness = mean, aoristic_sum = sum, dendro = moving_mean) %>%
+  filter(age < 8000 & age > 3000)
 
@@
 ```{r}
-write.csv(all_proxies, file = normalizePath(file.path(here(), "data","preprocessed_data", "all_proxies.csv"), mustWork = F))
+write.csv(all_proxies, file = normalizePath(
+  file.path(here(), "data","preprocessed_data", "all_proxies.csv"
+            ),
+  mustWork = F)
+  )
 ```
@@
 ```{r}
-all_proxies <- read.csv(file = normalizePath(file.path(here(), "data","preprocessed_data", "all_proxies.csv")), row.names = 1)
+all_proxies <- read.csv(file = normalizePath(
+  file.path(here(), "data","preprocessed_data", "all_proxies.csv")
+  ),
+  row.names = 1)
 ```
@@
 # Export common values for the cluster
-clusterExport(cl, c("model_code", "model_inits", "model_data", "model_constants", "model_monitors", "batches", "thinning"))
+clusterExport(cl,
+              c("model_code",
+                "model_inits",
+                "model_data",
+                "model_constants",
+                "model_monitors",
+                "batches",
+                "thinning"
+                )
+              )
 
@@
 ```{r}
-MCMCchains(mcmcSamples, params = "p") %>% as.data.frame %>% rename(sumcal = 'p[1]', openness = 'p[2]', aorist = 'p[3]', dendro = 'p[4]') %>% pivot_longer(everything()) %>% ggplot(aes(x = value, y = name)) + stat_halfeye()
+MCMCchains(mcmcSamples, params = "p") %>%
+  as.data.frame %>%
+  rename(sumcal = 'p[1]', openness = 'p[2]', aorist = 'p[3]', dendro = 'p[4]') %>%
+  pivot_longer(everything()) %>%
+  ggplot(aes(x = value, y = name)) + stat_halfeye()
 ```
@@
 ```{r fig.width=5,fig.height=12}
-MCMCsummary(mcmcSamples) %>% as.data.frame() %>% rownames_to_column() %>% ggplot() + geom_bar(aes(y=n.eff, x=rowname), stat = "identity") + coord_flip()
+MCMCsummary(mcmcSamples) %>%
+  as.data.frame() %>%
+  rownames_to_column() %>%
+  ggplot() + geom_bar(aes(y=n.eff, x=rowname), stat = "identity") + coord_flip()
 ```
@@
 
-  gc() # Garbage collector for more RAM
+  gc(verbose=F) # Garbage collector for more RAM
 
@@
 MCMCtrace(collector, params = "PopDens", priors = PopDens, pdf = F, ind = T, Rhat = T)
-gc()
+gc(verbose=F)
 ```
@@
 
-gc()
+gc(verbose=F)
 
@@
   geom_line(aes(x = 1950-age, y = scale(aoristic_sum), col="Aoristic Sum"), alpha = 0.5) +
-  geom_line(data = PopDens_summary, aes(x = rev(1950-all_proxies_orig$age), y=mean, col = "Estimation Result")) + 
-  geom_line(data = PopDens_summary, aes(x = rev(1950-all_proxies_orig$age), y=hpdi_l, col = "Estimation Result"), lty=2) + 
-  geom_line(data = PopDens_summary, aes(x = rev(1950-all_proxies_orig$age), y=hpdi_u, col = "Estimation Result"), lty=2) + theme_minimal() + 
-  scale_colour_manual(values = c("Openness" = "darkgreen", "Sum Calibration" = "red", "Aoristic Sum" = "gray", "Dendro Dated Settlements" = "brown", "Estimation" = "black"), name = "") + theme(legend.position="bottom") + ylab("Scaled values for Proxies,\n P/km² for Estimation") + xlab("cal BCE")
+  geom_line(data = PopDens_summary,
+            aes(x = rev(1950-all_proxies_orig$age), y=mean, col = "Estimation Result")) + 
+  geom_line(data = PopDens_summary,
+            aes(x = rev(1950-all_proxies_orig$age), y=hpdi_l, col = "Estimation Result"), lty=2) + 
+  geom_line(data = PopDens_summary,
+            aes(x = rev(1950-all_proxies_orig$age), y=hpdi_u, col = "Estimation Result"), lty=2) +
+  theme_minimal() + 
+  scale_colour_manual(
+    values = c("Openness" = "darkgreen",
+               "Sum Calibration" = "red",
+               "Aoristic Sum" = "gray",
+               "Dendro Dated Settlements" = "brown",
+               "Estimation" = "black"), name = "") +
+  theme(legend.position="bottom") +
+  ylab("Scaled values for Proxies,\n P/km² for Estimation") +
+  xlab("cal BCE")
 ```
@@
 ```{r}
-all_chains[,grep("^p\\[", params)] %>% as.data.frame %>% rename(sumcal = 'p[1]', openness = 'p[2]', aorist = 'p[3]', dendro = 'p[4]') %>% pivot_longer(everything()) %>% ggplot(aes(x = value, y = name)) + stat_halfeye()
+all_chains[,grep("^p\\[", params)] %>%
+  as.data.frame %>%
+  rename(sumcal = 'p[1]', openness = 'p[2]', aorist = 'p[3]', dendro = 'p[4]') %>%
+  pivot_longer(everything()) %>%
+  ggplot(aes(x = value, y = name)) +
+  stat_halfeye()
 ```

### 2022-04-08 analysis/analysis.Rmd
removed more optical clutter when loading pollen data
some improvements and additions

@@
 
--   ^14^C sum calibration, data sets from the XRONOS database as well as from the unpublished PhD thesis of Julian Laabs [\@laabs2019] and from the data collection of [\@martínez-grau2021].
+-   ^14^C sum calibration, data sets from the XRONOS database as well as from the unpublished PhD thesis of Julian Laabs [@laabs2019] and from the data collection of [@martínez-grau2021].
 
@@
 img <- magick::image_read(pollenmap)
+old.par <- par(mar = c(0, 0, 0, 0))
 plot(img)
+par(old.par)
 ```
@@
 
-```{r message=FALSE}
+```{r message=FALSE, warning=FALSE}
 pollen_data <- get_download(pollen_locations$neotoma_id)
@@
 all_output <- suppressWarnings(neotoma::compile_downloads(pollen_data_comp))
-
 ```
@@
 img <- magick::image_read(c14map)
+old.par <- par(mar = c(0, 0, 0, 0))
 plot(img)
+par(old.par)
 ```
@@
 img <- magick::image_read(sitesmap)
+old.par <- par(mar = c(0, 0, 0, 0))
 plot(img)
+par(old.par)
 ```

### 2022-04-08 manuscript/paper.Rmd
some improvements and additions

@@
 ---
-title: "Bayesian Hierarchical Model for estimating prehistoric population development on the Swiss Plateau"
+title: "Bayesian Hierarchical Model for estimating holocene prehistoric population development on the Swiss Plateau"
 author:
@@
       email: caroline.heitz@arch.ox.ac.uk
-      institute: [OX, IAW, OCCR]
+      institute: [IAW, OCCR, OX]
       correspondence: false
@@
   - OCCR: Oeschger Centre for Climate Change Research, University of Bern
+  - SFB1266: CRC 1266 - Scales of Transformation, University of Kiel
   - OX: Associated Researcher, School of Archaeology, University of Oxford
@@
   - IAMFAMU: Institute of Archaeology and Museology, Faculty of Arts, Masaryk University
-  - SFB1266: CRC 1266 - Scales of Transformation, University of Kiel
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
     - --lua-filter=../templates/pagebreak.lua
-  bookdown::pdf_book: default
+  bookdown::pdf_document2:
+    toc: no
+    pandoc_args:
+    - --lua-filter=../templates/scholarly-metadata.lua
+    - --lua-filter=../templates/author-info-blocks.lua
+    - --lua-filter=../templates/pagebreak.lua
 bibliography: references.bib
@@
 
-```{r setup, echo = FALSE}
+```{r setup, echo = FALSE, message = FALSE}
 knitr::opts_chunk$set(
@@
 )
+
+library(ggplot2)
+library(ggmap)
+library(sp)
+library(sf)
+library(ggsn)
+library(rnaturalearth)
+library(ggrepel)
+
+switzerland <- ne_countries(country = "Switzerland")
+bbox_ch <- bbox(switzerland) %>% as.vector()
+
+swiss_background <- get_stamenmap( bbox = bbox_ch, maptype = "terrain-background", color="bw", zoom = 8)
+
+my_basemap <- ggmap(swiss_background, darken = c(0.6, "white"))
+
+my_scalebar <- scalebar(transform = T, dist_unit = "km", dist=50, model = "WGS84", st.dist = 0.025, st.size = 3, x.min = bbox_ch[1], x.max = bbox_ch[3]-0.2, y.min = bbox_ch[2]+0.15, y.max = bbox_ch[4], border.size = 0.5)
 ```
@@
 
-\[Beispiele bringen? Bring some examples here...\]
+$$Beispiele bringen? Bring some examples here...$$
 
@@
 Bayesian demographic forecasting has been designed to solve all these problems, which also exist for demographic estimations using recent data.
-In their book on the subject, Bryant and Zhang \[BrynatandZhang\] state that Bayesian data modeling can be used to deal with exactly those problems that also affect archaeological data.
+In their book on the subject, Bryant and Zhang [@bryant2018] state that Bayesian data modeling can be used to deal with exactly those problems that also affect archaeological data.
 Especially for limited, unreliable and noisy data, Bayesian approaches are optimally suited.
@@
 
-# Methods
+# Materials and Methods
 
@@
 It is a univariate discrete probability distribution that can be used to model the number of events that occur independently of each other at a constant mean rate in a fixed time interval or spatial area.
-It is determined by a real parameter $\lambda$ >0, which describes the expected value and simultaneously the variance of the distribution.
+It is determined by a real parameter $\lambda$ \>0, which describes the expected value and simultaneously the variance of the distribution.
 Thus, the relationship shown above can also be rearranged as follows:
@@
 $$
+\begin{split}
 N_t \sim dpois(\lambda_t) \\
 \lambda_t = N_t
-$$
-If we now have information about the change in population development, we can use this to enter it into the model via a change in $\lambda$.
+\end{split}
+$$ If we now have information about the change in population development, we can use this to enter it into the model via a change in $\lambda$.
 This is done in the form of a regression: If $x \in R^n$ is a vector of independent variables, then the model takes the form
@@
 log({E} (Y\mid x))=\alpha + \beta' x
-$$
-In the case of population trends, the expected value would be equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables:
+$$ In the case of population trends, the expected value would be equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables:
 
@@
 $$
+\begin{split}
 max\_growth\_rate \sim dgamma(shape = 5, scale=0.05) \\
@@
 N_{t-1}/N_t < (max\_growth\_rate + 1)
+\end{split}
 $$
 
-The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$ (See also Figure X).
+The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$ (See also Figure \@ref(fig:gammadist)).
 
-```{r}
+```{r gammadist}
 plot(density(rgamma(100000,shape=5,scale = 0.05)+1), xlim=c(0,5), main ="gamma(shape = 5, scale=0.05) + 1" )
@@
 
-\[Prior Probability Simulation einbauen\]
-
 ## Data/Proxies
@@
 In principle, a large number of different data sources can be integrated into the overall model as observations, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created.
-A non-exhaustive list can be found in the following table:
+A non-exhaustive list can be found in the following table \@ref(tab:tableproxies):
 
@@
 
-: (#tab:table_proxies) A incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.
+: (\#tab:tableproxies) A incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.
 
 Our working region consists of the Swiss Plateau, which on the one hand offers excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking.
-If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also causes a research problem with regard to the ^14^C data often used as a proxy.
+If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also might cause a research problem with regard to the ^14^C data often used as a proxy.
+
+```{r echo=FALSE}
+dendro <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "raw_data", "dendro_years.csv"))
+```
+
+The dataset we use for the number of dendrodated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019].
+Details on the creation of this data series will be published at the referenced location.
+The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements.
+This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.
+
 Since many periods of the Swiss prehistoric dry land settlements are unknown, and archaeology is based primarily on the wet land settlements, the dendro data are used primarily here, since they are much more precise.
@@
 
-```{r}
-library(ggplot2)
+```{r echo=FALSE}
+c14 <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "raw_data", "14c_swiss_plateau.csv"))
+```
+
+The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database, supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of [@martínez-grau2021].
+It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites.
+The dates were selected so that their distribution area coincides with the catchment area of the pollen proxy explained below (see also Figure \@ref(fig:c14map)).
+The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` BP uncal.
+This time window extends beyond the study horizon in order to minimise boundary effects, which, however, cannot be completely avoided due to the Hallstatt plateau.
+But it was precisely to reduce its influence that the time window of the investigation was chosen to be 7000 -- 1000 BCE.
+
+We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites.
+For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [@crema2021] were used in the standard settings.
+
+```{r c14map, fig.cap="The location of the ^14^C dated sites in the dataset."}
+my_basemap + geom_point(data = c14, aes(x = lng, y = lat), color = "darkred", alpha = 0.75) + blank() + my_scalebar
+```
+
+```{r compdendroc14, fig.cap="Comparison of the scaled number of ^14^C and dendrodated sites over time in the dataset used."}
 joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))
 
-ggplot(joined_data) + geom_line(aes(x=1950-age, y=scale(sumcal, center = F), color = "red"), alpha = 0.75)  + geom_line(aes(x=1950-age, y=scale(dendro, center = F), color = "blue"), alpha = 0.75) + theme_minimal()  +
+ggplot(joined_data) + geom_line(aes(x=1950-age, y=scale(sumcal, center = F), color = "darkred"), alpha = 0.75)  + geom_line(aes(x=1950-age, y=scale(dendro, center = F), color = "darkblue"), alpha = 0.75) + theme_minimal()  +
     labs(x = "years BCE",
          y = "scaled values",
-         color = "Legend") + ggtitle("Number of dendro and 14C dated sites on the Swiss Plateau (scaled)") +
-    scale_color_manual(values = c("blue", "red"), labels = c("dendro dated", "14C dated"))
+         color = "Legend") + 
+    scale_color_manual(values = c("darkblue", "darkred"), labels = c("dendro dated", "14C dated"))+
+  theme(legend.position="bottom")
 ```
 
+We can now compare these two data sets (Figure \@ref(fig:compdendroc14)).
 In fact, there is a not uninteresting fit between the two data series.
@@
 
-We have integrated the ^14^C data as one of the proxies in the model, in the form of a sum calibration.
-The dataset consists of 1135 ^14^C dates from 246 sites, and dates between 10730 and 235 BP uncal.
-We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites.
-For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [!citation!] were used in the standard settings.
+```{r echo=FALSE}
+aorist <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "raw_data", "site_data_fuzzy.csv"))
+aorist_latlng_coords <- st_as_sf(aorist,coords = c("LV03_Rechtswert_fuzzy", "LV03_Hochwert_fuzzy"),crs = 21781) %>% sf::st_transform(
+  4326
+) %>% st_coordinates() %>% as.data.frame()
+```
 
-\[Histogramm 14C Datierungen\]
+```{r aoristmap, fig.cap="The location of the sites from the find reports of the cantonal archaeology (heritage management). The locations are fuzzified by ~1km."}
+my_basemap + geom_point(data = aorist_latlng_coords, aes(x = X, y = Y), color = "black", alpha = 0.75) + blank() + my_scalebar
+```
 
-Zugleich stand uns ein Datensatz über die Anzahl der Dendrodatierten Feuchtbodensiedlungen in der Drei-Seen-Region zur Verfügung, welcher durch Julian Laabs für seine PhD-These gesammelt worden ist und deren Hintergrunddaten am genannten Ort veröffentlicht werden.
-Die hier verwendete Zeitreihe läuft von X bis Y, und enthält die Anzahl von zeitliche verzeichneten Schlagphasen an einzelnen Siedlungen.
-Somit ergibt sich hier eine Zeitreihe, die die Besiedlung der Seeufer in der Zeit des Neolithikums und der Bronzezeit (?) wiederspiegelt.
-
-In order to add another indicator of archaeological evidence of occupation, we have included the data of the Cantonal Archaeology, and thus the Heritage Management, which are primarily derived from scattered surface finds, and which often have a low depth of information and thus dating accuracy.
+In order to add another indicator of archaeological evidence of occupation, we have included the data of the Cantonal Archaeology (Figure \@ref(fig:aoristmap)), and thus the Heritage Management, which are primarily derived from scattered surface finds, and which often have a low depth of information and thus dating accuracy.
 This information is incorporated into our model as a typologically dated aorist time series.
@@
 
-\[Abbildung Aoristische Kurve\]
+```{r aoristcurve, fig.cap="Comparison of the scaled number of ^14^C and dendrodated sites over time in the dataset used."}
+ggplot(joined_data) + geom_line(aes(x=1950-age, y=aoristic_sum, color = "darkgray"), alpha = 0.75) +
+  theme_minimal()  +
+    labs(x = "years BCE",
+         y = "aoristic sum",
+         color = "Legend") + 
+    scale_color_manual(values = "darkgray", labels = "Aoristic Sum")+
+  theme(legend.position="bottom")
+```
 
-Die durch die vielen sehen Natur räumlich gegebenen Gegebenheiten ermöglichen neben einer hoch präzisen Datierung von archäologischen Fundstellen gleichzeitig aber auch ein sehr dichtes Netz an pollenanalytischen Untersuchungen.
-Diesen Fakt machen wir uns zu Nutze, in dem wir aus den Pollendaten einem überregionalen Offenheitsanzeiger für den Bewuchs generieren.
-Genaueres dazu wird im Methoden Kapitel ausgeführt.
-Dieser Indikator hat spezifisch den Vorteil, dass er nicht an Überlieferungsbedingungen gebunden ist, wie ist die archäologischen Indikatoren sind.
-Damit wird er besonders wertvoll, um systematische Verzierungen anzuzeigen oder auszugleichen, die sich durch zeitlich spezifische Siedlungsweisen und Überlieferungsbedingungen ergeben.
+The natural conditions provided by the many lakes enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis.
+We make use of this fact by generating a supra-regional openness indicator for the vegetation from the pollen data (Figure \@ref(fig:pollensites)).
+This indicator has the specific advantage that it is not dependent on preservation conditions, as archaeological indicators are.
+This makes it particularly valuable for indicating or compensating for systematic distortions that result from temporally specific settlement patterns and archaeological preservation conditions.
+
+```{r pollensites, fig.cap="Location of the pollen profiles used for the openness indicator."}
+pollen_sites <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "raw_data", "pollen_locations.csv"))
+
+my_basemap +
+  geom_point(data = pollen_sites, aes(x = lng, y = lat), color = "darkgreen", alpha = 0.75) + blank() +
+  geom_text_repel(data = pollen_sites, aes(x = lng, y = lat, label = name), alpha = 0.75, nudge_x = 0.2, nudge_y=0.02) +
+  my_scalebar
+```
 
@@
 
-\[Abbildung Offenheitsanzeiger\]
+$$Abbildung Offenheitsanzeiger$$
 
@@
 
-\[Tabelle Expertenschätzungen\]
+$$Tabelle Expertenschätzungen$$
 

### 2022-04-09 analysis/analysis.Rmd
further additions to manuscript and analysis

@@
 
-```{r}
+```{r pollenproxy}
 res.open_smoothed <- res.open_smoothed[res.open_smoothed$age %in%
@@
 
-ggplot(all_proxies_orig) +
+popdens_plot <- ggplot(all_proxies_orig) +
   geom_line(aes(x = 1950-age, y = scale(openness), col="Openness Indicator"), alpha = 0.5) +
@@
   xlab("cal BCE")
+popdens_plot
+```
+
+Save as result image to the figures folder.
+
+```{r}
+ggsave(normalizePath(file.path(here(), "figures","popdens_estimation.pdf"), mustWork = F),
+       plot = popdens_plot, width = 21, height = 29.7/2, units = "cm")
 ```
@@
 ```{r}
-all_chains[,grep("^p\\[", params)] %>%
+p_plot <- all_chains[,grep("^p\\[", params)] %>%
   as.data.frame %>%
@@
   ggplot(aes(x = value, y = name)) +
-  stat_halfeye()
-```
+  stat_halfeye() + xlab("Percentage Influence") + ylab("Proxy")
+p_plot
+```
+
+Also save as result image.
+
+```{r}
+ggsave(normalizePath(file.path(here(), "figures","p_estimation.pdf"), mustWork = F),
+       plot = p_plot, width = 21, height = 29.7/2, units = "cm")
+```
+
+Also interesting is the variation coefficient from the estimation of the Population density:
+
+```{r}
+var_popdens <- data.frame(var_coeff = apply(all_chains[,2:100], 2, function(x) sd(x)/mean(x)),
+                          age = rev(all_proxies_orig$age))
+
+coeff <- 0.1
+
+popvar_plot <- ggplot(var_popdens) +
+  geom_line(aes(x = 1950-age, y = var_coeff/coeff, col="Variation Coefficient"), alpha = 0.5) +
+  geom_line(data = PopDens_summary,
+            aes(x = rev(1950-all_proxies_orig$age), y=mean, col = "Estimation Result")) + 
+  geom_line(data = PopDens_summary,
+            aes(x = rev(1950-all_proxies_orig$age), y=hpdi_l, col = "Estimation Result"), lty=2) + 
+  geom_line(data = PopDens_summary,
+            aes(x = rev(1950-all_proxies_orig$age), y=hpdi_u, col = "Estimation Result"), lty=2) +
+  theme_minimal() + 
+  scale_colour_manual(
+    values = c(
+               "Variation Coefficient" = "red",
+               "Estimation" = "black"), name = "") +
+  theme(legend.position="bottom") +
+  xlab("cal BCE") +
+
+  scale_y_continuous(
+
+    # Features of the first axis
+    name = "Scaled values for Proxies,\n P/km² for Estimation",
+
+    # Add a second axis and specify its features
+    sec.axis = sec_axis(~.*coeff, name="Variation Coefficient")
+  ) +
+
+  theme(
+    axis.title.y.right = element_text(color = "red"),
+    axis.text.y.right = element_text(color = "red"),
+     axis.line.y.right = element_line(color = "red"), 
+     axis.ticks.y.right = element_line(color = "red")
+  )
+
+popvar_plot
+```
+
+Also save as result image.
+
+```{r}
+ggsave(normalizePath(file.path(here(), "figures","popvar_plot.pdf"), mustWork = F),
+       plot = popvar_plot, width = 21, height = 29.7/2, units = "cm")
+```

### 2022-04-09 manuscript/paper.Rmd
further additions to manuscript and analysis

@@
 $$
-\begin{split}
-N_t \sim dpois(\lambda_t) \\
-\lambda_t = N_t
-\end{split}
-$$ If we now have information about the change in population development, we can use this to enter it into the model via a change in $\lambda$.
+\begin{aligned}
+N_t &\sim dpois(\lambda_t) \\
+\lambda_t &= N_t
+\end{aligned}
+$$If we now have information about the change in population development, we can use this to enter it into the model via a change in $\lambda$.
 This is done in the form of a regression: If $x \in R^n$ is a vector of independent variables, then the model takes the form
@@
 log({E} (Y\mid x))=\alpha + \beta' x
-$$ In the case of population trends, the expected value would be equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables:
+$$
+
+In the case of population trends, the expected value would be equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables:
 
 $$
-log(\lambda_t) = log(N_t) + \sum_{i=1}^n \beta_i x_{t,i}
+log(\lambda_t) = log(N_{t-1}) + \sum_{i=1}^n \beta_i x_{t,i}
 $$
@@
 $$
-\begin{split}
-max\_growth\_rate \sim dgamma(shape = 5, scale=0.05) \\
-N_t/N_{t-1} < (max\_growth\_rate + 1) \\
-N_{t-1}/N_t < (max\_growth\_rate + 1)
-\end{split}
+\begin{aligned}
+max\_growth\_rate &\sim dgamma(shape = 5, scale=0.05) \\
+N_t/N_{t-1} &< (max\_growth\_rate + 1) \\
+N_{t-1}/N_t &< (max\_growth\_rate + 1)
+\end{aligned}
 $$
@@
 
-```{r gammadist}
+```{r gammadist, fig.cap="The gamma-distributed maximum growth rate prior of the model."}
 plot(density(rgamma(100000,shape=5,scale = 0.05)+1), xlim=c(0,5), main ="gamma(shape = 5, scale=0.05) + 1" )
@@
 
-: (\#tab:tableproxies) A incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.
+: (#tab:tableproxies) A incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.
 
@@
 
-<!--# Ab hier muss noch umformuliert werden, ist cp aus dem Cambridge Artikel -->
+The utilisation of this proxy is based on the assumption that the higher the population density in an area, the greater the human influence on the natural environment [@lechterbeck2014], and that the agricultural space of an area is closely related to the population density [@zimmermann2004].
+Evidence of land clearing in pollen diagrams can therefore give further indications regarding population dynamics.
+The exact methodological procedure for obtaining this proxy from several, in our case 5, different pollen diagrams of sites mainly in the hinterland of the large Alpine lakes can be found in a previous publication [@heitz2021].
+The technical procedure is also documented in the accompanying R-script.
+The percentage pollen data based on a pollen sum of all terrestrial taxa of the individual sites were combined into one data set by means of a principal component analysis (Figure \@ref(fig:pollenproxy)).
+Only terrestrial pollen taxa with a frequency of more than 1/3 and, if present, with an average frequency of at least more than 0.1% were selected to reduce potential disturbance by rare species.
+Cereal pollen was explicitly retained as an important anthropogenic indicator.
+As each sample is absolutely dated, the data on the x-axis can be plotted against the openness value on the y-axis to obtain a time series time series for land clearing.
 
-Es ist eine allgemein akzeptierte Annahme, dass der menschliche Einfluss umso größer ist, je höher die Bevölkerungsdichte in einem Gebiet ist (Lechterbeck et al.
-2014), und dass die Bevölkerungsdichte eines Gebiets eng mit der landwirtschaftlichen Fläche zusammenhängt (Zimmermann et al.
-2004).
-Hinweise auf Landrodung in Pollendiagrammen können daher zusätzliche Anhaltspunkte für die Bevölkerungsdynamik liefern.
-Da ein einzelnes Pollenprofil eine Kombination von (außer-)lokalen und regionalen Signalen darstellt, haben wir für überregionale Analysen verschiedene Pollendiagramme von 5 kleinen Seen im Hinterland der großen Alpenseen kombiniert (Abb.
-7): Le Loclat (Hadorn 1992, 1994), Moossee (Rey et al.
-2019a, 2019b), Burgäschisee (Rey et al.
-2019a, 2019b), Soppensee (Hajdas und Michczynski 2010; Lotter 1999), Rotsee (Lotter 1988a, 1988b, 1990).
-Die Daten wurden aus der Neotoma-Datenbank extrahiert (Williams et al.
-2018; dort Datensatznummern 26632, 40454, 46280, 40955, 44723, 4382).
-Die prozentualen Pollendaten, die auf einer Pollensumme aller terrestrischen Taxa der einzelnen Standorte basieren, wurden zu einem Datensatz zusammengefasst (Abb.
-8a).
-Nur terrestrische Pollentaxa mit einer Häufigkeit von mehr als 1/3 und, falls vorhanden, mit einer durchschnittlichen Häufigkeit von mindestens mehr als 0,1 % wurden ausgewählt, um die potenzielle Störung durch seltene Arten zu reduzieren.
-Getreidepollen wurden ausdrücklich als wichtiger anthropogener Indikator beibehalten.
+```{r pollenproxy, fig.cap="Value on the first dimension of the PCA against dating of the samples for the individual pollen profiles and their combined average value as the openness indicator."}
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "analysis", "analysis_files", "figure-html", "pollenproxy-1.png"))
+```
 
-Wir wendeten eine Hauptkomponentenanalyse (PCA) nach der von Feeser et al.
-(2019) vorgeschlagenen Methode an und skalierten die relativen Pollenanteile der einzelnen Taxone durch z-Scores-Standardisierung.
-In der Paläoökologie können multivariate Ordinationsverfahren wie die PCA verwendet werden, um zugrundeliegende Gradienten innerhalb der Daten aufzudecken (Feeser et al.
-2019; Hinz et al.
-2012).
-Aufgrund der Verbreitung der Buche (Fagus) gibt es einen starken erklärbaren Gradienten im Datensatz, der das gewünschte Ergebnis verschleiern würde.
-Um dies zu eliminieren, entschieden wir uns, eine Partial Constrained PCA durchzuführen.
-Das Ergebnis der partiellen PCA, die die Pollendaten zwischen 7000 und 1 cal BCE einschließt, zeigt, dass 20,5 % der Variabilität im Datensatz durch die erste Achse erklärt wird (die zweite Achse macht 12,6 % aus, Abb.
-8b).
-Zu den Taxa mit hohen Werten auf der ersten Achse gehören neben Wildgrasarten auch klassische anthropogene Indikatoren (z.
-B.
-Cerealia, Urtica, Artemisa und Plantago); negative Werte werden in Übereinstimmung mit anderen Studien im Allgemeinen arborealen Taxa zugeschrieben (vgl.
-Behre 1981).
-Da dunkle mesophile Buchen-Tannenwälder als natürliche Vegetation unseres Untersuchungsgebiets angenommen werden können (vgl.
-Rey et al.
-2017), wird eine erhöhte Offenheit der Landschaft als Ausdruck des menschlichen Einflusses auf die natürliche Vegetation interpretiert.
-die natürliche Vegetation widerspiegelt.
-Da jede Probe absolut datiert ist, können die Daten auf der x-Achse gegen den Offenheitswert auf der y-Achse aufgetragen werden, um eine Zeitreihe Zeitreihe für die Landrodung zu erhalten.
+Unfortunately, in the Swiss Plateau there is no data from burials that could be usefully incorporated into such this model.
+Preservation conditions again simply prevent their use in this study.
+Nevertheless, we see a very high potential for other regions in the integration of demographic indicators from burial data in order to enlarge the canon of methods and the range of proxies.
 
-<!--# Ende cp aus dem Cambridge Artikel -->
+The values of the sum calibration, the openness index and the dendrodated settlements were smoothed by means of a moving average with a window of 50 years.
+Since the aoristic sum already had a very coarse temporal resolution, this was not applied for this measure.
+The range of the smoothing window corresponded to the sample interval, with which a unified resolution of 50 years was obtained for all proxies as time slices for the model.
+In addition, all data were restricted to a window of observation of 7000-1000 BCE.
+The upper limit is set by the post-glacial changes in the pollen spectra, which are hardly associated with human influence before 7000 in the working area, and which would consequently distort the openness indicator.
+The lower limit is defined by the Hallstatt Plateau (with buffer), which would have direct and indirect research influences on the ^14^C proxy.
 
-$$Abbildung Offenheitsanzeiger$$
+## Observational Model
 
-Leider fallen im Bereich des Schweizer Mittelland es Daten aus Bestattungen weg, die sich auch nutzbringend in ein solches Modell einbauen lassen würden.
-Hier sind es auch wieder schlicht Überlieferungsbedingungen, die eine Nutzung in dieser Untersuchung verwähren.
-Dennoch sehen wir für andere Regionen in der Integration von demographischen Indikatoren Ausgestaltungsdaten ein sehr hohes Potenzial, um den Methodenkanon und die Proxydichte zu erweitern.
+During the development of the overall model, we abandoned the implementation of dedicated observation models adapted to the conditions of the individual proxies and their generating processes.
+In previous implementations, which we have also presented at various conferences, the underdetermination by the currently usable data of the model with many degrees of freedom led to equifinality of the solutions and, thus, to a high path dependency of the individual model runs.
+Therefore, it was almost impossible to achieve convergence of the overall model.
+Nevertheless, we believe that for a future application of the model with more data, a larger geographical coverage and especially a regionalised approach with information transfer by means of partial pooling, this more specific approach will be feasible and a very useful approach.
 
-Letzt endlich haben wir uns auch dazu entschlossen, Expertenschätzungen in das Modell zu integrieren.
-These expert estimates are certainly very valuable, as they represent a quantification of - in the strict sense - hardly quantifiable holistic assessments and the intuitive combination of different variables.
-However, as they are very subjectively generated, they must be considered as relatively uncertain estimates and must at least be contrasted with each other and with independent data in order to achieve a minimum degree of verifiability and intersubjectively comprehensible methodology in data evaluation.
-Für unser Arbeitsgebiet gibt es eine begrenzte Anzahl von Expertenschätzungen, die verschiedene Zeit Epochen umfassen, und die sich deutlich in ihren Aussagen widersprechen.
-
-$$Tabelle Expertenschätzungen$$
-
-## Observational Models
-
-Für die einzelnen Datensätze existieren unabhängige Beobachtungsmodelle, die die unbeobachtbare latente Variable der Bevölkerungsdichte mit ihrem Effekt auf die beobachteten Proxies selbst bzw.
-spezifika darstellen.
-Dies geschieht einerseits in Hinsicht darauf, wie sich die hinter den Proxies liegenden Eigenschaften der Landschaft mit Bevölkerungsdichte und ihren Auswirkungen verbinden lassen, anderseits aber auch in Hinsicht darauf, wie die Beobachteten Werte selbst zu diesen Prozessen in Beziehung stehen (z.B.
-Überlieferungs- und Auffindungswahrscheinlichkeiten).
-Die implementierung der einzelnen Modelle ist teilweise sehr unterschiedlich, in Hinsicht auf die Zahl bekannter Archäologischer Fundstellen, welche hinter verschiedenen Proxies stehen, wiederum strukturell sehr ähnlich.
-
-### Expert Estimations
-
-Das einfachste Modell ergibt sich hinsichtlich der Expertenschätzungen.
-Bei diesen gehen wir davon aus, das zeitabhängig die Expertenschätzungen normalverteilt um den realen Wert der Bevölkerungsdichte schwanken.
-Die eigentlichen Werte der Schätzungen sind im gleichen Mass wie die zu schätzende versteckte latente Bevölkerungsdichte, die es zu schätzen gilt.
-Daher ist eine explizite Modellierung einer Transferfunktion nicht nötig, bzw.
-es ist kein eigentlicher "Messfehler" vorhanden.
-
-Da die Schätzungen im Allgemeinen einen konstanten Wert darstellen, welcher für eine bestimmte Zeitspanne gegeben ist, und der reale Wert der Bevölkerungsdichte in jedem Zeitschritt ein anderer sein dürfte, kann hier ein partielles Pooling nicht durchgeführt werden.
-Vielmehr ist der Fehler der Schätzung in jedem Zeitabschnitt ein anderer, und zwischen den einzelnen Experten kann auch keine Übertragung der Information stattfinden, da davon auszugehen ist, dass jeder Experte in unterschiedlicher Weise den realen Wert über- oder unterschätzt.
-Daraus ergibt sich eine experten- und Zeitschrittabhängige Formulierung des Verhältnisses von der Expertenschätzung zu den im Modell zu schätzenden Bevölkerungdichten.
+The implementation we introduce in this paper represents a Poisson regression in which the indicators used inform the change in the number of settlements from time step to time step.
+For this purpose, the individual proxies were z-normalised.
+The absolute differences from one time step to another were then computed from the resulting time series.
+Thus, if the value of the proxy increases, this results in a positive difference from the previous time step, and vice versa.
 
 $$
-Z_{i, t} \sim N(N_t, \sigma_{i, t}) \\
-\sigma_{i, t} \sim N(N_t, 1/N_t^2)
+\begin{aligned}
+z_t &= \frac{x_t - \bar{x}} {\sigma_x}\ |\ \sigma_x := Standard\ Deviation  \\
+\delta z{_t} &= z_t - z_{t-1}
+\end{aligned}
 $$
 
-$N_t$ stellt dabei die zum Zeitpunkt t geschätze Bevölkerungsgrösse dar.
-Dabei ist davon auszugehen, dass die Standardabweichung der Schätzung ($\sigma_{i,t}$) abhängig ist von der absolute Grösse des Schätzwertes sowie unabhängig je Experte.
-Diese wird dementsprechend aufgrund des im Modell vorhandenen Schätzwertes errechnet.
-
-### Openness Indicator
-
-Für den Offenheitsindikator haben wir die kombinierten überregionalen Werte, die sich aus den genutzten Pollendiagrammen bzw.
-deren zeitspezifischen Werten auf der ersten Dimension der PCA.
-Für diesen Wert nehmen wir einen Messfehler an, das Verhältnis zwischen den Beobachten Werten und der tatsächlichen Landöffnung der Umwelt zeitabhängig darstellt.
-Der Beobachtete Wert der Landöffung $V_t$ schwankt dabei normalverteilt um den tatsächlichen Wert $I_t$, wobei wir annehmen, dass hier eine zeitunabhängige Schätzungenauigkeit $\sigma_v$ besteht, über die ein partielles Pooling stattfinden kann.
-Insgesamt ergibt sich für diesen ersten Teil folgende Formel:
+The sum of the resulting differences between the time steps, together with the settlement number of the previous step as the expected value, then forms $\lambda_t$ as the expected value for the settlement number of the current time step.
 
 $$
-V_t \sim N(I_t, \sigma_v) \\
-\sigma_v \sim \Gamma(0.5, 0.5)
+log(\lambda_t) = log(N_{t-1}) + \sum_{i=1}^n \beta_i \delta z_{i, t}
 $$
 
-Somit ist $\sigma_v$ auf die gleiche Weise spezifiziert wie $\sigma_\lambda$ des Prozessmodells: eine Abweichung muss positiv sein, werte grösser als 1 sind eher unwahrscheinlich.
-
-In weiteren Hierarchieschritten wird nun die Beziehung der Landöffnung zur Menschlichen Besiedlung modelliert.
-Hierbei gehen wir davon aus, dass die Landöffnung linear zur Grösse der Bevölkerung innerhalb eines Zeitschrittes skaliert, das aber zwischen einzelnen Zeitschritten (durch technische Innovationen oder Veränderungen in den naturräumlichen Gegebenheiten oder kulturellen Präferenzen) eine unterschiedliche Intensität menschlicher Aktivitäten auf die Landöffnung möglich sind ($B_t$), wobei wir insgesamt gepooled davon ausgehen, das dieser Wert normalverteilt um einen Mittelwert schankt.
-Hinzu kommt eine Grundsätzliche Landöffnung, die wir zeitunabhängig annehmen, und die die natürlichen Landöffnungsprozesse (Feuer durch Blitzschlag u.ä., $BaseOpen$) wiederspiegelt.
-
-Insgesamt ergibt sich daraus folgendes Modell:
+Here, $\beta_i$ is a scaling factor that represents the influence of the respective proxy.
+It is a confidence value of the model for the respective proxy, so that the sum of all $\beta_i$ results in 1.
 
 $$
-I_t = N_t * B_t + BaseOpen \\
-BaseOpen \sim N(0,1000) \\
-B_t \sim N(\mu_b, \sigma_b) \\
-\mu_b \sim N(0.5,1) \\
-\sigma_b \sim \Gamma(0.5,0.5)
+\sum_{i=1}^n \beta_i = 1
 $$
 
-Für $\sigma_b$ ist wiederum eine Gamma-Verteilung mit Fokus auf positive Werte kleiner als 1 angenommen, für den Mittelwert $\mu_b$ nehmen wir eine breite Normalverteilung um 0.5 an, was bedeutet, das (in Zusammenspiel mit $BaseOpen$) darauf hinausläuft, dass wir im Mittel einen positiven Einfluss von menschlicher Bevölkerung auf Landöffnung annehmen, aber viel Spielraum für das Modell zur spezifikation des Wertes lassen (weakly informative prior).
+A probability distribution that can be used for this purpose in a hierarchical Bayesian model is the Dirichlet distribution, which is a multivariate generalization of the beta distribution, commonly used as prior distributions in Bayesian statistics.
+Its density function gives the probabilities of $i$ different exclusive events.
+It has a parameter vector $\alpha = (\alpha_1, ..., \alpha_i)\ |\ (\alpha_1, ..., \alpha_i) > 0$, for which we have chosen a weakly informative log-normal prior.
+The priors for the log-normal distribution in turn come from a weakly informative exponential distribution for the mean and a log-nomal distribution with $\mu$ of 1 and $\sigma_{log}$ of 0.1:
 
-### Aoristic Analysis
+$$
+\begin{aligned}
+\beta_i &\sim Dir(\alpha_{1-i}) \\
+\alpha_i &\sim LogNormal(\mu_{alpha_i}, \sigma_{alpha_i}) \\
+\mu_{alpha_i} &\sim Exp(1) \\
+\sigma_{alpha_i} &\sim LogNormal(1,0.1)
+\end{aligned}
+$$
+
+As an intuition, this means that we consider the sum of the proxies as determinant for the number of settlements.
+The estimation therefore assumes that all proxies together give the best possible estimation result, whereby the share of each individual proxy is considered variable and is estimated within the model.
+The error value is represented by the Poisson process in the process model, rather than directly as an estimation error for the individual proxies.
+Thus, our model does not correspond to a classical state space model, where the measured values are each considered to be error-prone.
+In the implementation, the model finds the best possible combination or compromise between the individual indicators to describe a settlement dynamic that is given by them.
+
+```{r expertestimations, fig.cap="Expert estimations for the population density on the Swiss Plateau from different authors."}
+expert_estimations <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "raw_data", "estimates.csv"))
+
+ggplot(expert_estimations) + geom_segment(aes(x=from, xend = to, y = estimate, yend = estimate, color = source)) + xlab("cal BCE") + ylab("Density p/km²") + theme_minimal()
+```
+
+In earlier implementations, expert estimates were also integrated into the model.
+However, since these are highly contradictory for the working area (Figure \@ref(fig:expertestimations), therefore have little influence on the model, and also lead to a significantly longer runtime, we have refrained from doing so in the current implementation.
+For future applications of the model with a larger geographical range and thus a higher information density, however, this would again be a factor that could be integrated in a useful way.
+
+## Model fitting
+
+The model was fitted using the R package nimble (version 0.11.1, R version 4.1.3).
+For this purpose, 4 chains were run in parallel.
+Achieving and ensuring convergence and sufficient effective samples (10000) for a reliable assessment of the highest posterior density interval was carried out in steps.
+
+In a first run, the model was initialised for each chain and run for 100000 iterations (with a thinning of 10).
+The computer used (Linux, Intel(R) Xeon(R) CPU E3-1240 v5 \@ 3.50GHz, 4 cores, 8 threads) needed approx.
+1 min for this.
+
+In a second step, the run was extended until convergence could be determined using Gelman and Rubin's convergence diagnostic, the criterion being that a potential scale reduction factor of less than 1.1 was achieved for all monitored variables.
+Convergence occurred after about 30 seconds.
+
+Due to the high correlation of the parameters and thus a low sampling efficiency, the collection of at least 10,000 effective samples for all parameters in the third step took about 5 hours.
+
+For the fitting process, a starting value for a population density of the Late Bronze Age (1000 BCE) was taken from the literature, which may represent a general average value for all prehistoric population estimates [@nikulka2016, 258].
+For the model, this was set as the mean of a normal distribution with a standard deviation of 0.5, which should give enough leeway for deviations resulting from the data.
+Nevertheless, especially the late part of the reconstruction is of course clearly influenced by this predefined value.
+
+For the traceplots and the prior-posterior overlap as well as the density functions of the posterior samples of the individual parameters, please refer to the supplementary material.
 
 # Results
+
+Within the model, the parameter p was estimated, which reflects the proportion of the individual proxies used in the estimation of the number of settlements, as well as the number of settlements or transforms the resulting population density under the assumption that per settlement a number of inhabitants lived poisson-distributed around a $\lambda$ of 50 persons.
+This parameter is variable, but has only a scaling influence on the final estimate of population density.
+
+```{r pdistribution, fig.cap="Distribution of the influence ratio of the different proxies on the final estimation of the number of sites."}
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "p_estimation.pdf"))
+```
+
+By looking at the distribution of posterior samples for the share of each proxy (Figure \@ref(fig:pdistribution)), it is clear that the model weights the openness indicator the highest.
+The average is slightly above 60%.
+The next most important indicator is the sum calibration value, which has an average of about 20%.
+The aoristic sum is slightly above 10%, whereas the importance of the dendrodated settlements is below 10%.
+
+The reason for the latter is certainly that there are no lakeshore settlements over large areas of the time window, and therefore the proxy achieves a low confidence value in comparison with the other estimators.
+In the case of the aoristic sum, it is certainly the fact that it is flat over large sections and has little structure, making it difficult to relate to the other estimators.
+The sum calibration shows very strong short-term fluctuations, which are presumably at least partly due to the effects of the calibration curve, and which also make this proxy seem ill-suited to reliably represent a continuous population trend.
+Nevertheless, its fluctuations do have an impact on the resulting overall estimate of the development of the number of settlements, albeit to a lesser extent.
+
+```{r popdensplot, fig.cap="The estimate of the population density resulting from the model and the three proxies, which are also plotted (scaled) for comparison."}
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popdens_estimation.pdf"))
+```
+
+The population density estimate (Figure \@ref(fig:popdensplot)) ranges on average between 0.2 p/km² for the beginning of the estimate (6000 BCE) and 4.8 p/km² for the end of the estimate (1000 BCE), reaching a maximum of 6.5 p/km² for the time slice 1250 BCE.
+Thus, the estimate remains within the values that are also considered plausible by the expert estimates.
+Clear peaks are reached around 1250 BCE (also the maximum value), as well as around 2750 BCE, which corresponds to the onset of the Corded Ware in the Swiss Plateau.
+However, we will go into further detail in the discussion on evaluations concerning the connections to cultural phenomena.
+
+```{r varcoeffplot, fig.cap="The variability of the estimate of the population density over time, with the estimate itself for reference."}
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popvar_plot.pdf"))
+```
+
+Furthermore, it may be relevant to look at the temporal distribution of the variability in the estimate (Figure \@ref(fig:varcoeffplot)).
+The coefficient of variation is 0.13 for the beginning and 0.1 for the end of the estimate, the greatest variability is reached around 2150 BCE with 0.4679342.
+The beginning and end of the time series are relatively clearly determined.
+The end results from the a priori setting of the parameter, but also here as at the beginning of the series the proxies are very uniform, which explains the low variability.
+Overall, the variability is relatively uniform over the entire course of the estimation and averages over all time slices at 33% of the respective mean.
 

### 2022-04-11 analysis/analysis.Rmd
updated analysis and final first draft paper

@@
 
-Finally, we can plot the result in comparison to the input data, superimposing them (scaled) on the estimation result in their original form rather than as difference data:
-
-```{r}
+Finally, we can plot the result in comparison to the input data, superimposing them (scaled) on the estimation result in their original form rather than as difference data.
+To do this, we first extract mean and HPDI from the posterior samples for all parameters:
+
+```{r}
+
+all_proxies_orig <- read.csv(file = normalizePath(file.path(here(), "data","preprocessed_data", "all_proxies.csv")), row.names = 1)
 
@@
 gc(verbose=F)
-
+```
+
+Then we obtain the population density estimates from these, and save them for further use:
+
+```{r}
 PopDens_summary <- this_mcmc_summary[grep("^PopDens", params),]
-
+PopDens_summary$age <- rev(all_proxies_orig$age)
+write.csv(PopDens_summary,
+          file = normalizePath(
+            file.path(here(), "data","preprocessed_data", "popdens_summary.csv"
+                      ),
+            mustWork = F)
+          )
+```
+
+Finally, compile the plot:
+
+```{r}
 all_proxies_orig <- read.csv(file = normalizePath(file.path(here(), "data","preprocessed_data", "all_proxies.csv")), row.names = 1)
@@
   geom_line(data = PopDens_summary,
-            aes(x = rev(1950-all_proxies_orig$age), y=mean, col = "Estimation Result")) + 
+            aes(x = 1950-age, y=mean, col = "Estimation Result")) + 
   geom_line(data = PopDens_summary,
-            aes(x = rev(1950-all_proxies_orig$age), y=hpdi_l, col = "Estimation Result"), lty=2) + 
+            aes(x = 1950-age, y=hpdi_l, col = "Estimation Result"), lty=2) + 
   geom_line(data = PopDens_summary,
-            aes(x = rev(1950-all_proxies_orig$age), y=hpdi_u, col = "Estimation Result"), lty=2) +
+            aes(x = 1950-age, y=hpdi_u, col = "Estimation Result"), lty=2) +
   theme_minimal() + 
@@
 
-And finally, the final estimate of the influence of the different proxies on the result:
-
-```{r}
+And than, the final estimate of the influence of the different proxies on the result:
+
+```{r}
+gc(verbose = F)
+
 p_plot <- all_chains[,grep("^p\\[", params)] %>%
@@
 ```
-
 Also interesting is the variation coefficient from the estimation of the Population density:
@@
 ```{r}
+gc(verbose = F)
+
 var_popdens <- data.frame(var_coeff = apply(all_chains[,2:100], 2, function(x) sd(x)/mean(x)),

### 2022-04-11 manuscript/paper.Rmd
updated analysis and final first draft paper

@@
 output:
-  bookdown::word_document2:
-    fig_caption: yes
-    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+  bookdown::pdf_document2:
+    toc: no
     pandoc_args:
@@
     - --lua-filter=../templates/pagebreak.lua
-  bookdown::pdf_document2:
-    toc: no
+  bookdown::word_document2:
+    fig_caption: yes
+    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
     pandoc_args:
@@
 This is precisely where it becomes very interesting for archaeological data, since they are mostly inaccurate, biased, sparse, simplified and often based on simplistic assumptions.
-  In this contribution, issues and possibilities of the application of Bayesian demography in the field of archaeological (prehistoric) population size and density estimation will be discussed and the results of a pilot study will be presented, which should lead to a more comprehensive project for the reconstruction of demographic developments in prehistory on a large scale.
+  In this contribution, a Bayesian hierarchical model, based on a Poisson Regression and a state space approach is applied to the field of archaeological (prehistoric) population size and density estimation.
+This represents a pilot study for the reconstruction of demographic developments in prehistory based on data assimilation with an underlying process model and a number of different proxies, which are combined and result in a more reliable estimation than a single proxy could deliver.
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  Bayesian demographic modelling; Multiproxy; Reconstruction of past settlement dynamics
 highlights: |
-  These are the highlights. 
+  - We provide a estimation based on the combination of different proxies.
+  - We estimate population development in terms of absolute settlement densities.
+  - We are developing a completely new method for the integration of heterogeneous information.
 ---
@@
 
+library(here)
 library(ggplot2)
@@
 
-$$Beispiele bringen? Bring some examples here...$$
-
 Archaeological data used to estimate population trends have the following characteristics:
@@
 
-Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on 14C data.Most notably are the recently published papers by Crema et al.
-However, the approach taken by Crema et al.
-is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of 14C data only.
-This is a clear step in the direction of a model-based, and thus scientific, analysis.However, on the one hand, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool.
+Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on ^14^C data.Most notably are the recently published papers by Crema and Shoda ([@crema2021a]).
+However, the approach taken there is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only.
+This is a clear step in the direction of a model-based, and thus scientific, analysis.
+However, on the one hand, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool.
 Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series.Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented.
@@
 
-# Conclusion
+The ambition in the development of the model was originally to base it on a state-space representation of the demographic process itself, and then to integrate the existing proxies by means of an observation model to inform this process.
+In the course of developing the model, however, we had to realise that the existing data or the data used in the modelling process for the Swiss Plateau alone were not sufficient to adequately determine this process or to adequately fix the degrees of freedom resulting from the process and the transfer model.
+Therefore, in this implementation we have turned to the Poisson regression approach.
+In this version, the model represents the best possible combination of the indicators used to describe the development of the number of sites based on them.
+
+In its present form, the reconstructed development already shows interesting aspects regarding the underlying proxies.
+Starting with the result of the sum calibration, which is currently the most frequently used indicator when attempting to reconstruct (relative) population changes.
+Its strong fluctuations are dampened by the connection of the other proxies, which is especially true for the first fluctuation around or shortly after 4000 BCE.
+The increase in the archaeological record with the process of Neolithisation is still clearly visible, but the curve of the overall estimate after the initial increase is much flatter than the sum calibration alone would suggest.
+The period between 3950 and 3700 in particular forms a noticeable plateau in the calibration curve, and this may lead to an overestimation of the ^14^C evidence.
+However, the effect of the calibration curve on the results of a cumulative calibration cannot yet be considered unambiguous.
+A second maximum, after 3000 BCE, is supported by the other proxies, and is consequently also much more clearly reflected in the overall estimate.
+Here, too, there is a smaller, albeit shorter plateau in the calibration curve, which is, however, much less pronounced than, for example, the one shortly before between 3350 and 3100.
+The rise towards the Middle and Late Bronze Age also receives support above all from the Aorist Sum, and therefore remains preserved in the combination of indicators.
+In this period, the calibration curve does not show any very clear, suspicious patterns.
+
+Overall, the model trusts the cumulative calibration with an average of about 20% significantly less than the openness indicator with 60%.
+After the initial increase, which can easily be linked to the spread of agricultural production, the latter tends to fluctuate less and thus has a dampening effect on the overall estimate.
+Nevertheless, by all appearances, the general trends in the cumulative calibration are quite well reflected in the land opening, even if such eyeballing should be interpreted with caution.
+We would suggest that the changes within the Neolithic and Bronze Age are rather gradual compared to the much more severe qualitative shifts to a productive economy as well as to the later developments outside our window of observation, which are suggestive of a much more large-scale reshaping of the landscape.
+The model itself is designed rather conservatively by the limit of maximum growth, which is estimated within the model but influences it by its very presence.
+Therefore, this proxy corresponds better and more homogeneously to a smooth increase in the number of settlements than do the strong fluctuations in the ^14^C data set.
+
+The aoristic sum remains relatively even over large areas of the observation window.
+It is not until the Middle and Late Bronze Age that their values rise significantly, which is also noticeable in the overall estimate by a clear increase.
+Overall, it remains to be seen to what extent a modelling of the taphonomic loss [@surovell2009] could be integrated.
+We have refrained from doing so in this first model, as this would have introduced further degrees of freedom - but we are aware that with a broader database this would be an interesting possibility, and that it would itself be a variable to be estimated, e.g.
+in connection with proxies that are not influenced by it (openness indicators, but also data from the demography of burial collectives).
+This would make it possible to estimate a value from original archaeological material independently of variables that have little to do with archaeological data, such as volcanic eruptions.
+
+The number of simultaneously existing lakeshore settlements is a very limited temporal and spatial estimator, but extremely reliable.
+Its limitations are reflected in the low overall confidence of the model, since its value is 0 over long stretches, while other indicators suggest clearly different patterns.
+However, where it has information potential, such as around and shortly after 3800 BCE, 3200 BCE or especially around 2750 BCE, its fluctuations have a noticeable influence on the overall estimate.
+The peak around 1600 BCE also leaves a noticeable impact.
+This highlights another potential of our approach: where a proxy has little structure and thus little significance, or where its trends cannot be linked to other indicators, it consequently has little influence.
+For periods in which it can provide information, however, this will also feed into the overall model, despite a low overall confidence in the estimator.
+
+In order to review the reconstruction against the background of established archaeological knowledge and narratives, we consider it useful to parallelise the archaeological phase boundaries with the changes in the estimate.
+It should be noted, however, that the estimate is not completely independent of the archaeological phases: due to the aoristic sum, which is itself strongly determined by this phase division, corresponding boundaries also influence the structure of the reconstruction.
+Nevertheless, it helps to check whether the estimate is in strong contradiction to the generally accepted picture or whether it is able to make a credible prediction within this context.
+
+```{r popestwithphases}
+PopDens_summary <- read.csv(
+  file = normalizePath(
+            file.path(here(), "data","preprocessed_data", "popdens_summary.csv"
+                      )
+            )
+  )
+
+chronology <- read.csv(
+  file = normalizePath(
+            file.path(here(), "data","raw_data", "chrono_rough.csv"
+                      )
+            )
+  )
+chronology <- chronology[order(chronology$dat_begin),]
+chronology$label <- factor(chronology$label, levels = chronology$label)
+
+ggplot(PopDens_summary) +
+  geom_line(aes(x = 1950-age, y = mean)) + 
+  geom_line(aes(x = 1950-age, y = hpdi_l), lty=2) + 
+  geom_line(aes(x = 1950-age, y = hpdi_u), lty=2) + 
+    theme_minimal() + 
+ylab("Scaled values for Proxies,\n P/km² for Estimation") +
+  xlab("cal BCE") +
+  geom_rect(data=chronology[-1,],aes(xmin=dat_begin,ymin=0,xmax=dat_end,ymax=Inf,fill=label),
+                    alpha=0.5,inherit.aes=FALSE) +
+    scale_fill_brewer(palette="Pastel1", guide = "none") +
+  geom_text(data=chronology[-1,],
+            aes(x=dat_end - 100,y = 0.25,label=label),
+            alpha=0.25,inherit.aes=FALSE, angle = 90, hjust = 0)
+```
+
+Figure \@ref(fig:popestwithphases) underlays the estimated development with the rough archaeological phases for the Swiss Plateau.
+The Older and Middle Neolithic phases are in fact hardly documented with known sites in Switzerland.
+Here we must assume a basically low level of settlement, probably mainly by hunter-gatherer groups.
+Isolated Neolithic sites of the LBK and later groups are known in the periphery of Switzerland, but they play a subordinate role.
+The Neolithic really began here in the so-called Upper Neolithic, connected with the cultural phenomena of Egolszwil and Cortaillod respectively Pfyn, as well as with the first lake shore settlements.
+In this period, we also see a clear increase in the estimated population in the model.
+In the transition to the Late Neolithic we know from the lakeshore settlements the so-called Horgen Gap.
+This is also visible as a slight decrease in the model.
+In another study [@heitz2021] we could show that this is in fact probably not a decline in population but rather a shift in settlement strategy.
+In the Late Neolithic, associated with the Horgen pottery, we then see a clear increase in the intensity of settlement, which reaches its peak and its break-off at the transition to the Corded Ware and thus to the Final Neolithic.
+In the second half of the Early Bronze Age, from which we again know lakeshore settlements, there is again a clear increase in population size according to the model, which continues until the Late Bronze Age.
+All in all, the estimate of the model corresponds to our expectations, although we must be aware that our expectations are not ground truth.
+Nevertheless, it can be said that the predictions of the model can be well integrated into the known narratives and make them more precise or provide them with a higher resolution, thus gaining credibility for the model predictions.
+
+# Conclusion and Outlook
+
+With this model and its predictions, we believe we can demonstrate how it is possible to achieve a true multiproxy analysis of prehistoric demographic processes.
+Previous attempts in this regard have so far only achieved a juxtaposition of different indicators, without the possibility of truly linking them or assessing the confidence in the individual indicators.
+Also, in previous approaches it was not possible to specify a confidence interval for the estimates.
+Within a Bayesian framework with an observational model based on Poisson regression, all these demands can be met.
+We thus believe we can provide a better basis than before for estimating population processes and changes in the number of settlements.
+
+The currently presented model is only the first step towards a more sophisticated approach.
+So far, we have trusted the individual indicators in their aggregation without modelling them with a measurement error, which they undoubtedly inherit.
+Moreover, our estimates are based on a limited number of indicators, almost all of which are subject to preservation conditions except for the openness indicator.
+Consequently, a transformation of the results into the range of absolute numbers for population density can only be achieved by means of predefined parameters for upper limit of population growth, settlement size and the initial value of the reconstruction.
+In the future application and refinement of the model, this must be overcome.
+For this purpose, further proxies independent of the immediate temporally changing conditions of the archaeological record have to be integrated.
+These can be data on settlement sizes, parameters for economic-ecological carrying capacity, demographic data from burial groups as well as palaeogenetic data on population sizes.
+Other possible indicators can still be identified.
+
+These data are available to varying degrees in different regions.
+In the Swiss Plateau, for example, there are too few data on human remains over large areas of prehistory to be able to integrate them meaningfully into a model.
+By means of large-scale modelling, however, it would be possible to supplement gaps in the data of individual regions with regions in which these data are available by means of regionalisation and partial transfer of information (partial pooling).
+Such an extension would be the next logical step in the improvement of the model, to which we hope to be able to contribute a further study in the near future.
 
 # Acknowledgements
+
+Large parts of the data collection were conducted as part of the project 'Beyond lake settlements' in the doctoral thesis of Julian Laabs, funded by the SNF (project number 152862, PI Albert Hafner).
+Further data collection was done as part of the XRONOS project, also funded by the SNSF (project number 198153, PI Martin Hinz).
+The development of the openness index took place within the framework of the project Time and Temporality in Archaeology (project number 194326, PI Caroline Heitz) and was inspired by the cooperation within the project QuantHum (project 169371, PI Marco Conedera), both also funded by the SNSF.
+We also thank the Institute of Archaeological Sciences of the University of Bern for its support and faith in the outcome of our modelling project.
+Finally, we thank (already) the unknown reviewers for their helpful comments, which will certainly improve this manuscript significantly.
 

### 2022-04-13 analysis/analysis.Rmd
implemented Julians suggestions and corrections

@@
 
-Running through the analysis can take a long time, especially the Bayesian model. We point out the long runtimes again at the appropriate place.
+Running through the analysis can take a long time, especially the Bayesian model.
+We point out the long runtimes again at the appropriate place.
+Also, you might need at least 64 GB of RAM to avoid crashes due to missing memory.
 
@@
 popdens_plot <- ggplot(all_proxies_orig) +
-  geom_line(aes(x = 1950-age, y = scale(openness), col="Openness Indicator"), alpha = 0.5) +
-  geom_line(aes(x = 1950-age, y = scale(sumcal), col="Sum Calibration"), alpha = 0.5) +
-  geom_line(aes(x = 1950-age, y = scale(dendro), col="Dendro Dated Settlements"), alpha = 0.5) +
-  geom_line(aes(x = 1950-age, y = scale(aoristic_sum), col="Aoristic Sum"), alpha = 0.5) +
+  geom_line(aes(x = 1950-age, y = scale(openness), col="Openness Indicator")) +
+  geom_line(aes(x = 1950-age, y = scale(sumcal), col="Sum Calibration")) +
+  geom_line(aes(x = 1950-age, y = scale(dendro), col="Dendro Dated Settlements")) +
+  geom_line(aes(x = 1950-age, y = scale(aoristic_sum), col="Aoristic Sum")) +
   geom_line(data = PopDens_summary,
@@
   ggplot(aes(x = value, y = name)) +
-  stat_halfeye() + xlab("Percentage Influence") + ylab("Proxy")
+  stat_halfeye() + xlab("Percentage Influence") + ylab("Proxy") + theme_minimal()
 p_plot

### 2022-04-13 manuscript/paper.Rmd
implemented Julians suggestions and corrections

@@
 output:
-  bookdown::pdf_document2:
-    toc: no
+  bookdown::word_document2:
+    fig_caption: yes
+    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
     pandoc_args:
@@
     - --lua-filter=../templates/pagebreak.lua
-  bookdown::word_document2:
-    fig_caption: yes
-    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+  bookdown::pdf_document2:
+    toc: no
     pandoc_args:
@@
 
-After research interest turned increasingly to other aspects with the emergence of post-process trends, a strong boom in the study of population sizes can be observed especially since the 2010s (or even earlier, [@shennan_population_2000]).
-This is probably due, on the one hand, to the new shift towards the study of human-environment relationships, for which an assessment of the size of human populations is indispensable.
+After research interest turned increasingly to other aspects with the emergence of post-process trends, a strong boom in the study of population sizes can be observed especially since the 2010s [or even earlier, @shennan_population_2000].
+This is probably not at least due to the new shift towards the study of human-environment relationships, for which an assessment of the size of human populations is indispensable.
 This is true especially for an evaluation of the human impact on the natural environment.
@@
 1.
-Proxy is based on a single line of evidence.
+Reliance on single line of evidence: Most investigations are using only one proxy.
 Although multiproxy approaches exist, in such cases the individual proxies only serve to support each other or the proxy used in the main case.
 There is no combination of results to increase the accuracy of each of the inaccurate and biased proxies.
-2.  The uncertainty of evidence is not adequately reflected. In most cases, individual curves are presented as estimates, but the size of the error bars for these estimates is almost never apparent.
+2.
+Uncertainty in measurements: In most cases, individual curves are presented as estimates, but the size of the error bars for these estimates is almost never apparent.
+This means, that the uncertainty of evidence and derived quantities is not adequately reflected.
 3.
-Nature of the archaeological data is taken into account only insufficiently.
-What is meant here is that they take little account of the inherent uncertainties of the proxies themselves, for lack of an evaluative framework, and that transfer functions from the proxy to the actual size to be observed, i.e.
+Lack of transfer functions: What is meant here is that they take little account of the inherent uncertainties of the proxies themselves, for lack of an evaluative framework, and that transfer functions from the proxy to the actual size to be observed, i.e.
 population size or density, are hardly ever evaluated more intensively.
+That means, that the nature of the archaeological data is taken into account only insufficiently.
 In the best case, a critical appraisal of the informative value is provided, but a quantification of the relationship between, for example, a doubling of a proxy value and the change in population density does not take place and cannot take place for lack of a suitable framework or external data for calibration.
 
-Archaeological data used to estimate population trends have the following characteristics:
+Archaeological data used to estimate population trends have the following characteristics, leading to these problems:
 
@@
 Preservation conditions, depositional behavior or social aspects should be mentioned here as an example.
--   Unreliable: Comprehensive information and derived data are influenced by research strategies. Systematic distortions are possible or rather the rule.
--   Highly heterogeneous: The number of artefacts at a site, the number of houses, the demographic conditions in cemeteries, the estimates of human impact based on pollen analyses: All these are data that may serve our purposes, but which are available in completely different spatio-temporal scales, granularities and information values, and simply represent technically very different data formats and scales.
+-   Unreliable: Comprehensive information and derived data are influenced by research strategies, research history and financial capability of the individual research projects.
+Systematic distortions are possible or rather the rule.
+-   Highly heterogeneous: The number of artefacts at a site, the number of houses, the demographic conditions in cemeteries, the estimates of human impact based on pollen analyses: All these are data that may serve our purposes, are available in completely different spatio-temporal scales, granularities and information values, and simply represent technically very different data formats and scales.
 -   Indirect: It is actually always proxy data that are collected as substitute values for the information that is actually needed (i.e.
@@
 
-Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on ^14^C data.Most notably are the recently published papers by Crema and Shoda ([@crema2021a]).
+Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on ^14^C data.
+Most notably are the recently published papers by Crema and Shoda ([@crema2021a]).
 However, the approach taken there is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only.
 This is a clear step in the direction of a model-based, and thus scientific, analysis.
-However, on the one hand, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool.
-Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series.Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented.
+However, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool.
+Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series.
+Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented.
 Therefore, we would like to better exploit the capabilities of Bayesian hierarchical models through a combination of inductive data analysis and model-passed data integration of different proxies.
@@
 
-# Background
+# Materials and Methods
+
+## Background
 
@@
 It is also in the nature of the approach that in real applications there is no point prediction, but in most cases a probability distribution for the prediction.
-Thus we simultaneously obtained a result and an estimate of the confidence intervals, or better, the credibility interval given the data.
+Thus, we simultaneously obtained a result and an estimate of the confidence intervals, or better, the credibility interval given the data.
 
@@
 ^14^C dates and stratigraphy are something completely different, but both can be mapped to the probability of older and younger and combined in this way.
-The same is also conceivable (and feasible) when it comes to the probability of population sizes or population densities or their derived dynamics (increase, decrease).
+The same is also feasible when it comes to the probability of population sizes or population densities or their derived dynamics.
 
@@
 In our example, these are the openness indicators from pollen data, which we can interpret primarily in terms of human influence and its intensity.
-On a more abstract level, we can also include expert estimates, as these also happen on (often unspecified) bases that are at least indirectly influenced by past population density.
-The specifics of the individual data we used in this modelling are discussed in the section \@ref(dataproxies).
-
-# Materials and Methods
-
-## Process Model
-
-The overall model for the estimation of demographic developments is broken down into several hierarchically interconnected individual elements in accordance with the basic structure of a state space model.
-The basis is a process model that represents the demographic development itself in terms of a structure model, without this model already being explicitly configured with data.
-
-In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods.
-In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.
-This results in the following very simple formula:
-
-$$
-N_t = N_{t-1} * \lambda_t
-$$
-
-A distribution that is particularly suitable for modelling frequencies is the Poisson distribution.
-It is a univariate discrete probability distribution that can be used to model the number of events that occur independently of each other at a constant mean rate in a fixed time interval or spatial area.
-It is determined by a real parameter $\lambda$ \>0, which describes the expected value and simultaneously the variance of the distribution.
-Thus, the relationship shown above can also be rearranged as follows:
-
-$$
-\begin{aligned}
-N_t &\sim dpois(\lambda_t) \\
-\lambda_t &= N_t
-\end{aligned}
-$$If we now have information about the change in population development, we can use this to enter it into the model via a change in $\lambda$.
-This is done in the form of a regression: If $x \in R^n$ is a vector of independent variables, then the model takes the form
-
-$$
-log({E} (Y\mid x))=\alpha + \beta' x
-$$
-
-In the case of population trends, the expected value would be equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables:
-
-$$
-log(\lambda_t) = log(N_{t-1}) + \sum_{i=1}^n \beta_i x_{t,i}
-$$
-
-The values for population size $N_t$ as well as for population change $\lambda_t$ are time-dependent.
-At each individual point in time in the time series, these variables can or will also take on different values.
-However, we can narrow down the structure of population change even further.
-We can assume that, considered overall over time, the population change the population change will not exceed certain limits over the entire time span, without it being possible here to specify this already.
-
-Thus, we can define the limits, the $max\_change\_rate$ as an time independent variable, again without already specifying them with fixed values at this stage.
-The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively.
-Overall, this represents a hierarchical model that can be noted as follows:
-
-$$
-\begin{aligned}
-max\_growth\_rate &\sim dgamma(shape = 5, scale=0.05) \\
-N_t/N_{t-1} &< (max\_growth\_rate + 1) \\
-N_{t-1}/N_t &< (max\_growth\_rate + 1)
-\end{aligned}
-$$
-
-The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$ (See also Figure \@ref(fig:gammadist)).
-
-```{r gammadist, fig.cap="The gamma-distributed maximum growth rate prior of the model."}
-plot(density(rgamma(100000,shape=5,scale = 0.05)+1), xlim=c(0,5), main ="gamma(shape = 5, scale=0.05) + 1" )
-```
-
-This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model.
-The interaction of these parameters results in the following prior probability distribution for $\lambda$ and thus the growth (or change) rate of the population:
+On a more abstract level, we could also include expert estimates, as these also happen on (often unspecified) bases that are at least indirectly influenced by past population density.
+The specifics of the individual data we used in this modelling are discussed in the next section.
 
@@
 This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.
-
-Since many periods of the Swiss prehistoric dry land settlements are unknown, and archaeology is based primarily on the wet land settlements, the dendro data are used primarily here, since they are much more precise.
-As a result, ^14^C data could be largely absent in this time period.
-If the amount of ^14^C data were now evaluated directly and naively over time, it would show that in the phases of particularly active settlement of the lakeshore, the ^14^C sum calibration shows minima.
 
@@
 Much more, these data provide an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by the same transmission filters and archaeological conditions as the evaluation of ^14^C data.
-Data from x sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy in archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.
+Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy in archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.
 
-```{r aoristcurve, fig.cap="Comparison of the scaled number of ^14^C and dendrodated sites over time in the dataset used."}
+```{r aoristcurve, fig.cap="Aoristic sum of the archaeological sites used in this analysis."}
 ggplot(joined_data) + geom_line(aes(x=1950-age, y=aoristic_sum, color = "darkgray"), alpha = 0.75) +
@@
 
+In our following modelling (see also \@ref(observational-model)) we now consider all these proxies as determinants for the change in the number of settlements.
+In the model, we therefore change the causality and shift the measurement error, which is certainly inherent in each of these indicators, to the process model, in which a Poisson process describes the number of simultaneous settlements.
+In doing so, we establish a likelihood that indicates how credible the data are, given the model.
+
+## Process Model
+
+The overall model for the estimation of demographic developments is broken down into several hierarchically interconnected individual elements in accordance with the basic structure of a state space model.
+The basis is a process model that represents the demographic development itself in terms of a structure model, without this model already being explicitly configured with data.
+
+In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods.
+In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.
+This results in the following very simple formula:
+
+$$
+N_t = N_{t-1} * \lambda_t
+$$
+
+A distribution that is particularly suitable for modelling frequencies is the Poisson distribution.
+It is a univariate discrete probability distribution that can be used to model the number of events that occur independently of each other at a constant mean rate in a fixed time interval or spatial area.
+It is determined by a real parameter $\lambda$ \>0, which describes the expected value and simultaneously the variance of the distribution.
+Thus, the relationship shown above can also be rearranged as follows:
+
+$$
+\begin{aligned}
+N_t &\sim dpois(\lambda_t) \\
+\lambda_t &= N_t
+\end{aligned}
+$$ If we now have information about the change in population development (the proxies), we can use this to enter it into the model via a change in $\lambda$.
+This is done in the form of a regression: For all proxy values as a vector of independent variables $x \in R^n$, $R^n$ as an n-dimensional Euclidean space, described in this case by the n dimensions of the n variables, then the model takes the form
+
+$$
+log({E} (Y\mid x))=\alpha + \beta' x
+$$
+
+Logarithmisation as a link function ensures that $\lambda$, which must always be positive for a Poisson process, can also be described by variables (proxies) that range in the space of real numbers and can therefore also be negative.
+$\beta$ can serve here as a slope factor, just as in a normal linear regression.
+In our case, it functions as a scaling factor for the individual proxies.
+$\alpha$, on the other hand, is to be understood as an intercept.
+If there were no change due to the variables, the regression would fall back to this value.
+This corresponds to the desired behaviour.
+In the case of population trends, the intercept would be equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables.
+If there is no change from these variables, then $\lambda$, and thus the expected value for the current time step would be equal to that from the previous time step:
+
+$$
+log(\lambda_t) = log(N_{t-1}) + \sum_{i=1}^n \beta_i x_{t,i}
+$$
+
+Since $\lambda$ and $N$ are essentially in the same range (e.g.
+if $lambda=1$, the expected value for $N$ would also be 1), $N_{t-1}$ must also be logarithmised in the above formula to obtain the congruence of both values.
+The values for population size $N_t$ as well as for population change $\lambda_t$ are time-dependent.
+At each individual point in time in the time series, these variables can or will also take on different values.
+However, we can narrow down the structure of population change even further.
+We can assume that, considered overall over time, the population change the population change will not exceed certain limits over the entire time span, without it being possible here to specify this already.
+
+Thus, we can define the limits, the $max\_change\_rate$ as an time independent variable, again without already specifying them with fixed values at this stage.
+The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively.
+Overall, this represents a hierarchical model that can be noted as follows:
+
+$$
+\begin{aligned}
+max\_growth\_rate &\sim dgamma(shape = 5, scale=0.05) \\
+N_t/N_{t-1} &< (max\_growth\_rate + 1) \\
+N_{t-1}/N_t &< (max\_growth\_rate + 1)
+\end{aligned}
+$$
+
+The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$ (See also Figure \@ref(fig:gammadist)).
+
+```{r gammadist, fig.cap="The gamma-distributed maximum growth rate prior of the model."}
+plot(density(rgamma(100000,shape=5,scale = 0.05)+1), xlim=c(0,5), main ="gamma(shape = 5, scale=0.05) + 1" )
+```
+
+This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model.
+The interaction of these parameters results in the following prior probability distribution for $\lambda$ and thus the growth (or change) rate of the population:
+
 ## Observational Model
@@
 During the development of the overall model, we abandoned the implementation of dedicated observation models adapted to the conditions of the individual proxies and their generating processes.
-In previous implementations, which we have also presented at various conferences, the underdetermination by the currently usable data of the model with many degrees of freedom led to equifinality of the solutions and, thus, to a high path dependency of the individual model runs.
+In previous implementations, the underdetermination by the currently usable data of the model with many degrees of freedom led to equifinality of the solutions and, thus, to a high path dependency of the individual model runs.
 Therefore, it was almost impossible to achieve convergence of the overall model.
@@
 As an intuition, this means that we consider the sum of the proxies as determinant for the number of settlements.
-The estimation therefore assumes that all proxies together give the best possible estimation result, whereby the share of each individual proxy is considered variable and is estimated within the model.
+The estimation therefore assumes that all proxies together give the best possible estimation result for contemporaneous sites at time t, whereby the share of each individual proxy is considered variable and is estimated within the model.
+This share is recorded within the model as the parameter p.
+
 The error value is represented by the Poisson process in the process model, rather than directly as an estimation error for the individual proxies.
 Thus, our model does not correspond to a classical state space model, where the measured values are each considered to be error-prone.
-In the implementation, the model finds the best possible combination or compromise between the individual indicators to describe a settlement dynamic that is given by them.
+In the implementation, the model finds the best possible combination or compromise between the individual indicators to describe a settlement dynamic that is given by them.In addition, the number of sites is converted into population density using some (certainly debatable) parameters that we have defined but which are only scaling factors for the intermediate value of number of settlements.
+For this, we assume that each site represents a number of people that is poisson distributed around the value 50.
+The number 50 represents a compromise, as both Mesolithic and Neolithic and Bronze Age settlement communities need to be represented.
+By means of a data series, which would represent an evidence-based estimate of the temporal development of settlement sizes, this specification could be made based on data.
+From the number of sites and the mean number of individuals represented in each case, a population density can be calculated using the area of the Swiss Plateau (12649 km²).
+The estimated result of the model is thus comparable with estimates from other sources or the literature.
 
@@
 
-For the fitting process, a starting value for a population density of the Late Bronze Age (1000 BCE) was taken from the literature, which may represent a general average value for all prehistoric population estimates [@nikulka2016, 258].
+For the fitting process, a starting value of 5 p/km² for a population density of the Late Bronze Age (1000 BCE) was taken from the literature, which may represent a general average value for all prehistoric population estimates [@nikulka2016, 258].
 For the model, this was set as the mean of a normal distribution with a standard deviation of 0.5, which should give enough leeway for deviations resulting from the data.
@@
 
-Within the model, the parameter p was estimated, which reflects the proportion of the individual proxies used in the estimation of the number of settlements, as well as the number of settlements or transforms the resulting population density under the assumption that per settlement a number of inhabitants lived poisson-distributed around a $\lambda$ of 50 persons.
+```{r popdensplot, fig.cap="The estimate of the population density resulting from the model and the three proxies, which are also plotted (scaled) for comparison."}
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popdens_estimation.pdf"))
+```
+
+The population density estimate (Figure \@ref(fig:popdensplot)) ranges on average between 0.2 p/km² for the beginning of the estimate (6000 BCE) and 4.8 p/km² for the end of the estimate (1000 BCE), reaching a maximum of 6.5 p/km² for the time slice 1250 BCE.
+Thus, the estimate remains within the values that are also considered plausible by the expert estimates.
+Clear peaks are reached around 1250 BCE, as well as around 2750 BCE, which corresponds to the transition to the ceramic style in the Swiss Plateau influence by the Corded Ware.
+
+```{r varcoeffplot, fig.cap="The variability of the estimate of the population density over time, with the estimate itself for reference."}
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popvar_plot.pdf"))
+```
+
+Furthermore, it may be relevant to look at the temporal distribution of the variability in the estimate (Figure \@ref(fig:varcoeffplot)), to estimate at which time steps the model attempts a more accurate estimate and at which the uncertainty is greater due to e.g.
+contradictions in the indicators.
+The coefficient of variation is 0.13 for the beginning and 0.1 for the end of the estimate, the greatest variability is reached around 2150 BCE with 0.47.
+The beginning and end of the time series are relatively clearly determined.
+The end results from the a priori setting of the parameter, but also here as at the beginning of the series the proxies are very uniform, which explains the low variability.
+Overall, the variability is relatively uniform over the entire course of the estimation and averages over all time slices at 33% of the respective mean.
+
+Within the model, the parameter p was estimated, which reflects the proportion of the individual proxies used in the estimation of the number of settlements.
 This parameter is variable, but has only a scaling influence on the final estimate of population density.
@@
 The aoristic sum is slightly above 10%, whereas the importance of the dendrodated settlements is below 10%.
-
 The reason for the latter is certainly that there are no lakeshore settlements over large areas of the time window, and therefore the proxy achieves a low confidence value in comparison with the other estimators.
@@
 Nevertheless, its fluctuations do have an impact on the resulting overall estimate of the development of the number of settlements, albeit to a lesser extent.
-
-```{r popdensplot, fig.cap="The estimate of the population density resulting from the model and the three proxies, which are also plotted (scaled) for comparison."}
-knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popdens_estimation.pdf"))
-```
-
-The population density estimate (Figure \@ref(fig:popdensplot)) ranges on average between 0.2 p/km² for the beginning of the estimate (6000 BCE) and 4.8 p/km² for the end of the estimate (1000 BCE), reaching a maximum of 6.5 p/km² for the time slice 1250 BCE.
-Thus, the estimate remains within the values that are also considered plausible by the expert estimates.
-Clear peaks are reached around 1250 BCE (also the maximum value), as well as around 2750 BCE, which corresponds to the onset of the Corded Ware in the Swiss Plateau.
-However, we will go into further detail in the discussion on evaluations concerning the connections to cultural phenomena.
-
-```{r varcoeffplot, fig.cap="The variability of the estimate of the population density over time, with the estimate itself for reference."}
-knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popvar_plot.pdf"))
-```
-
-Furthermore, it may be relevant to look at the temporal distribution of the variability in the estimate (Figure \@ref(fig:varcoeffplot)).
-The coefficient of variation is 0.13 for the beginning and 0.1 for the end of the estimate, the greatest variability is reached around 2150 BCE with 0.4679342.
-The beginning and end of the time series are relatively clearly determined.
-The end results from the a priori setting of the parameter, but also here as at the beginning of the series the proxies are very uniform, which explains the low variability.
-Overall, the variability is relatively uniform over the entire course of the estimation and averages over all time slices at 33% of the respective mean.
 
@@
 The increase in the archaeological record with the process of Neolithisation is still clearly visible, but the curve of the overall estimate after the initial increase is much flatter than the sum calibration alone would suggest.
-The period between 3950 and 3700 in particular forms a noticeable plateau in the calibration curve, and this may lead to an overestimation of the ^14^C evidence.
+The period between 3950 and 3700, which is contemporaneous with the first settlement cycle on the lakeshores of the Three Lakes area, in particular forms a noticeable plateau in the calibration curve, and this may lead to an overestimation of the ^14^C evidence.
 However, the effect of the calibration curve on the results of a cumulative calibration cannot yet be considered unambiguous.
@@
 
-With this model and its predictions, we believe we can demonstrate how it is possible to achieve a true multiproxy analysis of prehistoric demographic processes.
+An important result of the current model is the estimation of absolute population numbers with uncertainty over time.
+This estimate can be used as a basis for further studies where relative measures of population development are not helpful, such as long-term land use studies where modelling of large-scale socio-ecological systems based on archaeological data becomes possible and does have to rely on using deductive, asynchronous population models (eg.
+carrying capacity or ethnographic analogues).
+
+With this model and its predictions, we also believe we can demonstrate how it is possible to achieve a true multiproxy analysis of prehistoric demographic processes.
 Previous attempts in this regard have so far only achieved a juxtaposition of different indicators, without the possibility of truly linking them or assessing the confidence in the individual indicators.

### 2022-04-21 manuscript/paper.Rmd
Initial comments on the manuscript
As comments in paper.Rmd

@@
 abstract: |
-  In order to understand human-environment relationships in the past and the dynamics of socio-ecological systems, it is essential to have a robust estimate of population sizes.
+  In order to understand human-environment relationships in the past and the dynamics of socio-ecological systems, it is essential to have a robust estimate of population sizes<!-- JR: Could reference Shennan 2000 and/or Kintigh 2014 here?-->.
 Size and density of population are crucial factors for the reconstruction of group sizes, the character and range of political institutions for social organization, the constraints and possibilities of economic practice, the number of people available for collective activities, for economic and social exchange systems, and the creation and maintenance of collective identities.
@@
 
+<!-- JR: Might the following four paragraphs fit better in a background section ("Prehistoric demography")? -->
 Gordon Childe [-@childe_man_1936] was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution [@hassan_demographic_1981].
@@
 
+<!-- JR: This paragraph also sounds like background to me. -->
 Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on ^14^C data.
@@
 
+<!-- JR: Begin by defining, broadly, what Bayesian statistics is? -->
 At the heart of all Bayesian statistics is the concept of updating a given prior assumption with new data and expressing this in probabilities (refs.).
@@
 
+<!-- JR: The above is interesting, but maybe a bit general for this paper?
+It's also conspicuously lacking in citations for a background section.
+I'd suggest condensing and moving it to the introduction.
+-->
+
 A special class of Bayesian hierarchical models are so-called State Space Models (also known as Hidden Markov Models).
@@
 The specifics of the individual data we used in this modelling are discussed in the next section.
+
+<!-- JR: Ditto: maybe move the above to method and/or introduction? -->
 
@@
 
+<!-- JR: Non-Europeanists (like me!) might appreciate a brief summary of the geography and prehistory of the Swiss plateau here. -->
+
 Our working region consists of the Swiss Plateau, which on the one hand offers excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking.
@@
 ```
+
+<!-- JR: The rest of this section is a bit and I lost the thread a bit.
+Could we first have a brief summary of which proxies we will use, and why (helping to explain why the Swiss Plateau was selected), then further details on each proxy in labelled sub-sections?
+-->
 
@@
 
+<!-- JR: I think we should explain the choice of time window fully here, where it's first mentioned. -->
+
 We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites.
 For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [@crema2021] were used in the standard settings.
+
+<!-- JR: Would be nice to add rivers and lakes to the base map. -->
 
@@
 
+<!-- JR: I don't understand what "scaled values" means here. -->
+
 We can now compare these two data sets (Figure \@ref(fig:compdendroc14)).
@@
 ```
+
+<!-- JR: At the resolution the figure is rendered at, I don't think the 1 km jitter is perceptible. -->
 
@@
 
+<!-- JR: What preservation conditions? There are prehistoric burials and cemeteries in Switzerland, no? -->
+
 The values of the sum calibration, the openness index and the dendrodated settlements were smoothed by means of a moving average with a window of 50 years.
@@
 The lower limit is defined by the Hallstatt Plateau (with buffer), which would have direct and indirect research influences on the ^14^C proxy.
+
+<!-- JR: The last bit repeats an earlier point. -->
 
@@
 
-In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods.
+In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods <!-- JR: Can we gloss what autocorrelated means in plain language here, to make it clear that this is a trivial assumption?
+something like ", i.e.
+the number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on." -->.
 In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.
@@
 
-Thus, we can define the limits, the $max\_change\_rate$ as an time independent variable, again without already specifying them with fixed values at this stage.
+Thus, we can define the limits, the $max\_change\_rate$<!-- JR: no Greek letter?
+:( --> as an time independent variable, again without already specifying them with fixed values at this stage.
 The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively.
@@
 The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$ (See also Figure \@ref(fig:gammadist)).
+
+<!-- JR: A whole figure just to illustrate the gamma distribution seems like overkill.
+Could we add the prior distributions of other parameters?
+Or just get rid of it: you can always look up the shapes on Wikipedia :) -->
 
@@
 
+<!-- JR: I really like the idea of discussing "dead ends" in the model design process – very modern and reflexive! Maybe hightlight these in a section of their own? -->
+
 In earlier implementations, expert estimates were also integrated into the model.
@@
 
+<!-- JR: I don't know where it would fit best, but I'd find it very helpful to see a table of all the priors and fixed parameters with their initial values. -->
+
 # Results
+
+<!-- JR: I really think we need a section evaluating and validating the model fit before we discuss its output. -->
 
@@
 # Discussion
+
+<!-- JR: I would split this into subsections evaluation a) the merits of the approach and b) the specific results for Switzerland. -->
 

### 2022-04-25 manuscript/paper.Rmd
first batch of incorporating Joes suggestions

@@
 output:
+  bookdown::pdf_document2:
+    toc: no
+    pandoc_args:
+    - --lua-filter=../templates/scholarly-metadata.lua
+    - --lua-filter=../templates/author-info-blocks.lua
+    - --lua-filter=../templates/pagebreak.lua
   bookdown::word_document2:
@@
     - --lua-filter=../templates/pagebreak.lua
-  bookdown::pdf_document2:
-    toc: no
-    pandoc_args:
-    - --lua-filter=../templates/scholarly-metadata.lua
-    - --lua-filter=../templates/author-info-blocks.lua
-    - --lua-filter=../templates/pagebreak.lua
 bibliography: references.bib
@@
 abstract: |
+  In demography in general, the last decade has triggered an upswing in the application of Bayesian methods, so that a Bayesian demography has been announced.
+Bayesian demography is currently at the forefront of methodological developments in this area, but has reached such maturity that the United Nations has been using it since 2015 for population forecasts and projections.
+The Bayesian approach offers the possibility of combining heterogeneous data and at the same time qualifying them in terms of uncertainty and credibility.
+This is precisely where it becomes very interesting for archaeological data, since they are mostly inaccurate, biased, sparse, simplified and often based on simplistic assumptions.
+  In this contribution, a Bayesian hierarchical model, based on a Poisson Regression and a state space approach is applied to the field of archaeological (prehistoric) population size and density estimation.
+This represents a pilot study for the reconstruction of demographic developments in prehistory based on data assimilation with an underlying process model and a number of different proxies, which are combined and result in a more reliable estimation than a single proxy could deliver.
   In order to understand human-environment relationships in the past and the dynamics of socio-ecological systems, it is essential to have a robust estimate of population sizes<!-- JR: Could reference Shennan 2000 and/or Kintigh 2014 here?-->.
@@
 The figures in the archaeological discourses on population density in general currently come from vague sources and are often decades old.
-  In demography in general, the last decade has triggered an upswing in the application of Bayesian methods, so that a Bayesian demography has been announced.
-Bayesian demography is currently at the forefront of methodological developments in this area, but has reached such maturity that the United Nations has been using it since 2015 for population forecasts and projections.
-The Bayesian approach offers the possibility of combining heterogeneous data and at the same time qualifying them in terms of uncertainty and credibility.
-This is precisely where it becomes very interesting for archaeological data, since they are mostly inaccurate, biased, sparse, simplified and often based on simplistic assumptions.
-  In this contribution, a Bayesian hierarchical model, based on a Poisson Regression and a state space approach is applied to the field of archaeological (prehistoric) population size and density estimation.
-This represents a pilot study for the reconstruction of demographic developments in prehistory based on data assimilation with an underlying process model and a number of different proxies, which are combined and result in a more reliable estimation than a single proxy could deliver.
 keywords: |
@@
 
-```{r setup, echo = FALSE, message = FALSE}
+```{r setup, echo = FALSE, message = FALSE, warning=FALSE}
 knitr::opts_chunk$set(
@@
 library(ggmap)
+library(cowplot)
 library(sp)
@@
 
+switzerland_bb <- st_as_sfc(st_bbox(switzerland))
+
 swiss_background <- get_stamenmap( bbox = bbox_ch, maptype = "terrain-background", color="bw", zoom = 8)
 
-my_basemap <- ggmap(swiss_background, darken = c(0.6, "white"))
+rivers10 <- ne_download(scale = "large", type = 'rivers_lake_centerlines', category = 'physical') %>% st_as_sf()
+
+rivers10eu <- ne_download(scale = "large", type = 'rivers_europe', category = 'physical') %>% st_as_sf()
+
+lakes10 <- ne_download(scale = "large", type = 'lakes', category = 'physical') %>% st_as_sf()
+
+lakes10eu <- ne_download(scale = "large", type = 'lakes_europe', category = 'physical') %>% st_as_sf()
+
+my_basemap <- ggmap(swiss_background, darken = c(0.6, "white")) + coord_sf(default_crs = st_crs(4326)) + geom_sf(data = rivers10, color="lightblue4", inherit.aes = F)  + geom_sf(data = rivers10eu, color="lightblue4", inherit.aes = F)  + geom_sf(data = lakes10, fill="lightblue4", color=NA, inherit.aes = F)  + geom_sf(data = lakes10eu, fill="lightblue4", inherit.aes = F, color=NA)
 
 my_scalebar <- scalebar(transform = T, dist_unit = "km", dist=50, model = "WGS84", st.dist = 0.025, st.size = 3, x.min = bbox_ch[1], x.max = bbox_ch[3]-0.2, y.min = bbox_ch[2]+0.15, y.max = bbox_ch[4], border.size = 0.5)
+
+europe <- c(left = -12, bottom = 35, right = 30, top = 63) %>% as.vector() %>% +0.01
+
+europe_overview <- get_stamenmap( bbox = europe, maptype = "toner-background", color="bw", zoom = 3)
+
+europe_overview_map <- ggmap(europe_overview, darken = c(0.6, "white"))+ coord_sf(default_crs = st_crs(4326))
 ```
@@
 <!-- JR: Might the following four paragraphs fit better in a background section ("Prehistoric demography")? -->
+
 Gordon Childe [-@childe_man_1936] was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution [@hassan_demographic_1981].
@@
 <!-- JR: This paragraph also sounds like background to me. -->
+
 Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on ^14^C data.
@@
 <!-- JR: Begin by defining, broadly, what Bayesian statistics is? -->
+
 At the heart of all Bayesian statistics is the concept of updating a given prior assumption with new data and expressing this in probabilities (refs.).
@@
 
-<!-- JR: Non-Europeanists (like me!) might appreciate a brief summary of the geography and prehistory of the Swiss plateau here. -->
+```{r mapswissplateau, fig.cap="Location and extent of the Swiss Plateau."}
+fc <- sf::st_read("../data/fixed_online_data/BiogeographischeRegionen.gdb", layer = "N2020_Revision_BiogeoRegion")
+fc <- st_transform(fc, 4326)
+fc <- fc[fc$RegionNummer==2,] %>% st_union()
 
-Our working region consists of the Swiss Plateau, which on the one hand offers excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking.
+main_map <- my_basemap +
+  geom_sf(data = fc, inherit.aes = FALSE, alpha = 0.33, fill="darkred", color=NA) +
+  blank() +
+  my_scalebar
+
+inset_map <- europe_overview_map +
+  geom_sf(data = switzerland_bb, fill = NA, color = "darkred", inherit.aes = FALSE) +
+  blank()
+
+ggdraw() +
+  draw_plot(main_map) +
+  draw_plot(inset_map, x = 0, y = 0.7, width = 0.3, height = 0.3)
+```
+
+The Swiss Plateau (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland's territory and comprises the partly flat but largely hilly area between the Jura Mountains and the Alps.
+Its average altitude above sea level is between 400 and 600 m.
+Due to its situation as a basin, it represents a favourable area and is by far the most densely populated region of the Switzerland today.
+Although the Swiss Plateau is a basin, depending on the region it has a very diverse natural landscape.
+The Swiss Plateau was shaped by glaciers during the Ice Age, which led to the many lakes that characterise the landscape today, and which provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements.
+
+With a mild climate compared to the Jura and the Alps and a dense network of water bodies, the Swiss Plateau is a favourable area for human settlement in Switzerland.
+From the Neolithic onwards, settlement areas were concentrated along its rivers and lakes [@christianlüthi2009].
+Thus, our working region offers on the one hand excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking.
 If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also might cause a research problem with regard to the ^14^C data often used as a proxy.
@@
 
-<!-- JR: Would be nice to add rivers and lakes to the base map. -->
-
 ```{r c14map, fig.cap="The location of the ^14^C dated sites in the dataset."}
@@
 
-```{r compdendroc14, fig.cap="Comparison of the scaled number of ^14^C and dendrodated sites over time in the dataset used."}
+```{r compdendroc14, fig.cap="Comparison of the scaled (non-centered z-transformed) number of ^14^C and dendrodated sites over time in the dataset used."}
 joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))
@@
 ```
-
-<!-- JR: I don't understand what "scaled values" means here. -->
 
@@
 <!-- JR: At the resolution the figure is rendered at, I don't think the 1 km jitter is perceptible. -->
+
+<!-- But in the data table that accompanies the repository, they would be available -->
 
@@
 
-Unfortunately, in the Swiss Plateau there is no data from burials that could be usefully incorporated into such this model.
-Preservation conditions again simply prevent their use in this study.
+Unfortunately, in the Swiss Plateau there is no data from burials that could be usefully incorporated into such this model due to the absence of regular burials for larger parts of prehistory.
+Formation conditions, in this case most likely C-transformation [@schiffer1987], prevent their use in this study.
 Nevertheless, we see a very high potential for other regions in the integration of demographic indicators from burial data in order to enlarge the canon of methods and the range of proxies.
-
-<!-- JR: What preservation conditions? There are prehistoric burials and cemeteries in Switzerland, no? -->
 
@@
 
-In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods <!-- JR: Can we gloss what autocorrelated means in plain language here, to make it clear that this is a trivial assumption?
-something like ", i.e.
-the number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on." -->.
+In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods, i.e.
+the number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on.
 In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.
@@
 Thus, we can define the limits, the $max\_change\_rate$<!-- JR: no Greek letter?
-:( --> as an time independent variable, again without already specifying them with fixed values at this stage.
+:( MH: I could invent one, if you like?
+;-) --> as an time independent variable, again without already specifying them with fixed values at this stage.
 The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively.
@@
 
-The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$ (See also Figure \@ref(fig:gammadist)).
-
-<!-- JR: A whole figure just to illustrate the gamma distribution seems like overkill.
-Could we add the prior distributions of other parameters?
-Or just get rid of it: you can always look up the shapes on Wikipedia :) -->
-
-```{r gammadist, fig.cap="The gamma-distributed maximum growth rate prior of the model."}
-plot(density(rgamma(100000,shape=5,scale = 0.05)+1), xlim=c(0,5), main ="gamma(shape = 5, scale=0.05) + 1" )
-```
-
+The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$.
 This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model.

### 2022-04-25 manuscript/paper.Rmd
streamlined time window mentioning

@@
 But it was precisely to reduce its influence that the time window of the investigation was chosen to be 7000 -- 1000 BCE.
-
-<!-- JR: I think we should explain the choice of time window fully here, where it's first mentioned. -->
+The upper limit is set by the post-glacial changes in the pollen spectra, which are hardly associated with human influence before 7000 in the working area, and which would consequently distort the openness indicator.
+The lower limit is defined by the Hallstatt Plateau (with buffer), which would have direct and indirect research influences on the ^14^C proxy.
 
@@
 The range of the smoothing window corresponded to the sample interval, with which a unified resolution of 50 years was obtained for all proxies as time slices for the model.
-In addition, all data were restricted to a window of observation of 7000-1000 BCE.
-The upper limit is set by the post-glacial changes in the pollen spectra, which are hardly associated with human influence before 7000 in the working area, and which would consequently distort the openness indicator.
-The lower limit is defined by the Hallstatt Plateau (with buffer), which would have direct and indirect research influences on the ^14^C proxy.
-
-<!-- JR: The last bit repeats an earlier point. -->
+In addition, all data were restricted to the window of observation of 7000--1000 BCE.
 

### 2022-04-25 manuscript/paper.Rmd
added table of priors

@@
 This represents a pilot study for the reconstruction of demographic developments in prehistory based on data assimilation with an underlying process model and a number of different proxies, which are combined and result in a more reliable estimation than a single proxy could deliver.
-  In order to understand human-environment relationships in the past and the dynamics of socio-ecological systems, it is essential to have a robust estimate of population sizes<!-- JR: Could reference Shennan 2000 and/or Kintigh 2014 here?-->.
+  In order to understand human-environment relationships in the past and the dynamics of socio-ecological systems, it is essential to have a robust estimate of population sizes<!-- JR: Could reference Shennan 2000 and/or Kintigh 2014 here?
+MH: I would not like to have any citation in the abstract.
+What do you think?-->.
 Size and density of population are crucial factors for the reconstruction of group sizes, the character and range of political institutions for social organization, the constraints and possibilities of economic practice, the number of people available for collective activities, for economic and social exchange systems, and the creation and maintenance of collective identities.
@@
 
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
-
-<!-- With the following code you can access and display values from the yml header above. -->
-
 Keywords: `r rmarkdown::metadata$keywords`
@@
 Highlights: `r rmarkdown::metadata$highlights`
-
-<!-- The following code chunk defines some general settings how code chunks should behave. -->
 
@@
 
-<!-- The actual document text starts here: -->
+# Introduction
 
-# Introduction
+## Prehistoric demography as object of research
 
@@
 Although questions of population extent and distribution have been dominant fields since the beginning of archaeological research, the lack of access to this information, data on it, or methodological tools for the use of proxies has long made the statements made on this topic for prehistoric times vague and superficial [@hassan_demographic_1981], and they were often based on uncritical transfer of ethnographic parallels.
-
-<!-- JR: Might the following four paragraphs fit better in a background section ("Prehistoric demography")? -->
 
@@
 [-@kintigh_grand_2014] listed human influence, dominance and population size and growth as one of the key elements where methodological improvements are necessary in the article on the Grand Challenges for archaeology, in the area of 'human-environment interactions'.
+
+## Issues associated with existing approaches
 
@@
 
+## Strengths of a hierarchical Bayesian approach
+
 In recent years, a new method has been established in the field of demographic estimations.
@@
 
-<!-- JR: This paragraph also sounds like background to me. -->
+This modeling technique can thus be used to combine different lines of evidence horizontally and vertically and in this way combine their results into a common conclusion and estimate, which at the same time includes an estimation of their reliability: if the data contradict each other, the overall reliability will be lower.
+If they support each other, the confidence interval will be smaller.
+And if there is no systematic bias that affects all data sources to the same extent, it should be possible to arrive at the most reliable estimate possible through the most heterogeneous set of data sources.
 
@@
 
-<!-- JR: Begin by defining, broadly, what Bayesian statistics is? -->
+The basis of Bayesian statistics is the premise that there is always some prior assumption about the probability of an event, even if it may be completely vague.
+This assumption is then adjusted by observing the data.
+The concept of probability is not derived from theoretical, infinitely repeatable random experiments or distributions, but from the direct confrontation of the pre-assumptions (priors) with the available data.
+This involves checking how credible these prior assumptions are with regard to the available data [likelihood; see also @bryant2018, 66].
+A small amount of data leads to a broad probability distribution that is not strongly localised and restricted.
+The intuitive procedure is thereby the shifting of credibility through evidence.
 
-At the heart of all Bayesian statistics is the concept of updating a given prior assumption with new data and expressing this in probabilities (refs.).
+At the heart of all Bayesian statistics is therefore the concept of updating a given prior assumption with new data and expressing this in probabilities [cf.
+also @kruschke2015, especially 15--25].
 Our assumptions about the demographic development of the past must naturally be very conservative.
@@
 This Bayesian learning is iterative and sequential, so that the result of one Bayesian inference can form the prior of another, i.e.
-it is an additive process.
-Moreover, at the conceptual level, this allows different sources of information to be combined.
+it is an additive process [@kruschke2015, 17].
+Moreover, at the conceptual level, this allows different sources of information to be combined [@bryant2018, 219--224].
 This enables that they can be mapped to the same set and the same real-world domain, e.g.
 probabilities.
-This fact has long been exploited by archaeology in using stratigraphic information to make radiometric dating more accurate.
+This fact has long been exploited by archaeology in using stratigraphic information to make radiometric dating more accurate [@ramsey1995].
 ^14^C dates and stratigraphy are something completely different, but both can be mapped to the probability of older and younger and combined in this way.
@@
 
-A further characteristic of Bayesian modeling is that, due to the fact that results of a Bayesian inference can be regarded as a priority for the next one, a hierarchical formulation of problem domains is possible.
-Parameters that are necessary for an estimation, such as the relationship of population density to the clearing signal in pollen data, need not be specified explicitly, but can be given by probability distributions and then estimated in the model itself.
-The more data available, the more degrees of freedom can be estimated with a reasonable loss of confidence or a reasonable width of credibility intervals.
+A further characteristic of Bayesian modeling is that, due to the fact that results of a Bayesian inference can be regarded as a prior for the next one, a hierarchical formulation of problem domains is possible.
+Parameters that are necessary for an estimation, such as the relationship of population density to the clearing signal in pollen data, need not be specified explicitly, but can be given by probability distributions and then estimated in the model itself [@bryant2018, 186].
+The more data available, the more degrees of freedom can be estimated with a reasonable loss of confidence or a reasonable width of credibility intervals [@kruschke2015, 112].
 For the estimation of these parameters, in turn, submodels have to be created which describe the relationship of the data to the characteristics of the parameter.
-This can be carried out over several levels, depending on necessity.
-
-This modeling technique can thus be used to combine different lines of evidence horizontally and vertically and in this way combine their results into a common conclusion and estimate, which at the same time includes an estimation of their reliability: if the data contradict each other, the overall reliability will be lower.
-If they support each other, the confidence interval will be smaller.
-And if there is no systematic bias that affects all data sources to the same extent, it should be possible to arrive at the most reliable estimate possible through the most heterogeneous set of data sources.
-
-<!-- JR: The above is interesting, but maybe a bit general for this paper?
-It's also conspicuously lacking in citations for a background section.
-I'd suggest condensing and moving it to the introduction.
--->
-
-A special class of Bayesian hierarchical models are so-called State Space Models (also known as Hidden Markov Models).
-These are specifically designed for time series and follow two basic principles.
-First, a hidden or latent process is assumed, which represents the state of the variable of interest $x_t$ at all times.
-For the course of the states of this variable over time, it is assumed that every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$.
-At the same time, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$.
-This means that there is a relationship between the variable $x$ and the state of variable y over time.
-This implies that a relationship between the individual states of variable $y$ is generated over time via the hidden variable $x$, which itself is not observable.
-
-This basic structure of the model makes it particularly useful and suitable for the purpose of demographic reconstruction using archaeological, but also other data, which depend on population density in the past.
-This population density itself is not accessible or measurable by our means.
-All we have at our disposal are observations derived by unknown transfer functions.
-These can be of very different natures, such as number of archaeologically observable settlements, or other effects that can be observed through time series, and which are influenced by population in a given area.
-In our example, these are the openness indicators from pollen data, which we can interpret primarily in terms of human influence and its intensity.
-On a more abstract level, we could also include expert estimates, as these also happen on (often unspecified) bases that are at least indirectly influenced by past population density.
-The specifics of the individual data we used in this modelling are discussed in the next section.
-
-<!-- JR: Ditto: maybe move the above to method and/or introduction? -->
+This can be carried out over several levels, depending on necessity [@kruschke2015, 221--222].
 
@@
 
-<!-- JR: At the resolution the figure is rendered at, I don't think the 1 km jitter is perceptible. -->
-
-<!-- But in the data table that accompanies the repository, they would be available -->
+<!-- JR: At the resolution the figure is rendered at, I don't think the 1 km jitter is perceptible. MH: But in the data table that accompanies the repository, they would be available -->
 
@@
 ## Process Model
+
+A special class of Bayesian hierarchical models are so-called State Space Models (also known as Hidden Markov Models).
+These are specifically designed for time series and follow two basic principles.
+First, a hidden or latent process is assumed, which represents the state of the variable of interest $x_t$ at all times.
+For the course of the states of this variable over time, it is assumed that every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$.
+At the same time, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$.
+This means that there is a relationship between the variable $x$ and the state of variable y over time.
+This implies that a relationship between the individual states of variable $y$ is generated over time via the hidden variable $x$, which itself is not observable.
+
+This basic structure of the model makes it particularly useful and suitable for the purpose of demographic reconstruction using archaeological, but also other data, which depend on population density in the past.
+This population density itself is not accessible or measurable by our means.
+All we have at our disposal are observations derived by unknown transfer functions.
+These can be of very different natures, such as number of archaeologically observable settlements, or other effects that can be observed through time series, and which are influenced by population in a given area.
+In our example, these are the openness indicators from pollen data, which we can interpret primarily in terms of human influence and its intensity.
+On a more abstract level, we could also include expert estimates, as these also happen on (often unspecified) bases that are at least indirectly influenced by past population density.
 
@@
 <!-- JR: I don't know where it would fit best, but I'd find it very helpful to see a table of all the priors and fixed parameters with their initial values. -->
+
+```{r}
+x <- 1:100
+```
+
+```{r dpois50, dev='png', fig.show='hide', fig.height = 3, fig.width = 3}
+plot(x,dpois(x,50), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+```
+
+```{r dgammamaxgrowthrate, dev='png', fig.show='hide', fig.height = 3, fig.width = 3}
+plot(x/50+1,dgamma(x/50,shape = 5, scale = 0.05), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+```
+
+```{r mualpha, dev='png', fig.show='hide', fig.height = 3, fig.width = 3}
+plot(x/20,dlnorm(x/20,1,0.1), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+```
+
+```{r aalpha, dev='png', fig.show='hide', fig.height = 3, fig.width = 3}
+plot(x/10,dexp(x/10,1), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+```
+
+```{r alpha, dev='png', fig.show='hide', fig.height = 3, fig.width = 3}
+n_sample <- 1000000
+mu_a <- rlnorm(n_sample,1,0.1)
+a_a <- rexp(n_sample,1)
+alpha <- rlnorm(n_sample,mu_a,a_a)
+plot(density(alpha, from=0,to=100), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+```
+
+```{r p, dev='png', fig.show='hide', fig.height = 3, fig.width = 3, cache=TRUE}
+p <- matrix(nrow = 10000, ncol = 4)
+for (i in 1:nrow(p)) {
+ p[i,] <- nimble::rdirch(1, sample(alpha,4))
+}
+par(mfrow=c(2,2))
+oldmar <- par(mar = c(2, 2, 2, 2))
+plot(density(p[,1]), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+plot(density(p[,2]), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+plot(density(p[,3]), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+plot(density(p[,4]), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+par(mfrow=c(1,1))
+par(oldmar)
+```
+
+| **Priors**           | **Value**                            | **Plot/Comment**                                                               |
+|-----------------------------|--------------------------|-----------------|
+| MeanSiteSize         | dpois(50)                            | ![MeanSiteSize](`r knitr::fig_chunk('dpois50', 'png')`)                |
+| max_growth_rate      | dgamma(shape = 5, scale=0.05) + 1    | ![max_growth_rate](`r knitr::fig_chunk('dgammamaxgrowthrate','png')`) |
+| mu_alpha             | dlnorm(1,sdlog=0.1)                  | ![mu_alpha](`r knitr::fig_chunk('mualpha','png')`)                    |
+| a_alpha              | dexp(1)                              | ![a_alpha](`r knitr::fig_chunk('aalpha','png')`)                      |
+| alpha                | dlnorm(mu_alpha[j],sdlog=a_alpha[j]) | ![alpha](`r knitr::fig_chunk('alpha','png')`)                         |
+| p                    | ddirch(alpha[1:4])                   | ![p](`r knitr::fig_chunk('p','png')`)                                 |
+| **Parameters**       |                                      |                                                                                |
+| nEnd                 | 5                                    |                                                                                |
+| AreaSwissPlateau     | 12649 km²                            |                                                                                |
+| **Initial Values**   |                                      |                                                                                |
+| lambda$_{1:nYears}$  | $log(1-10^{\frac{1}{nYears-1}})$     | exponential increase of the factor 10                                          |
+| PopDens$_{1:nYears}$ | nEnd (=5)                            |                                                                                |
+| nSites$_{1:nYears}$  | 50                                   |                                                                                |
+
+: (#tab:priorsandparameters) Priors and fixed parameters used in the model.
 
@@
 
-<!-- The following line inserts a page break  -->
-
 \newpage
@@
 # References
-
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
 

### 2022-04-27 analysis/analysis.Rmd
implemented or commented Joes suggestions

@@
     to=10000, by=50),]
-ggplot(res.open_smoothed,aes(x=1950-age,y=moving_mean)) +
+pollenproxy_plot <- ggplot(res.open_smoothed,aes(x=1950-age,y=moving_mean)) +
   geom_point(aes(color=site), alpha =.5) +
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
+pollenproxy_plot
+```
+We save this plot for later use:
+
+```{r}
+ggsave(normalizePath(file.path(here(), "figures","pollenproxy.pdf"), mustWork = F),
+       plot = pollenproxy_plot, width = 21, height = 29.7/2, units = "cm")
 ```

### 2022-04-27 manuscript/paper.Rmd
implemented or commented Joes suggestions

@@
 
-```{r setup, echo = FALSE, message = FALSE, warning=FALSE}
+```{r setup, echo = FALSE, message = FALSE, warning=FALSE, results='hide'}
 knitr::opts_chunk$set(
@@
 europe_overview_map <- ggmap(europe_overview, darken = c(0.6, "white"))+ coord_sf(default_crs = st_crs(4326))
+
+fc <- sf::st_read("../data/fixed_online_data/BiogeographischeRegionen.gdb", layer = "N2020_Revision_BiogeoRegion")
+fc <- st_transform(fc, 4326)
+fc <- fc[fc$RegionNummer==2,] %>% st_union()
+
+plot_path <- normalizePath(
+  file.path(
+    here(), "figures"
+    )
+  )
 ```
@@
 
+The Swiss Plateau (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland's territory and comprises the partly flat but largely hilly area between the Jura Mountains and the Alps.
+Its average altitude above sea level is between 400 and 600 m.
+Due to its situation as a basin, it represents a favourable area and is by far the most densely populated region of the Switzerland today.
+Although the Swiss Plateau is a basin, depending on the region it has a very diverse natural landscape.
+The Swiss Plateau was shaped by glaciers during the Ice Age, which led to the many lakes that characterise the landscape today, and which provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, as well as a rich pool of sources for vegetation reconstructions by means of pollen analyses.
+A very active and efficient archaeological research and heritage management also offers a rich repository of archaeological data, known sites and ^14^C data.
+
 In principle, a large number of different data sources can be integrated into the overall model as observations, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created.
-A non-exhaustive list can be found in the following table \@ref(tab:tableproxies):
+A non-exhaustive list can be found in the following table \@ref(tab:tableproxies).
+For our analysis of the Swiss Plateau we used an open land indicator, an aoristic sum of sites based on typological dating, a sum calibration and the number of dendro dated lakeshore settlements in the Three Lakes area.
+The individual proxies will be discussed in more detail below.
 
@@
 |------------------------------------------------|
-| **Expert estimates**                           |
+| Expert estimates                           |
 | Ethnographic Analogies                         |
@@
 ```{r mapswissplateau, fig.cap="Location and extent of the Swiss Plateau."}
-fc <- sf::st_read("../data/fixed_online_data/BiogeographischeRegionen.gdb", layer = "N2020_Revision_BiogeoRegion")
-fc <- st_transform(fc, 4326)
-fc <- fc[fc$RegionNummer==2,] %>% st_union()
-
 main_map <- my_basemap +
@@
 
-The Swiss Plateau (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland's territory and comprises the partly flat but largely hilly area between the Jura Mountains and the Alps.
-Its average altitude above sea level is between 400 and 600 m.
-Due to its situation as a basin, it represents a favourable area and is by far the most densely populated region of the Switzerland today.
-Although the Swiss Plateau is a basin, depending on the region it has a very diverse natural landscape.
-The Swiss Plateau was shaped by glaciers during the Ice Age, which led to the many lakes that characterise the landscape today, and which provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements.
+### Dendro Dated Lakeshore Settlements
 
@@
 
-<!-- JR: The rest of this section is a bit and I lost the thread a bit.
-Could we first have a brief summary of which proxies we will use, and why (helping to explain why the Swiss Plateau was selected), then further details on each proxy in labelled sub-sections?
--->
-
 The dataset we use for the number of dendrodated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019].
@@
 This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.
+
+### ^14^C sum calibration
 
@@
 
+### Aoristic Sum
+
 ```{r echo=FALSE}
@@
 
+### Openness Indicator
+
 The natural conditions provided by the many lakes enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis.
@@
 ```{r pollenproxy, fig.cap="Value on the first dimension of the PCA against dating of the samples for the individual pollen profiles and their combined average value as the openness indicator."}
-knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "analysis", "analysis_files", "figure-html", "pollenproxy-1.png"))
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "pollenproxy.pdf"))
 ```
@@
 
-<!-- JR: I really like the idea of discussing "dead ends" in the model design process – very modern and reflexive! Maybe hightlight these in a section of their own? -->
+<!-- JR: I really like the idea of discussing "dead ends" in the model design process – very modern and reflexive!
+Maybe hightlight these in a section of their own?
+MH: I do not feel that this part would provide enough material to justify an own section.
+-->
 
@@
 
-<!-- JR: I don't know where it would fit best, but I'd find it very helpful to see a table of all the priors and fixed parameters with their initial values. -->
-
-```{r}
+```{r, results='hide'}
 x <- 1:100
@@
 
-```{r dpois50, dev='png', fig.show='hide', fig.height = 3, fig.width = 3}
+```{r dpois50, results='hide'}
+pdf(file = file.path(plot_path, "dpois50.pdf"), width = 2, height = 2)
+oldmar <- par(mar = c(2, 2, 2, 2))
 plot(x,dpois(x,50), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+par(oldmar)
+dev.off()
 ```
 
-```{r dgammamaxgrowthrate, dev='png', fig.show='hide', fig.height = 3, fig.width = 3}
+```{r dgammamaxgrowthrate, results='hide'}
+pdf(file = file.path(plot_path, "dgammamaxgrowthrate.pdf"), width = 2, height = 2)
+oldmar <- par(mar = c(2, 2, 2, 2))
 plot(x/50+1,dgamma(x/50,shape = 5, scale = 0.05), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+par(oldmar)
+dev.off()
 ```
 
-```{r mualpha, dev='png', fig.show='hide', fig.height = 3, fig.width = 3}
+```{r mualpha, results='hide'}
+pdf(file = file.path(plot_path, "mualpha.pdf"), width = 2, height = 2)
+oldmar <- par(mar = c(2, 2, 2, 2))
 plot(x/20,dlnorm(x/20,1,0.1), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+par(oldmar)
+dev.off()
 ```
 
-```{r aalpha, dev='png', fig.show='hide', fig.height = 3, fig.width = 3}
+```{r aalpha, results='hide'}
+pdf(file = file.path(plot_path, "aalpha.pdf"), width = 2, height = 2)
+oldmar <- par(mar = c(2, 2, 2, 2))
 plot(x/10,dexp(x/10,1), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+par(oldmar)
+dev.off()
 ```
 
-```{r alpha, dev='png', fig.show='hide', fig.height = 3, fig.width = 3}
+```{r alpha, cache=TRUE, results='hide'}
 n_sample <- 1000000
@@
 alpha <- rlnorm(n_sample,mu_a,a_a)
+pdf(file = file.path(plot_path, "alpha.pdf"), width = 2, height = 2)
+oldmar <- par(mar = c(2, 2, 2, 2))
 plot(density(alpha, from=0,to=100), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+par(oldmar)
+dev.off()
 ```
 
-```{r p, dev='png', fig.show='hide', fig.height = 3, fig.width = 3, cache=TRUE}
+```{r p, cache=TRUE, results='hide'}
 p <- matrix(nrow = 10000, ncol = 4)
@@
 }
+pdf(file = file.path(plot_path, "p.pdf"), width = 2.5, height = 2.5)
+oldmar <- par(mar = c(2, 2, 2, 2))
 par(mfrow=c(2,2))
-oldmar <- par(mar = c(2, 2, 2, 2))
 plot(density(p[,1]), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
@@
 plot(density(p[,4]), type="l", ann = FALSE, axes = FALSE, lwd=4, col = "skyblue");axis(1)
+par(oldmar)
 par(mfrow=c(1,1))
-par(oldmar)
+dev.off()
 ```
@@
 |-----------------------------|--------------------------|-----------------|
-| MeanSiteSize         | dpois(50)                            | ![MeanSiteSize](`r knitr::fig_chunk('dpois50', 'png')`)                |
-| max_growth_rate      | dgamma(shape = 5, scale=0.05) + 1    | ![max_growth_rate](`r knitr::fig_chunk('dgammamaxgrowthrate','png')`) |
-| mu_alpha             | dlnorm(1,sdlog=0.1)                  | ![mu_alpha](`r knitr::fig_chunk('mualpha','png')`)                    |
-| a_alpha              | dexp(1)                              | ![a_alpha](`r knitr::fig_chunk('aalpha','png')`)                      |
-| alpha                | dlnorm(mu_alpha[j],sdlog=a_alpha[j]) | ![alpha](`r knitr::fig_chunk('alpha','png')`)                         |
-| p                    | ddirch(alpha[1:4])                   | ![p](`r knitr::fig_chunk('p','png')`)                                 |
+| MeanSiteSize         | dpois(50)                            | ![MeanSiteSize](../figures/dpois50.pdf)                |
+| max_growth_rate      | dgamma(shape = 5, scale=0.05) + 1    | ![max_growth_rate](../figures/dgammamaxgrowthrate.pdf) |
+| mu_alpha             | dlnorm(1,sdlog=0.1)                  | ![mu_alpha](../figures/mualpha.pdf)                    |
+| a_alpha              | dexp(1)                              | ![a_alpha](../figures/aalpha.pdf)                      |
+| alpha                | dlnorm(mu_alpha[j],sdlog=a_alpha[j]) | ![alpha](../figures/alpha.pdf)                         |
+| p                    | ddirch(alpha[1:4])                   | ![p](../figures/p.pdf)                                 |
 | **Parameters**       |                                      |                                                                                |
@@
 
-<!-- JR: I really think we need a section evaluating and validating the model fit before we discuss its output. -->
+<!-- JR: I really think we need a section evaluating and validating the model fit before we discuss its output.
+MH: that was the intention of the Model fitting section before, but I think this rather belongs to methods than to results!?-->
 
@@
 
-<!-- JR: I would split this into subsections evaluation a) the merits of the approach and b) the specific results for Switzerland. -->
+<!-- JR: I would split this into subsections evaluation a) the merits of the approach and b) the specific results for Switzerland.
+MH: I thought I already have done this, discussion more the results for switzerland, conclusion more the generel merrits!?
+-->
 

### 2022-05-18 analysis/analysis.Rmd
included comments and corrections of Jan

@@
 popdens_plot <- ggplot(all_proxies_orig) +
-  geom_line(aes(x = 1950-age, y = scale(openness), col="Openness Indicator")) +
-  geom_line(aes(x = 1950-age, y = scale(sumcal), col="Sum Calibration")) +
-  geom_line(aes(x = 1950-age, y = scale(dendro), col="Dendro Dated Settlements")) +
-  geom_line(aes(x = 1950-age, y = scale(aoristic_sum), col="Aoristic Sum")) +
+  geom_line(aes(x = (1950-age) * -1, y = scale(openness), col="Openness Indicator")) +
+  geom_line(aes(x = (1950-age) * -1, y = scale(sumcal), col="Sum Calibration")) +
+  geom_line(aes(x = (1950-age) * -1, y = scale(dendro), col="Dendro Dated Settlements")) +
+  geom_line(aes(x = (1950-age) * -1, y = scale(aoristic_sum), col="Aoristic Sum")) +
   geom_line(data = PopDens_summary,
-            aes(x = 1950-age, y=mean, col = "Estimation Result")) + 
+            aes(x = (1950-age) * -1, y=mean, col = "Estimation Result")) + 
   geom_line(data = PopDens_summary,
-            aes(x = 1950-age, y=hpdi_l, col = "Estimation Result"), lty=2) + 
+            aes(x = (1950-age) * -1, y=hpdi_l, col = "Estimation Result"), lty=2) + 
   geom_line(data = PopDens_summary,
-            aes(x = 1950-age, y=hpdi_u, col = "Estimation Result"), lty=2) +
-  theme_minimal() + 
+            aes(x = (1950-age) * -1, y=hpdi_u, col = "Estimation Result"), lty=2) +
+  theme_minimal() + scale_x_reverse() + 
   scale_colour_manual(
@@
 popvar_plot <- ggplot(var_popdens) +
-  geom_line(aes(x = 1950-age, y = var_coeff/coeff, col="Variation Coefficient"), alpha = 0.5) +
+  geom_line(aes(x = (1950-age) * -1, y = var_coeff/coeff, col="Variation Coefficient"), alpha = 0.5) +
   geom_line(data = PopDens_summary,
-            aes(x = rev(1950-all_proxies_orig$age), y=mean, col = "Estimation Result")) + 
+            aes(x = rev((1950-all_proxies_orig$age)*-1), y=mean, col = "Estimation Result")) + 
   geom_line(data = PopDens_summary,
-            aes(x = rev(1950-all_proxies_orig$age), y=hpdi_l, col = "Estimation Result"), lty=2) + 
+            aes(x = rev((1950-all_proxies_orig$age)*-1), y=hpdi_l, col = "Estimation Result"), lty=2) + 
   geom_line(data = PopDens_summary,
-            aes(x = rev(1950-all_proxies_orig$age), y=hpdi_u, col = "Estimation Result"), lty=2) +
-  theme_minimal() + 
+            aes(x = rev((1950-all_proxies_orig$age)*-1), y=hpdi_u, col = "Estimation Result"), lty=2) +
+  theme_minimal() + scale_x_reverse() + 
   scale_colour_manual(

### 2022-05-18 manuscript/paper.Rmd
also reknitted pdf
included comments and corrections of Jan

@@
 Size and density of population are crucial factors for the reconstruction of group sizes, the character and range of political institutions for social organization, the constraints and possibilities of economic practice, the number of people available for collective activities, for economic and social exchange systems, and the creation and maintenance of collective identities.
-And, of course, the human impact on the environment.
+And, of course, the human influcence on the environment.
 Most estimates of population sizes currently work on limited spatial scales or use proxies that hardly allow for absolute numbers (e.g.
@@
 At the end of the 1960s, however, with the emergence of the processual archaeology, such investigations became increasingly popular.
-It was the rise of the ecological paradigm that gave even greater weight to population studies, especially those dealing with population ecology [@hassan_demographic_1981].
+It was the increased focus on human-environment relations paradigm that increased the emphasis on population studies, especially those dealing with population ecology [@hassan_demographic_1981].
 
-After research interest turned increasingly to other aspects with the emergence of post-process trends, a strong boom in the study of population sizes can be observed especially since the 2010s [or even earlier, @shennan_population_2000].
+After research interest turned increasingly to other aspects with the emergence of post-process trends, a strong boom in the study of population sizes can be observed especially since the 2010s [@riede_climate_2009 and others in that special issue of Human Biology, or even earlier, @shennan_population_2000].
 This is probably not at least due to the new shift towards the study of human-environment relationships, for which an assessment of the size of human populations is indispensable.
 This is true especially for an evaluation of the human impact on the natural environment.
-Another reason is certainly the emergence of the 'dates as data' approach [@rick_dates_1987], which only really took off with the publications around the London UCL group.
+Another reason is certainly the emergence of the 'dates as data' approach [@rick_dates_1987], which only really took off with the publications around the London UCL group [@shennan_regional_2013], but also others [].
 Moreover, Kintigh et al.
 [-@kintigh_grand_2014] listed human influence, dominance and population size and growth as one of the key elements where methodological improvements are necessary in the article on the Grand Challenges for archaeology, in the area of 'human-environment interactions'.
+Methodology continues to evolve, but we believe that essential problems underlying summation calibration cannot be solved by methodological refinements in this area alone [@french_manifesto_2021, @schmidt_approaching_2021].
 
@@
 
-Müller and Diachenko [@muller_tracing_2019] has compiled a list of currently used approaches and proxies for estimating prehistoric population sizes.
+Müller and Diachenko [@muller_tracing_2019] have compiled a list of currently used approaches and proxies for estimating prehistoric population sizes.
 These can be roughly divided into ethnographic parallels, ecologic-economic deductive estimates, and the counting and spatial interpolation of archaeological features (settlements, houses, individual find types).
@@
 In recent years, a new method has been established in the field of demographic estimations.
-Bayesian demographic forecasting has been designed to solve all these problems, which also exist for demographic estimations using recent data.
+The Bayesian population forecast was developed to provide a solution to these problems, which also exist in demographic estimates with current data.
 In their book on the subject, Bryant and Zhang [@bryant2018] state that Bayesian data modeling can be used to deal with exactly those problems that also affect archaeological data.
@@
 This is well-accepted in archaeological applications through the Bayesian calibration and modelling of stratigraphic conditions and radiometric data.
-Furthermore, spatially and temporally missing data can be taken into account: Where these data are missing, the uncertainty automatically increases, but this does not prevent general modelling and estimation.
+Furthermore, spatially and temporally incomplete data can be taken into account: Where these data are missing, the uncertainty automatically increases, but this does not prevent general modelling and estimation.
 Finally, hierarchically structured model suites, in which sub-models are created for the utilisation of individual proxies, can also be used to estimate transfer functions between individual proxies and the value to be modelled, thanks to the interaction of a large number of data sources and evidence.
@@
 It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites.
-The dates were selected so that their distribution area coincides with the catchment area of the pollen proxy explained below (see also Figure \@ref(fig:c14map)).
+The dates were selected so that their distribution area coincides with the catchment area of the pollen proxy (see also Figure \@ref(fig:c14map)).
 The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` BP uncal.
@@
 
-ggplot(joined_data) + geom_line(aes(x=1950-age, y=scale(sumcal, center = F), color = "darkred"), alpha = 0.75)  + geom_line(aes(x=1950-age, y=scale(dendro, center = F), color = "darkblue"), alpha = 0.75) + theme_minimal()  +
+ggplot(joined_data) + geom_line(aes(x=(1950-age)*-1, y=scale(sumcal, center = F), color = "darkred"), alpha = 0.75)  + geom_line(aes(x=(1950-age)*-1, y=scale(dendro, center = F), color = "darkblue"), alpha = 0.75) +
+  theme_minimal() + scale_x_reverse() +
     labs(x = "years BCE",
@@
 
-In order to add another indicator of archaeological evidence of occupation, we have included the data of the Cantonal Archaeology (Figure \@ref(fig:aoristmap)), and thus the Heritage Management, which are primarily derived from scattered surface finds, and which often have a low depth of information and thus dating accuracy.
+In order to add another indicator of archaeological evidence of occupation, we have included the data of the Cantonal Archaeology (Figure \@ref(fig:aoristmap)), and thus the Heritage Management, which are primarily derived from scattered surface finds, and which often have a low dating accuracy.
 This information is incorporated into our model as a typologically dated aorist time series.
-Although the dating accuracy is very low, the advantage here lies in the fact that we are not bound to the conditions and problems of radiocarbon dates, and thus involved in the issue of sum calibration.
-Much more, these data provide an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by the same transmission filters and archaeological conditions as the evaluation of ^14^C data.
+The dating accuracy is only in the range of archaeological periods, but the advantage is that we are not bound to the conditions and problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided.
+Much more, these data provide an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data.
 Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy in archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.
@@
 ```{r aoristcurve, fig.cap="Aoristic sum of the archaeological sites used in this analysis."}
-ggplot(joined_data) + geom_line(aes(x=1950-age, y=aoristic_sum, color = "darkgray"), alpha = 0.75) +
-  theme_minimal()  +
+ggplot(joined_data) + geom_line(aes(x=(1950-age)*-1, y=aoristic_sum, color = "darkgray"), alpha = 0.75) +
+  theme_minimal()  +  scale_x_reverse() +
     labs(x = "years BCE",
@@
 The utilisation of this proxy is based on the assumption that the higher the population density in an area, the greater the human influence on the natural environment [@lechterbeck2014], and that the agricultural space of an area is closely related to the population density [@zimmermann2004].
-Evidence of land clearing in pollen diagrams can therefore give further indications regarding population dynamics.
+Evidence of land clearing in pollen diagrams can therefore provide further indications of population dynamics where humans are the main driver of this process, which is the case in much of Europe.
 The exact methodological procedure for obtaining this proxy from several, in our case 5, different pollen diagrams of sites mainly in the hinterland of the large Alpine lakes can be found in a previous publication [@heitz2021].
@@
 
-The implementation we introduce in this paper represents a Poisson regression in which the indicators used inform the change in the number of settlements from time step to time step.
+The implementation we introduce in this paper represents a Poisson regression where the proxies are used to inform the change in the number of settlements from time step to time step.
 For this purpose, the individual proxies were z-normalised.
@@
 Thus, our model does not correspond to a classical state space model, where the measured values are each considered to be error-prone.
-In the implementation, the model finds the best possible combination or compromise between the individual indicators to describe a settlement dynamic that is given by them.In addition, the number of sites is converted into population density using some (certainly debatable) parameters that we have defined but which are only scaling factors for the intermediate value of number of settlements.
+In the implementation, the model finds the best possible combination or compromise between the individual proxies to describe a settlement dynamic that is given by them.
+In addition, the number of sites is converted into population density using some (certainly debatable) parameters that we have defined but which are only scaling factors for the intermediate value of number of settlements.
 For this, we assume that each site represents a number of people that is poisson distributed around the value 50.
@@
 
-ggplot(expert_estimations) + geom_segment(aes(x=from, xend = to, y = estimate, yend = estimate, color = source)) + xlab("cal BCE") + ylab("Density p/km²") + theme_minimal()
+ggplot(expert_estimations) + geom_segment(aes(x=from * -1, xend = to * -1, y = estimate, yend = estimate, color = source)) + xlab("cal BCE") + ylab("Density p/km²") + theme_minimal() + scale_x_reverse()
 ```
@@
 
-```{r popdensplot, fig.cap="The estimate of the population density resulting from the model and the three proxies, which are also plotted (scaled) for comparison."}
+```{r popdensplot, fig.cap="The estimate of the population density resulting from the model and the four proxies, which are also plotted (scaled) for comparison."}
 knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popdens_estimation.pdf"))
@@
 
-```{r popestwithphases}
+```{r popestwithphases, fig.cap="The estimate of the population density in relation to the chronology for the Swiss Plateau."}
 PopDens_summary <- read.csv(
@@
                       )
-            )
+            ), row.names = 1
   )
@@
 ggplot(PopDens_summary) +
-  geom_line(aes(x = 1950-age, y = mean)) + 
-  geom_line(aes(x = 1950-age, y = hpdi_l), lty=2) + 
-  geom_line(aes(x = 1950-age, y = hpdi_u), lty=2) + 
-    theme_minimal() + 
+  geom_line(aes(x = (1950-age)*-1, y = mean)) + 
+  geom_line(aes(x = (1950-age)*-1, y = hpdi_l), lty=2) + 
+  geom_line(aes(x = (1950-age)*-1, y = hpdi_u), lty=2) + 
+    theme_minimal() + scale_x_reverse() +  
 ylab("Scaled values for Proxies,\n P/km² for Estimation") +
   xlab("cal BCE") +
-  geom_rect(data=chronology[-1,],aes(xmin=dat_begin,ymin=0,xmax=dat_end,ymax=Inf,fill=label),
+  geom_rect(data=chronology[-1,],aes(xmin=dat_begin*-1,ymin=0,xmax=dat_end*-1,ymax=Inf,fill=label),
                     alpha=0.5,inherit.aes=FALSE) +
@@
   geom_text(data=chronology[-1,],
-            aes(x=dat_end - 100,y = 0.25,label=label),
+            aes(x=(dat_end - 100)* -1,y = 0.25,label=label),
             alpha=0.25,inherit.aes=FALSE, angle = 90, hjust = 0)
@@
 In the second half of the Early Bronze Age, from which we again know lakeshore settlements, there is again a clear increase in population size according to the model, which continues until the Late Bronze Age.
+The general trends fit very well with the previous reconstructions of population development for Switzerland [see eg.
+@lechterbeck2014].
 All in all, the estimate of the model corresponds to our expectations, although we must be aware that our expectations are not ground truth.
@@
 So far, we have trusted the individual indicators in their aggregation without modelling them with a measurement error, which they undoubtedly inherit.
-Moreover, our estimates are based on a limited number of indicators, almost all of which are subject to preservation conditions except for the openness indicator.
+Moreover, our estimates are based on a limited number of indicators, almost all of which are subject to archaeological preservation conditions except for the openness indicator.
 Consequently, a transformation of the results into the range of absolute numbers for population density can only be achieved by means of predefined parameters for upper limit of population growth, settlement size and the initial value of the reconstruction.
@@
 In the Swiss Plateau, for example, there are too few data on human remains over large areas of prehistory to be able to integrate them meaningfully into a model.
+Also, for different regions proxies used in this paper (eg.
+the palynological openness indicator) have to be adapted to fit the local conditions.
 By means of large-scale modelling, however, it would be possible to supplement gaps in the data of individual regions with regions in which these data are available by means of regionalisation and partial transfer of information (partial pooling).

### 2022-05-19 manuscript/paper.Rmd
Suggest new title and alternatives

@@
 ---
-title: "Bayesian Hierarchical Model for estimating holocene prehistoric population development on the Swiss Plateau"
+title: "Bayesian inference of prehistoric demography from multiple proxies: a case study from the Swiss plateau"
 author:
@@
 ```
+
+<!--
+[JR] suggested alternative titles:
+* A Bayesian hierarchical model of prehistoric population dynamics on the Swiss Plateau
+* Multi-proxy estimation of prehistoric population dynamics on the Swiss Plateau using Bayesian hierarchical modelling
+* Combining multiple proxies to estimate prehistoric population dynamics on the Swiss Plateau using Bayesian hierarchical modelling
+* A high-resolution estimate of prehistoric population dynamics on the Swiss Plateau using multi-proxy Bayesian hierarchical modelling
+* Prepend "Beyond summed radiocarbon:" to any of the above
+* Bayesian inference of prehistoric demography from multiple proxies: a case study from the Swiss plateau
+-->
 

### 2022-05-19 manuscript/paper.Rmd
Edits and suggestions to introduction

@@
 
-## Prehistoric demography as object of research
+## Prehistoric demography
 
-The study of population size and population density as important demographic factors has a long history in archaeology.
-Although questions of population extent and distribution have been dominant fields since the beginning of archaeological research, the lack of access to this information, data on it, or methodological tools for the use of proxies has long made the statements made on this topic for prehistoric times vague and superficial [@hassan_demographic_1981], and they were often based on uncritical transfer of ethnographic parallels.
+Prehistorians have long recognised demography as a fundamental force in human cultural evolution [@childe_man_1936].
+However, despite decades of interest in the demographic dynamics of prehistoric societies, concrete estimates of population size and population density before written records remain elusive.
+Though the archaeological record provides multiple possible demographic proxies [@hassan_demographic_1981], lack of access to this data, and a lack of methodological tools for turning it into quantitative estimates, has often left the conclusions drawn from it vague and superficial [@hassan_demographic_1981].
+As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence.<!--REF?
+Turchin?
+Morris?-->
 
-Gordon Childe [-@childe_man_1936] was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution [@hassan_demographic_1981].
+Early attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack [-@hack_changing_1942], Colton [-@colton_prehistoric_1949], Frankfort [-@frankfort_town_1950] and Cook [-@cook_reconsideration_1946].
+Cook and Heizer (1965, 1968) and Naroll (1962)<!--Are these in the bibliography?--> provided rules derived from ethnographic contexts for estimating the size of prehistoric populations.
+Cook and Heizer (1966) also made the first attempt to estimate the population growth rate during the Neolithic [@hassan_demographic_1981].
+These approaches became increasingly common after the 1960s, with the emergence of processual archaeology and its focus on human–environment relations and population ecology [@hassan_demographic_1981].
 
-Attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack [-@hack_changing_1942], Colton [-@colton_prehistoric_1949], Frankfort [-@frankfort_town_1950] and Cook [-@cook_reconsideration_1946].
-Cook and Heizer (1965, 1968) and Naroll (1962) provided rules derived from ethnographic contexts for estimating the size of prehistoric populations.
-Cook and Heizer (1966) made the first attempt to estimate the population growth rate during the Neolithic [@hassan_demographic_1981].
+Following a spell out of fashion with the post-processual turn, prehistoric demography has experienced a revival in recent years as part of the cultural evolution paradigm [@shennan_population_2000; @riede_climate_2009 and others in same issue].
+Part of this resurgence can be explained by a renewed interest in human–environment relations and human impact on the natural environment, which necessarily requires an assessment of population size.
+Kintigh et al.
+[-@kintigh_grand_2014], for example, list human influence, dominance, population size, and population growth amongst their "grand challenges" for archaeology in the 21st century.
+But methodological advances have certainly also played a role.
+In the last decade, the 'dates as data' technique [@rick_dates_1987], which uses the frequency of radiocarbon dates as a proxy for population dynamics, was taken up and significantly developed in a series of publications from the UCL group [e.g.
+@shennan_regional_2013], and has since been widely applied to archaeological contexts around the world [@crema_review_2022].
 
-At the end of the 1960s, however, with the emergence of the processual archaeology, such investigations became increasingly popular.
-It was the increased focus on human-environment relations paradigm that increased the emphasis on population studies, especially those dealing with population ecology [@hassan_demographic_1981].
+The 'dates as data' or summed radiocarbon approach has contributed greatly to our understanding of prehistoric demography, but it is not without its critics [@attenbrow_hiscock_2015, @carleton_groucutt_2020].
+While the methodology continues to evolve and to an extent accommodate these critiques [@crema_review_2022], there are certain essential problems that cannot be solved with methodological refinements in this area alone [@french_manifesto_2021, @schmidt_approaching_2021].<!-- [JR] What problems?
+-->
 
-After research interest turned increasingly to other aspects with the emergence of post-process trends, a strong boom in the study of population sizes can be observed especially since the 2010s [@riede_climate_2009 and others in that special issue of Human Biology, or even earlier, @shennan_population_2000].
-This is probably not at least due to the new shift towards the study of human-environment relationships, for which an assessment of the size of human populations is indispensable.
-This is true especially for an evaluation of the human impact on the natural environment.
-Another reason is certainly the emergence of the 'dates as data' approach [@rick_dates_1987], which only really took off with the publications around the London UCL group [@shennan_regional_2013], but also others [].
-Moreover, Kintigh et al.
-[-@kintigh_grand_2014] listed human influence, dominance and population size and growth as one of the key elements where methodological improvements are necessary in the article on the Grand Challenges for archaeology, in the area of 'human-environment interactions'.
-Methodology continues to evolve, but we believe that essential problems underlying summation calibration cannot be solved by methodological refinements in this area alone [@french_manifesto_2021, @schmidt_approaching_2021].
-
-## Issues associated with existing approaches
+## Evaluation of existing approaches
 

### 2022-05-23 manuscript/paper.Rmd
Clarify our approach as an extension of SPD

@@
 The 'dates as data' or summed radiocarbon approach has contributed greatly to our understanding of prehistoric demography, but it is not without its critics [@attenbrow_hiscock_2015, @carleton_groucutt_2020].
-While the methodology continues to evolve and to an extent accommodate these critiques [@crema_review_2022], there are certain essential problems that cannot be solved with methodological refinements in this area alone [@french_manifesto_2021, @schmidt_approaching_2021].<!-- [JR] What problems?
--->
+While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to certain fundamental problems common to all approaches that rely on a single proxy [@french_manifesto_2021, @schmidt_approaching_2021].
+We believe that these problems cannot be overcome by methodological refinements in this area alone.
+Instead, the Bayesian approach we develop in this paper offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon.
 

### 2022-05-23 manuscript/paper.Rmd
Restructure section headings
* Created a background section
* Split materials and methods and gave them more informative titles
* Made case/terminology consistent between headings
Remove old comments

@@
 
-## Evaluation of existing approaches
+# Background
+
+## Existing approaches to population estimation
 
@@
 
-## Strengths of a hierarchical Bayesian approach
+## Hierarchical Bayesian demographic models
 
@@
 
-# Materials and Methods
-
-## Background
+## Assumptions of a Bayesian approach
 
@@
 
-## Data/Proxies
+# <!-- Materials -->Population proxy data
 
@@
 A very active and efficient archaeological research and heritage management also offers a rich repository of archaeological data, known sites and ^14^C data.
+
+```{r mapswissplateau, fig.cap="Location and extent of the Swiss Plateau."}
+main_map <- my_basemap +
+  geom_sf(data = fc, inherit.aes = FALSE, alpha = 0.33, fill="darkred", color=NA) +
+  blank() +
+  my_scalebar
+
+inset_map <- europe_overview_map +
+  geom_sf(data = switzerland_bb, fill = NA, color = "darkred", inherit.aes = FALSE) +
+  blank()
+
+ggdraw() +
+  draw_plot(main_map) +
+  draw_plot(inset_map, x = 0, y = 0.7, width = 0.3, height = 0.3)
+```
 
@@
 
-```{r mapswissplateau, fig.cap="Location and extent of the Swiss Plateau."}
-main_map <- my_basemap +
-  geom_sf(data = fc, inherit.aes = FALSE, alpha = 0.33, fill="darkred", color=NA) +
-  blank() +
-  my_scalebar
-
-inset_map <- europe_overview_map +
-  geom_sf(data = switzerland_bb, fill = NA, color = "darkred", inherit.aes = FALSE) +
-  blank()
-
-ggdraw() +
-  draw_plot(main_map) +
-  draw_plot(inset_map, x = 0, y = 0.7, width = 0.3, height = 0.3)
-```
-
-### Dendro Dated Lakeshore Settlements
+## Dendro-dated lakeshore settlements
 
@@
 
-### ^14^C sum calibration
+## Summed radiocarbon
 
@@
 
-### Aoristic Sum
+## Aoristic sum
 
@@
 
-<!-- JR: At the resolution the figure is rendered at, I don't think the 1 km jitter is perceptible. MH: But in the data table that accompanies the repository, they would be available -->
-
-```{r aoristmap, fig.cap="The location of the sites from the find reports of the cantonal archaeology (heritage management). The locations are fuzzified by ~1km."}
+```{r aoristmap, fig.cap="The location of the sites from the find reports of the cantonal archaeology (heritage management). The locations are 'fuzzed' by ~1km."}
 my_basemap + geom_point(data = aorist_latlng_coords, aes(x = X, y = Y), color = "black", alpha = 0.75) + blank() + my_scalebar
@@
 
-### Openness Indicator
+## Landscape openness
 
@@
 
-## Process Model
+# <!-- Methods -->Bayesian model
+
+## Process model
 
@@
 
-Thus, we can define the limits, the $max\_change\_rate$<!-- JR: no Greek letter?
-:( MH: I could invent one, if you like?
-;-) --> as an time independent variable, again without already specifying them with fixed values at this stage.
+Thus, we can define the limits, the $max\_change\_rate$ as an time independent variable, again without already specifying them with fixed values at this stage.
 The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively.
@@
 
-## Observational Model
+## Observational model
 
@@
 ```
-
-<!-- JR: I really like the idea of discussing "dead ends" in the model design process – very modern and reflexive!
-Maybe hightlight these in a section of their own?
-MH: I do not feel that this part would provide enough material to justify an own section.
--->
 
@@
 # Results
-
-<!-- JR: I really think we need a section evaluating and validating the model fit before we discuss its output.
-MH: that was the intention of the Model fitting section before, but I think this rather belongs to methods than to results!?-->
 
@@
 
-# Conclusion and Outlook
+# Conclusions
 

### 2022-05-23 manuscript/paper.Rmd
Copyedit background section

@@
 
-Müller and Diachenko [@muller_tracing_2019] have compiled a list of currently used approaches and proxies for estimating prehistoric population sizes.
-These can be roughly divided into ethnographic parallels, ecologic-economic deductive estimates, and the counting and spatial interpolation of archaeological features (settlements, houses, individual find types).
-From this short literature review three basic problems are common to all these approaches and the studies based on them:
+Müller and Diachenko [@muller_tracing_2019] have compiled a list proxies currently used for the estimation of population size in prehistory.
+These can be roughly divided into three groups: ethnographic analogies; deductive estimates from ecological/economic factors; and the interpolation of frequencies of archaeological features (e.g.
+settlements, structures, individual finds).
+Three basic problems are common to all these approaches:
 
 1.
-Reliance on single line of evidence: Most investigations are using only one proxy.
-Although multiproxy approaches exist, in such cases the individual proxies only serve to support each other or the proxy used in the main case.
-There is no combination of results to increase the accuracy of each of the inaccurate and biased proxies.
+**Reliance on single proxy**: Most investigations are using only one source of evidence.
+Although multi-proxy approaches exist, in such cases the individual proxies only serve to support each other or the proxy used as the main estimator.
+There is no explicit combination of results to reduce the inaccuracy and biases inherent to single proxies.
 2.
-Uncertainty in measurements: In most cases, individual curves are presented as estimates, but the size of the error bars for these estimates is almost never apparent.
-This means, that the uncertainty of evidence and derived quantities is not adequately reflected.
+**Uncertainty in measurements**: All archaeological evidence comes with an inherent uncertainty which is carried through to derived measurements.
+However, in most studies of prehistoric population, individual curves are presented as estimates, and the potential error associated with these estimates is almost never specified.
 3.
-Lack of transfer functions: What is meant here is that they take little account of the inherent uncertainties of the proxies themselves, for lack of an evaluative framework, and that transfer functions from the proxy to the actual size to be observed, i.e.
-population size or density, are hardly ever evaluated more intensively.
-That means, that the nature of the archaeological data is taken into account only insufficiently.
-In the best case, a critical appraisal of the informative value is provided, but a quantification of the relationship between, for example, a doubling of a proxy value and the change in population density does not take place and cannot take place for lack of a suitable framework or external data for calibration.
+**Lack of a transfer function**: By 'transfer function', we mean something that allows for the proxy data to be interpreted in terms of actual population size or density.
+This could be absolute, i.e.
+a numerical estimate of population, or relative, i.e.
+a means of scaling changes in the proxy value to changes in population.
+Lack of a suitable frameworks and 'calibration' data means that this is rarely presented alongside proxy estimates.
+In the best case scenario, there will be a qualitative assessment of the informative value of the proxy, but this does not sufficiently account for the complex nature of archaeological data.
 
-Archaeological data used to estimate population trends have the following characteristics, leading to these problems:
+Furthermore, the types of archaeological data commonly used as population proxies share a number of problematic characteristics. They are:
 
--   Limited: We have only incomplete data that can be used for these purposes, and they are usually very low in information value.
--   Unevenly distributed: Although there is a good data on settlement frequencies for some regions, and these are sometimes in very high temporal resolution, such favorable situations are very unevenly distributed over time and space.
--   Noisy: Often other influences, not related to population development, affect the characteristics of individual proxies.
-Preservation conditions, depositional behavior or social aspects should be mentioned here as an example.
--   Unreliable: Comprehensive information and derived data are influenced by research strategies, research history and financial capability of the individual research projects.
-Systematic distortions are possible or rather the rule.
--   Highly heterogeneous: The number of artefacts at a site, the number of houses, the demographic conditions in cemeteries, the estimates of human impact based on pollen analyses: All these are data that may serve our purposes, are available in completely different spatio-temporal scales, granularities and information values, and simply represent technically very different data formats and scales.
--   Indirect: It is actually always proxy data that are collected as substitute values for the information that is actually needed (i.e.
-population density or size).
-The transfer functions that link the collected data with the data to be obtained are unknown and are at best established by observing the (also ethnological) presence.
-However, since living conditions, circumstances and economic parameters differ greatly between the present time and prehistory, these transfer functions can only be evaluated as a first approximation.
--   Contradicting: Particularly when considering several proxies, differences in transfer functions and data quality and noisiness inevitably lead to contrasts.
-However, this usually only leads to the rejection of a proxy for a certain time period, without mutually supporting or contradictory information leading to a quantitative evaluation of the proxy and its uncertainties.
+-   **Limited**: We have only incomplete data that can be used for these purposes, and they are usually not very informative.
+-   **Unevenly distributed**: For example, although there is a good data on settlement frequencies for some regions, and these are sometimes have a very high temporal resolution, these regions are very unevenly distributed over time and space.
+-   **Noisy**: Frequently individual proxies are strongly influenced by factors unrelated to population, for example taphonomic conditions or depositional biases.
+-   **Unreliable**: Research strategies, research history and varying levels of resources available to researchers strongly affect the nature of compiled datasets.
+As a result, systematic distortions are the rule rather than the exception.
+-   **Heterogeneous**: A wide range of proxies \@ref(tab:tableproxies) can potentially inform us about population dynamics, all with different spatio-temporal scales, granularity, information value, scales, and data formats.
+-   **Indirect**: We will never have direct data on prehistoric population; only proxy data that is thought to be a reasonable substitute for it.
+The transfer functions that link the proxy data with the desired quantity (population) are unknown.
+At best, they may be estimated by comparison with historical examples or the (ethnographic) present.
+However, since living conditions, circumstances and economic parameters differ greatly between the recent history and prehistory, these transfer functions can only be considered a rough first approximation.
+-   **Contradictory**: When considering several proxies, differences in transfer functions, data quality and noisiness inevitably lead to different results.
+The most common way to deal with this is to reject certain proxies for a certain time periods, but this is not performed on a consistent or quantitative basis.
+
+Here we argue that many, if not all, of these problems can be ameliorated through a) the explicit, quantitative integration of multiple proxies; and b) the use of a Bayesian approach to take account of and estimate uncertainty.
 
@@
 
-In recent years, a new method has been established in the field of demographic estimations.
-The Bayesian population forecast was developed to provide a solution to these problems, which also exist in demographic estimates with current data.
-In their book on the subject, Bryant and Zhang [@bryant2018] state that Bayesian data modeling can be used to deal with exactly those problems that also affect archaeological data.
-Especially for limited, unreliable and noisy data, Bayesian approaches are optimally suited.
-Various even contradictory data can be brought into a common framework and support each other.
-Similarly, unlikelihood and uncertainty of a model approach can be quantified.
-This is well-accepted in archaeological applications through the Bayesian calibration and modelling of stratigraphic conditions and radiometric data.
-Furthermore, spatially and temporally incomplete data can be taken into account: Where these data are missing, the uncertainty automatically increases, but this does not prevent general modelling and estimation.
-Finally, hierarchically structured model suites, in which sub-models are created for the utilisation of individual proxies, can also be used to estimate transfer functions between individual proxies and the value to be modelled, thanks to the interaction of a large number of data sources and evidence.
+Many of the problems with archaeological population proxies are also encountered in contemporary demography.
+In response, demographers have increasing turned to Bayesian methods to estimate and forecast contemporary population dynamics.
+For example, Bryant and Zhang [@bryant2018] consider Bayesian data modelling a solution to exactly the kind of problems that affect archaeological data.
+Specifically, Bayesian approaches are well suited for limited, unreliable and noisy data.
+Various sources of data, even contradictory data, can be brought into a common framework and used to support one another.
+These methods also provide a quantitative estimate of the likelihood and uncertainty of the model's resulting predictions (or in our case retro-dictions).
+Bayesian approaches are also capable of accounting for spatially and temporally incomplete data: where this data is missing, the uncertainty automatically increases, but this does not prevent general modelling and estimation.
+Finally, hierarchically-structured model suites, in which sub-models are created for each individual proxy, can be used to estimate transfer functions between individual proxies and the value to be modelled, thanks to the interaction of a large number of data sources and evidence.
 
@@
 
-Bayesian modelling techniques have also been used recently as a tool for hypothesis testing of demographic trends or underlying models based on ^14^C data.
-Most notably are the recently published papers by Crema and Shoda ([@crema2021a]).
-However, the approach taken there is clearly different from the one presented in this paper.In these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only.
-This is a clear step in the direction of a model-based, and thus scientific, analysis.
-However, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool.
+The well-established application of a Bayesian approach in radiocarbon calibration, where it is used to model radiometric uncertainty based on stratigraphic information, represents a very similar use case in archaeology.
+More recently, archaeologists have also used Bayesian modelling techniques as a tool for testing hypotheses relating to demographic trends or underlying models based on ^14^C data[e.g.
+@crema2021a].
+This approach differs from the one presented in this paper in that, in these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only.
+This is a clear step forward in the direction of a model-based, and thus scientific, analysis.
+However, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of the inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool.
 Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series.
 Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented.
-Therefore, we would like to better exploit the capabilities of Bayesian hierarchical models through a combination of inductive data analysis and model-passed data integration of different proxies.
+We would like to better exploit the capabilities of Bayesian hierarchical models through a combination of inductive data analysis and model-passed data integration of different proxies.
 
-This contribution now represents an attempt to make these techniques usable for archaeological reconstructions.
+Here, we attempt to make Bayesian hierarchical techniques usable for archaeological reconstructions.
 In addition to a presentation of the basics and possible procedures, we want to show in the following, in a reproducible and practical form using a case study, how Bayesian methods can also make a decisive contribution to a better assessment of population development.
@@
 
-The basis of Bayesian statistics is the premise that there is always some prior assumption about the probability of an event, even if it may be completely vague.
+The basis of Bayesian statistics is the premise that there is always some prior assumption about the probability of an event, even if it may be very rough.
 This assumption is then adjusted by observing the data.
@@
 In the logic of the basic Bayesian equation, these assumptions represent the prior (probability).
-In conjunction with the data that are included as the likelihood of the prior, a posterior (probability) results, which represents the Bayesian learning from data.
+In conjunction with the data that are included as the likelihood of the prior, a posterior (probability) results, which represents the Bayesian model learning from data.
 It is also in the nature of the approach that in real applications there is no point prediction, but in most cases a probability distribution for the prediction.
-Thus, we simultaneously obtained a result and an estimate of the confidence intervals, or better, the credibility interval given the data.
+Thus, we simultaneously obtained a result and an estimate of the confidence intervals, or rather the credibility interval, given the data.
 

### 2022-05-23 manuscript/paper.Rmd
Copyedit materials section

@@
 
-The Swiss Plateau (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland's territory and comprises the partly flat but largely hilly area between the Jura Mountains and the Alps.
-Its average altitude above sea level is between 400 and 600 m.
-Due to its situation as a basin, it represents a favourable area and is by far the most densely populated region of the Switzerland today.
-Although the Swiss Plateau is a basin, depending on the region it has a very diverse natural landscape.
-The Swiss Plateau was shaped by glaciers during the Ice Age, which led to the many lakes that characterise the landscape today, and which provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, as well as a rich pool of sources for vegetation reconstructions by means of pollen analyses.
+The Swiss Plateau (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland's territory and comprises the partly flat, but largely hilly area, between the Jura Mountains and the Alps.
+It averages an altitude of between 400 and 600 m a.s.l.
+It is a favourable area for settlement and agriculture and is by far the most densely populated region of the Switzerland today.
+Although the Swiss Plateau is a basin, depending on the region it has a very diverse natural landscape: shaped by glaciers during the Ice Age, the many lakes that characterise the landscape today also provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, as well as a rich pool of sources for vegetation reconstructions by means of pollen analyses.
 A very active and efficient archaeological research and heritage management also offers a rich repository of archaeological data, known sites and ^14^C data.
@@
 
-In principle, a large number of different data sources can be integrated into the overall model as observations, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created.
-A non-exhaustive list can be found in the following table \@ref(tab:tableproxies).
-For our analysis of the Swiss Plateau we used an open land indicator, an aoristic sum of sites based on typological dating, a sum calibration and the number of dendro dated lakeshore settlements in the Three Lakes area.
-The individual proxies will be discussed in more detail below.
+Our case study targets the period between 7000–1000 BCE.
+The lower limit of this window was chosen to avoid the so-called 'Halstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, which would cause difficulties for the ^14^C proxy data.
+The upper limit was chosen to coincide with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence before 7000 in the working area.
+
+In principle, a large number of different data sources can be integrated into the overall model as proxies, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created.
+\@ref(tab:tableproxies) provides a non-exhaustive list.
+For our case study of the Swiss Plateau, we used: a landscape openness indicator; an aoristic sum of sites based on typological dating; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes area.
 
@@
 
-With a mild climate compared to the Jura and the Alps and a dense network of water bodies, the Swiss Plateau is a favourable area for human settlement in Switzerland.
+With a mild climate compared to the Jura and the Alps, and a dense network of water bodies, the Swiss Plateau has always been a favourable area for human settlement in Switzerland.
 From the Neolithic onwards, settlement areas were concentrated along its rivers and lakes [@christianlüthi2009].
@@
 
-The dataset we use for the number of dendrodated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019].
+The dataset we use for the number of dendro-dated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019].
 Details on the creation of this data series will be published at the referenced location.
@@
 
-The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database, supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of [@martínez-grau2021].
+The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of [@martínez-grau2021].
 It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites.
 The dates were selected so that their distribution area coincides with the catchment area of the pollen proxy (see also Figure \@ref(fig:c14map)).
-The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` BP uncal.
-This time window extends beyond the study horizon in order to minimise boundary effects, which, however, cannot be completely avoided due to the Hallstatt plateau.
-But it was precisely to reduce its influence that the time window of the investigation was chosen to be 7000 -- 1000 BCE.
-The upper limit is set by the post-glacial changes in the pollen spectra, which are hardly associated with human influence before 7000 in the working area, and which would consequently distort the openness indicator.
-The lower limit is defined by the Hallstatt Plateau (with buffer), which would have direct and indirect research influences on the ^14^C proxy.
+The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` uncal BP.
+This time window extends beyond the study horizon in order to minimise boundary effects.
 
 We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites.
-For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [@crema2021] were used in the standard settings.
+For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.
 
@@
 
-In order to add another indicator of archaeological evidence of occupation, we have included the data of the Cantonal Archaeology (Figure \@ref(fig:aoristmap)), and thus the Heritage Management, which are primarily derived from scattered surface finds, and which often have a low dating accuracy.
-This information is incorporated into our model as a typologically dated aorist time series.
+To add another archaeological indicator of occupation, we include relative dating information obtained from the Swiss cantonal archaeology/heritage management authorities (Figure \@ref(fig:aoristmap)).
+These are primarily derived from scattered surface finds, which often have a low dating accuracy.
+We incorporate this data into our model as a typologically dated aorist time series.
 The dating accuracy is only in the range of archaeological periods, but the advantage is that we are not bound to the conditions and problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided.
-Much more, these data provide an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data.
-Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy in archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.
+Furthermore, this data provides an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data.
+Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy offered by archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.
 
@@
 
-The natural conditions provided by the many lakes enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis.
+The natural conditions of the lakes of the Swiss plateau enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis.
 We make use of this fact by generating a supra-regional openness indicator for the vegetation from the pollen data (Figure \@ref(fig:pollensites)).
-This indicator has the specific advantage that it is not dependent on preservation conditions, as archaeological indicators are.
+This proxy has the specific advantage that it is not dependent on preservation conditions, as archaeological indicators are.
 This makes it particularly valuable for indicating or compensating for systematic distortions that result from temporally specific settlement patterns and archaeological preservation conditions.
@@
 
-The utilisation of this proxy is based on the assumption that the higher the population density in an area, the greater the human influence on the natural environment [@lechterbeck2014], and that the agricultural space of an area is closely related to the population density [@zimmermann2004].
-Evidence of land clearing in pollen diagrams can therefore provide further indications of population dynamics where humans are the main driver of this process, which is the case in much of Europe.
-The exact methodological procedure for obtaining this proxy from several, in our case 5, different pollen diagrams of sites mainly in the hinterland of the large Alpine lakes can be found in a previous publication [@heitz2021].
-The technical procedure is also documented in the accompanying R-script.
-The percentage pollen data based on a pollen sum of all terrestrial taxa of the individual sites were combined into one data set by means of a principal component analysis (Figure \@ref(fig:pollenproxy)).
-Only terrestrial pollen taxa with a frequency of more than 1/3 and, if present, with an average frequency of at least more than 0.1% were selected to reduce potential disturbance by rare species.
+The utilisation of this proxy is based on the assumption that the higher the population density in an area, the greater the human influence on the natural environment [@lechterbeck2014], and that the extensiveness of agricultural activity in an area is related to human population density [@zimmermann2004].
+Evidence of land clearance in pollen diagrams can therefore provide further indications of population dynamics where humans can be assumed to be the main driver of this process, which is the case in much of Europe.
+The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2020].
+In this study, five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes were used.
+The technical steps are also documented in the accompanying R compendium.
+The percentage pollen data, based on a pollen sum of all terrestrial taxa of the individual sites, was combined into one data set by means of a principal component analysis (Figure \@ref(fig:pollenproxy)).
+Only terrestrial pollen taxa with a frequency of more than one third and, if present, with an average frequency of at least 0.1% were selected to reduce potential disturbance by rare species.
 Cereal pollen was explicitly retained as an important anthropogenic indicator.
-As each sample is absolutely dated, the data on the x-axis can be plotted against the openness value on the y-axis to obtain a time series time series for land clearing.
+As each sample is absolutely dated, the data on the x-axis can be plotted against the openness value on the y-axis to obtain a time series time series for land clearance.
 
@@
 
-Unfortunately, in the Swiss Plateau there is no data from burials that could be usefully incorporated into such this model due to the absence of regular burials for larger parts of prehistory.
+## Other proxies
+
+Unfortunately, there is no burial data available the Swiss plateau that could be usefully incorporated into such this model due to the absence of regular burials for larger parts of prehistory.
 Formation conditions, in this case most likely C-transformation [@schiffer1987], prevent their use in this study.
@@
 
-The values of the sum calibration, the openness index and the dendrodated settlements were smoothed by means of a moving average with a window of 50 years.
+# <!-- Methods -->Bayesian model
+
+The sum calibration, openness and the dendro-dated settlement data was smoothed by means of a moving average with a window of 50 years.
 Since the aoristic sum already had a very coarse temporal resolution, this was not applied for this measure.
@@
 
-In our following modelling (see also \@ref(observational-model)) we now consider all these proxies as determinants for the change in the number of settlements.
-In the model, we therefore change the causality and shift the measurement error, which is certainly inherent in each of these indicators, to the process model, in which a Poisson process describes the number of simultaneous settlements.
-In doing so, we establish a likelihood that indicates how credible the data are, given the model.
-
-# <!-- Methods -->Bayesian model
+In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements in the Swiss plateau.
+Consequently, we shift the causality and consideration of measurement error, which is certainly inherent in each of these proxies, to a preceding 'process model', in which a Poisson process describes the number of simultaneous settlements.
+In doing so, we establish a likelihood that estimates how credible the data is, given the model.
 

### 2022-05-24 manuscript/paper.Rmd
Copyedit results, discussion, conclusion

@@
 
-```{r compdendroc14, fig.cap="Comparison of the scaled (non-centered z-transformed) number of ^14^C and dendrodated sites over time in the dataset used."}
+```{r compdendroc14, fig.cap="Comparison of the scaled (non-centered z-transformed) number of ^14^C and dendro-dated sites over time in the dataset used."}
 joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))
@@
 The sum calibration, openness and the dendro-dated settlement data was smoothed by means of a moving average with a window of 50 years.
-Since the aoristic sum already had a very coarse temporal resolution, this was not applied for this measure.
+The aoristic sum was not smoothed, because it already has a very coarse temporal resolution.
 The range of the smoothing window corresponded to the sample interval, with which a unified resolution of 50 years was obtained for all proxies as time slices for the model.
-In addition, all data were restricted to the window of observation of 7000--1000 BCE.
+In addition, all data was restricted to the window of observation of 7000--1000 BCE.
 
@@
 These are specifically designed for time series and follow two basic principles.
-First, a hidden or latent process is assumed, which represents the state of the variable of interest $x_t$ at all times.
-For the course of the states of this variable over time, it is assumed that every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$.
+First, a hidden or latent process is assumed, which represents the state of the variable of interest $x_t$ through the entire time series.
+Over time, it is assumed that every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$.
 At the same time, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$.
@@
 
-This basic structure of the model makes it particularly useful and suitable for the purpose of demographic reconstruction using archaeological, but also other data, which depend on population density in the past.
+This basic structure of this class of model makes it particularly useful and suitable for the purpose of demographic reconstruction using archaeological and other data which are reflective of population density in the past.
 This population density itself is not accessible or measurable by our means.
@@
 These can be of very different natures, such as number of archaeologically observable settlements, or other effects that can be observed through time series, and which are influenced by population in a given area.
-In our example, these are the openness indicators from pollen data, which we can interpret primarily in terms of human influence and its intensity.
+In our case study, these are the openness indicators from pollen data, which we can interpret primarily in terms of human influence and its intensity.
 On a more abstract level, we could also include expert estimates, as these also happen on (often unspecified) bases that are at least indirectly influenced by past population density.
@@
 
-In this process model we assume that the latent variable number of sites is strongly autocorrelated across different time periods, i.e.
+In this process model we assume that the latent variable 'number of sites' is strongly autocorrelated across different time periods, i.e.
 the number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on.
 In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.
-This results in the following very simple formula:
+This gives us the basic formula:
 
@@
 
-A distribution that is particularly suitable for modelling frequencies is the Poisson distribution.
+The Poisson distribution is particularly suitable for modelling frequencies.
 It is a univariate discrete probability distribution that can be used to model the number of events that occur independently of each other at a constant mean rate in a fixed time interval or spatial area.
@@
 \end{aligned}
-$$ If we now have information about the change in population development (the proxies), we can use this to enter it into the model via a change in $\lambda$.
-This is done in the form of a regression: For all proxy values as a vector of independent variables $x \in R^n$, $R^n$ as an n-dimensional Euclidean space, described in this case by the n dimensions of the n variables, then the model takes the form
+$$
+
+If we now have information about the change in population development (the proxies), we can use this to enter it into the model via a change in $\lambda$.
+This is done in the form of a regression: for all proxy values—represented as a vector of independent variables $x \in R^n$, with $R^n$ as an n-dimensional Euclidean space, described in this case by the n dimensions of the n variables—then the model takes the form:
 
@@
 
-Logarithmisation as a link function ensures that $\lambda$, which must always be positive for a Poisson process, can also be described by variables (proxies) that range in the space of real numbers and can therefore also be negative.
+Using the logarithm as a link function ensures that $\lambda$, which must always be positive for a Poisson process, can also be described by variables (proxies) that range in the space of real numbers and can therefore also be negative.
 $\beta$ can serve here as a slope factor, just as in a normal linear regression.
@@
 Since $\lambda$ and $N$ are essentially in the same range (e.g.
-if $lambda=1$, the expected value for $N$ would also be 1), $N_{t-1}$ must also be logarithmised in the above formula to obtain the congruence of both values.
+if $lambda=1$, the expected value for $N$ would also be 1), $N_{t-1}$ must also be log-transformed in the above formula to obtain the congruence of both values.
 The values for population size $N_t$ as well as for population change $\lambda_t$ are time-dependent.
@@
 However, we can narrow down the structure of population change even further.
-We can assume that, considered overall over time, the population change the population change will not exceed certain limits over the entire time span, without it being possible here to specify this already.
+We can assume that, considered overall over time, the population change will not exceed certain limits over the entire timespan, though it is not possible to specify this at this point.
 
-Thus, we can define the limits, the $max\_change\_rate$ as an time independent variable, again without already specifying them with fixed values at this stage.
+Thus, we can define the limits, the $max\_change\_rate$ as a time independent variable, again without specifying them with fixed values at this stage.
 The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively.
@@
 The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$.
-This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model.
+This prevents the number of sites from explosively increasing between two time periods, which would be unrealistic given our sampling interval (50 years), and would lead to problems for the convergence of the model.
 The interaction of these parameters results in the following prior probability distribution for $\lambda$ and thus the growth (or change) rate of the population:
+
+<!-- Is there supposed to be something here?? -->
 
@@
 
-During the development of the overall model, we abandoned the implementation of dedicated observation models adapted to the conditions of the individual proxies and their generating processes.
-In previous implementations, the underdetermination by the currently usable data of the model with many degrees of freedom led to equifinality of the solutions and, thus, to a high path dependency of the individual model runs.
-Therefore, it was almost impossible to achieve convergence of the overall model.
-Nevertheless, we believe that for a future application of the model with more data, a larger geographical coverage and especially a regionalised approach with information transfer by means of partial pooling, this more specific approach will be feasible and a very useful approach.
+During the development of our model, we experimented with but ultimately abandoned the idea of implementing dedicated observation models adapted to the conditions of the individual proxies and their generating processes.
+We found that underdetermination of currently usable data in the model, having many degrees of freedom, produced a high degree of equifinality in the solutions and thus led to a high path dependency in individual model runs.
+As a result, it was almost impossible to achieve convergence of the overall model.
+Nevertheless, we believe that in future applications of the model with more data, a larger geographical coverage and/or, especially, a regionalised approach with information transfer by means of partial pooling, this more specific approach would be a feasible and a very useful approach.
 
-The implementation we introduce in this paper represents a Poisson regression where the proxies are used to inform the change in the number of settlements from time step to time step.
+The implementation we present in this paper instead represents a Poisson regression, where the proxies are used to inform the change in the number of settlements from time step to time step.
 For this purpose, the individual proxies were z-normalised.
@@
 In earlier implementations, expert estimates were also integrated into the model.
-However, since these are highly contradictory for the working area (Figure \@ref(fig:expertestimations), therefore have little influence on the model, and also lead to a significantly longer runtime, we have refrained from doing so in the current implementation.
-For future applications of the model with a larger geographical range and thus a higher information density, however, this would again be a factor that could be integrated in a useful way.
+However, since these are highly contradictory for the working area (Figure \@ref(fig:expertestimations), we found they had little influence on the model and also significantly increased its runtime.
+We therefore excluded them in the implementation presented here, but again, given a larger geographical range and thus a higher information density, this proxy could be usefully integrated into future applications.
 
@@
 
-The model was fitted using the R package nimble (version 0.11.1, R version 4.1.3).
+The model was fitted using the R package *nimble* (version 0.11.1, R version 4.1.3).
 For this purpose, 4 chains were run in parallel.
@@
 In a first run, the model was initialised for each chain and run for 100000 iterations (with a thinning of 10).
-The computer used (Linux, Intel(R) Xeon(R) CPU E3-1240 v5 \@ 3.50GHz, 4 cores, 8 threads) needed approx.
-1 min for this.
+On a reasonably capable computer (Linux, Intel(R) Xeon(R) CPU E3-1240 v5 \@ 3.50GHz, 4 cores, 8 threads), this takes approximately a minute.
 
 In a second step, the run was extended until convergence could be determined using Gelman and Rubin's convergence diagnostic, the criterion being that a potential scale reduction factor of less than 1.1 was achieved for all monitored variables.
-Convergence occurred after about 30 seconds.
+Convergence occurred after about thirty seconds.
 
-Due to the high correlation of the parameters and thus a low sampling efficiency, the collection of at least 10,000 effective samples for all parameters in the third step took about 5 hours.
+Due to the high correlation of the parameters and thus a low sampling efficiency, the collection of at least 10,000 effective samples for all parameters in the third step took about five hours.
 
@@
 
+__<!-- Clear syntax highlighting. Remove Joe! -->
+
 # Results
@@
 
-The population density estimate (Figure \@ref(fig:popdensplot)) ranges on average between 0.2 p/km² for the beginning of the estimate (6000 BCE) and 4.8 p/km² for the end of the estimate (1000 BCE), reaching a maximum of 6.5 p/km² for the time slice 1250 BCE.
+The population density estimated by the model (Figure \@ref(fig:popdensplot)) ranges on between 0.2 p/km² for the beginning of the estimate (6000 BCE) and 4.8 p/km² for the end of the estimate (1000 BCE), reaching a maximum of 6.5 p/km² for the time slice 1250 BCE.
 Thus, the estimate remains within the values that are also considered plausible by the expert estimates.
@@
 
-Furthermore, it may be relevant to look at the temporal distribution of the variability in the estimate (Figure \@ref(fig:varcoeffplot)), to estimate at which time steps the model attempts a more accurate estimate and at which the uncertainty is greater due to e.g.
-contradictions in the indicators.
+The temporal distribution of the variability in the estimate (Figure \@ref(fig:varcoeffplot)) allows us assess at which time steps the model attempts a more accurate estimate, and at which the uncertainty is greater due to e.g.
+contradictions in the proxies.
 The coefficient of variation is 0.13 for the beginning and 0.1 for the end of the estimate, the greatest variability is reached around 2150 BCE with 0.47.
 The beginning and end of the time series are relatively clearly determined.
-The end results from the a priori setting of the parameter, but also here as at the beginning of the series the proxies are very uniform, which explains the low variability.
+The end results from the *a priori* setting of the parameter, but also here, as at the beginning of the series, the proxies are very uniform, which explains the low variability.
 Overall, the variability is relatively uniform over the entire course of the estimation and averages over all time slices at 33% of the respective mean.
 
-Within the model, the parameter p was estimated, which reflects the proportion of the individual proxies used in the estimation of the number of settlements.
+Within the model, the parameter $p$ was estimated, which reflects the relative weight given to the individual proxies used in the estimation of the number of settlements.
 This parameter is variable, but has only a scaling influence on the final estimate of population density.
@@
 The next most important indicator is the sum calibration value, which has an average of about 20%.
-The aoristic sum is slightly above 10%, whereas the importance of the dendrodated settlements is below 10%.
+The aoristic sum is slightly above 10%, whereas the importance of the dendro-dated settlements is below 10%.
 The reason for the latter is certainly that there are no lakeshore settlements over large areas of the time window, and therefore the proxy achieves a low confidence value in comparison with the other estimators.
@@
 
-<!-- JR: I would split this into subsections evaluation a) the merits of the approach and b) the specific results for Switzerland.
-MH: I thought I already have done this, discussion more the results for switzerland, conclusion more the generel merrits!?
+Our original ambition in the development of this model was to base it on a state-space representation of the demographic process itself, and then to integrate the existing proxies by means of an observation model to inform this process.
+In the course of developing the model, we realised that the data we had available for the Swiss Plateau alone was not sufficient to adequately determine this process, or to adequately fix the degrees of freedom resulting from the process and the transfer model.
+Therefore, in implementation presented here, we have turned to a Poisson regression approach: the model represents the best possible combination of the indicators used to describe the development of the number of sites based on them.
+
+## Comparison of individual proxies
+
+Comparing the model's overall estimate with trends indicated by individual proxies provide gives us new insights into the quality of these records.
+Sum calibration, currently the most frequently used proxy for (relative) population changes in prehistory, has its large fluctuations dampened when considered alongside other proxies.
+This is especially true for the first fluctuation, shortly after 4000 BCE.
+The expected increase in archaeological remains with the onset of Neolithisation is still clearly visible, but the curve of the overall estimate, after the initial increase, is much flatter than sum calibration alone indicates.
+The period between 3950 and 3700, which is contemporaneous with the first settlement cycle on the lakeshores of the Three Lakes area, coincides with a noticeable plateau on the calibration curve, which be producing an overestimation of the ^14^C density.
+However, the effect of the calibration curve on the results of a cumulative calibration cannot yet be considered unambiguous.
+A second maximum, after 3000 BCE, is supported by the other proxies, and is consequently also much more clearly reflected in our overall estimate.
+Here, too plateau in the calibration curve — albeit a smaller, shorter one which is much less pronounced than, for example, the one shortly before between 3350 and 3100.
+The rise towards the Middle and Late Bronze Age is also supported by the other proxies, especially the aoristic sum, and therefore is preserved in the overall estimate.
+In this period, the calibration curve does not show any very clear, significant patterns.
+We may conclude that the model is successful in using information from other proxies to sift 'real' fluctuations in the summed radiocarbon record from artefacts of the calibration curve.
+
+On average, the model weights the sum calibration at about 20%, significantly less than the 60% afforded to the openness indicator.
+After an initial increase, which is easily explained by spread of agriculture, the openness indicator tends to fluctuate less and thus has a dampening effect on the overall estimate.
+Nevertheless, it appears the general trends in the sum calibration are well reflected in land openness – even if such 'eyeballing' should be interpreted with caution.
+In contrast to this dramatic trend, changes within the Neolithic and Bronze Age are more gradual, suggestive of larger scale transformations of the landscape.
+The model is designed rather conservatively by the limit of maximum growth, which is estimated within the model but influences it by its very presence.
+Therefore, this proxy corresponds better and more homogeneously to a smooth increase in the number of settlements than do the strong fluctuations in the ^14^C data set.
+
+The aoristic sum remains relatively even over long spans of time.
+It is not until the Middle and Late Bronze Age that we see a significant rise, which is also apparent in the model's overall estimate.
+It remains to be seen to what extent modelling of the taphonomic loss [@surovell2009] could be integrated in this approach.
+We have refrained from doing so in this first model, as this would have introduced further degrees of freedom - but we are aware that with a broader database this would be an interesting possibility, and that it would itself be a variable to be estimated, e.g.
+in connection with proxies that are not influenced by it (openness indicators, but also data from the demography of burial collectives).
+This would make it possible to estimate a value from original archaeological material independently of variables that have little to do with archaeological data, such as volcanic eruptions.<!-- JR: I don't follow...
+where do eruptions come in?
 -->
 
-The ambition in the development of the model was originally to base it on a state-space representation of the demographic process itself, and then to integrate the existing proxies by means of an observation model to inform this process.
-In the course of developing the model, however, we had to realise that the existing data or the data used in the modelling process for the Swiss Plateau alone were not sufficient to adequately determine this process or to adequately fix the degrees of freedom resulting from the process and the transfer model.
-Therefore, in this implementation we have turned to the Poisson regression approach.
-In this version, the model represents the best possible combination of the indicators used to describe the development of the number of sites based on them.
-
-In its present form, the reconstructed development already shows interesting aspects regarding the underlying proxies.
-Starting with the result of the sum calibration, which is currently the most frequently used indicator when attempting to reconstruct (relative) population changes.
-Its strong fluctuations are dampened by the connection of the other proxies, which is especially true for the first fluctuation around or shortly after 4000 BCE.
-The increase in the archaeological record with the process of Neolithisation is still clearly visible, but the curve of the overall estimate after the initial increase is much flatter than the sum calibration alone would suggest.
-The period between 3950 and 3700, which is contemporaneous with the first settlement cycle on the lakeshores of the Three Lakes area, in particular forms a noticeable plateau in the calibration curve, and this may lead to an overestimation of the ^14^C evidence.
-However, the effect of the calibration curve on the results of a cumulative calibration cannot yet be considered unambiguous.
-A second maximum, after 3000 BCE, is supported by the other proxies, and is consequently also much more clearly reflected in the overall estimate.
-Here, too, there is a smaller, albeit shorter plateau in the calibration curve, which is, however, much less pronounced than, for example, the one shortly before between 3350 and 3100.
-The rise towards the Middle and Late Bronze Age also receives support above all from the Aorist Sum, and therefore remains preserved in the combination of indicators.
-In this period, the calibration curve does not show any very clear, suspicious patterns.
-
-Overall, the model trusts the cumulative calibration with an average of about 20% significantly less than the openness indicator with 60%.
-After the initial increase, which can easily be linked to the spread of agricultural production, the latter tends to fluctuate less and thus has a dampening effect on the overall estimate.
-Nevertheless, by all appearances, the general trends in the cumulative calibration are quite well reflected in the land opening, even if such eyeballing should be interpreted with caution.
-We would suggest that the changes within the Neolithic and Bronze Age are rather gradual compared to the much more severe qualitative shifts to a productive economy as well as to the later developments outside our window of observation, which are suggestive of a much more large-scale reshaping of the landscape.
-The model itself is designed rather conservatively by the limit of maximum growth, which is estimated within the model but influences it by its very presence.
-Therefore, this proxy corresponds better and more homogeneously to a smooth increase in the number of settlements than do the strong fluctuations in the ^14^C data set.
-
-The aoristic sum remains relatively even over large areas of the observation window.
-It is not until the Middle and Late Bronze Age that their values rise significantly, which is also noticeable in the overall estimate by a clear increase.
-Overall, it remains to be seen to what extent a modelling of the taphonomic loss [@surovell2009] could be integrated.
-We have refrained from doing so in this first model, as this would have introduced further degrees of freedom - but we are aware that with a broader database this would be an interesting possibility, and that it would itself be a variable to be estimated, e.g.
-in connection with proxies that are not influenced by it (openness indicators, but also data from the demography of burial collectives).
-This would make it possible to estimate a value from original archaeological material independently of variables that have little to do with archaeological data, such as volcanic eruptions.
-
 The number of simultaneously existing lakeshore settlements is a very limited temporal and spatial estimator, but extremely reliable.
-Its limitations are reflected in the low overall confidence of the model, since its value is 0 over long stretches, while other indicators suggest clearly different patterns.
+Its limitations are reflected in the low overall confidence of the model, since its value is zero over long stretches, while other indicators suggest clearly different patterns.
 However, where it has information potential, such as around and shortly after 3800 BCE, 3200 BCE or especially around 2750 BCE, its fluctuations have a noticeable influence on the overall estimate.
@@
 
-In order to review the reconstruction against the background of established archaeological knowledge and narratives, we consider it useful to parallelise the archaeological phase boundaries with the changes in the estimate.
-It should be noted, however, that the estimate is not completely independent of the archaeological phases: due to the aoristic sum, which is itself strongly determined by this phase division, corresponding boundaries also influence the structure of the reconstruction.
+## Prehistoric population dynamics on the Swiss plateau
+
+In order to review the reconstruction against the background of established archaeological knowledge, it is useful to overlay conventionally-defined archaeological phase boundaries on the results of our model \@ref(fig:popestwithphases).
+It should be noted, however, that the model's estimate is not completely independent of these phases: due to the aoristic sum, which is itself strongly determined by this phase division, corresponding boundaries also influence the structure of the reconstruction.
 Nevertheless, it helps to check whether the estimate is in strong contradiction to the generally accepted picture or whether it is able to make a credible prediction within this context.
@@
 
-Figure \@ref(fig:popestwithphases) underlays the estimated development with the rough archaeological phases for the Swiss Plateau.
 The Older and Middle Neolithic phases are in fact hardly documented with known sites in Switzerland.
@@
 @lechterbeck2014].
-All in all, the estimate of the model corresponds to our expectations, although we must be aware that our expectations are not ground truth.
-Nevertheless, it can be said that the predictions of the model can be well integrated into the known narratives and make them more precise or provide them with a higher resolution, thus gaining credibility for the model predictions.
+All in all, the estimate of the model corresponds to our expectations, although we must be aware that these expectations are not 'ground truth'.
+Nevertheless, we can claim that the model predictions are credible because they accord well with existing knowledge, while offering higher precision and higher resolution.
 
@@
 
-An important result of the current model is the estimation of absolute population numbers with uncertainty over time.
-This estimate can be used as a basis for further studies where relative measures of population development are not helpful, such as long-term land use studies where modelling of large-scale socio-ecological systems based on archaeological data becomes possible and does have to rely on using deductive, asynchronous population models (eg.
+An key advance in the model we present here is the ability to estimate, in absolute terms, past population sizes and the uncertainty accompanying our present knowledge.
+These estimates can be used as a basis for further studies where relative measures of population development are not helpful, such as long-term land use studies, where modelling of large-scale socio-ecological systems based on archaeological data becomes possible and does have to rely on using deductive, asynchronous population models (e.g.
 carrying capacity or ethnographic analogues).
 
-With this model and its predictions, we also believe we can demonstrate how it is possible to achieve a true multiproxy analysis of prehistoric demographic processes.
-Previous attempts in this regard have so far only achieved a juxtaposition of different indicators, without the possibility of truly linking them or assessing the confidence in the individual indicators.
-Also, in previous approaches it was not possible to specify a confidence interval for the estimates.
-Within a Bayesian framework with an observational model based on Poisson regression, all these demands can be met.
-We thus believe we can provide a better basis than before for estimating population processes and changes in the number of settlements.
+We have also demonstrated that, with Bayesian hierarchical modelling, it is possible to achieve a true multi-proxy analysis of prehistoric demographic processes – as opposed to a juxtaposition of different indicators.
+This opens up the possibility of quantitatively linking different records and assessing their credibility.
+In addition, and in contrast to existing approaches, we are able to specifying a confidence interval for the overall estimate.
+The result is a firmer basis for reconstructing population dynamics and settlement patterns in prehistory.
 
-The currently presented model is only the first step towards a more sophisticated approach.
-So far, we have trusted the individual indicators in their aggregation without modelling them with a measurement error, which they undoubtedly inherit.
-Moreover, our estimates are based on a limited number of indicators, almost all of which are subject to archaeological preservation conditions except for the openness indicator.
-Consequently, a transformation of the results into the range of absolute numbers for population density can only be achieved by means of predefined parameters for upper limit of population growth, settlement size and the initial value of the reconstruction.
-In the future application and refinement of the model, this must be overcome.
-For this purpose, further proxies independent of the immediate temporally changing conditions of the archaeological record have to be integrated.
-These can be data on settlement sizes, parameters for economic-ecological carrying capacity, demographic data from burial groups as well as palaeogenetic data on population sizes.
-Other possible indicators can still be identified.
+Nevertheless, we consider the model present here as only the first step towards a more sophisticated Bayesian approach.
+We have trusted the individual proxies in aggregate, without considering measurement error.
+Moreover, our estimates are based on a limited number of sources of evidence, almost all of which are subject to taphonomic biases in the archaeological record (except for the openness indicator).
+Consequently, we can only transform the model's prediction into an absolute estimate of population density with predefined parameters: the upper limit of population growth, settlement size and the initial value of the reconstruction.
+Overcoming this limitation would represent a major refinement of our approach.
 
-These data are available to varying degrees in different regions.
-In the Swiss Plateau, for example, there are too few data on human remains over large areas of prehistory to be able to integrate them meaningfully into a model.
-Also, for different regions proxies used in this paper (eg.
-the palynological openness indicator) have to be adapted to fit the local conditions.
-By means of large-scale modelling, however, it would be possible to supplement gaps in the data of individual regions with regions in which these data are available by means of regionalisation and partial transfer of information (partial pooling).
-Such an extension would be the next logical step in the improvement of the model, to which we hope to be able to contribute a further study in the near future.
+Incorporating additional proxies that are independent of the immediate, time-dependent conditions of the archaeological record could be one way to achieve this.
+These could be data on settlement sizes, parameters for economic-ecological carrying capacity, demographic data from burial groups, archaeogenetic data on population sizes, or other records as yet unidentified.
+This data is available to varying degrees in different regions.
+On the Swiss plateau, for example, we have too little data on human remains over large spans of prehistory to be able to integrate them meaningfully into a model.
+
+To apply the model to different regions, the proxies used we use here (e.g.
+the palynological openness indicator) would have to be adapted to fit local conditions and research histories.
+By means of large-scale modelling, however, it would be possible to supplement gaps in the data in one region with data from other regions by means of regionalisation and a partial transfer of information (partial pooling).
+Such an extension would be the next logical step in the improvement of the model, to which end we hope to be able to contribute a further study in the near future.<!-- Maybe for tactical reasons we shouldn't spoil the sequel?
+:) -->
 

### 2022-05-26 manuscript/paper.Rmd

@@
 ---
-title: "Bayesian inference of prehistoric demography from multiple proxies: a case study from the Swiss plateau"
+title: "Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the Swiss plateau"
 author:
@@
       email: joe@joeroe.io
-      institute: [IAW, OCCR]
+      institute: [IAW]
       correspondence: false
@@
 abstract: |
-  In demography in general, the last decade has triggered an upswing in the application of Bayesian methods, so that a Bayesian demography has been announced.
-Bayesian demography is currently at the forefront of methodological developments in this area, but has reached such maturity that the United Nations has been using it since 2015 for population forecasts and projections.
-The Bayesian approach offers the possibility of combining heterogeneous data and at the same time qualifying them in terms of uncertainty and credibility.
-This is precisely where it becomes very interesting for archaeological data, since they are mostly inaccurate, biased, sparse, simplified and often based on simplistic assumptions.
-  In this contribution, a Bayesian hierarchical model, based on a Poisson Regression and a state space approach is applied to the field of archaeological (prehistoric) population size and density estimation.
-This represents a pilot study for the reconstruction of demographic developments in prehistory based on data assimilation with an underlying process model and a number of different proxies, which are combined and result in a more reliable estimation than a single proxy could deliver.
-  In order to understand human-environment relationships in the past and the dynamics of socio-ecological systems, it is essential to have a robust estimate of population sizes<!-- JR: Could reference Shennan 2000 and/or Kintigh 2014 here?
-MH: I would not like to have any citation in the abstract.
-What do you think?-->.
-Size and density of population are crucial factors for the reconstruction of group sizes, the character and range of political institutions for social organization, the constraints and possibilities of economic practice, the number of people available for collective activities, for economic and social exchange systems, and the creation and maintenance of collective identities.
-And, of course, the human influcence on the environment.
-Most estimates of population sizes currently work on limited spatial scales or use proxies that hardly allow for absolute numbers (e.g.
-sum calibration).
-The figures in the archaeological discourses on population density in general currently come from vague sources and are often decades old.
+  The archaeological record provides several potential proxies of human population, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage.
+Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely ona single proxy, and are rarely able to infer population size or density in absolute terms.
+In contemporary demography, it is becoming increasingly common to use Bayesian statistics to estimate population trends and project them into the future.
+The Bayesian approach is popular because offers the possibility of combining heterogenous data, and at the same time qualifying the uncertainty and credibility attached to forecasts.
+These same characteristics make it well-suited to applications to archaeological data in paleodemographic studies.
+  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density.
+Using the prehistoric Swiss plateau (7000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
+The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research.
+Our prediction of population development on the Swiss plateau accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed radiocarbon method.
+  Robust estimates of population are essential to the study of human–environment relations and socio-ecological dynamics in the past.
+Population size and density can directly inform reconstructions of prehistoric group size, social organisation, economic constraints, exchange, and political and social institutions.
+In this pilot study, we present an approach that we believe can be usefully transferred to other regions, as well as refined and extended to greatly advance our understanding of prehistoric demography.
 keywords: |
-  Bayesian demographic modelling; Multiproxy; Reconstruction of past settlement dynamics
+  Prehistoric demography; Bayesian modelling; Multi-proxy; Settlement dynamics
 highlights: |
-  - We provide a estimation based on the combination of different proxies.
-  - We estimate population development in terms of absolute settlement densities.
-  - We are developing a completely new method for the integration of heterogeneous information.
+  - Bayesian modelling can integrate multiple, heterogeneous population proxies from the archaeological record
+  - Our initial model produces more robust, high-resolution estimates of past population dynamics than previous, single-proxy approaches
+  - We provide absolute estimates of population size and density on the prehistoric Swiss plateau (7000–1000 BCE)
 ---
@@
 
-__<!-- Clear syntax highlighting. Remove Joe! -->
-
 # Results

### 2022-05-26 README.Rmd
small enhancement
updated readme

@@
 # Please put your title here to include it in the file below.
-Title <- "Estimation of Population Density for the prehistoric Swiss Plateau using a Hierarchical Bayesian Poisson Regression Model"
+Title <- "Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the Swiss plateau"
 ```
@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/MartinHinz/bayesian.demographic.reconstruction.2021/master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/MartinHinz/bayesian.demographic.reconstruction.2022/master?urlpath=rstudio)
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Hinz, M., Roe, J., Laabs, J., Heitz, C., Kolar, J. (2022). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Hinz, M., Roe, J., Laabs, J., Heitz, C., Kolar, J. (2022). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Hinz, M., Roe, J., Laabs, J., Heitz, C., Kolar, J., (`r format(Sys.Date(), "%Y")`).
+_Compendium of R code and data for `r Title`_.
+Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/xxx/xxx>
 
@@
     for the actual analysis. Includes code to reproduce the figures and tables
-    generated by the analysis.
+    generated by the analysis. You can look at the [Github-Markdown Version](/analysis/analysis.md)
+    for a rendered version of the analytical run incl. figures.
   - [:file\_folder: code](/code): Functions used in the analysis.
@@
 
+To perform the actual analysis, we recommend a powerful computer with a
+multi-core processor, ideally with the Linux operating system, which has
+at least 64GB RAM memory. In addition, a hard disk space of at least 5GB
+should be reserved for the process.
+
 ### Licenses
 
-**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)
-
-**Code :** See the [DESCRIPTION](DESCRIPTION) file
+**Text, code and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)
 

### 2022-05-26 README.Rmd
Updated Readme once more

@@
     for manuscript.
-  - [:file\_folder: paper](/paper): A rendered version, of the manuscript
-    `paper.docx` and `paper.pdf`, suitable for reading (the code is replaced by figures
+  - [:file\_folder: paper](/paper): A rendered version, of the manuscript as
+    [`paper.docx`](/paper/paper.docx) and [`paper.pdf`](/paper/paper.pdf), suitable for reading (the code is replaced by figures
     and tables in this file)

### 2022-05-26 manuscript/paper.Rmd
reknitted manuscript

@@
 abstract: |
-  The archaeological record provides several potential proxies of human population, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage.
-Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely ona single proxy, and are rarely able to infer population size or density in absolute terms.
-In contemporary demography, it is becoming increasingly common to use Bayesian statistics to estimate population trends and project them into the future.
-The Bayesian approach is popular because offers the possibility of combining heterogenous data, and at the same time qualifying the uncertainty and credibility attached to forecasts.
-These same characteristics make it well-suited to applications to archaeological data in paleodemographic studies.
   Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density.
@@
 In this pilot study, we present an approach that we believe can be usefully transferred to other regions, as well as refined and extended to greatly advance our understanding of prehistoric demography.
+  The archaeological record provides several potential proxies of human population, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage.
+Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely ona single proxy, and are rarely able to infer population size or density in absolute terms.
+In contemporary demography, it is becoming increasingly common to use Bayesian statistics to estimate population trends and project them into the future.
+The Bayesian approach is popular because offers the possibility of combining heterogenous data, and at the same time qualifying the uncertainty and credibility attached to forecasts.
+These same characteristics make it well-suited to applications to archaeological data in paleodemographic studies.
 keywords: |
@@
 
-<!--
-[JR] suggested alternative titles:
-* A Bayesian hierarchical model of prehistoric population dynamics on the Swiss Plateau
-* Multi-proxy estimation of prehistoric population dynamics on the Swiss Plateau using Bayesian hierarchical modelling
-* Combining multiple proxies to estimate prehistoric population dynamics on the Swiss Plateau using Bayesian hierarchical modelling
-* A high-resolution estimate of prehistoric population dynamics on the Swiss Plateau using multi-proxy Bayesian hierarchical modelling
-* Prepend "Beyond summed radiocarbon:" to any of the above
-* Bayesian inference of prehistoric demography from multiple proxies: a case study from the Swiss plateau
--->
-
 # Introduction
@@
 Though the archaeological record provides multiple possible demographic proxies [@hassan_demographic_1981], lack of access to this data, and a lack of methodological tools for turning it into quantitative estimates, has often left the conclusions drawn from it vague and superficial [@hassan_demographic_1981].
-As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence.<!--REF?
-Turchin?
-Morris?-->
+As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence [@morris_measure_2013, @turchin_seshat_2015].
 
 Early attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack [-@hack_changing_1942], Colton [-@colton_prehistoric_1949], Frankfort [-@frankfort_town_1950] and Cook [-@cook_reconsideration_1946].
-Cook and Heizer (1965, 1968) and Naroll (1962)<!--Are these in the bibliography?--> provided rules derived from ethnographic contexts for estimating the size of prehistoric populations.
+Cook and Heizer [-@cook_quantitative_1965, @cook_relationships_1968] and Naroll [-@naroll_floor_1962] provided rules derived from ethnographic contexts for estimating the size of prehistoric populations.
 Cook and Heizer (1966) also made the first attempt to estimate the population growth rate during the Neolithic [@hassan_demographic_1981].
@@
 [-@kintigh_grand_2014], for example, list human influence, dominance, population size, and population growth amongst their "grand challenges" for archaeology in the 21st century.
+
 But methodological advances have certainly also played a role.
@@
 
-# <!-- Materials -->Population proxy data
+# Materials: Population proxy data
 
@@
 Evidence of land clearance in pollen diagrams can therefore provide further indications of population dynamics where humans can be assumed to be the main driver of this process, which is the case in much of Europe.
-The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2020].
+The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2021].
 In this study, five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes were used.
@@
 
-# <!-- Methods -->Bayesian model
+# Methods: Bayesian model
 
@@
 This prevents the number of sites from explosively increasing between two time periods, which would be unrealistic given our sampling interval (50 years), and would lead to problems for the convergence of the model.
-The interaction of these parameters results in the following prior probability distribution for $\lambda$ and thus the growth (or change) rate of the population:
-
-<!-- Is there supposed to be something here?? -->
+The interaction of these parameters results in the following prior probability distribution for $\lambda$ and thus the growth (or change) rate of the population.
 
@@
 in connection with proxies that are not influenced by it (openness indicators, but also data from the demography of burial collectives).
-This would make it possible to estimate a value from original archaeological material independently of variables that have little to do with archaeological data, such as volcanic eruptions.<!-- JR: I don't follow...
-where do eruptions come in?
--->
+This would make it possible to estimate a value from original archaeological material independently of variables that have little to do with archaeological data, such as volcanic eruptions [@ballenger_temporal_2011].
 
@@
 By means of large-scale modelling, however, it would be possible to supplement gaps in the data in one region with data from other regions by means of regionalisation and a partial transfer of information (partial pooling).
-Such an extension would be the next logical step in the improvement of the model, to which end we hope to be able to contribute a further study in the near future.<!-- Maybe for tactical reasons we shouldn't spoil the sequel?
-:) -->
+Such an extension would be the next logical step in the improvement of the model, to which end we hope to be able to contribute a further study in the near future.
 

### 2022-05-28 analysis/analysis.Rmd
included Jans latest suggestions

@@
 
-popdens_plot <- ggplot(all_proxies_orig) +
-  geom_line(aes(x = (1950-age) * -1, y = scale(openness), col="Openness Indicator")) +
-  geom_line(aes(x = (1950-age) * -1, y = scale(sumcal), col="Sum Calibration")) +
-  geom_line(aes(x = (1950-age) * -1, y = scale(dendro), col="Dendro Dated Settlements")) +
-  geom_line(aes(x = (1950-age) * -1, y = scale(aoristic_sum), col="Aoristic Sum")) +
-  geom_line(data = PopDens_summary,
-            aes(x = (1950-age) * -1, y=mean, col = "Estimation Result")) + 
-  geom_line(data = PopDens_summary,
-            aes(x = (1950-age) * -1, y=hpdi_l, col = "Estimation Result"), lty=2) + 
-  geom_line(data = PopDens_summary,
-            aes(x = (1950-age) * -1, y=hpdi_u, col = "Estimation Result"), lty=2) +
-  theme_minimal() + scale_x_reverse() + 
+proxies_and_result <- rbind(melt(all_proxies_orig, id.vars = "age"),
+                            melt(PopDens_summary[,2:5], id.vars = "age")
+)
+
+variable_full_name = c(
+      sumcal = "Sum Calibration",
+      openness = "Openness",
+      aoristic_sum = "Aoristic Sum",
+      dendro = "Dendro Dated Settlements",
+      mean = "Estimation"
+      )
+
+popdens_plot <- ggplot() +
+  geom_line(data = subset(
+    proxies_and_result,
+    variable %in% c("mean",
+                    "openness",
+                    "sumcal",
+                    "dendro",
+                    "aoristic_sum")
+    ),
+    aes(x = (1950-age) * -1, y = value, color = variable)) +
+    geom_ribbon(data = subset(
+      proxies_and_result,
+      variable %in% c("mean")
+      ),
+      aes(x = (1950-PopDens_summary$age) * -1,
+          ymin = PopDens_summary$hpdi_l,
+          ymax = PopDens_summary$hpdi_u),
+      fill = "grey70", alpha = 0.5) +
+  facet_wrap(. ~ variable,
+             scales="free_y",
+             ncol = 1,
+             labeller = labeller(variable = variable_full_name)) +
+  theme_minimal() + scale_x_reverse() + theme(legend.position="false") +
+  ylab("Scaled values for Proxies,\n P/km² for Estimation") +
+  xlab("cal BCE") + 
   scale_colour_manual(
-    values = c("Openness" = "darkgreen",
-               "Sum Calibration" = "red",
+    values = c("sumcal" = "red",
+      "openness" = "darkgreen",
                "Aoristic Sum" = "gray",
-               "Dendro Dated Settlements" = "brown",
-               "Estimation" = "black"), name = "") +
-  theme(legend.position="bottom") +
-  ylab("Scaled values for Proxies,\n P/km² for Estimation") +
-  xlab("cal BCE")
-popdens_plot
+               "dendro" = "brown",
+               "mean" = "black")
+    )
+
+popdens_plot_adjusted <- ggplotGrob(popdens_plot)
+
+popdens_plot_adjusted$heights[[28]] <- unit(4, 'null')
+
+grid.newpage()
+grid.draw(popdens_plot_adjusted)
 ```
@@
 ggsave(normalizePath(file.path(here(), "figures","popdens_estimation.pdf"), mustWork = F),
-       plot = popdens_plot, width = 21, height = 29.7/2, units = "cm")
+       plot = popdens_plot_adjusted, width = 21, height = 29.7/2, units = "cm")
 ```

### 2022-05-28 manuscript/paper.Rmd
included Jans latest suggestions

@@
 ---
-title: "Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the Swiss plateau"
+title: "Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the North of the Swiss Alps"
 author:
@@
       email: caroline.heitz@arch.ox.ac.uk
-      institute: [IAW, OCCR, OX]
+      institute: [SFB1266]
       correspondence: false
@@
   - SFB1266: CRC 1266 - Scales of Transformation, University of Kiel
-  - OX: Associated Researcher, School of Archaeology, University of Oxford
   - IBOT: Department of Vegetation Ecology, Institute of Botany of the Czech Academy of Sciences
@@
 output:
-  bookdown::pdf_document2:
-    toc: no
-    pandoc_args:
-    - --lua-filter=../templates/scholarly-metadata.lua
-    - --lua-filter=../templates/author-info-blocks.lua
-    - --lua-filter=../templates/pagebreak.lua
   bookdown::word_document2:
@@
     - --lua-filter=../templates/pagebreak.lua
+  bookdown::pdf_document2:
+    toc: no
+    pandoc_args:
+    - --lua-filter=../templates/scholarly-metadata.lua
+    - --lua-filter=../templates/author-info-blocks.lua
+    - --lua-filter=../templates/pagebreak.lua
 bibliography: references.bib
@@
 abstract: |
-  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density.
-Using the prehistoric Swiss plateau (7000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
-The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research.
-Our prediction of population development on the Swiss plateau accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed radiocarbon method.
   Robust estimates of population are essential to the study of human–environment relations and socio-ecological dynamics in the past.
@@
 In this pilot study, we present an approach that we believe can be usefully transferred to other regions, as well as refined and extended to greatly advance our understanding of prehistoric demography.
-  The archaeological record provides several potential proxies of human population, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage.
+  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density.
+Using the  area North of the main ridge of the Swiss Alps in prehistoric times (7000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
+The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research.
+Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.
+  The archaeological record provides several potential proxies of human population dynamics, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage.
 Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely ona single proxy, and are rarely able to infer population size or density in absolute terms.
@@
   - Our initial model produces more robust, high-resolution estimates of past population dynamics than previous, single-proxy approaches
-  - We provide absolute estimates of population size and density on the prehistoric Swiss plateau (7000–1000 BCE)
+  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (7000–1000 BCE)
 ---
@@
 Early attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack [-@hack_changing_1942], Colton [-@colton_prehistoric_1949], Frankfort [-@frankfort_town_1950] and Cook [-@cook_reconsideration_1946].
-Cook and Heizer [-@cook_quantitative_1965, @cook_relationships_1968] and Naroll [-@naroll_floor_1962] provided rules derived from ethnographic contexts for estimating the size of prehistoric populations.
+Cook and Heizer [-@cook_quantitative_1965, -@cook_relationships_1968] and Naroll [-@naroll_floor_1962] provided rules derived from ethnographic contexts for estimating the size of prehistoric populations.
 Cook and Heizer (1966) also made the first attempt to estimate the population growth rate during the Neolithic [@hassan_demographic_1981].
@@
 Kintigh et al.
-[-@kintigh_grand_2014], for example, list human influence, dominance, population size, and population growth amongst their "grand challenges" for archaeology in the 21st century.
+[-@kintigh_grand_2014], for example, list human influence, dominance, population size, and population growth amongst their 'grand challenges' for archaeology in the 21st century.
 
@@
 We believe that these problems cannot be overcome by methodological refinements in this area alone.
-Instead, the Bayesian approach we develop in this paper offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon.
+Instead, the Bayesian approach we develop in this paper offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.
 
@@
 
-Müller and Diachenko [@muller_tracing_2019] have compiled a list proxies currently used for the estimation of population size in prehistory.
+Müller and Diachenko [-@muller_tracing_2019] have compiled a list proxies currently used for the estimation of population size in prehistory.
 These can be roughly divided into three groups: ethnographic analogies; deductive estimates from ecological/economic factors; and the interpolation of frequencies of archaeological features (e.g.
@@
 As a result, systematic distortions are the rule rather than the exception.
--   **Heterogeneous**: A wide range of proxies \@ref(tab:tableproxies) can potentially inform us about population dynamics, all with different spatio-temporal scales, granularity, information value, scales, and data formats.
+-   **Heterogeneous**: A wide range of proxies can potentially inform us about population dynamics, all with different spatio-temporal scales, granularity, information value, scales, and data formats.
 -   **Indirect**: We will never have direct data on prehistoric population; only proxy data that is thought to be a reasonable substitute for it.
@@
 In response, demographers have increasing turned to Bayesian methods to estimate and forecast contemporary population dynamics.
-For example, Bryant and Zhang [@bryant2018] consider Bayesian data modelling a solution to exactly the kind of problems that affect archaeological data.
+For example, Bryant and Zhang [-@bryant2018] consider Bayesian data modelling a solution to exactly the kind of problems that affect archaeological data.
 Specifically, Bayesian approaches are well suited for limited, unreliable and noisy data.
@@
 The well-established application of a Bayesian approach in radiocarbon calibration, where it is used to model radiometric uncertainty based on stratigraphic information, represents a very similar use case in archaeology.
-More recently, archaeologists have also used Bayesian modelling techniques as a tool for testing hypotheses relating to demographic trends or underlying models based on ^14^C data[e.g.
+More recently, archaeologists have also used Bayesian modelling techniques as a tool for testing hypotheses relating to demographic trends or underlying models based on ^14^C data [e.g.
 @crema2021a].
@@
 The concept of probability is not derived from theoretical, infinitely repeatable random experiments or distributions, but from the direct confrontation of the pre-assumptions (priors) with the available data.
-This involves checking how credible these prior assumptions are with regard to the available data [likelihood; see also @bryant2018, 66].
+This involves checking how credible these prior assumptions are with regard to the available data [likelihood, see also @bryant2018, 66].
 A small amount of data leads to a broad probability distribution that is not strongly localised and restricted.
@@
 A further characteristic of Bayesian modeling is that, due to the fact that results of a Bayesian inference can be regarded as a prior for the next one, a hierarchical formulation of problem domains is possible.
-Parameters that are necessary for an estimation, such as the relationship of population density to the clearing signal in pollen data, need not be specified explicitly, but can be given by probability distributions and then estimated in the model itself [@bryant2018, 186].
+Parameters that are necessary for an estimation, such as the relationship of population density to the deforestation signal in pollen data, need not be specified explicitly, but can be given by probability distributions and then estimated in the model itself [@bryant2018, 186].
 The more data available, the more degrees of freedom can be estimated with a reasonable loss of confidence or a reasonable width of credibility intervals [@kruschke2015, 112].
@@
 
-The Swiss Plateau (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland's territory and comprises the partly flat, but largely hilly area, between the Jura Mountains and the Alps.
+The case study area north of the Swiss Alps (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland’s territory and comprises the partly flat, but largely hilly area, between the Jura Mountains and the Alps.
 It averages an altitude of between 400 and 600 m a.s.l.
-It is a favourable area for settlement and agriculture and is by far the most densely populated region of the Switzerland today.
-Although the Swiss Plateau is a basin, depending on the region it has a very diverse natural landscape: shaped by glaciers during the Ice Age, the many lakes that characterise the landscape today also provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, as well as a rich pool of sources for vegetation reconstructions by means of pollen analyses.
-A very active and efficient archaeological research and heritage management also offers a rich repository of archaeological data, known sites and ^14^C data.
+It is a favourable area for settlement and agriculture and especially the Swiss Plateau is by far the most densely populated region of the Switzerland today.
+The Swiss Plateau between Lake Zurich and Lake Geneva will serve as a core region in our case study, because archaeological data from there is most abundant and accessible.
+Although it is a basin, depending on the region it has a very diverse natural landscape: shaped by glaciers during the ice ages, the many lakes and bogs that characterise the landscape today also provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, as well as a rich pool of sources for vegetation reconstructions by means of pollen analyses.
+Thanks to the very active and efficient archaeological research and heritage management there is an abundance of archaeological information, including known sites as well as dendrochronological and ^14^C data.
 
-```{r mapswissplateau, fig.cap="Location and extent of the Swiss Plateau."}
+```{r mapswissplateau, fig.cap="Location and extent of the Swiss Plateau as biogeographical region (based on swisstopo) including additional low altitude areas in the north of Switzerland (regions along the High Rhine between Schaffhausen and Basel)."}
 main_map <- my_basemap +
@@
 Our case study targets the period between 7000–1000 BCE.
-The lower limit of this window was chosen to avoid the so-called 'Halstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, which would cause difficulties for the ^14^C proxy data.
+The lower limit of this window was chosen to avoid the so-called 'Hallstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, which would cause difficulties for the ^14^C proxy data.
 The upper limit was chosen to coincide with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence before 7000 in the working area.
@@
 In principle, a large number of different data sources can be integrated into the overall model as proxies, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created.
-\@ref(tab:tableproxies) provides a non-exhaustive list.
-For our case study of the Swiss Plateau, we used: a landscape openness indicator; an aoristic sum of sites based on typological dating; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes area.
+Table \@ref(tab:tableproxies) provides a non-exhaustive list.
+For our case study, we used: a landscape openness indicator; an aoristic sum of sites based on typological dating; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).
 
@@
 
-With a mild climate compared to the Jura and the Alps, and a dense network of water bodies, the Swiss Plateau has always been a favourable area for human settlement in Switzerland.
+With the vast areas of arable soils with small slope gradient compared to the Jura and the Alps, and a dense network of water bodies, the Swiss Plateau has always been a particularly favourable area for human settlement in Switzerland.
 From the Neolithic onwards, settlement areas were concentrated along its rivers and lakes [@christianlüthi2009].
@@
 
-The natural conditions of the lakes of the Swiss plateau enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis.
+The natural conditions of the lakes of the Swiss Plateau enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis.
 We make use of this fact by generating a supra-regional openness indicator for the vegetation from the pollen data (Figure \@ref(fig:pollensites)).
@@
 
-Unfortunately, there is no burial data available the Swiss plateau that could be usefully incorporated into such this model due to the absence of regular burials for larger parts of prehistory.
-Formation conditions, in this case most likely C-transformation [@schiffer1987], prevent their use in this study.
+Unfortunately, there is no burial data available in the case study area that could be usefully incorporated into such this model due to the absence of regular burials for larger parts of prehistory.
+Formation conditions, in this case most likely C-transformation [@schiffer1987], the active cultural choice to not use archaeological traceable burial pratices, prevent their use in this study.
 Nevertheless, we see a very high potential for other regions in the integration of demographic indicators from burial data in order to enlarge the canon of methods and the range of proxies.
@@
 
-In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements in the Swiss plateau.
-Consequently, we shift the causality and consideration of measurement error, which is certainly inherent in each of these proxies, to a preceding 'process model', in which a Poisson process describes the number of simultaneous settlements.
+In the construction of our ‘observational model’, we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps.
+Consequently, we shift the causality and consideration of measurement error, which is certainly inherent in each of these proxies, to a preceding ‘process model’, in which a Poisson process describes the number of simultaneous settlements.
 In doing so, we establish a likelihood that estimates how credible the data is, given the model.
@@
 In our case, it functions as a scaling factor for the individual proxies.
-$\alpha$, on the other hand, is to be understood as an intercept.
+$\alpha$, is to be understood as an intercept.
 If there were no change due to the variables, the regression would fall back to this value.
@@
 By means of a data series, which would represent an evidence-based estimate of the temporal development of settlement sizes, this specification could be made based on data.
-From the number of sites and the mean number of individuals represented in each case, a population density can be calculated using the area of the Swiss Plateau (12649 km²).
+From the number of sites and the mean number of individuals represented in each case, a population density can be calculated using the case study area (12649 km²).
 The estimated result of the model is thus comparable with estimates from other sources or the literature.
@@
 Thus, the estimate remains within the values that are also considered plausible by the expert estimates.
-Clear peaks are reached around 1250 BCE, as well as around 2750 BCE, which corresponds to the transition to the ceramic style in the Swiss Plateau influence by the Corded Ware.
+Clear peaks are reached around 1250 BCE, as well as around 2750 BCE, which corresponds to the transition to a ceramic style as observed for sites of the Swiss Plateau influence by the Corded Ware ceramic style [@hafner_vom_2004].
 
@@
 The coefficient of variation is 0.13 for the beginning and 0.1 for the end of the estimate, the greatest variability is reached around 2150 BCE with 0.47.
+This is not astonishing as there are fewer archaeological contexts recorded connected to the earlier phase of the Early Bronze Age c.
+2200-1800 BCE, this picture changes from c.
+1800 BCE onwards [@hafner_fruhe_1995, @david-elbiali_suisse_2000].
 The beginning and end of the time series are relatively clearly determined.
@@
 Our original ambition in the development of this model was to base it on a state-space representation of the demographic process itself, and then to integrate the existing proxies by means of an observation model to inform this process.
-In the course of developing the model, we realised that the data we had available for the Swiss Plateau alone was not sufficient to adequately determine this process, or to adequately fix the degrees of freedom resulting from the process and the transfer model.
+In the course of developing the model, we realised that the data we had available for the case study area north of the Swiss Alps  alone was not sufficient to adequately determine this process, or to adequately fix the degrees of freedom resulting from the process and the transfer model.
 Therefore, in implementation presented here, we have turned to a Poisson regression approach: the model represents the best possible combination of the indicators used to describe the development of the number of sites based on them.
@@
 
-Comparing the model's overall estimate with trends indicated by individual proxies provide gives us new insights into the quality of these records.
+Comparing the model's overall estimate with trends indicated by individual proxies gives us new insights into the quality of these records.
 Sum calibration, currently the most frequently used proxy for (relative) population changes in prehistory, has its large fluctuations dampened when considered alongside other proxies.
@@
 The expected increase in archaeological remains with the onset of Neolithisation is still clearly visible, but the curve of the overall estimate, after the initial increase, is much flatter than sum calibration alone indicates.
-The period between 3950 and 3700, which is contemporaneous with the first settlement cycle on the lakeshores of the Three Lakes area, coincides with a noticeable plateau on the calibration curve, which be producing an overestimation of the ^14^C density.
+The period between 3950 and 3700 BCE, which is contemporaneous with the first larger settlement cycle on the lakeshores of the Three Lakes region, coincides with a noticeable plateau on the calibration curve, which be producing an overestimation of the ^14^C density.
 However, the effect of the calibration curve on the results of a cumulative calibration cannot yet be considered unambiguous.
@@
 
-## Prehistoric population dynamics on the Swiss plateau
+## Prehistoric population dynamics north of the Swiss Alps
 
-In order to review the reconstruction against the background of established archaeological knowledge, it is useful to overlay conventionally-defined archaeological phase boundaries on the results of our model \@ref(fig:popestwithphases).
+In order to review the reconstruction against the background of established archaeological knowledge, it is useful to overlay conventionally-defined archaeological phase boundaries [@hafner_neolithikum_2005] on the results of our model (Figure \@ref(fig:popestwithphases)).
 It should be noted, however, that the model's estimate is not completely independent of these phases: due to the aoristic sum, which is itself strongly determined by this phase division, corresponding boundaries also influence the structure of the reconstruction.
@@
 
-```{r popestwithphases, fig.cap="The estimate of the population density in relation to the chronology for the Swiss Plateau."}
+```{r popestwithphases, fig.cap="The estimate of the population density in relation to the chronology for the case study area north of the Swiss Alps."}
 PopDens_summary <- read.csv(
@@
 Here we must assume a basically low level of settlement, probably mainly by hunter-gatherer groups.
-Isolated Neolithic sites of the LBK and later groups are known in the periphery of Switzerland, but they play a subordinate role.
-The Neolithic really began here in the so-called Upper Neolithic, connected with the cultural phenomena of Egolszwil and Cortaillod respectively Pfyn, as well as with the first lake shore settlements.
+Isolated Neolithic sites of the LBK and later groups are known in the periphery of Switzerland, but they play a subordinate role [@ebersbach_nutzung_2012].
+The evidence of the Neolithic especially what the Swiss Plateau is concerned is dense from the so-called Upper Neolithic onwards, connected with the typo-chronological units based on pottery that are called Egolzwil (late 5th M BCE) and Cortaillod respectively Pfyn (first half of the 4th M BCE).
+The first lake shore settlements north of the Alps date to this time too.
 In this period, we also see a clear increase in the estimated population in the model.
-In the transition to the Late Neolithic we know from the lakeshore settlements the so-called Horgen Gap.
+In the transition to the Late Neolithic we know from the lakeshore settlements the so-called Horgen Gap [@hafner_neolithikum_2005].
 This is also visible as a slight decrease in the model.
-In another study [@heitz2021] we could show that this is in fact probably not a decline in population but rather a shift in settlement strategy.
-In the Late Neolithic, associated with the Horgen pottery, we then see a clear increase in the intensity of settlement, which reaches its peak and its break-off at the transition to the Corded Ware and thus to the Final Neolithic.
-In the second half of the Early Bronze Age, from which we again know lakeshore settlements, there is again a clear increase in population size according to the model, which continues until the Late Bronze Age.
+In another study [@heitz2021] we could show that this is in fact probably not a decline in population.
+Rather communities relocated their settlements to the hinterland of the large lakes in times of lake level rises of higher magnitudes due to climatic changes.
+In the Late Neolithic, associated with the Horgen pottery, we then see a clear increase in the intensity of settlement, which reaches its peak and its break-off at the transition to the Corded Ware and thus to the Final Neolithic [@hafner_vom_2004].
+In the second half of the Early Bronze Age, during which – after a break from the Final Neolithic to the very early Bronze Age – lakeshores were resettled, but to a smaller extend.
+There is again a clear increase in population size according to the model, which continues until the Late Bronze Age.
 The general trends fit very well with the previous reconstructions of population development for Switzerland [see eg.
@@
 
-An key advance in the model we present here is the ability to estimate, in absolute terms, past population sizes and the uncertainty accompanying our present knowledge.
+The key advance in the model we present here is the ability to estimate, in absolute terms, past population sizes and the uncertainty accompanying our present knowledge.
 These estimates can be used as a basis for further studies where relative measures of population development are not helpful, such as long-term land use studies, where modelling of large-scale socio-ecological systems based on archaeological data becomes possible and does have to rely on using deductive, asynchronous population models (e.g.
@@
 This data is available to varying degrees in different regions.
-On the Swiss plateau, for example, we have too little data on human remains over large spans of prehistory to be able to integrate them meaningfully into a model.
+On the Swiss Plateau, for example, we have too little data on human remains over large spans of prehistory compared to the abundance of wetland settlements to be able to integrate them meaningful into a model.
 
-To apply the model to different regions, the proxies used we use here (e.g.
+To apply the model to different regions, the proxies we use here (e.g.
 the palynological openness indicator) would have to be adapted to fit local conditions and research histories.
@@
 The development of the openness index took place within the framework of the project Time and Temporality in Archaeology (project number 194326, PI Caroline Heitz) and was inspired by the cooperation within the project QuantHum (project 169371, PI Marco Conedera), both also funded by the SNSF.
+Jan Kolář was supported by a long-term research development project (RV67985939) and by a grant from the Czech Science Foundation (19-20970Y).
 We also thank the Institute of Archaeological Sciences of the University of Bern for its support and faith in the outcome of our modelling project.

### 2022-05-28 manuscript/paper_shorter.Rmd
added shorter 5000 word version

--- 2022-05-28 manuscript/paper.Rmd
@@
   Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density.
-Using the  area North of the main ridge of the Swiss Alps in prehistoric times (7000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
+Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
 The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research.
@@
   - Our initial model produces more robust, high-resolution estimates of past population dynamics than previous, single-proxy approaches
-  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (7000–1000 BCE)
+  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (6000–1000 BCE)
 ---
@@
 Prehistorians have long recognised demography as a fundamental force in human cultural evolution [@childe_man_1936].
-However, despite decades of interest in the demographic dynamics of prehistoric societies, concrete estimates of population size and population density before written records remain elusive.
-Though the archaeological record provides multiple possible demographic proxies [@hassan_demographic_1981], lack of access to this data, and a lack of methodological tools for turning it into quantitative estimates, has often left the conclusions drawn from it vague and superficial [@hassan_demographic_1981].
+Despite decades of interest in the demographic dynamics of prehistoric societies, concrete estimates of population size and density before written records remain elusive.
+Though the archaeological record provides multiple possible demographic proxies [@muller_tracing_2019], a lack of access to this data and methodological tools for turning it into quantitative estimates has left the conclusions drawn from it vague and superficial [@hassan_demographic_1981].
 As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence [@morris_measure_2013, @turchin_seshat_2015].
 
-Early attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack [-@hack_changing_1942], Colton [-@colton_prehistoric_1949], Frankfort [-@frankfort_town_1950] and Cook [-@cook_reconsideration_1946].
-Cook and Heizer [-@cook_quantitative_1965, -@cook_relationships_1968] and Naroll [-@naroll_floor_1962] provided rules derived from ethnographic contexts for estimating the size of prehistoric populations.
-Cook and Heizer (1966) also made the first attempt to estimate the population growth rate during the Neolithic [@hassan_demographic_1981].
-These approaches became increasingly common after the 1960s, with the emergence of processual archaeology and its focus on human–environment relations and population ecology [@hassan_demographic_1981].
+Prehistoric demography has experienced a resurgence in recent years [@shennan_population_2000; @riede_climate_2009 and others in same issue], partly explained by a renewed interest in human–environment relations and human impact, necessarily requiring an assessment of population size.
+Kintigh et al.
+[-@kintigh_grand_2014] list human influence, dominance, population size, and population growth amongst their 'grand challenges' for archaeology in the 21st century.
 
-Following a spell out of fashion with the post-processual turn, prehistoric demography has experienced a revival in recent years as part of the cultural evolution paradigm [@shennan_population_2000; @riede_climate_2009 and others in same issue].
-Part of this resurgence can be explained by a renewed interest in human–environment relations and human impact on the natural environment, which necessarily requires an assessment of population size.
-Kintigh et al.
-[-@kintigh_grand_2014], for example, list human influence, dominance, population size, and population growth amongst their 'grand challenges' for archaeology in the 21st century.
-
-But methodological advances have certainly also played a role.
-In the last decade, the 'dates as data' technique [@rick_dates_1987], which uses the frequency of radiocarbon dates as a proxy for population dynamics, was taken up and significantly developed in a series of publications from the UCL group [e.g.
-@shennan_regional_2013], and has since been widely applied to archaeological contexts around the world [@crema_review_2022].
-
-The 'dates as data' or summed radiocarbon approach has contributed greatly to our understanding of prehistoric demography, but it is not without its critics [@attenbrow_hiscock_2015, @carleton_groucutt_2020].
-While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to certain fundamental problems common to all approaches that rely on a single proxy [@french_manifesto_2021, @schmidt_approaching_2021].
+In the last decade, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, was taken up and significantly developed [e.g.
+@shennan_regional_2013], and has since been widely applied to archaeological contexts worldwide [@crema_review_2022].
+This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015, @carleton_groucutt_2020].
+While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems common to all approaches relying on a single proxy [@french_manifesto_2021, @schmidt_approaching_2021].
 We believe that these problems cannot be overcome by methodological refinements in this area alone.
-Instead, the Bayesian approach we develop in this paper offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.
+Instead, a Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.
 
@@
 
-Müller and Diachenko [-@muller_tracing_2019] have compiled a list proxies currently used for the estimation of population size in prehistory.
-These can be roughly divided into three groups: ethnographic analogies; deductive estimates from ecological/economic factors; and the interpolation of frequencies of archaeological features (e.g.
+Proxies currently used for the estimation of population size in prehistory [following @muller_tracing_2019] can roughly be divided into three groups: ethnographic analogies; deductive estimates from ecological/economic factors; and the interpolation of frequencies of archaeological features (e.g.
 settlements, structures, individual finds).
@@
 **Reliance on single proxy**: Most investigations are using only one source of evidence.
-Although multi-proxy approaches exist, in such cases the individual proxies only serve to support each other or the proxy used as the main estimator.
-There is no explicit combination of results to reduce the inaccuracy and biases inherent to single proxies.
+Although multi-proxy approaches exist, there the individual proxies only serve to support each other or the main estimator, without explicitly combining them to reduce the inaccuracy and biases inherent to single proxies.
 2.
-**Uncertainty in measurements**: All archaeological evidence comes with an inherent uncertainty which is carried through to derived measurements.
-However, in most studies of prehistoric population, individual curves are presented as estimates, and the potential error associated with these estimates is almost never specified.
+**Uncertainty in measurements**: All archaeological evidence is inherently uncertain which is carried through to derived measurements.
+However, in most studies, single curves are presented as estimates, and the potential error associated is almost never specified.
 3.
 **Lack of a transfer function**: By 'transfer function', we mean something that allows for the proxy data to be interpreted in terms of actual population size or density.
-This could be absolute, i.e.
+These could be absolute, i.e.
 a numerical estimate of population, or relative, i.e.
 a means of scaling changes in the proxy value to changes in population.
-Lack of a suitable frameworks and 'calibration' data means that this is rarely presented alongside proxy estimates.
-In the best case scenario, there will be a qualitative assessment of the informative value of the proxy, but this does not sufficiently account for the complex nature of archaeological data.
+Lack of suitable frameworks and 'calibration' data means that this is rarely presented alongside proxy estimates.
+In the best cases, there is a qualitative assessment of the informative value of the proxy, not sufficiently accounting for the complex nature of archaeological data.
 
-Furthermore, the types of archaeological data commonly used as population proxies share a number of problematic characteristics. They are:
+Furthermore, the types of archaeological data commonly used as population proxies share a number of problematic characteristics, being:
 
--   **Limited**: We have only incomplete data that can be used for these purposes, and they are usually not very informative.
--   **Unevenly distributed**: For example, although there is a good data on settlement frequencies for some regions, and these are sometimes have a very high temporal resolution, these regions are very unevenly distributed over time and space.
+-   **Limited**: We have only incomplete data, and they are usually not very informative.
+-   **Unevenly distributed**: For example, although there is a good data on settlement frequencies for some regions, these regions are very unevenly distributed over time and space.
 -   **Noisy**: Frequently individual proxies are strongly influenced by factors unrelated to population, for example taphonomic conditions or depositional biases.
 -   **Unreliable**: Research strategies, research history and varying levels of resources available to researchers strongly affect the nature of compiled datasets.
-As a result, systematic distortions are the rule rather than the exception.
--   **Heterogeneous**: A wide range of proxies can potentially inform us about population dynamics, all with different spatio-temporal scales, granularity, information value, scales, and data formats.
--   **Indirect**: We will never have direct data on prehistoric population; only proxy data that is thought to be a reasonable substitute for it.
-The transfer functions that link the proxy data with the desired quantity (population) are unknown.
-At best, they may be estimated by comparison with historical examples or the (ethnographic) present.
-However, since living conditions, circumstances and economic parameters differ greatly between the recent history and prehistory, these transfer functions can only be considered a rough first approximation.
+Systematic distortions are the rule rather than the exception.
+-   **Heterogeneous**: All potential proxies have different spatio-temporal scales, granularity, information value, scales, and data formats.
+-   **Indirect**: We will never have direct data on prehistoric population; only proxy data that is thought to be a reasonable substitute.
+The transfer functions linking the proxy data with the desired quantity (population) are unknown.
 -   **Contradictory**: When considering several proxies, differences in transfer functions, data quality and noisiness inevitably lead to different results.
-The most common way to deal with this is to reject certain proxies for a certain time periods, but this is not performed on a consistent or quantitative basis.
 
-Here we argue that many, if not all, of these problems can be ameliorated through a) the explicit, quantitative integration of multiple proxies; and b) the use of a Bayesian approach to take account of and estimate uncertainty.
+We argue that many, if not all, of these problems can be ameliorated through a) the explicit, quantitative integration of multiple proxies; and b) the use of a Bayesian approach to take account of and estimate uncertainty.
 
@@
 
-Many of the problems with archaeological population proxies are also encountered in contemporary demography.
-In response, demographers have increasing turned to Bayesian methods to estimate and forecast contemporary population dynamics.
+Many of the problems with archaeological population proxies are shared with contemporary demography.
+In response, demographers have increasingly turned to Bayesian methods to estimate and forecast contemporary population dynamics.
 For example, Bryant and Zhang [-@bryant2018] consider Bayesian data modelling a solution to exactly the kind of problems that affect archaeological data.
-Specifically, Bayesian approaches are well suited for limited, unreliable and noisy data.
-Various sources of data, even contradictory data, can be brought into a common framework and used to support one another.
+Bayesian approaches are well suited for limited, unreliable and noisy data.
+Various data sources, even contradictory data, can be brought into a common framework and used to support one another.
 These methods also provide a quantitative estimate of the likelihood and uncertainty of the model's resulting predictions (or in our case retro-dictions).
-Bayesian approaches are also capable of accounting for spatially and temporally incomplete data: where this data is missing, the uncertainty automatically increases, but this does not prevent general modelling and estimation.
-Finally, hierarchically-structured model suites, in which sub-models are created for each individual proxy, can be used to estimate transfer functions between individual proxies and the value to be modelled, thanks to the interaction of a large number of data sources and evidence.
+Bayesian approaches are also capable of accounting for spatially and temporally incomplete data: where this data is missing, the uncertainty increases, but this does not prevent general modelling and estimation.
+Finally, hierarchically-structured model suites, with sub-models for each individual proxy, can be used to estimate transfer functions between them and the value to be modelled, thanks to the interaction of a large number of evidence.
 
-This modeling technique can thus be used to combine different lines of evidence horizontally and vertically and in this way combine their results into a common conclusion and estimate, which at the same time includes an estimation of their reliability: if the data contradict each other, the overall reliability will be lower.
-If they support each other, the confidence interval will be smaller.
-And if there is no systematic bias that affects all data sources to the same extent, it should be possible to arrive at the most reliable estimate possible through the most heterogeneous set of data sources.
+This modeling technique can thus be used to join different lines of evidence horizontally and vertically and combine their results into a overall estimate, including an assessment of their reliability: contradicting data lead to a lower overall reliability, while a mutual support to smaller confidence intervals.
+If there is no systematic bias that affects all data sources to the same extent, this results in the most reliable estimate possible through the most heterogeneous set of data sources.
 
 The well-established application of a Bayesian approach in radiocarbon calibration, where it is used to model radiometric uncertainty based on stratigraphic information, represents a very similar use case in archaeology.
-More recently, archaeologists have also used Bayesian modelling techniques as a tool for testing hypotheses relating to demographic trends or underlying models based on ^14^C data [e.g.
+More recently, archaeologists have also used Bayesian modelling techniques for testing hypotheses relating to demographic models based on ^14^C data [e.g.
 @crema2021a].
-This approach differs from the one presented in this paper in that, in these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only.
-This is a clear step forward in the direction of a model-based, and thus scientific, analysis.
-However, the use of only one proxy, and its use exclusively for testing hypotheses developed independently, creates problems comparable to those of the inductive approaches used so far: due to the lack of a combination with other indicators, one is limited to the problems and conditions of sum calibration as a tool.
-Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of the time series.
-Thus, the credibility of a model can only be checked as a whole, without the dynamic developments that can arise in the course of demographic processes being represented.
-We would like to better exploit the capabilities of Bayesian hierarchical models through a combination of inductive data analysis and model-passed data integration of different proxies.
+This approach differs from the one presented here in that, in these analyses, deductive models are generated and their plausibility is tested on the basis of ^14^C data only.
+This is a clear step forward to a model-based, scientific analysis.
+However, the use of only one proxy, exclusively for testing hypotheses developed independently, creates problems comparable to those of the inductive approaches used so far: lacking a combination with other indicators, one is limited to the problems and conditions of sum calibration.
+Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of time series.
 
-Here, we attempt to make Bayesian hierarchical techniques usable for archaeological reconstructions.
-In addition to a presentation of the basics and possible procedures, we want to show in the following, in a reproducible and practical form using a case study, how Bayesian methods can also make a decisive contribution to a better assessment of population development.
-These assessments are crucial for the reconstruction of the human past, even in for periods for which we only have very patchy, noisy and unreliable data.
+We attempt to make Bayesian hierarchical techniques usable for archaeological reconstructions.
+We want to show, in a reproducible and practical form using a case study, how Bayesian methods can make a decisive contribution to a better assessment of population development, crucial for the reconstruction of the human past, even in for periods for which we only have very patchy, noisy and unreliable data.
 
@@
 
-The basis of Bayesian statistics is the premise that there is always some prior assumption about the probability of an event, even if it may be very rough.
-This assumption is then adjusted by observing the data.
-The concept of probability is not derived from theoretical, infinitely repeatable random experiments or distributions, but from the direct confrontation of the pre-assumptions (priors) with the available data.
-This involves checking how credible these prior assumptions are with regard to the available data [likelihood, see also @bryant2018, 66].
-A small amount of data leads to a broad probability distribution that is not strongly localised and restricted.
-The intuitive procedure is thereby the shifting of credibility through evidence.
+Bayesian statistics rely on the premise that there is always some prior assumption, even if very rough, about the probability of an event.
+This assumption is adjusted by observing data, by checking how credible these priors are [likelihood, see also @bryant2018, 66].
+This is Bayesian updating [cf.
+also @kruschke2015, especially 15--25], resulting in the posterior probability distribution, which represents not a point prediction.
+Small amounts of data lead to a broad distribution not strongly localised and restricted.
+Thus, we simultaneously obtained a result and an estimate the credibility interval, given the data.
 
-At the heart of all Bayesian statistics is therefore the concept of updating a given prior assumption with new data and expressing this in probabilities [cf.
-also @kruschke2015, especially 15--25].
-Our assumptions about the demographic development of the past must naturally be very conservative.
-In the logic of the basic Bayesian equation, these assumptions represent the prior (probability).
-In conjunction with the data that are included as the likelihood of the prior, a posterior (probability) results, which represents the Bayesian model learning from data.
-It is also in the nature of the approach that in real applications there is no point prediction, but in most cases a probability distribution for the prediction.
-Thus, we simultaneously obtained a result and an estimate of the confidence intervals, or rather the credibility interval, given the data.
+This Bayesian learning is iterative and sequential, so that the result of one Bayesian inference can form the prior of another [@kruschke2015, 17].
+This allows different information to be combined [@bryant2018, 219--224], as it has long been exploited by archaeology in using stratigraphic information to make radiometric dating more accurate [@ramsey1995].
 
-This Bayesian learning is iterative and sequential, so that the result of one Bayesian inference can form the prior of another, i.e.
-it is an additive process [@kruschke2015, 17].
-Moreover, at the conceptual level, this allows different sources of information to be combined [@bryant2018, 219--224].
-This enables that they can be mapped to the same set and the same real-world domain, e.g.
-probabilities.
-This fact has long been exploited by archaeology in using stratigraphic information to make radiometric dating more accurate [@ramsey1995].
-^14^C dates and stratigraphy are something completely different, but both can be mapped to the probability of older and younger and combined in this way.
-The same is also feasible when it comes to the probability of population sizes or population densities or their derived dynamics.
-
-A further characteristic of Bayesian modeling is that, due to the fact that results of a Bayesian inference can be regarded as a prior for the next one, a hierarchical formulation of problem domains is possible.
+This also makes a hierarchical formulation of problem domains possible.
 Parameters that are necessary for an estimation, such as the relationship of population density to the deforestation signal in pollen data, need not be specified explicitly, but can be given by probability distributions and then estimated in the model itself [@bryant2018, 186].
-The more data available, the more degrees of freedom can be estimated with a reasonable loss of confidence or a reasonable width of credibility intervals [@kruschke2015, 112].
-For the estimation of these parameters, in turn, submodels have to be created which describe the relationship of the data to the characteristics of the parameter.
-This can be carried out over several levels, depending on necessity [@kruschke2015, 221--222].
+The more data available, the more degrees of freedom can be estimated with a reasonable width of credibility intervals [@kruschke2015, 112].
+For the estimation of these parameters, submodels have to be created describing the relationship of the data to the characteristics of the parameter [@kruschke2015, 221--222].
 
@@
 
-The case study area north of the Swiss Alps (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland’s territory and comprises the partly flat, but largely hilly area, between the Jura Mountains and the Alps.
-It averages an altitude of between 400 and 600 m a.s.l.
-It is a favourable area for settlement and agriculture and especially the Swiss Plateau is by far the most densely populated region of the Switzerland today.
-The Swiss Plateau between Lake Zurich and Lake Geneva will serve as a core region in our case study, because archaeological data from there is most abundant and accessible.
-Although it is a basin, depending on the region it has a very diverse natural landscape: shaped by glaciers during the ice ages, the many lakes and bogs that characterise the landscape today also provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, as well as a rich pool of sources for vegetation reconstructions by means of pollen analyses.
+Our case study area north of the Swiss Alps (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland’s territory and comprises the partly flat, but largely hilly area between the Jura Mountains and the Alps.
+It is favourable for settlement and agriculture, especially the Swiss Plateau between Lake Zurich and Lake Geneva, by far the most densely populated region of the Switzerland today.
+This serves as core region in our case study, because here archaeological data is most abundant and accessible.
+Depending on the region, it has a very diverse natural landscape: shaped by glaciers during the ice ages, the many lakes and bogs provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, and a rich source for vegetation reconstructions by means of pollen analyses.
 Thanks to the very active and efficient archaeological research and heritage management there is an abundance of archaeological information, including known sites as well as dendrochronological and ^14^C data.
@@
 
-Our case study targets the period between 7000–1000 BCE.
-The lower limit of this window was chosen to avoid the so-called 'Hallstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, which would cause difficulties for the ^14^C proxy data.
-The upper limit was chosen to coincide with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence before 7000 in the working area.
+Our case study targets the time window between 6000–1000 BCE.
+Its lower limit was chosen to avoid the so-called 'Hallstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, causing difficulties for the ^14^C proxy.
+The upper limit coincides with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence.
 
-In principle, a large number of different data sources can be integrated into the overall model as proxies, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created.
+A large number of different proxies can be integrated into an overall model, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created.
 Table \@ref(tab:tableproxies) provides a non-exhaustive list.
-For our case study, we used: a landscape openness indicator; an aoristic sum of sites based on typological dating; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).
+For our case study, we used: a landscape openness indicator; an aoristic sum of typological dated sites; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).
 
@@
 
-With the vast areas of arable soils with small slope gradient compared to the Jura and the Alps, and a dense network of water bodies, the Swiss Plateau has always been a particularly favourable area for human settlement in Switzerland.
-From the Neolithic onwards, settlement areas were concentrated along its rivers and lakes [@christianlüthi2009].
-Thus, our working region offers on the one hand excellent data for demographic estimation, but on the other hand poses very specific problems for such an undertaking.
-If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also might cause a research problem with regard to the ^14^C data often used as a proxy.
+From the Neolithic onwards, known settlement areas in Switzerland concentrate along its rivers and lakes [@christianlüthi2009].
+Thus, our working region offers excellent data for demographic estimation, but poses very specific problems for such an undertaking.
+We have high-resolution information on the temporal sequence of individual lakeside settlements by means of dendro data.
+In these cases, ^14^C data are not as abundant simply because they are inferior to dendro dating.
 
@@
 The dataset we use for the number of dendro-dated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019].
-Details on the creation of this data series will be published at the referenced location.
 The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements.
-This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.
 
@@
 
-The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of [@martínez-grau2021].
-It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites.
-The dates were selected so that their distribution area coincides with the catchment area of the pollen proxy (see also Figure \@ref(fig:c14map)).
+The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of Martínez-Grau et al.
+[-@martínez-grau2021].
+It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites (see Figure \@ref(fig:c14map)).
 The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` uncal BP.
@@
 We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites.
-For the creation of the cumulative calibration, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.
+For the creation of the sum calibration, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.
 
@@
 ```
-
-```{r compdendroc14, fig.cap="Comparison of the scaled (non-centered z-transformed) number of ^14^C and dendro-dated sites over time in the dataset used."}
-joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))
-
-ggplot(joined_data) + geom_line(aes(x=(1950-age)*-1, y=scale(sumcal, center = F), color = "darkred"), alpha = 0.75)  + geom_line(aes(x=(1950-age)*-1, y=scale(dendro, center = F), color = "darkblue"), alpha = 0.75) +
-  theme_minimal() + scale_x_reverse() +
-    labs(x = "years BCE",
-         y = "scaled values",
-         color = "Legend") + 
-    scale_color_manual(values = c("darkblue", "darkred"), labels = c("dendro dated", "14C dated"))+
-  theme(legend.position="bottom")
-```
-
-We can now compare these two data sets (Figure \@ref(fig:compdendroc14)).
-In fact, there is a not uninteresting fit between the two data series.
-However, it must be assumed that the two dating methods, even if they would contradict each other, actually complement each other, and thus allow a better overall unified picture of the actual settlement density than each of the individual proxies would allow on their own.
 
@@
 
-To add another archaeological indicator of occupation, we include relative dating information obtained from the Swiss cantonal archaeology/heritage management authorities (Figure \@ref(fig:aoristmap)).
-These are primarily derived from scattered surface finds, which often have a low dating accuracy.
-We incorporate this data into our model as a typologically dated aorist time series.
-The dating accuracy is only in the range of archaeological periods, but the advantage is that we are not bound to the conditions and problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided.
-Furthermore, this data provides an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data.
-Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy offered by archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.
+We include relative dating information obtained from the Swiss cantonal heritage management (Figure \@ref(fig:aoristmap)).
+These are primarily derived from scattered surface finds, often with a low dating accuracy (only in the range of archaeological periods), incorporate into our model as a typologically dated aorist time series.
+It is not bound to the problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided.
+Data from `r nrow(aorist)` sites were included in the aoristic sum.
 
 ```{r aoristcurve, fig.cap="Aoristic sum of the archaeological sites used in this analysis."}
+joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))
 ggplot(joined_data) + geom_line(aes(x=(1950-age)*-1, y=aoristic_sum, color = "darkgray"), alpha = 0.75) +
@@
 The natural conditions of the lakes of the Swiss Plateau enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis.
-We make use of this fact by generating a supra-regional openness indicator for the vegetation from the pollen data (Figure \@ref(fig:pollensites)).
-This proxy has the specific advantage that it is not dependent on preservation conditions, as archaeological indicators are.
-This makes it particularly valuable for indicating or compensating for systematic distortions that result from temporally specific settlement patterns and archaeological preservation conditions.
+We make use of this by generating a supra-regional openness indicator for the vegetation from the pollen data (Figure \@ref(fig:pollensites)).
+This proxy has the specific advantage that it is not dependent on archaeological preservation conditions, making it particularly valuable for compensating systematic distortions that result from archaeological taphonomy and temporally specific settlement patterns.
 
@@
 
-The utilisation of this proxy is based on the assumption that the higher the population density in an area, the greater the human influence on the natural environment [@lechterbeck2014], and that the extensiveness of agricultural activity in an area is related to human population density [@zimmermann2004].
-Evidence of land clearance in pollen diagrams can therefore provide further indications of population dynamics where humans can be assumed to be the main driver of this process, which is the case in much of Europe.
+We assume that the higher the population density in an area, the greater the human influence on the natural environment [@zimmermann2004, @lechterbeck2014].
+Evidence of deforestation can therefore provide indications of population dynamics.
 The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2021].
-In this study, five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes were used.
-The technical steps are also documented in the accompanying R compendium.
-The percentage pollen data, based on a pollen sum of all terrestrial taxa of the individual sites, was combined into one data set by means of a principal component analysis (Figure \@ref(fig:pollenproxy)).
-Only terrestrial pollen taxa with a frequency of more than one third and, if present, with an average frequency of at least 0.1% were selected to reduce potential disturbance by rare species.
-Cereal pollen was explicitly retained as an important anthropogenic indicator.
-As each sample is absolutely dated, the data on the x-axis can be plotted against the openness value on the y-axis to obtain a time series time series for land clearance.
+Here, we use five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes.
 
@@
 
-## Other proxies
-
-Unfortunately, there is no burial data available in the case study area that could be usefully incorporated into such this model due to the absence of regular burials for larger parts of prehistory.
-Formation conditions, in this case most likely C-transformation [@schiffer1987], the active cultural choice to not use archaeological traceable burial pratices, prevent their use in this study.
-Nevertheless, we see a very high potential for other regions in the integration of demographic indicators from burial data in order to enlarge the canon of methods and the range of proxies.
-
 # Methods: Bayesian model
 
-The sum calibration, openness and the dendro-dated settlement data was smoothed by means of a moving average with a window of 50 years.
+The sum calibration, openness and the dendro-dated settlement data was smoothed by a moving average with a 50 years window, corresponding to the unified sampling interval for all proxies.
 The aoristic sum was not smoothed, because it already has a very coarse temporal resolution.
-The range of the smoothing window corresponded to the sample interval, with which a unified resolution of 50 years was obtained for all proxies as time slices for the model.
-In addition, all data was restricted to the window of observation of 7000--1000 BCE.
-
 In the construction of our ‘observational model’, we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps.
-Consequently, we shift the causality and consideration of measurement error, which is certainly inherent in each of these proxies, to a preceding ‘process model’, in which a Poisson process describes the number of simultaneous settlements.
-In doing so, we establish a likelihood that estimates how credible the data is, given the model.
+Population development is simulated in a ‘process model’ using a Poisson process.
 
@@
 
-A special class of Bayesian hierarchical models are so-called State Space Models (also known as Hidden Markov Models).
-These are specifically designed for time series and follow two basic principles.
-First, a hidden or latent process is assumed, which represents the state of the variable of interest $x_t$ through the entire time series.
-Over time, it is assumed that every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$.
-At the same time, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$.
-This means that there is a relationship between the variable $x$ and the state of variable y over time.
+A special class of Bayesian hierarchical models are so-called State Space Models, specifically designed for time series following two principles.
+First, a hidden or latent process is assumed, representing the state of the variable of interest $x_t$ through the entire time series.
+Every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$.
+Second, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$.
 This implies that a relationship between the individual states of variable $y$ is generated over time via the hidden variable $x$, which itself is not observable.
 
-This basic structure of this class of model makes it particularly useful and suitable for the purpose of demographic reconstruction using archaeological and other data which are reflective of population density in the past.
-This population density itself is not accessible or measurable by our means.
+This structure makes these models particularly suitable for demographic reconstruction using archaeological and other data.
+The population density itself is not directly measurable.
 All we have at our disposal are observations derived by unknown transfer functions.
-These can be of very different natures, such as number of archaeologically observable settlements, or other effects that can be observed through time series, and which are influenced by population in a given area.
-In our case study, these are the openness indicators from pollen data, which we can interpret primarily in terms of human influence and its intensity.
-On a more abstract level, we could also include expert estimates, as these also happen on (often unspecified) bases that are at least indirectly influenced by past population density.
 
-The overall model for the estimation of demographic developments is broken down into several hierarchically interconnected individual elements in accordance with the basic structure of a state space model.
-The basis is a process model that represents the demographic development itself in terms of a structure model, without this model already being explicitly configured with data.
-
-In this process model we assume that the latent variable 'number of sites' is strongly autocorrelated across different time periods, i.e.
-the number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on.
-In principle, one can represent a population development in such a way that the population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.
-This gives us the basic formula:
+The overall model for the estimation of demographic developments is broken down into several hierarchically interconnected individual elements.
+The a process model represents the demographic development itself, without already being explicitly parameterised with data.
+Here we assume that the latent variable 'number of sites' is strongly autocorrelated across different time periods.
+The number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on.
+The population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.
 
@@
 
-The Poisson distribution is particularly suitable for modelling frequencies.
-It is a univariate discrete probability distribution that can be used to model the number of events that occur independently of each other at a constant mean rate in a fixed time interval or spatial area.
-It is determined by a real parameter $\lambda$ \>0, which describes the expected value and simultaneously the variance of the distribution.
-Thus, the relationship shown above can also be rearranged as follows:
+The univariate discrete Poisson distribution is particularly suitable for modelling frequencies, numbers of events that occur independently of each other at a constant mean rate in a fixed time interval or spatial area.
+It is determined by a real parameter $\lambda$ \>0, describing the expected value and the variance.
+Thus, the relationship shown above can be rearranged as follows:
 
@@
 
-If we now have information about the change in population development (the proxies), we can use this to enter it into the model via a change in $\lambda$.
-This is done in the form of a regression: for all proxy values—represented as a vector of independent variables $x \in R^n$, with $R^n$ as an n-dimensional Euclidean space, described in this case by the n dimensions of the n variables—then the model takes the form:
+If we now have information about the change in population development (the proxies), this can enter into the model via a change in $\lambda$ in form of a regression: for all proxy values — represented as a vector of independent variables $x \in R^n$, with $R^n$ as an n-dimensional Euclidean space defined by the n variables — the model takes the form:
 
@@
 
-Using the logarithm as a link function ensures that $\lambda$, which must always be positive for a Poisson process, can also be described by variables (proxies) that range in the space of real numbers and can therefore also be negative.
-$\beta$ can serve here as a slope factor, just as in a normal linear regression.
-In our case, it functions as a scaling factor for the individual proxies.
-$\alpha$, is to be understood as an intercept.
-If there were no change due to the variables, the regression would fall back to this value.
-This corresponds to the desired behaviour.
-In the case of population trends, the intercept would be equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables.
-If there is no change from these variables, then $\lambda$, and thus the expected value for the current time step would be equal to that from the previous time step:
+Using the logarithm as a link function ensures that $\lambda$, which must always be positive, can also be described by variables (proxies) that may also be negative.
+$\beta$ serves as slope factor, as in a normal linear regression.
+Here, it functions as a scaling factor for the individual proxies.
+$\alpha$ is to be understood as an intercept, representing a baseline when there were no change due to the variables.
+This is the desired behaviour: $\lambda$ is equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables.
 
@@
 Since $\lambda$ and $N$ are essentially in the same range (e.g.
-if $lambda=1$, the expected value for $N$ would also be 1), $N_{t-1}$ must also be log-transformed in the above formula to obtain the congruence of both values.
-The values for population size $N_t$ as well as for population change $\lambda_t$ are time-dependent.
-At each individual point in time in the time series, these variables can or will also take on different values.
-However, we can narrow down the structure of population change even further.
-We can assume that, considered overall over time, the population change will not exceed certain limits over the entire timespan, though it is not possible to specify this at this point.
-
-Thus, we can define the limits, the $max\_change\_rate$ as a time independent variable, again without specifying them with fixed values at this stage.
-The estimation of these parameters for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data, respectively.
-Overall, this represents a hierarchical model that can be noted as follows:
+if $lambda=1$, the expected value for $N$ would also be 1), $N_{t-1}$ must also be log-transformed for the congruence of both values.
+Population size $N_t$ as well as population change $\lambda_t$ are time-dependent.
+At each individual point in time, these variables will take on different values.
+But we can assume that the population change will not exceed certain limits ($max\_change\_rate$), though it is not possible to specify this at this point.
 
@@
 
-The gamma distribution used centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$.
-This prevents the number of sites from explosively increasing between two time periods, which would be unrealistic given our sampling interval (50 years), and would lead to problems for the convergence of the model.
-The interaction of these parameters results in the following prior probability distribution for $\lambda$ and thus the growth (or change) rate of the population.
+The gamma distribution centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$.
+This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model.
+The estimation of this parameter for the entire model, as well as the estimation of the respective population change per time section, results from the modelling and the interaction with the data.
 
@@
 
-During the development of our model, we experimented with but ultimately abandoned the idea of implementing dedicated observation models adapted to the conditions of the individual proxies and their generating processes.
-We found that underdetermination of currently usable data in the model, having many degrees of freedom, produced a high degree of equifinality in the solutions and thus led to a high path dependency in individual model runs.
-As a result, it was almost impossible to achieve convergence of the overall model.
-Nevertheless, we believe that in future applications of the model with more data, a larger geographical coverage and/or, especially, a regionalised approach with information transfer by means of partial pooling, this more specific approach would be a feasible and a very useful approach.
-
-The implementation we present in this paper instead represents a Poisson regression, where the proxies are used to inform the change in the number of settlements from time step to time step.
-For this purpose, the individual proxies were z-normalised.
-The absolute differences from one time step to another were then computed from the resulting time series.
-Thus, if the value of the proxy increases, this results in a positive difference from the previous time step, and vice versa.
+The implementation of the observational model in this paper is a Poisson regression, where the proxies are used to inform the change in the number of settlements between time steps.
+The individual proxies were z-normalised, absolute differences between time steps were then computed.
+If the value of the proxy increases, this results in a positive difference from the previous time step, and vice versa.
 
@@
 
-A probability distribution that can be used for this purpose in a hierarchical Bayesian model is the Dirichlet distribution, which is a multivariate generalization of the beta distribution, commonly used as prior distributions in Bayesian statistics.
+A Dirichlet distribution, a multivariate generalization of the beta distribution, is commonly used for this purpose in a hierarchical Bayesian model.
 Its density function gives the probabilities of $i$ different exclusive events.
@@
 
-As an intuition, this means that we consider the sum of the proxies as determinant for the number of settlements.
-The estimation therefore assumes that all proxies together give the best possible estimation result for contemporaneous sites at time t, whereby the share of each individual proxy is considered variable and is estimated within the model.
+As an intuition, we consider the sum of the proxies as determinant for the number of settlements, whereby the share of each individual proxy is variable and is estimated within the model.
 This share is recorded within the model as the parameter p.
 
-The error value is represented by the Poisson process in the process model, rather than directly as an estimation error for the individual proxies.
-Thus, our model does not correspond to a classical state space model, where the measured values are each considered to be error-prone.
-In the implementation, the model finds the best possible combination or compromise between the individual proxies to describe a settlement dynamic that is given by them.
-In addition, the number of sites is converted into population density using some (certainly debatable) parameters that we have defined but which are only scaling factors for the intermediate value of number of settlements.
-For this, we assume that each site represents a number of people that is poisson distributed around the value 50.
-The number 50 represents a compromise, as both Mesolithic and Neolithic and Bronze Age settlement communities need to be represented.
-By means of a data series, which would represent an evidence-based estimate of the temporal development of settlement sizes, this specification could be made based on data.
-From the number of sites and the mean number of individuals represented in each case, a population density can be calculated using the case study area (12649 km²).
-The estimated result of the model is thus comparable with estimates from other sources or the literature.
+The error value is represented by the Poisson process in the process model.
+In this implementation, the model finds the best possible combination between the individual proxies to describe a settlement dynamic.
+The number of sites is converted into population density using (certainly debatable) parameters defined by us, but which are only scaling factors for the intermediate value of number of settlements.
+We assume that each site represents a number of people that is poisson distributed around the value 50, a compromise, as both Mesolithic and Neolithic and Bronze Age settlement communities need to be represented.
+An evidence-based estimate data series of the temporal development of settlement sizes could enhance this specification.
+From the number of sites and the mean number of individuals a population density can be calculated using the case study area (12649 km²), making the models estimate comparable with estimates from other sources or the literature.
 
@@
 
-In earlier implementations, expert estimates were also integrated into the model.
-However, since these are highly contradictory for the working area (Figure \@ref(fig:expertestimations), we found they had little influence on the model and also significantly increased its runtime.
-We therefore excluded them in the implementation presented here, but again, given a larger geographical range and thus a higher information density, this proxy could be usefully integrated into future applications.
-
 ## Model fitting
 
-The model was fitted using the R package *nimble* (version 0.11.1, R version 4.1.3).
-For this purpose, 4 chains were run in parallel.
+The model was fitted using the R package *nimble* (version 0.11.1, R version 4.1.3), using 4 parallel chains.
 Achieving and ensuring convergence and sufficient effective samples (10000) for a reliable assessment of the highest posterior density interval was carried out in steps.
 
-In a first run, the model was initialised for each chain and run for 100000 iterations (with a thinning of 10).
+1) the model was initialised for each chain and run for 100000 iterations (with a thinning of 10).
 On a reasonably capable computer (Linux, Intel(R) Xeon(R) CPU E3-1240 v5 \@ 3.50GHz, 4 cores, 8 threads), this takes approximately a minute.
 
-In a second step, the run was extended until convergence could be determined using Gelman and Rubin's convergence diagnostic, the criterion being that a potential scale reduction factor of less than 1.1 was achieved for all monitored variables.
+2) the run was extended until convergence could be determined using Gelman and Rubin's convergence diagnostic, the criterion being that a potential scale reduction factor of less than 1.1 was achieved for all monitored variables.
 Convergence occurred after about thirty seconds.
 
-Due to the high correlation of the parameters and thus a low sampling efficiency, the collection of at least 10,000 effective samples for all parameters in the third step took about five hours.
+3) Due to the high correlation of the parameters and thus a low sampling efficiency, the collection of at least 10,000 effective samples for all parameters took about five hours.
 
-For the fitting process, a starting value of 5 p/km² for a population density of the Late Bronze Age (1000 BCE) was taken from the literature, which may represent a general average value for all prehistoric population estimates [@nikulka2016, 258].
+A starting value of 5 p/km² for a population density of the Late Bronze Age (1000 BCE) was taken from the literature, which may represent a general average value for all prehistoric population estimates [@nikulka2016, 258].
 For the model, this was set as the mean of a normal distribution with a standard deviation of 0.5, which should give enough leeway for deviations resulting from the data.
@@
 
-The population density estimated by the model (Figure \@ref(fig:popdensplot)) ranges on between 0.2 p/km² for the beginning of the estimate (6000 BCE) and 4.8 p/km² for the end of the estimate (1000 BCE), reaching a maximum of 6.5 p/km² for the time slice 1250 BCE.
-Thus, the estimate remains within the values that are also considered plausible by the expert estimates.
-Clear peaks are reached around 1250 BCE, as well as around 2750 BCE, which corresponds to the transition to a ceramic style as observed for sites of the Swiss Plateau influence by the Corded Ware ceramic style [@hafner_vom_2004].
+The population density estimated by the model (Figure \@ref(fig:popdensplot)) ranges between 0.2 p/km² for the beginning (6000 BCE) and 4.8 p/km² for the end of the estimate (1000 BCE), reaching a maximum of 6.5 p/km² for around 1250 BCE, remaining within the values considered plausible by the expert estimates.
+Clear peaks are around 1250 BCE, as well as around 2750 BCE, which corresponds to the beginning of the influence of the Corded Ware ceramic style [@hafner_vom_2004].
 
@@
 
-The temporal distribution of the variability in the estimate (Figure \@ref(fig:varcoeffplot)) allows us assess at which time steps the model attempts a more accurate estimate, and at which the uncertainty is greater due to e.g.
+The temporal distribution of the variability in the estimate (Figure \@ref(fig:varcoeffplot)) allows us assess at which time steps the uncertainty is greater due to e.g.
 contradictions in the proxies.
@@
 2200-1800 BCE, this picture changes from c.
-1800 BCE onwards [@hafner_fruhe_1995, @david-elbiali_suisse_2000].
-The beginning and end of the time series are relatively clearly determined.
-The end results from the *a priori* setting of the parameter, but also here, as at the beginning of the series, the proxies are very uniform, which explains the low variability.
-Overall, the variability is relatively uniform over the entire course of the estimation and averages over all time slices at 33% of the respective mean.
-
-Within the model, the parameter $p$ was estimated, which reflects the relative weight given to the individual proxies used in the estimation of the number of settlements.
-This parameter is variable, but has only a scaling influence on the final estimate of population density.
+1800 BCE onwards [@hafner_fruhe_1995; @david_elbiali_suisse_2000].
+The beginning and end of the time series are relatively clearly determined, resulting from the *a priori* setting of final population density, but also from the uniformity of the proxies during these periods.
+Overall, the variability is relatively stable over the entire estimation and averages overall at 33% of the respective mean.
 
@@
 
-By looking at the distribution of posterior samples for the share of each proxy (Figure \@ref(fig:pdistribution)), it is clear that the model weights the openness indicator the highest.
-The average is slightly above 60%.
-The next most important indicator is the sum calibration value, which has an average of about 20%.
+The parameter $p$ reflects the relative weight given to the individual proxies.
+Its posterior distribution (Figure \@ref(fig:pdistribution)) shows that the model weights the openness indicator the highest, averaging slightly above 60%, followed by the sum calibration, with an average of about 20%.
 The aoristic sum is slightly above 10%, whereas the importance of the dendro-dated settlements is below 10%.
-The reason for the latter is certainly that there are no lakeshore settlements over large areas of the time window, and therefore the proxy achieves a low confidence value in comparison with the other estimators.
-In the case of the aoristic sum, it is certainly the fact that it is flat over large sections and has little structure, making it difficult to relate to the other estimators.
-The sum calibration shows very strong short-term fluctuations, which are presumably at least partly due to the effects of the calibration curve, and which also make this proxy seem ill-suited to reliably represent a continuous population trend.
-Nevertheless, its fluctuations do have an impact on the resulting overall estimate of the development of the number of settlements, albeit to a lesser extent.
+The reason for the latter is certainly that there are no lakeshore settlements over large areas of the time window, and therefore the overall confidence in the proxy is low.
+The aoristic sum is flat over large sections, making it difficult to relate to the other estimators.
+The sum calibration shows very strong short-term fluctuations, at least partly due to the calibration curve, which also make this proxy seem ill-suited to reliably represent a continuous population trend.
+Its fluctuations have an impact on the overall estimate, albeit to a lesser extent.
 
@@
 
-Our original ambition in the development of this model was to base it on a state-space representation of the demographic process itself, and then to integrate the existing proxies by means of an observation model to inform this process.
-In the course of developing the model, we realised that the data we had available for the case study area north of the Swiss Alps  alone was not sufficient to adequately determine this process, or to adequately fix the degrees of freedom resulting from the process and the transfer model.
-Therefore, in implementation presented here, we have turned to a Poisson regression approach: the model represents the best possible combination of the indicators used to describe the development of the number of sites based on them.
-
 ## Comparison of individual proxies
 
-Comparing the model's overall estimate with trends indicated by individual proxies gives us new insights into the quality of these records.
-Sum calibration, currently the most frequently used proxy for (relative) population changes in prehistory, has its large fluctuations dampened when considered alongside other proxies.
-This is especially true for the first fluctuation, shortly after 4000 BCE.
-The expected increase in archaeological remains with the onset of Neolithisation is still clearly visible, but the curve of the overall estimate, after the initial increase, is much flatter than sum calibration alone indicates.
-The period between 3950 and 3700 BCE, which is contemporaneous with the first larger settlement cycle on the lakeshores of the Three Lakes region, coincides with a noticeable plateau on the calibration curve, which be producing an overestimation of the ^14^C density.
-However, the effect of the calibration curve on the results of a cumulative calibration cannot yet be considered unambiguous.
-A second maximum, after 3000 BCE, is supported by the other proxies, and is consequently also much more clearly reflected in our overall estimate.
-Here, too plateau in the calibration curve — albeit a smaller, shorter one which is much less pronounced than, for example, the one shortly before between 3350 and 3100.
-The rise towards the Middle and Late Bronze Age is also supported by the other proxies, especially the aoristic sum, and therefore is preserved in the overall estimate.
-In this period, the calibration curve does not show any very clear, significant patterns.
+Comparing the model's overall estimate with the individual proxies enables insights into the quality of these records.
+The sum calibration, currently the most frequently used proxy for (relative) population changes in prehistory, has its large fluctuations dampened when considered alongside other proxies, especially the first fluctuation shortly after 4000 BCE.
+The expected increase in archaeological remains with the onset of Neolithisation is still clearly visible, but the overall curve is much flatter than the sum calibration itself.
+The period between 3950 and 3700 BCE, contemporaneous with the first major settlement of the Three Lakes regions' lakeshores, coincides with a noticeable plateau in the calibration curve, producing an overestimation of the ^14^C density.
+A second maximum, after 3000 BCE, is supported by the other proxies, and is consequently much more reflected in the overall estimate, coinciding with a smaller and shorter plateau.
+The rise towards the Middle and Late Bronze Age is also supported by the other proxies, without a significant pattern in the calibration curve.
 We may conclude that the model is successful in using information from other proxies to sift 'real' fluctuations in the summed radiocarbon record from artefacts of the calibration curve.
@@
 After an initial increase, which is easily explained by spread of agriculture, the openness indicator tends to fluctuate less and thus has a dampening effect on the overall estimate.
-Nevertheless, it appears the general trends in the sum calibration are well reflected in land openness – even if such 'eyeballing' should be interpreted with caution.
-In contrast to this dramatic trend, changes within the Neolithic and Bronze Age are more gradual, suggestive of larger scale transformations of the landscape.
-The model is designed rather conservatively by the limit of maximum growth, which is estimated within the model but influences it by its very presence.
-Therefore, this proxy corresponds better and more homogeneously to a smooth increase in the number of settlements than do the strong fluctuations in the ^14^C data set.
+In general, this trend in the sum calibration is well reflected in land openness, while changes within the Neolithic and Bronze Age are more gradual.
 
-The aoristic sum remains relatively even over long spans of time.
+The aoristic sum remains flat over long spans of time.
 It is not until the Middle and Late Bronze Age that we see a significant rise, which is also apparent in the model's overall estimate.
 It remains to be seen to what extent modelling of the taphonomic loss [@surovell2009] could be integrated in this approach.
-We have refrained from doing so in this first model, as this would have introduced further degrees of freedom - but we are aware that with a broader database this would be an interesting possibility, and that it would itself be a variable to be estimated, e.g.
-in connection with proxies that are not influenced by it (openness indicators, but also data from the demography of burial collectives).
-This would make it possible to estimate a value from original archaeological material independently of variables that have little to do with archaeological data, such as volcanic eruptions [@ballenger_temporal_2011].
 
-The number of simultaneously existing lakeshore settlements is a very limited temporal and spatial estimator, but extremely reliable.
-Its limitations are reflected in the low overall confidence of the model, since its value is zero over long stretches, while other indicators suggest clearly different patterns.
-However, where it has information potential, such as around and shortly after 3800 BCE, 3200 BCE or especially around 2750 BCE, its fluctuations have a noticeable influence on the overall estimate.
-The peak around 1600 BCE also leaves a noticeable impact.
+The number of simultaneously existing lakeshore settlements is a temporally and spatially limited estimator, but extremely reliable.
+Its limitations are reflected in the low overall confidence of the model, since its value is zero over long stretches.
+However, where it has information potential, such as around and shortly after 3800 BCE, 3200 BCE, 1600 BCE or especially around 2750 BCE, its fluctuations have a noticeable influence on the overall estimate.
 This highlights another potential of our approach: where a proxy has little structure and thus little significance, or where its trends cannot be linked to other indicators, it consequently has little influence.
@@
 In order to review the reconstruction against the background of established archaeological knowledge, it is useful to overlay conventionally-defined archaeological phase boundaries [@hafner_neolithikum_2005] on the results of our model (Figure \@ref(fig:popestwithphases)).
-It should be noted, however, that the model's estimate is not completely independent of these phases: due to the aoristic sum, which is itself strongly determined by this phase division, corresponding boundaries also influence the structure of the reconstruction.
-Nevertheless, it helps to check whether the estimate is in strong contradiction to the generally accepted picture or whether it is able to make a credible prediction within this context.
 
@@
   file = normalizePath(
-            file.path(here(), "data","raw_data", "chrono_rough.csv"
+            file.path(here(), "data","raw_data", "chrono_rough_de.csv"
                       )
@@
     theme_minimal() + scale_x_reverse() +  
-ylab("Scaled values for Proxies,\n P/km² for Estimation") +
+ylab("P/km²") +
   xlab("cal BCE") +
@@
 
-The Older and Middle Neolithic phases are in fact hardly documented with known sites in Switzerland.
-Here we must assume a basically low level of settlement, probably mainly by hunter-gatherer groups.
+The Early and Middle Neolithic are hardly documented with known sites in Switzerland.
+We must assume a low level of settlement, probably mainly by mobile groups.
 Isolated Neolithic sites of the LBK and later groups are known in the periphery of Switzerland, but they play a subordinate role [@ebersbach_nutzung_2012].
-The evidence of the Neolithic especially what the Swiss Plateau is concerned is dense from the so-called Upper Neolithic onwards, connected with the typo-chronological units based on pottery that are called Egolzwil (late 5th M BCE) and Cortaillod respectively Pfyn (first half of the 4th M BCE).
+The evidence of the Neolithic is dense from the so-called Upper Neolithic onwards, connected with the typochronological pottery phases of Egolzwil (late 5th M BCE) and Cortaillod respectively Pfyn (first half of the 4th M BCE).
 The first lake shore settlements north of the Alps date to this time too.
-In this period, we also see a clear increase in the estimated population in the model.
+Here we see a clear increase in the estimated population in the model.
 In the transition to the Late Neolithic we know from the lakeshore settlements the so-called Horgen Gap [@hafner_neolithikum_2005].
@@
 In another study [@heitz2021] we could show that this is in fact probably not a decline in population.
-Rather communities relocated their settlements to the hinterland of the large lakes in times of lake level rises of higher magnitudes due to climatic changes.
-In the Late Neolithic, associated with the Horgen pottery, we then see a clear increase in the intensity of settlement, which reaches its peak and its break-off at the transition to the Corded Ware and thus to the Final Neolithic [@hafner_vom_2004].
-In the second half of the Early Bronze Age, during which – after a break from the Final Neolithic to the very early Bronze Age – lakeshores were resettled, but to a smaller extend.
-There is again a clear increase in population size according to the model, which continues until the Late Bronze Age.
+Rather communities relocated their settlements to the hinterland of the large lakes in times of stronger lake level rises due to climatic changes.
+In the Late Neolithic, associated with the Horgen pottery, we then see a clear increase in the settlement intensity, which peaks and breaks off at the transition to the Final Neolithic [@hafner_vom_2004].
+In the second half of the Early Bronze Age, during which lakeshores were resettled to a smaller extend, there is again a clear increase in population size according to the model, continuing until the Late Bronze Age.
 The general trends fit very well with the previous reconstructions of population development for Switzerland [see eg.
-@lechterbeck2014].
-All in all, the estimate of the model corresponds to our expectations, although we must be aware that these expectations are not 'ground truth'.
-Nevertheless, we can claim that the model predictions are credible because they accord well with existing knowledge, while offering higher precision and higher resolution.
+@lechterbeck2014], while offering higher precision and higher resolution.
 
@@
 
-The key advance in the model we present here is the ability to estimate, in absolute terms, past population sizes and the uncertainty accompanying our present knowledge.
-These estimates can be used as a basis for further studies where relative measures of population development are not helpful, such as long-term land use studies, where modelling of large-scale socio-ecological systems based on archaeological data becomes possible and does have to rely on using deductive, asynchronous population models (e.g.
+The key advance in the model we presented is the ability to estimate, in absolute terms, past population sizes and the uncertainty accompanying our present knowledge.
+These estimates can be a basis for further studies where relative measures of population development are not helpful, such as long-term land use studies.
+Modelling of large-scale socio-ecological systems based on archaeological data does not have to rely deductive, asynchronous population models (e.g.
 carrying capacity or ethnographic analogues).
 
-We have also demonstrated that, with Bayesian hierarchical modelling, it is possible to achieve a true multi-proxy analysis of prehistoric demographic processes – as opposed to a juxtaposition of different indicators.
+We have also demonstrated that, with Bayesian hierarchical modelling, it is possible to achieve a true multi-proxy analysis – as opposed to a juxtaposition of different indicators.
 This opens up the possibility of quantitatively linking different records and assessing their credibility.
-In addition, and in contrast to existing approaches, we are able to specifying a confidence interval for the overall estimate.
+We are also able to specifying a confidence interval for the overall estimate.
 The result is a firmer basis for reconstructing population dynamics and settlement patterns in prehistory.
 
-Nevertheless, we consider the model present here as only the first step towards a more sophisticated Bayesian approach.
-We have trusted the individual proxies in aggregate, without considering measurement error.
-Moreover, our estimates are based on a limited number of sources of evidence, almost all of which are subject to taphonomic biases in the archaeological record (except for the openness indicator).
-Consequently, we can only transform the model's prediction into an absolute estimate of population density with predefined parameters: the upper limit of population growth, settlement size and the initial value of the reconstruction.
+Nevertheless, we consider this model as only the first step towards a more sophisticated Bayesian approach.
+We have trusted the individual proxies in aggregate, without individualised measurement error.
+Our estimates are based on a limited number of sources, almost all of which are subject to taphonomic biases in the archaeological record.
+Consequently, we can only transform the model's prediction into an absolute estimate of population density with predefined parameters: settlement size and the initial value of the reconstruction.
 Overcoming this limitation would represent a major refinement of our approach.
 
-Incorporating additional proxies that are independent of the immediate, time-dependent conditions of the archaeological record could be one way to achieve this.
-These could be data on settlement sizes, parameters for economic-ecological carrying capacity, demographic data from burial groups, archaeogenetic data on population sizes, or other records as yet unidentified.
+Incorporating additional proxies independent of the immediate, time-dependent conditions of the archaeological record could be one way to achieve this.
+These could be data on settlement sizes, parameters for economic-ecological carrying capacity, demographic data from burial groups or archaeogenetic data on population sizes.
 This data is available to varying degrees in different regions.
-On the Swiss Plateau, for example, we have too little data on human remains over large spans of prehistory compared to the abundance of wetland settlements to be able to integrate them meaningful into a model.
+On the Swiss Plateau, for example, we have too little data on human remains over large spans of prehistory compared to the abundance of wetland settlements.
 
-To apply the model to different regions, the proxies we use here (e.g.
-the palynological openness indicator) would have to be adapted to fit local conditions and research histories.
-By means of large-scale modelling, however, it would be possible to supplement gaps in the data in one region with data from other regions by means of regionalisation and a partial transfer of information (partial pooling).
+To apply the model to different regions, the proxies we use here would have to be adapted to fit local conditions and research histories.
+By means of large-scale modelling, however, it would be possible to supplement gaps in the data in one region with data from another by regionalisation and a partial transfer of information (partial pooling).
 Such an extension would be the next logical step in the improvement of the model, to which end we hope to be able to contribute a further study in the near future.
@@
 
-Large parts of the data collection were conducted as part of the project 'Beyond lake settlements' in the doctoral thesis of Julian Laabs, funded by the SNF (project number 152862, PI Albert Hafner).
-Further data collection was done as part of the XRONOS project, also funded by the SNSF (project number 198153, PI Martin Hinz).
-The development of the openness index took place within the framework of the project Time and Temporality in Archaeology (project number 194326, PI Caroline Heitz) and was inspired by the cooperation within the project QuantHum (project 169371, PI Marco Conedera), both also funded by the SNSF.
+Data collection were conducted as part of the project 'Beyond lake settlements' in the doctoral thesis of Julian Laabs, funded by the SNF (project number 152862, PI Albert Hafner) and as part of the XRONOS project, also funded by the SNSF (project number 198153, PI Martin Hinz).
+The development of the openness index took place within the framework of the project Time and Temporality in Archaeology (project number 194326, PI Caroline Heitz), inspired by the cooperation within the project QuantHum (project 169371, PI Marco Conedera), both also funded by the SNSF.
 Jan Kolář was supported by a long-term research development project (RV67985939) and by a grant from the Czech Science Foundation (19-20970Y).

### 2022-05-28 manuscript/paper_shorter.Rmd
Further copyedits

@@
 
-## Prehistoric demography
-
 Prehistorians have long recognised demography as a fundamental force in human cultural evolution [@childe_man_1936].
-Despite decades of interest in the demographic dynamics of prehistoric societies, concrete estimates of population size and density before written records remain elusive.
+Despite decades of interest in the population dynamics of prehistoric societies, concrete estimates of population size and density before written records remain elusive.
 Though the archaeological record provides multiple possible demographic proxies [@muller_tracing_2019], a lack of access to this data and methodological tools for turning it into quantitative estimates has left the conclusions drawn from it vague and superficial [@hassan_demographic_1981].
@@
 
-Prehistoric demography has experienced a resurgence in recent years [@shennan_population_2000; @riede_climate_2009 and others in same issue], partly explained by a renewed interest in human–environment relations and human impact, necessarily requiring an assessment of population size.
+Prehistoric demography has experienced a resurgence in interest in recent years [@shennan_population_2000; @riede_climate_2009 and others in same issue], partly explained by a renewed interest in human–environment relations and human impact, necessarily requiring an assessment of population size.
 Kintigh et al.
@@
 
-In the last decade, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, was taken up and significantly developed [e.g.
-@shennan_regional_2013], and has since been widely applied to archaeological contexts worldwide [@crema_review_2022].
+In particular, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, has been significantly developed in the last decade [e.g.
+@shennan_regional_2013] and widely applied to archaeological contexts worldwide [@crema_review_2022].
 This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015, @carleton_groucutt_2020].
@@
 
-## Existing approaches to population estimation
+## Population estimation in prehistory
 
@@
 1.
-**Reliance on single proxy**: Most investigations are using only one source of evidence.
-Although multi-proxy approaches exist, there the individual proxies only serve to support each other or the main estimator, without explicitly combining them to reduce the inaccuracy and biases inherent to single proxies.
+**Reliance on single proxy**: Most investigations use only one source of evidence.
+Although multi-proxy approaches exist, the individual proxies only serve to support each other or the main estimator, without explicitly combining them.
 2.
@@
 **Lack of a transfer function**: By 'transfer function', we mean something that allows for the proxy data to be interpreted in terms of actual population size or density.
-These could be absolute, i.e.
+This could be absolute, i.e.
 a numerical estimate of population, or relative, i.e.
@@
 
--   **Limited**: We have only incomplete data, and they are usually not very informative.
+-   **Limited**: We have only incomplete data, and it is usually not very informative.
 -   **Unevenly distributed**: For example, although there is a good data on settlement frequencies for some regions, these regions are very unevenly distributed over time and space.
@@
 
-We argue that many, if not all, of these problems can be ameliorated through a) the explicit, quantitative integration of multiple proxies; and b) the use of a Bayesian approach to take account of and estimate uncertainty.
+Many, if not all, of these problems can be ameliorated through a) the explicit, quantitative integration of multiple proxies; and b) the use of a Bayesian approach to take account of and estimate uncertainty.
 
@@
 
-The well-established application of a Bayesian approach in radiocarbon calibration, where it is used to model radiometric uncertainty based on stratigraphic information, represents a very similar use case in archaeology.
+Bayesian radiocarbon calibration is a similar, well-established application in archaeology, where radiometric uncertainty is modelled based on prior stratigraphic information.
 More recently, archaeologists have also used Bayesian modelling techniques for testing hypotheses relating to demographic models based on ^14^C data [e.g.
@@
 
-## Assumptions of a Bayesian approach
+## The Bayesian approach
 
-Bayesian statistics rely on the premise that there is always some prior assumption, even if very rough, about the probability of an event.
+Bayesian statistics relies on the premise that there is always some prior assumption, even if very rough, about the probability of an event.
 This assumption is adjusted by observing data, by checking how credible these priors are [likelihood, see also @bryant2018, 66].
@@
 
-# Materials: Population proxy data
+# Materials: population proxy data
 
 Our case study area north of the Swiss Alps (Figure \@ref(fig:mapswissplateau)) covers about one third of Switzerland’s territory and comprises the partly flat, but largely hilly area between the Jura Mountains and the Alps.
-It is favourable for settlement and agriculture, especially the Swiss Plateau between Lake Zurich and Lake Geneva, by far the most densely populated region of the Switzerland today.
-This serves as core region in our case study, because here archaeological data is most abundant and accessible.
-Depending on the region, it has a very diverse natural landscape: shaped by glaciers during the ice ages, the many lakes and bogs provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, and a rich source for vegetation reconstructions by means of pollen analyses.
+It is favourable for settlement and agriculture; the Swiss Plateau between Lake Zurich and Lake Geneva is by far the most densely populated region of the Switzerland today.
+This serves as our core region of interest because it is here that archaeological data is most abundant and accessible.
+The region has a very diverse natural landscape: shaped by glaciers during the ice ages, the many lakes and bogs provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, and a rich source for vegetation reconstructions by means of pollen analyses.
 Thanks to the very active and efficient archaeological research and heritage management there is an abundance of archaeological information, including known sites as well as dendrochronological and ^14^C data.
@@
 
-Our case study targets the time window between 6000–1000 BCE.
-Its lower limit was chosen to avoid the so-called 'Hallstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, causing difficulties for the ^14^C proxy.
+Our case study targets the period between 6000–1000 BCE.
+The lower limit of this time window was chosen to avoid the so-called 'Hallstatt plateau' in the Northern Hemisphere radiocarbon calibration curve, which causes difficulties for the ^14^C proxy.
 The upper limit coincides with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence.
 
-A large number of different proxies can be integrated into an overall model, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created.
+A large number of different proxies can be integrated into a model of this type, provided that these observations a) can be understood as dependent on the population density in the past, and b) a model-like description of this dependence can be created.
 Table \@ref(tab:tableproxies) provides a non-exhaustive list.
-For our case study, we used: a landscape openness indicator; an aoristic sum of typological dated sites; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).
+For our case study, we used a landscape openness indicator; an aoristic sum of typological dated sites; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).
 
@@
 
-```{r aoristmap, fig.cap="The location of the sites from the find reports of the cantonal archaeology (heritage management). The locations are 'fuzzed' by ~1km."}
+```{r aoristmap, fig.cap="Location of the sites from the find reports of cantonal archaeology (heritage management) authorities. Locations are 'fuzzed' by approximately 1 km."}
 my_basemap + geom_point(data = aorist_latlng_coords, aes(x = X, y = Y), color = "black", alpha = 0.75) + blank() + my_scalebar
@@
 
-We include relative dating information obtained from the Swiss cantonal heritage management (Figure \@ref(fig:aoristmap)).
-These are primarily derived from scattered surface finds, often with a low dating accuracy (only in the range of archaeological periods), incorporate into our model as a typologically dated aorist time series.
-It is not bound to the problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided.
+We include relative dating information obtained from the heritage authorities of the Swiss cantons (Figure \@ref(fig:aoristmap)).
+These are primarily derived from scattered surface finds, often with a low dating accuracy (only in the range of archaeological periods), incorporated into our model as a typologically-dated, aorist time series.
+However, it is not dependent radiocarbon dating and thus it avoids the methodological issues of sum calibration.
 Data from `r nrow(aorist)` sites were included in the aoristic sum.
 
-```{r aoristcurve, fig.cap="Aoristic sum of the archaeological sites used in this analysis."}
+```{r aoristcurve, fig.cap="Aoristic sum of archaeological sites used in the analysis."}
 joined_data <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "preprocessed_data", "all_proxies.csv"))
@@
 
-The natural conditions of the lakes of the Swiss Plateau enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis.
+Natural conditions in the Swiss lakes enable not only highly precise dating of archaeological sites, but also a very dense network of pollen analysis.
 We make use of this by generating a supra-regional openness indicator for the vegetation from the pollen data (Figure \@ref(fig:pollensites)).
-This proxy has the specific advantage that it is not dependent on archaeological preservation conditions, making it particularly valuable for compensating systematic distortions that result from archaeological taphonomy and temporally specific settlement patterns.
+This proxy has the specific advantage that it is not dependent on archaeological preservation conditions, making it particularly valuable for compensating systematic distortions that result from archaeological taphonomy and period-specific settlement patterns.
 
@@
 
-The sum calibration, openness and the dendro-dated settlement data was smoothed by a moving average with a 50 years window, corresponding to the unified sampling interval for all proxies.
+Sum calibration, openness and the dendro-dated settlement data was smoothed by a moving average with a 50 years window, corresponding to the unified sampling interval for all proxies.
 The aoristic sum was not smoothed, because it already has a very coarse temporal resolution.
-In the construction of our ‘observational model’, we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps.
-Population development is simulated in a ‘process model’ using a Poisson process.
+In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps.
+Population development is simulated in a 'process model' using a Poisson process.
 
@@
 
-A special class of Bayesian hierarchical models are so-called State Space Models, specifically designed for time series following two principles.
+A special class of Bayesian hierarchical models are so-called 'state space models', specifically designed for time series.
+They follow two principles.
 First, a hidden or latent process is assumed, representing the state of the variable of interest $x_t$ through the entire time series.
@@
 Second, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$.
-This implies that a relationship between the individual states of variable $y$ is generated over time via the hidden variable $x$, which itself is not observable.
+This implies that a relationship between the individual states of variable $y$ is generated over time via the hidden variable $x$, which is not directly observable.
 
 This structure makes these models particularly suitable for demographic reconstruction using archaeological and other data.
-The population density itself is not directly measurable.
-All we have at our disposal are observations derived by unknown transfer functions.
+Population density itself is not directly measurable: all we have at our disposal are observations derived by unknown transfer functions.
 
-The overall model for the estimation of demographic developments is broken down into several hierarchically interconnected individual elements.
+Our overall model is broken down into several hierarchically-connected individual elements.
 The a process model represents the demographic development itself, without already being explicitly parameterised with data.
@@
 
-The univariate discrete Poisson distribution is particularly suitable for modelling frequencies, numbers of events that occur independently of each other at a constant mean rate in a fixed time interval or spatial area.
+A univariate discrete Poisson distribution is particularly suitable for modelling frequencies, numbers of events that occur independently of each other at a constant mean rate in a fixed time interval or spatial area.
 It is determined by a real parameter $\lambda$ \>0, describing the expected value and the variance.
@@
 
-Using the logarithm as a link function ensures that $\lambda$, which must always be positive, can also be described by variables (proxies) that may also be negative.
+Using the logarithm as a link function ensures that $\lambda$, which must always be positive, can also be described by variables that may also be negative.
 $\beta$ serves as slope factor, as in a normal linear regression.
@@
 
-The gamma distribution centres the probability in a range $[0,1[$, adding 1 makes this range $[1-2[$.
+A gamma distribution centres probability in the range $[0,1[$; adding 1 makes this range $[1-2[$.
 This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model.
@@
 
-The implementation of the observational model in this paper is a Poisson regression, where the proxies are used to inform the change in the number of settlements between time steps.
-The individual proxies were z-normalised, absolute differences between time steps were then computed.
+In this initial implementation, the observational model is essentially a Poisson regression, where the proxies are used to inform the change in the number of settlements between time steps.
+The individual proxies were z-normalised and absolute differences between time steps were then computed.
 If the value of the proxy increases, this results in a positive difference from the previous time step, and vice versa.
@@
 
-The sum of the resulting differences between the time steps, together with the settlement number of the previous step as the expected value, then forms $\lambda_t$ as the expected value for the settlement number of the current time step.
+The sum of the resulting differences between the time steps, together with the settlement number of the previous step as the expected value, then forms $\lambda_t$: the expected value for the settlement number of the current time step.
 
@@
 
-A Dirichlet distribution, a multivariate generalization of the beta distribution, is commonly used for this purpose in a hierarchical Bayesian model.
+A Dirichlet distribution—a multivariate generalization of the beta distribution—is commonly used for this purpose in hierarchical Bayesian modelling.
 Its density function gives the probabilities of $i$ different exclusive events.
@@
 
-As an intuition, we consider the sum of the proxies as determinant for the number of settlements, whereby the share of each individual proxy is variable and is estimated within the model.
+Intuitively, we consider the sum of the proxies as determinant of the number of settlements.
+That is, the share of each individual proxy is variable and is estimated within the model.
 This share is recorded within the model as the parameter p.
@@
 
-```{r expertestimations, fig.cap="Expert estimations for the population density on the Swiss Plateau from different authors."}
+```{r expertestimations, fig.cap="Expert estimate of population density on the Swiss Plateau."}
 expert_estimations <- read.csv(file.path(rstudioapi::getActiveProject(), "data", "raw_data", "estimates.csv"))
@@
 
-A starting value of 5 p/km² for a population density of the Late Bronze Age (1000 BCE) was taken from the literature, which may represent a general average value for all prehistoric population estimates [@nikulka2016, 258].
+A starting value of 5 p/km² for the population density of the Late Bronze Age (1000 BCE) was taken from the literature, which may represent a general average value for all prehistoric population estimates [@nikulka2016, 258].
 For the model, this was set as the mean of a normal distribution with a standard deviation of 0.5, which should give enough leeway for deviations resulting from the data.
-Nevertheless, especially the late part of the reconstruction is of course clearly influenced by this predefined value.
+Nevertheless, it should be noted that our resulting estimate is strongly conditioned by this predefined value, especially in the later sections.
 
-For the traceplots and the prior-posterior overlap as well as the density functions of the posterior samples of the individual parameters, please refer to the supplementary material.
+For traceplots and the prior-posterior overlap, as well as density functions of the posterior samples of the individual parameters, please refer to the supplementary material.
 
@@
 
-```{r popdensplot, fig.cap="The estimate of the population density resulting from the model and the four proxies, which are also plotted (scaled) for comparison."}
+```{r popdensplot, fig.cap="Estimate of population density predicted by the model. The four input proxies are also plotted (scaled) for comparison."}
 knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popdens_estimation.pdf"))
@@
 
-The population density estimated by the model (Figure \@ref(fig:popdensplot)) ranges between 0.2 p/km² for the beginning (6000 BCE) and 4.8 p/km² for the end of the estimate (1000 BCE), reaching a maximum of 6.5 p/km² for around 1250 BCE, remaining within the values considered plausible by the expert estimates.
-Clear peaks are around 1250 BCE, as well as around 2750 BCE, which corresponds to the beginning of the influence of the Corded Ware ceramic style [@hafner_vom_2004].
+The population density estimated by the model (Figure \@ref(fig:popdensplot)) ranges between 0.2 p/km² for the beginning (6000 BCE) and 4.8 p/km² for the end of the estimate (1000 BCE), reaching a maximum of 6.5 p/km² for around 1250 BCE.
+This remains within the range considered plausible according to expert estimates.
+There are clear peaks around 1250 BCE and around 2750 BCE, which corresponds to the beginning of the influence of Corded Ware ceramic styles [@hafner_vom_2004].
 
-```{r varcoeffplot, fig.cap="The variability of the estimate of the population density over time, with the estimate itself for reference."}
+```{r varcoeffplot, fig.cap="Variability of the model estimate of population density over time, with the estimate itself for reference."}
 knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popvar_plot.pdf"))
@@
 
-The temporal distribution of the variability in the estimate (Figure \@ref(fig:varcoeffplot)) allows us assess at which time steps the uncertainty is greater due to e.g.
+The temporal distribution of variability in the estimate (Figure \@ref(fig:varcoeffplot)) allows us assess at which time steps the uncertainty is greater due to e.g.
 contradictions in the proxies.
-The coefficient of variation is 0.13 for the beginning and 0.1 for the end of the estimate, the greatest variability is reached around 2150 BCE with 0.47.
-This is not astonishing as there are fewer archaeological contexts recorded connected to the earlier phase of the Early Bronze Age c.
-2200-1800 BCE, this picture changes from c.
+The coefficient of variation is 0.13 for the beginning and 0.1 for the end of the estimate, with the greatest variability (0.47) seen around 2150 BCE.
+This is not surprising as there are fewer archaeological contexts recorded from the earlier phase of the Early Bronze Age, c.
+2200-1800 BCE.
+This picture changes from c.
 1800 BCE onwards [@hafner_fruhe_1995; @david_elbiali_suisse_2000].
 The beginning and end of the time series are relatively clearly determined, resulting from the *a priori* setting of final population density, but also from the uniformity of the proxies during these periods.
-Overall, the variability is relatively stable over the entire estimation and averages overall at 33% of the respective mean.
+Overall, the variability is relatively stable over the entire estimation and averages 33% of the respective mean.
 
-```{r pdistribution, fig.cap="Distribution of the influence ratio of the different proxies on the final estimation of the number of sites."}
+```{r pdistribution, fig.cap="Distribution of influence ratios of proxies on model's final estimation of number of sites."}
 knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "p_estimation.pdf"))
@@
 The reason for the latter is certainly that there are no lakeshore settlements over large areas of the time window, and therefore the overall confidence in the proxy is low.
-The aoristic sum is flat over large sections, making it difficult to relate to the other estimators.
-The sum calibration shows very strong short-term fluctuations, at least partly due to the calibration curve, which also make this proxy seem ill-suited to reliably represent a continuous population trend.
-Its fluctuations have an impact on the overall estimate, albeit to a lesser extent.
+The aoristic sum is flat for long periods, making it difficult to integrate with other proxies.
+The sum calibration shows very strong short-term fluctuations, at least partly due to the calibration curve, which suggests that it does not reliably represent a continuous population trend.
+Its fluctuations have an impact on the model's estimate, albeit to a lesser extent than the general trend.
 
@@
 
-## Comparison of individual proxies
+## Reliability of individual proxies
 
-Comparing the model's overall estimate with the individual proxies enables insights into the quality of these records.
-The sum calibration, currently the most frequently used proxy for (relative) population changes in prehistory, has its large fluctuations dampened when considered alongside other proxies, especially the first fluctuation shortly after 4000 BCE.
+Comparing the model's overall estimate with the individual proxies provides several insights into the quality of these records.
+The sum calibration, currently the most frequently used proxy for (relative) population change in prehistory, has its large fluctuations dampened when considered alongside other proxies.
+This is especially ture of the first fluctuation shortly after 4000 BCE.
 The expected increase in archaeological remains with the onset of Neolithisation is still clearly visible, but the overall curve is much flatter than the sum calibration itself.
@@
 
-```{r popestwithphases, fig.cap="The estimate of the population density in relation to the chronology for the case study area north of the Swiss Alps."}
+```{r popestwithphases, fig.cap="Estimate of population density in relation to the established chronology of the case study area north of the Swiss Alps."}
 PopDens_summary <- read.csv(
@@
 
-The Early and Middle Neolithic are hardly documented with known sites in Switzerland.
+The Early and Middle Neolithic are hardly documented in Switzerland.
 We must assume a low level of settlement, probably mainly by mobile groups.
 Isolated Neolithic sites of the LBK and later groups are known in the periphery of Switzerland, but they play a subordinate role [@ebersbach_nutzung_2012].
-The evidence of the Neolithic is dense from the so-called Upper Neolithic onwards, connected with the typochronological pottery phases of Egolzwil (late 5th M BCE) and Cortaillod respectively Pfyn (first half of the 4th M BCE).
+The evidence of the Neolithic is dense from the so-called Upper Neolithic onwards, connected with the typochronological pottery phases of Egolzwil (late 5th millennium BCE) and Cortaillod respectively Pfyn (first half of the 4th millennium BCE).
 The first lake shore settlements north of the Alps date to this time too.
 Here we see a clear increase in the estimated population in the model.
-In the transition to the Late Neolithic we know from the lakeshore settlements the so-called Horgen Gap [@hafner_neolithikum_2005].
+In the transition to the Late Neolithic, we know from the lakeshore settlements the so-called Horgen Gap [@hafner_neolithikum_2005].
 This is also visible as a slight decrease in the model.
-In another study [@heitz2021] we could show that this is in fact probably not a decline in population.
+In another study [@heitz2021] we demonstrated that this is in fact probably not a decline in population.
 Rather communities relocated their settlements to the hinterland of the large lakes in times of stronger lake level rises due to climatic changes.
@@
 
-The key advance in the model we presented is the ability to estimate, in absolute terms, past population sizes and the uncertainty accompanying our present knowledge.
+The key advance in the model we present is the ability to estimate, in absolute terms, past population sizes and the uncertainty accompanying our present knowledge.
 These estimates can be a basis for further studies where relative measures of population development are not helpful, such as long-term land use studies.
@@
 This data is available to varying degrees in different regions.
-On the Swiss Plateau, for example, we have too little data on human remains over large spans of prehistory compared to the abundance of wetland settlements.
+On the Swiss Plateau, for example, we have little data on human remains over large spans of prehistory, in contrast to the abundance of wetland settlements.
 
-To apply the model to different regions, the proxies we use here would have to be adapted to fit local conditions and research histories.
+To apply this approach to other regions, the proxies we use here would have to be adapted to fit local conditions and research histories.
 By means of large-scale modelling, however, it would be possible to supplement gaps in the data in one region with data from another by regionalisation and a partial transfer of information (partial pooling).

### 2022-05-30 manuscript/paper_extended_version.Rmd
renamed paper versions

--- 2022-05-28 manuscript/paper.Rmd
@@
   bookdown::pdf_document2:
+    keep_md: yes
+    includes:
+        in_header: preamble.tex
     toc: no
@@
   Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density.
-Using the  area North of the main ridge of the Swiss Alps in prehistoric times (7000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
+Using the  area North of the main ridge of the Swiss Alps in prehistoric times (–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
 The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research.
@@
   - Our initial model produces more robust, high-resolution estimates of past population dynamics than previous, single-proxy approaches
-  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (7000–1000 BCE)
+  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (–1000 BCE)
 ---
@@
 
-Our case study targets the period between 7000–1000 BCE.
+Our case study targets the period between 6000–1000 BCE.
 The lower limit of this window was chosen to avoid the so-called 'Hallstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, which would cause difficulties for the ^14^C proxy data.
@@
 2200-1800 BCE, this picture changes from c.
-1800 BCE onwards [@hafner_fruhe_1995, @david-elbiali_suisse_2000].
+1800 BCE onwards [@hafner_fruhe_1995; @david_elbiali_suisse_2000].
 The beginning and end of the time series are relatively clearly determined.
@@
 
-### Colophon
+# Author contributions
+- *Martin Hinz*: Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing - Original Draft, Writing - Review & Editing, Visualization
+- *Joe Roe*: Software, Validation, Writing - Review & Editing
+- *Julian Laabs*: Investigation, Data Curation, Writing - Review & Editing
+- *Caroline Heitz*: Conceptualization, Investigation, Writing - Review & Editing
+- *Jan Kolář*: Conceptualization, Writing - Review & Editing
+
+# Colophon
 

### 2022-05-30 manuscript/paper_submission.Rmd
renamed paper versions

--- 2022-05-28 manuscript/paper_shorter.Rmd
@@
     - --lua-filter=../templates/pagebreak.lua
-  bookdown::word_document2:
-    fig_caption: yes
-    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+  bookdown::pdf_document2:
+    keep_md: yes
+    includes:
+        in_header: preamble.tex
+    toc: no
     pandoc_args:

### 2022-05-30 README.Rmd
Updated Readme

--- 2022-05-26 README.Rmd
@@
 # Please put your title here to include it in the file below.
-Title <- "Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the Swiss plateau"
+Title <- "Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the North of the Swiss Alps"
 ```
@@
 
+## Authors
+- Martin Hinz (<martin.hinz@iaw.unibe.ch>), Institute of Archaeological Sciences & Oeschger Centre for Climate Change Research, University of Bern [![ORCiD](https://img.shields.io/badge/ORCiD-0000--0002--9904--6548-green.svg)](http://orcid.org/0000-0002-9904-6548)
+- Joe Roe (<joe@joeroe.io>), Institute of Archaeological Sciences, University of Bern [![ORCiD](https://img.shields.io/badge/ORCiD-0000--0002--1011--1244-green.svg)](http://orcid.org/0000-0002-1011-1244)
+- Julian Laabs (<julian.laabs@ufg.uni-kiel.de>), CRC 1266 - Scales of Transformation, University of Kiel [![ORCiD](https://img.shields.io/badge/ORCiD-0000--0003--1378--3908-green.svg)](http://orcid.org/0000-0003-1378-3908)
+- Caroline Heitz (<caroline.heitz@iaw.unibe.ch>), CRC 1266 - Scales of Transformation, University of Kiel [![ORCiD](https://img.shields.io/badge/ORCiD-0000--0001--7188--6775-green.svg)](http://orcid.org/0000-0001-7188-6775)
+- Jan Kolář (<jan.kolar@ibot.cas.cz>), Department of Vegetation Ecology, Institute of Botany of the Czech Academy of Sciences & Institute of Archaeology and Museology, Faculty of Arts, Masaryk University [![ORCiD](https://img.shields.io/badge/ORCiD-0000--0001--8013--6992-green.svg)](http://orcid.org/0000-0001-8013-6992)
+
+## Abstract
+Robust estimates of population are essential to the study of human–environment relations and socio-ecological dynamics in the past.
+Population size and density can directly inform reconstructions of prehistoric group size, social organisation, economic constraints, exchange, and political and social institutions.
+In this pilot study, we present an approach that we believe can be usefully transferred to other regions, as well as refined and extended to greatly advance our understanding of prehistoric demography.
+
+  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density.
+Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
+The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research.
+Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.
+
+  The archaeological record provides several potential proxies of human population dynamics, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage.
+Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely ona single proxy, and are rarely able to infer population size or density in absolute terms.
+In contemporary demography, it is becoming increasingly common to use Bayesian statistics to estimate population trends and project them into the future.
+The Bayesian approach is popular because offers the possibility of combining heterogenous data, and at the same time qualifying the uncertainty and credibility attached to forecasts.
+These same characteristics make it well-suited to applications to archaeological data in paleodemographic studies.
+
+## Highlights
+- Bayesian modelling can integrate multiple, heterogeneous population proxies from the archaeological record
+- Our initial model produces more robust, high-resolution estimates of past population dynamics than previous, single-proxy approaches
+- We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (6000–1000 BCE)
+
+## Keywords
+Prehistoric demography; Bayesian modelling; Multi-proxy; Settlement dynamics
+
 ## Contents
@@
     for manuscript.
-  - [:file\_folder: paper](/paper): A rendered version, of the manuscript
-    `paper.docx` and `paper.pdf`, suitable for reading (the code is replaced by figures
-    and tables in this file)
+  - [:file\_folder: paper](/paper): A rendered version, of the submitted manuscript as
+    [`paper_submission.docx`](/paper/paper_submission.docx) and [`paper_submission.pdf`](/paper/paper_submission.pdf), suitable for reading (the code is replaced by figures
+    and tables in this file).
+You also find (as an editors cut) those versions before we trimmed them down to the word count of the journal ([`paper_extended_version.docx`](/paper/paper_extended_version.docx) and [`paper_extended_version.pdf`](/paper/paper_extended_version.pdf) respectively).
   - [:file\_folder: analysis](/analysis): R Markdown source document

### 2022-05-30 README.Rmd
corrected date format

@@
 ```{r, echo = FALSE}
+invisible(Sys.setlocale("LC_TIME",'C'))
 knitr::opts_chunk$set(

### 2022-05-30 README.Rmd
added zenodo badge

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/MartinHinz/bayesian.demographic.reconstruction.2022/master?urlpath=rstudio)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/MartinHinz/bayesian.demographic.reconstruction.2022/master?urlpath=rstudio) [![DOI](https://zenodo.org/badge/477622472.svg)](https://zenodo.org/badge/latestdoi/477622472)
 

### 2022-05-30 manuscript/paper_submission.Rmd
removed (useless) md versions of the paper
post zenodo deposition update of the submission version

@@
 output:
-  bookdown::word_document2:
-    fig_caption: yes
-    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+  bookdown::pdf_document2:
+    includes:
+        in_header: preamble.tex
+    toc: no
     pandoc_args:
@@
     - --lua-filter=../templates/pagebreak.lua
-  bookdown::pdf_document2:
-    keep_md: yes
-    includes:
-        in_header: preamble.tex
-    toc: no
+  bookdown::word_document2:
+    fig_caption: yes
+    reference_docx: "../templates/template.docx" # Insert path for the DOCX file
     pandoc_args:
@@
 
+# Code availability
+The computer code used to generate the Bayesian Population model is provided in full in the Supplementary Information, together with information about the program and version used.
+The R code and Data are available online at https://github.com/MartinHinz/bayesian.demographic.reconstruction.2022 and is archived at https://doi.org/10.5281/zenodo.6594498.
+
 \newpage

### 2022-11-24 analysis/sensitivity_analysis.Rmd
pushed sensitivity analysis v0.1

---
title: "Sensitivity Analysis"
author: "Martin Hinz"
date: "2022-11-20"
output:
  pdf_document: default
  html_document:
    keep_md: yes
bibliography: references.bib
---

# Setup

The model has 2 predefined parameters whose effect we want to check in a sensitivity analysis. First we load the necessary libraries.

```{r}
library(here)
library(tidyverse)
library(parallel)
library(doParallel)
library(nimble)
library(coda)
library(MCMCvis)

```

Then we set up the basic framework for the model run. To do this, we load the input data and prepare it according to the model (for details on these steps please consult the actual analysis):

```{r}
all_proxies <- read.csv(file = normalizePath(
  file.path(here(), "data","preprocessed_data", "all_proxies.csv")
  ),
  row.names = 1)
all_proxies <- all_proxies %>% arrange(desc(age))
all_proxies[,2:5] <- all_proxies[,2:5] %>% scale() %>% diff() %>% rbind(0,.)
model_data <- all_proxies
```

Next, we incorporate the model, making all potentially sensitive parameters accessible from the outside:

```{r}
model_code <- nimbleCode( {
  # ---- Process Model ----

  # Estimate the initial state vector of population abundances
  nEnd ~ dnorm(nSites[nYears] * MeanSiteSize / AreaSwissPlateau, sd=0.5)

  # Autoregressive process for remaining years
  for(t in 2:(nYears)) {
    # The actual number of sites per year
    nSites[t] ~ dpois(lambda[t])

    # limiting the change to a maximum value, estimated in the model
    constraint_lambda_lower[t] ~ dconstraint(
      nSites[t]/nSites[t-1] < (max_growth_rate + 1)
      )
    constraint_lambda_upper[t] ~ dconstraint(
      nSites[t-1]/nSites[t] < (max_growth_rate + 1)
      )
  }

  # ---- Observational Model ----

  # For all but the first year
  for(t in 2:(nYears)) {
    # lambda depends on the number of sites at the previous year, plus
    # changes in relation to the proxies
    log(lambda[t]) <- log(nSites[t-1]) + (
      p[1] * sumcal[t]  +
        p[2] * openness[t] +
        p[3] * aoristic_sum[t] + 
        p[4] * dendro[t]
      )
  }
  # ---- Priors ----
  # Relevance of the proxies is estimated as Dirichlet distribution
  p[1:4] ~ ddirch(alpha[1:4])

  # The parameters for the Dirichlet distribution have a weakly informative prior
  for(j in 1:4) {
    alpha[j] ~ dlnorm(mu_alpha[j],sdlog=a_alpha[j])
    a_alpha[j] ~ dexp(1)
    mu_alpha[j] ~ dlnorm(1,sdlog=0.1)
  }

  # The maximum growth rate has a prior gamma distributed between 0 and 1
  # by adding 1 in the process model, this becomes 1-2[
  max_growth_rate ~ dgamma(shape = 5, scale=0.05)

  # The mean site size
  MeanSiteSize ~ dpois(50)

  # ---- transformed data ----

  # Population density and total population as function of site number
  PopDens[1:(nYears)] <- PopTotal[1:(nYears)] / AreaSwissPlateau
  PopTotal[1:nYears] <- nSites[1:nYears] * MeanSiteSize
})
```

# Parameter Sweep

We will run 20 parameterisations with a range whose centre is our proposed value. To do this, we will run the model until convergence. Afterwards, we will compare the results of each run.

We have externalised the function to run the model, we load it first:

```{r}
source(file = normalizePath(file.path(here(), "code", "sensitivity_helpers.R")))
```

## Mean Site Size

First, we vary the mean site size:

```{r}

mean_site_size_sweep <- seq(5,100,length.out = 20)

model_constants_df <- data.frame(
  nEnd = rep(5, length(mean_site_size_sweep)),
  nYears = rep(nrow(model_data), length(mean_site_size_sweep)),
  AreaSwissPlateau = rep(12649, length(mean_site_size_sweep)),
  ParamMeanSiteSize = mean_site_size_sweep
  )

model_constants_list <- purrr::transpose(model_constants_df)

```

Next, we can run the actual analysis and save the results. In the current version this is deactivated to speed up the rendering of the representation of the analysis. The data is loaded from a previous run. However, you are welcome to turn it on to run it yourself.

```{r, eval=FALSE}
sweep_results <- sweep_run(model_constants_list = model_constants_list,
                           this_model_data = model_data,
                           model_code = model_code)

saveRDS(sweep_results, file = normalizePath(
  file.path(here(), "data","preprocessed_data", "sensitivity_meansitesize.RDS"
            ),
  mustWork = F)
  )    
```

Then we reload the results (this is the entry point to avoid having to do the whole calculation all over again):

```{r}
sweep_results <- readRDS(file = normalizePath(
  file.path(here(), "data","preprocessed_data", "sensitivity_meansitesize.RDS"
            ))
  )
```

Now we can visualise the results. We will plot the individual mean values on top of each other:

```{r}
res_for_plot <- as.data.frame(sapply(sweep_results, function(x) x$mean))

res_for_plot$age <- model_data$age

res_for_plot <- pivot_longer(res_for_plot, !age)

ggplot(res_for_plot) + geom_line(aes(x = age, y = value, color = name))
```

The specification of a medium settlement size does not have a strong discernible effect on the result of the estimation.

## Mean Site Size

Next, we check the dependence on the proposed final population density.

```{r}
nEnd_sweep <- seq(0.5,10,length.out = 20)

model_constants_df <- data.frame(
  nEnd = nEnd_sweep,
  nYears = rep(nrow(model_data), length(nEnd_sweep)),
  AreaSwissPlateau = rep(12649, length(nEnd_sweep)),
  ParamMeanSiteSize = rep(50, length(nEnd_sweep))
  )

model_constants_list <- purrr::transpose(model_constants_df)
```

Here, too, the actual run is currently deactivated in order to speed up the rendering. However, you are welcome to perform it yourself.

```{r, eval=FALSE}
sweep_results <- sweep_run(model_constants_list = model_constants_list,
                           this_model_data = model_data,
                           model_code = model_code)

saveRDS(sweep_results, file = normalizePath(
  file.path(here(), "data","preprocessed_data", "sensitivity_nend.RDS"
            ),
  mustWork = F)
  )    
```

And now (especially if the analysis has not gone through), the results will load in again:

```{r}
sweep_results <- readRDS(file = normalizePath(
  file.path(here(), "data","preprocessed_data", "sensitivity_nend.RDS"
            ))
  )
```

Now we can visualise the results. We will plot the individual mean values on top of each other:

```{r}
res_for_plot <- as.data.frame(sapply(sweep_results, function(x) x$mean))

res_for_plot$age <- model_data$age

res_for_plot <- pivot_longer(res_for_plot, !age)

ggplot(res_for_plot) + geom_line(aes(x = age, y = value, color = name))
```

As expected, this parameter has a decisive influence on the result.

### 2022-11-24 manuscript/paper_revision_1.Rmd
added first revisions of the paper to a separate rmd file

--- 2022-05-30 manuscript/paper_submission.Rmd
@@
   Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density.
+At its core, the statistical model is as follows: if the proxies as a whole have a positive delta, a Poisson draw for the new number of settlements is biased by a positive amount (and similarly for negative deltas).
 Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
@@
   The archaeological record provides several potential proxies of human population dynamics, but individually they are inaccurate, biased, and sparse in their spatial and temporal coverage.
-Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely ona single proxy, and are rarely able to infer population size or density in absolute terms.
+Similarly, current methods for estimating past population dynamics are often simplistic: they work on limited spatial scales, tend to rely on a single proxy, and are rarely able to infer population size or density in absolute terms.
 In contemporary demography, it is becoming increasingly common to use Bayesian statistics to estimate population trends and project them into the future.
-The Bayesian approach is popular because offers the possibility of combining heterogenous data, and at the same time qualifying the uncertainty and credibility attached to forecasts.
+The Bayesian approach is popular because it offers the possibility of combining heterogenous data, and at the same time qualifying the uncertainty and credibility attached to forecasts.
 These same characteristics make it well-suited to applications to archaeological data in paleodemographic studies.
@@
 Though the archaeological record provides multiple possible demographic proxies [@muller_tracing_2019], a lack of access to this data and methodological tools for turning it into quantitative estimates has left the conclusions drawn from it vague and superficial [@hassan_demographic_1981].
-As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence [@morris_measure_2013, @turchin_seshat_2015].
+As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence [@morris_measure_2013; @turchin_seshat_2015].
 
@@
 @shennan_regional_2013] and widely applied to archaeological contexts worldwide [@crema_review_2022].
-This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015, @carleton_groucutt_2020].
-While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems common to all approaches relying on a single proxy [@french_manifesto_2021, @schmidt_approaching_2021].
+This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015; @carleton_groucutt_2020; @price_end__end_2021].
+While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems common to all approaches relying on a single proxy [@french_manifesto_2021; @schmidt_approaching_2021].
 We believe that these problems cannot be overcome by methodological refinements in this area alone.
 Instead, a Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.
+
+In the following article we present our methodology and demonstrate its application with a case study in the area north of the Swiss Alps.
+We use a type of Bayesian Generalised Linear Model [@kruschke2015] to combine the different proxies.
+Their assessment (weight and credibility intervals) results from the modelling and the combination of the data itself and is not defined a priori.
+Furthermore, we pursue the concept of a state space model [@auger-methe_guide_2021], a time series model in which a time series is interpreted as the result of a noisy observation of a stochastic process.
+In this sense, it is not a deductive model for testing a hypothesis, but a reconstructive (abductive) one, such it is also used in environmental reconstructions [@inkpen_explaining_2009].
+In this respect, we hope to convince the readers that such an approach is far better suited to realising reliable reconstructions of past population developments than any of the previous approaches.
 
@@
 The region has a very diverse natural landscape: shaped by glaciers during the ice ages, the many lakes and bogs provide excellent preservation conditions for the numerous Neolithic and Bronze Age lakeside settlements, and a rich source for vegetation reconstructions by means of pollen analyses.
-Thanks to the very active and efficient archaeological research and heritage management there is an abundance of archaeological information, including known sites as well as dendrochronological and ^14^C data.
+Thanks to very active and efficient archaeological research and heritage management there is an abundance of archaeological information, including known sites as well as dendrochronological and ^14^C data.
 
@@
 
-: (#tab:tableproxies) A incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.
+: (#tab:tableproxies) An incomplete list of possible observation that can be linked to population developments in the past. Proxies used in this study are highlighted.
+
+For all proxies used, it is a common assumption that they are positively correlated with past population levels.
+In some cases we have reprocessed them in this respect in order to emphasise this relationship even more strongly in our opinion (example: pollen data).
+This is of course a strong assumption, but we believe it is justified, as this has been widely accepted in the scientific literature, and these proxies are used accordingly (each in isolation).
+Our ambition with this article is to present an abductive model to help improve the current practice of proxy use.
+Our ambition is not to justify or verify the use of these proxies itself.
+This can only be done through empirical, hypothesis-testing approaches with known response observations.
+What our method can do, however, is to infer, through a weighting resulting from the modelling, which indicators seem to be better suited in terms of their variability and fit with other indicators to trace a population development in the past.
 
@@
 ## Summed radiocarbon
+
+Summing up the probabilities of radiocarbon dates can undoubtedly be considered the standard approach in demographic reconstruction in archaeology today.
+The logic for this can be derived from the Poisson binomial distribution: such a distribution can be described as the sum of independent Bernoulli trials as it represents the probability estimate for the existence or use of a site per unit time.
+The expected value per time unit for the number of simultaneously existing sites/settlements is the sum of the individual probabilities for the sites, i.e.
+the area under the curve of the probabilities of 14C data.
+Incidentally, the same logic applies to the aoristic sum (shown below).
+The main problem with this proxy is therefore not so much its statistical nature, but mainly the systematic biases that arise from the conservation and finding practicalities in the research process.
+To counter these, and to detect and mitigate the systematic biases, hierarchical modelling such as that carried out in this study is appropriate.
 
@@
 These are primarily derived from scattered surface finds, often with a low dating accuracy (only in the range of archaeological periods), incorporated into our model as a typologically-dated, aorist time series.
-However, it is not dependent radiocarbon dating and thus it avoids the methodological issues of sum calibration.
+However, it is not dependent on radiocarbon dating and thus it avoids the methodological issues of sum calibration.
 Data from `r nrow(aorist)` sites were included in the aoristic sum.
@@
 
-We assume that the higher the population density in an area, the greater the human influence on the natural environment [@zimmermann2004, @lechterbeck2014].
+We assume that the higher the population density in an area, the greater the human influence on the natural environment [@zimmermann2004; @lechterbeck2014].
 Evidence of deforestation can therefore provide indications of population dynamics.
@@
 Our overall model is broken down into several hierarchically-connected individual elements.
-The a process model represents the demographic development itself, without already being explicitly parameterised with data.
+The process model represents the demographic development itself, without already being explicitly parameterised with data.
 Here we assume that the latent variable 'number of sites' is strongly autocorrelated across different time periods.
@@
 
+## Reliability of the overall estimation
+
+We consider each of the proxies used to be intrisically flawed and biased.
+Therefore, we assume that none of them alone is suitable to generate a population estimate with sufficient credibility.
+Especially since the use of a single proxy makes it impossible to specify a confidence interval, as its value must be used as a direct indication of a change in population size.
+
+Rather, we assume that all proxies used are influenced, to a greater or lesser extent, by a proportion of random noise as well as by one of the actual signal.
+As long as there is no ground truth that can be used for calibration, this cannot be fully differentiated.
+What we assume, however, is that the noise between the individual proxies is uncorrelated, but the signal must essentially produce a coherent pattern.
+
+However, other reasons are also conceivable.
+For example, a correlation between two parameters can also be caused by the fact that they are affected by the same confounding influence.
+All archaeologically derived proxies are subject to preservation conditions.
+If no signs of human activity are preserved for a certain space at a certain time based on the surviving archaeological evidence, then all these indicators will also show a settlement gap.
+It is therefore all the more important to use independent proxies that are not affected by the same distorting influences for a robust reconstruction.
+This is the reason why the openness indicator is an essential part of this study, as it is not confounded by the archaeological record.
+Other similarly independent indicators are to be integrated in future applications of our approach to make the result even more robust.
+
+Moreover, our modelling approach in terms of an abductive reconstructive model is also such that in its present form the model provides the best possible summary of the proxies used and thus the best possible reconstruction given the available information.
+It must be emphasised that such a model does not attempt to directly represent the process of the relationship between proxy and latent variable.
+We treat all proxies as equal (false), and leave it to coherence with respect to the underlying process (the latent population development) to provide a weighting.
+In our view, this corresponds to a (pragmatic) implementation of the general scientific process.
+However, this also means that we prefer a useful model to a less useful one without wanting to (or being able to) make a statement about the concept of truth of a model.Especially in comparison to the uncritical application of a purely radiocarbon data-based proxy, as it is currently used in large scale, we see the significance of such a model and the philosophy behind it as more than justified.
+
 ## Reliability of individual proxies

### 2022-12-08 analysis/analysis.Rmd
post revision fix neotoma data retrieval

@@
 knitr::opts_chunk$set(echo = TRUE)
+Sys.setenv(OPENSSL_CONF="/etc/ssl/")
 if (!webshot::is_phantomjs_installed()) webshot::install_phantomjs()
@@
 library(fmcmc)
+library(grid)
 ```
@@
 ```
+Also, neotoma is not longer available on cran, and it has been replaced by neotoma2, so we had to replace it. That is why we have to load it again:
+
+```{r}
+if(!require('neotoma2')) devtools::install_github('NeotomaDB/neotoma2')
+require(neotoma2)
+```
 
@@
 ```{r message=FALSE, warning=FALSE}
-pollen_data <- get_download(pollen_locations$neotoma_id)
+pollen_data <- get_downloads(pollen_locations$neotoma_id, all_data=T)
 ```
@@
 ```{r}
-pollen_data_filtered <- filter_by_unit(filter_by_element(pollen_data, "pollen"), "NISP")
-class(pollen_data_filtered) <- append(class(pollen_data_filtered), "download_list")
+pollen_data.samples <- samples(pollen_data)
+
+pollen_data.samples.filtered <- pollen_data.samples %>% subset(element == "pollen" & units == "NISP")
 ```
@@
 ```{r}
-pollen_data_comp <- transform_to_percent(pollen_data_filtered)
+pollen_data_comp <- transform_to_percent(pollen_data.samples.filtered)
 ```
@@
 ```{r}
-all_output <- suppressWarnings(neotoma::compile_downloads(pollen_data_comp))
+all_output <- pollen_data_comp
 ```
@@
 all_output <- subset(all_output, 1950-age > timeframe[1] & 1950-age < timeframe[2])
+all_output[is.na(all_output)] <- 0
 ```
@@
 ```{r}
-all_data <- all_output[,11:ncol(all_output)]
-all_data[is.na(all_data)] <- 0
+all_data <- all_output[, !(colnames(all_output) %in%  c("age", "sampleid", "sitename"))]
 ```
@@
 res.open$age <- all_output$age
-res.open$site <- all_output$site.name
+res.open$site <- all_output$sitename
 ggplot(res.open) + geom_line(aes(x = 1950-age, y = score, color = site))
@@
   rename(age = calBP, sumcal = PrDens, openness = mean, aoristic_sum = sum, dendro = moving_mean) %>%
-  filter(age < 8000 & age > 3000)
+  dplyr::filter(age < 8000 & age > 3000)
 

### 2022-12-08 manuscript/paper_revision_1.Rmd
further update manuscript

@@
 Furthermore, we pursue the concept of a state space model [@auger-methe_guide_2021], a time series model in which a time series is interpreted as the result of a noisy observation of a stochastic process.
+We evaluate the number of settlements at time t as a function of the number of settlements at time t-1, modified according to the changes in the proxies.
+For this purpose, we also use these as the difference between these points in time and transform them accordingly.
 In this sense, it is not a deductive model for testing a hypothesis, but a reconstructive (abductive) one, such it is also used in environmental reconstructions [@inkpen_explaining_2009].
@@
 They follow two principles.
-First, a hidden or latent process is assumed, representing the state of the variable of interest $x_t$ through the entire time series.
-Every state of variable x in the future, as well as in the past, is bound by a Markov process to the state of variable $x$ at time $t$.
-Second, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $x$ at time $t$.
-This implies that a relationship between the individual states of variable $y$ is generated over time via the hidden variable $x$, which is not directly observable.
+First, a hidden or latent process is assumed, representing the state of the variable of interest $z$ through the entire time series.
+Every state of variable z as $z_t$ at a specific time is bound by a Markov process to the state of variable $z$ at time before, at $t-1$.
+Second, it is assumed that certain observations, represented in variable y, are dependent on the state of variable $z$ at time $t$.
+This implies that a relationship between the individual manifestation of variable $y$ is generated over time via the hidden variable $z$, which is not directly observable.
+
+In the following I reproduce the mathematical representation according to @auger-methe_guide_2021:
+
+$$
+z_t = \beta z_{t-1} + \epsilon_t
+\tag{1}
+$$
+
+$$
+y_t = \alpha z_t + \eta_t
+\tag{2}
+$$
+
+"The autocorrelation in the states is captured by the parameter $\beta$.
+The observations are a function of the states only and the parameter $\alpha$ allows the observation at time t to be a biased estimate of the state at time t." [@auger-methe_guide_2021, 4-5].
+Dabei sind $\epsilon$ die process variation, und $\eta$ der observation error, welche mit einer passenden Verteilung modelliert werden können.
+
+Um die Struktur der Abhängigkeiten zu verdeutlichen, kann man das Modell auch in terms of probability distributions darstellen:
+
+$$
+f(z_t | z_{t-1}, \theta_p), \quad t = 1, ... , T
+\tag{3}
+$$
+
+$$
+f(y_t | z_{t}, \theta_o), \quad t = 1, ... , T
+\tag{4}
+$$
+
+Dabei sind f and g are two normal probability density functions, and $\theta_p$ and $\theta_o$ are vectors of parameters associated with each equation.
+Eq.
+3 describes the autocorrelation in state values as a first-order Markov process, and Eq.
+4 describes how observations depend simply on the states.
+This definition also demonstrates that states are random variables and thus that SSMs are a type of hierarchical mode [@auger-methe_guide_2021, 5].
 
@@
 The process model represents the demographic development itself, without already being explicitly parameterised with data.
-Here we assume that the latent variable 'number of sites' is strongly autocorrelated across different time periods.
+Here we assume that the latent variable 'number of sites' ($z$) is strongly autocorrelated across different time periods ($z_t$).
 The number of sites in 3000 BCE is strongly conditioned by the number of sites in 3050 BCE, and so on.
-The population at time $t$ results from the population at time $t-1$ times a parameter $\lambda$, which represents the population change at this time.
-
-$$
-N_t = N_{t-1} * \lambda_t
-$$
+The population at time $t$ results from the population at time $t-1$ times a parameter $\beta$, which represents the population change at this time (see Equation 1).
 
@@
 $$
-\begin{aligned}
-N_t &\sim dpois(\lambda_t) \\
-\lambda_t &= N_t
-\end{aligned}
+z_t \sim Poisson(\lambda_t)
+\tag{5}
+$$
+where
+
+$$
+log(\lambda_t) = log(z_{t-1}) + \sum_{i=1}^n \beta_i x_{t,i}
+\tag{6}
+$$
+which can be rearranged to
+
+$$
+\lambda_t = z_{t-1} e^{\mathbf{\beta^T \delta \mathbf{y}_t}}
+\tag{7}
 $$
 
-If we now have information about the change in population development (the proxies), this can enter into the model via a change in $\lambda$ in form of a regression: for all proxy values — represented as a vector of independent variables $x \in R^n$, with $R^n$ as an n-dimensional Euclidean space defined by the n variables — the model takes the form:
-
-$$
-log({E} (Y\mid x))=\alpha + \beta' x
-$$
+$\mathbf{\beta}$ is a parameter vector of weightings that sums to $1$, a vector transpose is indicated by $^T$, and $y_t$ is a vector of proxy difference values.
 
@@
 Here, it functions as a scaling factor for the individual proxies.
-$\alpha$ is to be understood as an intercept, representing a baseline when there were no change due to the variables.
+$z_t$ is to be understood as an intercept, representing a baseline when there were no change due to the variables.
 This is the desired behaviour: $\lambda$ is equal to the value of the population in the previous time period, plus or minus the changes resulting from the variables.
 
-$$
-log(\lambda_t) = log(N_{t-1}) + \sum_{i=1}^n \beta_i x_{t,i}
-$$
-
-Since $\lambda$ and $N$ are essentially in the same range (e.g.
-if $lambda=1$, the expected value for $N$ would also be 1), $N_{t-1}$ must also be log-transformed for the congruence of both values.
-Population size $N_t$ as well as population change $\lambda_t$ are time-dependent.
+Since $\lambda$ and $z$ are essentially in the same range (e.g.
+if $lambda=1$, the expected value for $z$ would also be 1), $z_{t-1}$ must also be log-transformed for the congruence of both values.
+Population size $z_t$ as well as population change $\lambda_t$ are time-dependent.
 At each individual point in time, these variables will take on different values.
@@
 \begin{aligned}
-max\_growth\_rate &\sim dgamma(shape = 5, scale=0.05) \\
-N_t/N_{t-1} &< (max\_growth\_rate + 1) \\
-N_{t-1}/N_t &< (max\_growth\_rate + 1)
+max\_growth\_rate &\sim Gamma(shape = 5, scale=0.05) \\
+z_t/z_{t-1} &< (max\_growth\_rate + 1) \\
+z_{t-1}/z_t &< (max\_growth\_rate + 1)
 \end{aligned}
+\tag{8}
 $$
-
 A gamma distribution centres probability in the range $[0,1[$; adding 1 makes this range $[1-2[$.
@@
 \begin{aligned}
-z_t &= \frac{x_t - \bar{x}} {\sigma_x}\ |\ \sigma_x := Standard\ Deviation  \\
-\delta z{_t} &= z_t - z_{t-1}
+y'_t &= \frac{y_t - \bar{y}} {\sigma^2_y}\ |\ \sigma^2_y := Standard\ Deviation  \\
+\delta y'_t &= y'_t - y'_{t-1}
 \end{aligned}
+\tag{9}
 $$
 
-The sum of the resulting differences between the time steps, together with the settlement number of the previous step as the expected value, then forms $\lambda_t$: the expected value for the settlement number of the current time step.
+The fact that we use the difference here, and not the actual value, may need some explanation.
 
-$$
-log(\lambda_t) = log(N_{t-1}) + \sum_{i=1}^n \beta_i \delta z_{i, t}
-$$
+The sum of the resulting differences between the time steps, together with the settlement number of the previous step as the expected value, then forms $\lambda_t$: the expected value for the settlement number of the current time step (see Equation 6).
 
+This means that we consider the development of population size and thus the number of settlements as autocorrelated per se.
+There are two reasons for this: First, the proxies we use are obviously autocorrelated.
+This can be seen as empirical confirmation that such an autocorrelation can also be regarded as probable for the latent variable of population size behind it.
+On the other hand, a model that does not assume such an autocorrelation would mean that population development is primarily controlled by migration, since the population already present in an area would play no role in this.
+We think it more likely that it is primarily births and death events, together with migration, that are responsible for population numbers and their change.
+Therefore, we view the population at time t as a function of the population at time t-1 plus a change that can be estimated from the change in the proxies.
+In this way, we make all time steps interdependent, which constrains the model in a comprehensible and justifiable way, and thus narrows the probability range for the prediction by providing additional information in the model.
 Here, $\beta_i$ is a scaling factor that represents the influence of the respective proxy.
@@
 \sum_{i=1}^n \beta_i = 1
+\tag{10}
 $$
@@
 \begin{aligned}
-\beta_i &\sim Dir(\alpha_{1-i}) \\
-\alpha_i &\sim LogNormal(\mu_{alpha_i}, \sigma_{alpha_i}) \\
+\beta_i &\sim Dirichlet(\alpha_{1-i}) \\
+\alpha_i &\sim Lognormal(\mu_{alpha_i}, \sigma^2_{alpha_i}) \\
 \mu_{alpha_i} &\sim Exp(1) \\
-\sigma_{alpha_i} &\sim LogNormal(1,0.1)
+\sigma^2_{alpha_i} &\sim Lognormal(1,0.1)
 \end{aligned}
+\tag{11}
 $$
 
-Intuitively, we consider the sum of the proxies as determinant of the number of settlements.
+Intuitively, we consider the sum of the proxies as determinant of the change in number of settlements.
 That is, the share of each individual proxy is variable and is estimated within the model.
@@
 
+Another direction of development is to reconsider and improve some decisions currently made for practical reasons regarding the modelling approach.
+For example, one proposed alternative is to use the values of the proxies directly instead of using them via normalisation and differencing as an indicator of the change in a population from time t-1 to time t.
+This would require that explicit data models and observation models for the individual proxies become part of the hierarchical model in order to show in detail their dependence on the respective population numbers and to have their parameters estimated as probability distributions in the model.
+Consequently, we would have to compensate for the resulting degrees of freedom with additional information (more proxies, regionalisation), and we would have to cope with the increased numerical complexity and the resulting higher computational effort.
+This can be done, for example, by using a high performance computing environment, and is intended for the extension of the model.
+
 To apply this approach to other regions, the proxies we use here would have to be adapted to fit local conditions and research histories.

### 2022-12-08 analysis/alt_analysis.Rmd

--- 2022-05-28 analysis/analysis.Rmd
@@
 bibliography: references.bib
+editor_options: 
+  chunk_output_type: inline
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
-if (!webshot::is_phantomjs_installed()) webshot::install_phantomjs()
-library(magick)
 ```
@@
 
-Running through the analysis can take a long time, especially the Bayesian model.
+In the review of the article, the model and the associated analysis, it was noted that the model we use is too constrained by the Poisson process.
+We are convinced that the Poisson process is the right choice to model a settlement number as a process autocorrelated over time.
+Nevertheless, we are happy to modify the model accordingly to show the differences.
+This document contains the corresponding alternative version of the model and its results.
+
+As with the original model: Running through the analysis can take a long time, especially the Bayesian model.
 We point out the long runtimes again at the appropriate place.
@@
 library(here)
-library(RcppRoll)
-library(sf)
-library(mapview)
-library(vegan)
-library(neotoma)
-library(analogue)
-library(rcarbon)
+#library(RcppRoll)
+#library(sf)
+#library(mapview)
+#library(vegan)
+#library(neotoma)
+#library(analogue)
+#library(rcarbon)
 library(doParallel)
@@
 
-For the aorist analysis we also use the package aoristAAR, which is not available on CRAN, but would have to be installed directly from GitHub:
+## Loading and transforming the preprocessed data
 
-```{r message=FALSE}
-if(!require('devtools')) install.packages('devtools')
-library(devtools)
-install_github('ISAAKiel/aoristAAR')
-require(aoristAAR)
-```
-
-## Data collection
-
-The data used in this analysis are the following proxies obtained from the following sources:
-
--   An openness index generated by a PCA of five pollen profiles in the Swiss Plateau from the neotoma database [@williams2018].
-
--   ^14^C sum calibration, data sets from the XRONOS database as well as from the unpublished PhD thesis of Julian Laabs [@laabs2019] and from the data collection of [@martínez-grau2021].
-
--   Sites from the database of the Cantonal Archaeology <!--# which! --> as the basis of an aoristic analysis
-
--   The number of dated fell phases at lakeshore settlements in the Three Lakes Region from the doctoral thesis by Julian Laabs [\@laabs2019].
-
-### Openness index
-
-We start with the openness index, as this will set the spatial framework for our analyses. The five pollen profiles are the following:
-
-```{r}
-pollen_locations <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "pollen_locations.csv")))
-pollen_locations
-```
-
-We can also map these. To do this, we first transform the coordinate system into the Swiss national coordinate system, and then display them:
-
-```{r}
-pollen_locations.sf <- st_as_sf(pollen_locations,
-                                coords = c("lng", "lat"), # coordinate columns
-                                crs = 4326)  # coordinate references system in epsg))
-pollen_locations.sf.lv95 <- st_transform(pollen_locations.sf, crs = 2056)
-
-pollenmap <- tempfile(fileext = ".png")
-mapshot(
-  mapview(pollen_locations.sf),
-  file = pollenmap
-  )
-
-img <- magick::image_read(pollenmap)
-old.par <- par(mar = c(0, 0, 0, 0))
-plot(img)
-par(old.par)
-```
-
-We download the data from neotoma:
-
-```{r message=FALSE, warning=FALSE}
-pollen_data <- get_download(pollen_locations$neotoma_id)
-```
-
-Then we load a few help scripts which can be found under "code/neotoma_helpers.R".
-
-```{r}
-source(file = normalizePath(file.path(here(), "code", "neotoma_helpers.R")))
-```
-
-Now we filter the data sets so that we only retain the pollen data that record "NISP".
-
-```{r}
-pollen_data_filtered <- filter_by_unit(filter_by_element(pollen_data, "pollen"), "NISP")
-class(pollen_data_filtered) <- append(class(pollen_data_filtered), "download_list")
-```
-
-Then we convert the NISP into percentages:
-
-```{r}
-pollen_data_comp <- transform_to_percent(pollen_data_filtered)
-```
-
-Furthermore, we need a (handwritten) concordance list for the taxa names:
-
-```{r}
-common_names <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "species.csv")), row.names = 1)
-common_names
-```
-
-With this concordance table we combine the species with respect to their genus:
-
-```{r}
-pollen_data_comp <- aggregate_by_joint_name(pollen_data_comp, common_names = common_names)
-```
-
-Now we can combine the individual pollen profiles into an overall table, where each row represents a sample:
-
-```{r}
-all_output <- suppressWarnings(neotoma::compile_downloads(pollen_data_comp))
-```
-
-We restrict the data set to the Holocene so that we no longer have the pre-Holocene developments in the principal component analysis:
-
-```{r}
-timeframe <- c(-7000,-1000)
-all_output <- subset(all_output, 1950-age > timeframe[1] & 1950-age < timeframe[2])
-```
-
-Then we extract the actual data:
-
-```{r}
-all_data <- all_output[,11:ncol(all_output)]
-all_data[is.na(all_data)] <- 0
-```
-
-We then filter the data. Only those taxa are considered which are present in at least 1/3 of all data series:
-
-```{r}
-all_data <- all_data[ , colSums(all_data) != 0]
-all_data <- all_data[,colSums(all_data>0)>(nrow(all_data)/3)]
-```
-
-In addition, only those with an average frequency greater than 0.1% are included:
-
-```{r}
-all_data <- all_data[,colMeans(all_data)>0.1]
-```
-
-However, we add the *cerealia* back in, regardless of their frequency:
-
-```{r}
-if(!("Cerealia" %in% colnames(all_data)))
-{
-  all_data <- data.frame(all_data, Cerealia = all_output[,"Cerealia"])  
-}
-```
-
-These data can now be evaluated by means of a principal component analysis:
-
-```{r}
-res.pca <- rda(all_data, scale = TRUE)
-biplot(res.pca, type="text")
-```
-
-This calculation already shows the familiar structure that we have the known open land indicators at one end of the spectrum and the forest species (tree pollen) at the other.
-However, the result is also influenced by a gradient which has nothing to do with the effect we are interested in: The immigration of beech.
-To eliminate this gradient, we use a partial RDA in which only the residuals for Fagus are evaluated:
-
-```{r}
-res.pca <- rda(all_data ~ Condition(Fagus), data=all_data, scale = TRUE)
-biplot(res.pca, type="text")
-```
-
-From this principal component analysis, we now extract the first dimension as the openness index, and supplement it with the datings of the individual samples.
-In doing so, we ensure that open land indicators end up on the positive side of the result vector:
-
-```{r}
-orientation <- sign(scores(res.pca)$species["Cerealia",1])
-res.open <- data.frame(scores(res.pca, choices = 1, display = "si")*orientation)
-colnames(res.open) <- "score"
-res.open$age <- all_output$age
-res.open$site <- all_output$site.name
-ggplot(res.open) + geom_line(aes(x = 1950-age, y = score, color = site))
-```
-
-The individual profiles show a very similar development over time.
-We now summarise the common trend and smooth it with a 50-year moving window.
-For this, however, we first have to interpolate shared points in time, which we do via a linear approximation:
-
-```{r}
-
-sw <- 50
-
-full_range <- full_seq(res.open$age, 1)
-
-res.open.interpolated <- by(res.open, res.open$site, function(x) approx(x$age,x$score, full_range)$y)
-
-res.open.interpolated <- data.frame(do.call(cbind, res.open.interpolated))
-
-res.open.interpolated$age <- full_range
-
-res.open.interpolated <- melt(res.open.interpolated, id.vars=c("age"))
-
-res.open.interpolated$site <- res.open.interpolated$variable
-
-res.open.interpolated$variable <- NULL
-
-res.open_smoothed <- res.open.interpolated %>% 
-  group_by(site) %>%
-  # RcppRoll::roll_mean() is written in C++ for speed 
-  mutate(moving_mean = roll_mean(value, sw, fill = NA, na.rm = TRUE))
-
-res.open_smoothed$value<-NULL
-res.open_smoothed <- na.omit(res.open_smoothed)
-```
-
-Then we pick every 50th date:
-
-```{r pollenproxy}
-res.open_smoothed <- res.open_smoothed[res.open_smoothed$age %in%
-seq(from=1000,
-    to=10000, by=50),]
-pollenproxy_plot <- ggplot(res.open_smoothed,aes(x=1950-age,y=moving_mean)) +
-  geom_point(aes(color=site), alpha =.5) +
-  geom_smooth(span = 0.05, color = "black") +
-  scale_x_continuous(breaks=seq(timeframe[1],timeframe[2],200), limits = timeframe) +
-  labs(y = "Openness score [PCA]",x = "Date",colour = "Sites") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-pollenproxy_plot
-```
-We save this plot for later use:
-
-```{r}
-ggsave(normalizePath(file.path(here(), "figures","pollenproxy.pdf"), mustWork = F),
-       plot = pollenproxy_plot, width = 21, height = 29.7/2, units = "cm")
-```
-
-Finally, we calculate the average of all profiles as the final supra-regional openness index.
-
-```{r}
-res.open_final <- res.open_smoothed %>%
-  group_by(age) %>%
-  mutate(mean=mean(moving_mean),
-         sd = sd(moving_mean)) %>% 
-  select(age,mean,sd) %>%
-  unique()  %>% arrange(desc(age))
-
-ggplot(res.open_final,aes(x=1950-age)) +
-  geom_line(aes(y=mean)) +
-  geom_segment(aes(y=mean-sd, yend =mean+sd, xend=1950-age), alpha = .25)
-```
-
-### ^14^C Summenkalibration
-
-The data set for the 14C sum calibration is composed of the above-mentioned sources.
-
-```{r}
-c14 <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "14c_swiss_plateau.csv")))
-```
-
-We can show this on the map:
-
-```{r}
-c14map <- tempfile(fileext = ".png")
-mapshot(
-  mapview(st_as_sf(c14,
-                   coords = c("lng", "lat"), 
-                   crs = 4326),layer.name = "14C Data"
-          ),
-  file = c14map
-  )
-
-img <- magick::image_read(c14map)
-old.par <- par(mar = c(0, 0, 0, 0))
-plot(img)
-par(old.par)
-```
-
-Next, we calibrate the 14C data using rcarbon: <!--# uncal vs cal? -->
-
-```{r}
-c14.calibrated <- calibrate(x = c14$bp, errors = c14$std, verbose = F)
-```
-
-Subsequently, we perform the sum calibration, whereby we perform a binning at the level of the sites:
-
-```{r}
-sumcal <- spd(c14.calibrated, timeRange = 1950-timeframe, bins = c14$site, runm = sw, verbose = F)
-```
-
-Below is the visualisation of the sum calibration:
-
-```{r}
-plot(1950-sumcal$grid$calBP, sumcal$grid$PrDens,type="l", main = paste0("n=", length(c14.calibrated)))
-
-```
-
-Finally, we extract the values every 50 years:
-
-```{r}
-res.sumcal <- sumcal$grid[sumcal$grid$calBP %in%
-seq(from=1000,
-    to=10000, by=50),]
-```
-
-### Aoristic Site Count
-
-The third proxy is the aoristic sum of the number of sites from the cantonal archaeology database.
-For these, we have obscured the coordinates to avoid the well-known problems that can arise from clear site coordinates.
-
-```{r}
-sites <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "site_data_fuzzy.csv")))
-```
-
-We can depict this on the map:
-
-```{r}
-sitesmap <- tempfile(fileext = ".png")
-mapshot(
-  mapview(st_as_sf(sites,
-                   coords = c("LV03_Rechtswert_fuzzy", "LV03_Hochwert_fuzzy"),
-                   crs = 21781),
-          layer.name = "Sites"
-  ),
-  file = sitesmap
-)
-img <- magick::image_read(sitesmap)
-old.par <- par(mar = c(0, 0, 0, 0))
-plot(img)
-par(old.par)
-```
-
-The dates come in two categories, rough and very rough We use the rough datings where they are available.
-Where they are not, we fall back on the very rough dates.
-But first we load the concordance table:
-
-```{r}
-
-chrono_rough <- read.csv(file = normalizePath(
-  file.path(here(), "data","raw_data", "chrono_rough.csv")
-  ),
-  row.names = 1)
-
-chrono_vrough <- read.csv(file = normalizePath(
-  file.path(here(), "data","raw_data", "chrono_very_rough.csv")
-  ),
-  row.names = 1)
-```
-
-Next, we prepare the beginning and end as a column:
-
-```{r}
-sites$begin <- NA
-sites$end <- NA
-```
-
-And then we first fill these columns with the data from the very rough dating:
-
-```{r}
-for(i in 1:nrow(chrono_vrough)){
-  this_phase <- chrono_vrough[i,]
-  sites$begin[sites$chrono_very_rough == this_phase[,1]] <- this_phase$dat_begin
-  sites$end[sites$chrono_very_rough == this_phase[,1]] <- this_phase$dat_end
-}
-```
-
-Where available, we overwrite the very rough dating with the rough dating:
-
-```{r}
-for(i in 1:nrow(chrono_rough)){
-  this_phase <- chrono_rough[i,]
-  sites$begin[sites$chrono_rough == this_phase[,1]] <- this_phase$dat_begin
-  sites$end[sites$chrono_rough == this_phase[,1]] <- this_phase$dat_end
-}
-```
-
-However, this does not allow for a very precise dating:
-
-```{r}
-table(sites$begin)
-```
-
-Nevertheless, we next use the aorist sum as a mapping of the expected value for simultaneously existing settlements:
-
-```{r}
-res.aorist <- aorist(sites, from = "begin", to = "end", method = "weight")
-
-res.aorist$age <- 1950-res.aorist$date
-res.aorist <- subset(res.aorist, date >= timeframe[1] & date <= timeframe[2])
-```
-
-Again, the curve of the proxy is plotted here also:
-
-```{r}
-plot(res.aorist$date, res.aorist$sum,type="l", main = paste0("n=", nrow(sites)))
-```
-
-Once more, we extract the values every 50 years:
-
-```{r}
-res.aorist <- res.aorist[res.aorist$age %in%
-seq(from=1000,
-    to=10000, by=50),]
-```
-
-### Dendro Dated Settlements
-
-This year-by-year dataset maps the number of lakeshore settlements in the Three Lakes Region.
-Details can be found in the publication by [@laabs2019].
-To match the resolution of the other datasets, this dataset is also sampled at 50-year intervals, and smoothed accordingly beforehand using a moving average.
-
-First we load the data:
-
-```{r}
-lss <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "dendro_years.csv")))
-```
-
-Then we complete the data set so that it covers the entire period under consideration:
-
-```{r}
-lss <- rbind(data.frame(date=-7050:min(lss$date)-1,n_settlements=0),
-             lss,
-             data.frame(date=max(lss$date)+1:-1,n_settlements=0))
-```
-
-Finally, we run a moving average over it with the same window as for the other data sets:
-
-```{r}
-res.lss <- lss %>% 
-  mutate(moving_mean = roll_mean(n_settlements, sw, fill = NA, na.rm = TRUE))
-```
-
-Here is yet another plot of the resulting curve:
-
-```{r}
-plot(res.lss$date, res.lss$moving_mean,type="l")
-```
-
-Picking the data points every 50 years again:
-
-```{r}
-res.lss <- res.lss[res.lss$date %in% seq(from=timeframe[1],
-                                         to=timeframe[2], by=50),
-        ]
-res.lss$age <- 1950-res.lss$date
-```
-
-### Combining the proxies
-
-Thus we have all proxies available, and can display them (scaled) all together:
-
-```{r}
-ggplot() +
-  geom_line(data = res.open_final, aes(x = 1950-age, y = scale(mean)), col="darkgreen", alpha = 0.75) +
-  geom_line(data = res.sumcal, aes(x = 1950-calBP, y = scale(PrDens)), col="red", alpha = 0.75) +
-  geom_line(data = res.aorist, aes(x = date, y = scale(sum)), col="gray", alpha = 0.75) + 
-  geom_line(data = res.lss, aes(x = date, y = scale(moving_mean)), col="brown", alpha = 0.75) +
-  scale_colour_manual(values = c("darkgreen", "red", "gray", "brown"),
-                      labels = c("Openness Indicator", "Sum Calibration", "Aoristic Sum", "Dendro Dated Settlements")
-                      )
-```
-
-We merge the dataset and restrict it to our observation window:
-
-```{r}
-all_proxies <- merge(res.sumcal, res.open_final, by.x = "calBP", by.y = "age") %>%
-  merge(res.aorist, by.x = "calBP", by.y = "age") %>%
-  merge(res.lss, by.x = "calBP", by.y = "age") %>%
-  select(calBP, PrDens, mean, sum, moving_mean) %>%
-  rename(age = calBP, sumcal = PrDens, openness = mean, aoristic_sum = sum, dendro = moving_mean) %>%
-  filter(age < 8000 & age > 3000)
-
-all_proxies
-```
-
-We might like to store the result for the actual analysis:
-
-```{r}
-write.csv(all_proxies, file = normalizePath(
-  file.path(here(), "data","preprocessed_data", "all_proxies.csv"
-            ),
-  mustWork = F)
-  )
-```
-
-## Transforming the Data
-
-If the data has already been created in a previous run, we can also reload it directly at this point:
+First we load the pre-processed data:
 
@@
 ```
+
+Then we carry out the transformation steps as we did in the actual analysis.
 
@@
 
-The actual model consists of a nimble code block that describes the relationships and sets the priors for the parameters:
+The actual model consists of a nimble code block that describes the relationships and sets the priors for the parameters.
+It should be noted that we have changed the function for the number of settlements, and thus the proposal distribution.
+It is now a Negative Binomial Distribution, where both the expected value and the variance can vary.
+The expected value is still derived from $\lambda_t$, but this is integrated into the model via a success parameter $prob_t$, which results as follows:
+
+$$
+prob_t = \frac{r}{r+\lambda_t}
+$$
+
+The parameter $r$ used here and also in the distribution itself is the one that determines the overdispersion and thus makes the variance variable compared to a Poisson distribution.
+If this parameter approaches infinity, the Negative Binomial Distribution approaches a Poisson distribution.
+In practice, values above 30 lead to convergence.
+$r$ is estimated in the model itself.
 
@@
     # The actual number of sites per year
-    nSites[t] ~ dpois(lambda[t])
+    nSites[t] ~ dnegbin(prob[t],r)
+    prob[t] <- r/(r+lambda[t])
 
@@
   }
+
+  #r ~ dunif(0,10000)
+  #r ~ dgamma(0.5,0.5)
+  #r ~ dlnorm(6,1)
+  r.cauchy ~ dt(mu = 0, sigma = 5, df = 1)
+  r <- abs(r.cauchy)
 
@@
 ```{r}
-model_monitors <- c("PopDens", "p")
+model_monitors <- c("PopDens", "p", "r")
 ```
@@
 
-# Spare 4 cores from the calculation
-ncore <- detectCores() - 4
+# Spare 2 cores from the calculation
+ncore <- detectCores() - 2
 ncore <- max(ncore, 1)
@@
 converged <- F
+progress <- data.frame(tick = vector(),
+                        mean_psrf = vector(),
+                        max_psrf = vector())
+tick <- 1
 
@@
 
+  tick <- tick + 1
   # run the model for another batch, with resetting the values (burn-in)
@@
 
+  progress <- rbind(progress,
+                     data.frame(tick = tick,
+                        mean_psrf = mean(psrf),
+                        max_psrf = max(psrf)))
   # optional: create an mcmc traceplot for visual inspection
@@
 
+It is already clear here that the basic structure and the (mean) values of the estimation do not differ significantly from the Poisson model.
+As expected, the credibility intervals are larger because an additional degree of freedom has been introduced by the free variance.
+
 This converged run is sufficient to check some initial parameter values. For example, we can see how the proportional importance of the proxies is distributed:
@@
 
-The model trusts mainly the openness index, the aoristic sum and the dendro data play a subordinate role.
-The sum calibration has a higher weight.
-However, the shape of the distributions also indicates that we may have too few effective samples to make a reliable statement about the highest posterior density intervals.
+The model still trusts mainly the openness index, the aoristic sum and the dendro data play a subordinate role. The sum calibration has a higher weight.
+
+Der Parameter für die Überdispersion $r$ ist ja auch im Modell geschätzt worden. Folgende Graphik gibt seine Verteilung wieder:
+
+```{r}
+MCMCchains(mcmcSamples, params = "r") %>%
+  as.data.frame %>%
+  pivot_longer(everything()) %>%
+  ggplot(aes(x = value, y = name)) + stat_halfeye() + xlim(0,1000)
+```
+
 Let's take a look at the effective samples:
@@
 
+  ness <- ness[!(names(ness) %in% c("r", "r.cauchy")) ]
+
   # we have enough samples, when all parameters have more than the minimum ESS
@@
   # just in case: store the result in a RDS file for later inspection  
-  saveRDS(collector, file = "mcmc_collector.RDS")
+  saveRDS(collector, file = normalizePath(
+  file.path(here(), "data","result_data", "mcmc_collector_neg_binomial.RDS"), mustWork = F
+  ))
 }
@@
 ```
+
+Please note that I excluded 'r' and 'r.cauchy' from the list of parameters that need at least 10000 effective samples.
+Having that in prolongues the calculation by a factor of 4, which might not be necessary in this context.
 
@@
 ```{r}
-collector <- readRDS("mcmc_collector.RDS")
+collector <- readRDS(file = normalizePath(
+  file.path(here(), "data","result_data", "mcmc_collector_neg_binomial.RDS")
+  ))
 ```
@@
 # MCMCvis methods are rather memory intensive. We implement the summary from scratch
-all_chains <- as.matrix(collector)
+all_chains <- collector[[1]]
+for(i in seq_along(collector)) {
+  if (i==1) next
+  all_chains <- rbind(all_chains, collector[[i]])
+}
 
@@
           file = normalizePath(
-            file.path(here(), "data","preprocessed_data", "popdens_summary.csv"
+            file.path(here(), "data","preprocessed_data", "popdens_summary_neg_poisson.csv"
                       ),
@@
 proxies_and_result <- rbind(melt(all_proxies_orig, id.vars = "age"),
-                            melt(PopDens_summary[,2:5], id.vars = "age")
+                            melt(PopDens_summary[,1:4], id.vars = "age")
 )
@@
 
+It can be seen that the result is essentially very similar to the poisson model, only that the credibility intervalls are slightly bigger.
+
 Save as result image to the figures folder.
@@
 ```{r}
-ggsave(normalizePath(file.path(here(), "figures","popdens_estimation.pdf"), mustWork = F),
+ggsave(normalizePath(file.path(here(), "figures","popdens_estimation_neg_binomial.pdf"), mustWork = F),
        plot = popdens_plot_adjusted, width = 21, height = 29.7/2, units = "cm")
@@
 ```{r}
-ggsave(normalizePath(file.path(here(), "figures","p_estimation.pdf"), mustWork = F),
+ggsave(normalizePath(file.path(here(), "figures","p_estimation_neg_binomial.pdf"), mustWork = F),
        plot = p_plot, width = 21, height = 29.7/2, units = "cm")
@@
 ```{r}
-ggsave(normalizePath(file.path(here(), "figures","popvar_plot.pdf"), mustWork = F),
+ggsave(normalizePath(file.path(here(), "figures","popvar_plot_neg_binomial.pdf"), mustWork = F),
        plot = popvar_plot, width = 21, height = 29.7/2, units = "cm")
 ```
+
+And finally, how "Poisson-like" is the Negative Binomial, that means, how large is 'r'?
+If r tends to infinity, than the Negative Binomial Distribution approaches the Poisson Distribution.
+Values over 50 are considered to be essential binomial.
+
+```{r}
+gc(verbose = F)
+
+r_plot <- 
+  all_chains[,"r"] %>%
+  as.data.frame %>%
+  pivot_longer(everything()) %>%
+  ggplot(aes(x = value, y = name)) + stat_halfeye() +
+  xlim(0,1000) + xlab("Value of R") + ylab("") + theme_minimal()
+r_plot
+```
+
+Or, to give the values of r more numerical:
+
+```{r}
+this_mcmc_summary["r",]
+```
+
+There is quite a range, since there is no real difference once the value surpasses the threshold to become "Poisson-like".
+But it indicates quite clearly, that even when we model with variable variance, the model tends strongly to converge on a poisson-like solution.
+We can take this as indication that a more restrictive Poisson model is not an unreasonable choice.

### 2022-12-08 analysis/sensitivity_analysis.Rmd

@@
 output:
-  pdf_document: default
   html_document:
     keep_md: yes
+  pdf_document: default
 bibliography: references.bib

### 2022-12-08 analysis/analysis.Rmd
updated names for variables according to the mathematical model

--- 2022-05-28 analysis/analysis.Rmd
@@
   # Estimate the initial state vector of population abundances
-  nEnd ~ dnorm(nSites[nYears] * MeanSiteSize / AreaSwissPlateau, sd=0.5)
+  nEnd ~ dnorm(z[nYears] * Meapize / AreaSwissPlateau, sd=0.5)
 
@@
     # The actual number of sites per year
-    nSites[t] ~ dpois(lambda[t])
+    z[t] ~ dpois(lambda[t])
 
@@
     constraint_lambda_lower[t] ~ dconstraint(
-      nSites[t]/nSites[t-1] < (max_growth_rate + 1)
+      z[t]/z[t-1] < (max_growth_rate + 1)
       )
     constraint_lambda_upper[t] ~ dconstraint(
-      nSites[t-1]/nSites[t] < (max_growth_rate + 1)
+      z[t-1]/z[t] < (max_growth_rate + 1)
       )
@@
     # changes in relation to the proxies
-    log(lambda[t]) <- log(nSites[t-1]) + (
-      p[1] * sumcal[t]  +
-        p[2] * openness[t] +
-        p[3] * aoristic_sum[t] + 
-        p[4] * dendro[t]
+    log(lambda[t]) <- log(z[t-1]) + (
+      beta[1] * sumcal[t]  +
+        beta[2] * openness[t] +
+        beta[3] * aoristic_sum[t] + 
+        beta[4] * dendro[t]
       )
@@
   # Relevance of the proxies is estimated as Dirichlet distribution
-  p[1:4] ~ ddirch(alpha[1:4])
+  beta[1:4] ~ ddirch(alpha[1:4])
 
@@
   PopDens[1:(nYears)] <- PopTotal[1:(nYears)] / AreaSwissPlateau
-  PopTotal[1:nYears] <- nSites[1:nYears] * MeanSiteSize
+  PopTotal[1:nYears] <- z[1:nYears] * MeanSiteSize
 })
@@
   PopDens = rep(model_constants$nEnd,redux_nrow),
-  nSites = rep(50,redux_nrow)
+  z = rep(50,redux_nrow)
 )
@@
 Finally, we have to define which parameters (apart from the top-level stochastic nodes of the model) are to be observed and stored.
-We also include the population density as a core parameter and the relative proportion of proxies p:
-
-```{r}
-model_monitors <- c("PopDens", "p")
+We also include the population density as a core parameter and the relative proportion of proxies beta:
+
+```{r}
+model_monitors <- c("PopDens", "beta")
 ```
@@
 ```{r}
-MCMCchains(mcmcSamples, params = "p") %>%
+MCMCchains(mcmcSamples, params = "beta") %>%
   as.data.frame %>%
-  rename(sumcal = 'p[1]', openness = 'p[2]', aorist = 'p[3]', dendro = 'p[4]') %>%
+  rename(sumcal = 'beta[1]', openness = 'beta[2]', aorist = 'beta[3]', dendro = 'beta[4]') %>%
   pivot_longer(everything()) %>%
@@
 
-For comparison with the priors, we simulate from these, first for p:
+For comparison with the priors, we simulate from these, first for beta:
 
@@
 
-p_mat <- matrix(ncol = 4, nrow = n_prior)
+beta_mat <- matrix(ncol = 4, nrow = n_prior)
 for(i in 1:n_prior){
- p_mat[i,] <- rdirch(1, c(alpha[[1]][i], alpha[[2]][i], alpha[[3]][i], alpha[[4]][i]))
-}
-MCMCtrace(collector, params = "p", priors = p_mat, pdf = F, ind = T, Rhat = T)
-```
-
-And then for lambda and nSites (respectively derived for PopDens):
-
-```{r}
-nSites <- matrix(ncol = model_constants$nYears, nrow = n_prior)
+ beta_mat[i,] <- rdirch(1, c(alpha[[1]][i], alpha[[2]][i], alpha[[3]][i], alpha[[4]][i]))
+}
+MCMCtrace(collector, params = "beta", priors = beta_mat, pdf = F, ind = T, Rhat = T)
+```
+
+And then for lambda and z (respectively derived for PopDens):
+
+```{r}
+z <- matrix(ncol = model_constants$nYears, nrow = n_prior)
 MeanSiteSize <- rpois(n_prior, 50)
@@
 
-nSites[,model_constants$nYears] <- rpois(n_prior,nEnd / MeanSiteSize * model_constants$AreaSwissPlateau)
+z[,model_constants$nYears] <- rpois(n_prior,nEnd / MeanSiteSize * model_constants$AreaSwissPlateau)
 
 for(i in (model_constants$nYears-1):1) {
-  nSites[,i] <- rpois(n_prior, nSites[,i+1])
-}
-
-PopDens <- nSites * MeanSiteSize / model_constants$AreaSwissPlateau
+  z[,i] <- rpois(n_prior, z[,i+1])
+}
+
+PopDens <- z * MeanSiteSize / model_constants$AreaSwissPlateau
 
@@
 
-p_plot <- all_chains[,grep("^p\\[", params)] %>%
+beta_plot <- all_chains[,grep("^beta\\[", params)] %>%
   as.data.frame %>%
-  rename(sumcal = 'p[1]', openness = 'p[2]', aorist = 'p[3]', dendro = 'p[4]') %>%
+  rename(sumcal = 'beta[1]', openness = 'beta[2]', aorist = 'beta[3]', dendro = 'beta[4]') %>%
   pivot_longer(everything()) %>%
@@
   stat_halfeye() + xlab("Percentage Influence") + ylab("Proxy") + theme_minimal()
-p_plot
+beta_plot
 ```
@@
 ```{r}
-ggsave(normalizePath(file.path(here(), "figures","p_estimation.pdf"), mustWork = F),
-       plot = p_plot, width = 21, height = 29.7/2, units = "cm")
+ggsave(normalizePath(file.path(here(), "figures","beta_estimation.pdf"), mustWork = F),
+       plot = beta_plot, width = 21, height = 29.7/2, units = "cm")
 ```

### 2022-12-08 manuscript/paper_revision_1.Rmd
revision 1 martin done for the time being

@@
 In this pilot study, we present an approach that we believe can be usefully transferred to other regions, as well as refined and extended to greatly advance our understanding of prehistoric demography.
-  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density.
-At its core, the statistical model is as follows: if the proxies as a whole have a positive delta, a Poisson draw for the new number of settlements is biased by a positive amount (and similarly for negative deltas).
+  Here, we present a Bayesian hierarchical model suite that uses Negative Binomial linear model and state-space representation to produce absolute estimates of past population size and density.
+At its core, the statistical model is as follows: if the proxies as a whole have a positive delta, a Negative Binomial draw for the new number of settlements is biased by a positive amount (and similarly for negative deltas).
 Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
@@
 In contemporary demography, it is becoming increasingly common to use Bayesian statistics to estimate population trends and project them into the future.
-The Bayesian approach is popular because it offers the possibility of combining heterogenous data, and at the same time qualifying the uncertainty and credibility attached to forecasts.
+Bayesian approach is the natural and principled one for fusing data while appropriately propagating uncertainty.
+This makes it possible to qualifying the uncertainty and credibility attached to forecasts.
 These same characteristics make it well-suited to applications to archaeological data in paleodemographic studies.
@@
   - Our initial model produces more robust, high-resolution estimates of past population dynamics than previous, single-proxy approaches
-  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (6000–1000 BCE)
+  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (6000–1000 BCE) based on an initial value
 ---
@@
 This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015; @carleton_groucutt_2020; @price_end__end_2021].
-While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems common to all approaches relying on a single proxy [@french_manifesto_2021; @schmidt_approaching_2021].
+While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems.
 We believe that these problems cannot be overcome by methodological refinements in this area alone.
-Instead, a Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.
+One of those problems is, that most analysis of this kind rely on a single proxy [@french_manifesto_2021; @schmidt_approaching_2021].
+Other data for demographic processes, such as age composition of grave groups [@bocquet_appel_explaining_2008; @kohler_long_2014] exist, as does the use of environmental data [@lechterbeck2014].
+The combination of a large number of indicators can compensate for the weaknesses of the individual proxies used and open up completely new possibilities for insights [@wolpert_past_2021, 10].
+A Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.
+Problems such as statistical dispersion of sum calibration and effects of the calibration curve will be mitigated respectively detected by this combination with indicators that are not subject to these effects.
+Moreover, in our approach, the mutual evaluation of the proxies in terms of concordance penalises divergent behaviour of an indicator (due to intrinsic biases) by reducing its relevance for the resulting estimate.
+While this does not eliminate these problems entirely, it does reduce them in a numerically manageable way for large data sets.
 
@@
 **Uncertainty in measurements**: All archaeological evidence is inherently uncertain which is carried through to derived measurements.
-However, in most studies, single curves are presented as estimates, and the potential error associated is almost never specified.
+However, in most studies, single curves are presented as estimates, and no uncertainty measurement exists.
 3.
-**Lack of a transfer function**: By 'transfer function', we mean something that allows for the proxy data to be interpreted in terms of actual population size or density.
+**Lack of a linking function**: By 'linking function', we mean something that allows for the proxy data to be interpreted in terms of actual population size or density.
 This could be absolute, i.e.
@@
 In the best cases, there is a qualitative assessment of the informative value of the proxy, not sufficiently accounting for the complex nature of archaeological data.
+
+Our current implementation, with an abductive understanding of the data and the need to set a determinant initial value for the reconstruction, represents a first step towards a fully developed end-to-end hierachical Bayesian model capable of solving these problems.
 
@@
 -   **Indirect**: We will never have direct data on prehistoric population; only proxy data that is thought to be a reasonable substitute.
-The transfer functions linking the proxy data with the desired quantity (population) are unknown.
--   **Contradictory**: When considering several proxies, differences in transfer functions, data quality and noisiness inevitably lead to different results.
+The functions linking the proxy data with the desired quantity (population) are unknown.
+-   **Contradictory**: When considering several proxies, differences in linking functions, data quality and noisiness inevitably lead to different results.
 
@@
 Bayesian approaches are also capable of accounting for spatially and temporally incomplete data: where this data is missing, the uncertainty increases, but this does not prevent general modelling and estimation.
-Finally, hierarchically-structured model suites, with sub-models for each individual proxy, can be used to estimate transfer functions between them and the value to be modelled, thanks to the interaction of a large number of evidence.
+Finally, hierarchically-structured model suites, with sub-models for each individual proxy, can be used to estimate linking functions between them and the value to be modelled, thanks to the interaction of a large number of evidence.
 
@@
 If there is no systematic bias that affects all data sources to the same extent, this results in the most reliable estimate possible through the most heterogeneous set of data sources.
+Even though data fusion can be achieved with different techniques, the Bayesian approach seems to us to be excellently suited for our purposes due to its flexibility and robustness.
 
@@
 This is a clear step forward to a model-based, scientific analysis.
-However, the use of only one proxy, exclusively for testing hypotheses developed independently, creates problems comparable to those of the inductive approaches used so far: lacking a combination with other indicators, one is limited to the problems and conditions of sum calibration.
-Furthermore, this approach loses significant potential information that would be gained by a direct evaluation of time series.
+However, the use of only one proxy, exclusively for testing hypotheses developed independently, creates problems comparable to those of the inductive approaches used so far: without a combination with other indicators, one is largely limited to the problems and conditions of summation calibration.
+However, recent work [@price_end__end_2021; @carleton_groucutt_2020] shows a possible way out here, and might be a promising alternative to abductive approaches like ours.
 
@@
 Our case study targets the period between 6000–1000 BCE.
-The lower limit of this time window was chosen to avoid the so-called 'Hallstatt plateau' in the Northern Hemisphere radiocarbon calibration curve, which causes difficulties for the ^14^C proxy.
+The lower limit of this time window was chosen to avoid the so-called 'Hallstatt plateau' in the Northern Hemisphere radiocarbon calibration curve, which causes difficulties for the using sum calibration and similar approaches.
 The upper limit coincides with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence.
@@
 Table \@ref(tab:tableproxies) provides a non-exhaustive list.
-For our case study, we used a landscape openness indicator; an aoristic sum of typological dated sites; a sum calibration; and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).
+Furthermore, it is possible to combine analytical deductive approaches with the inductive proxies, such as ethnographic analogies and economic modelling.
+For our case study, we used a landscape openness indicator; an aoristic sum of typological dated sites; a radiocarbon based settlement count (sum calibration); and frequency data for dendro-dated lakeshore settlements in the Three Lakes region (western Swiss Plateau).
 
-| Proxies                                        |
+| Possible Proxies                               |
 |------------------------------------------------|
 | Expert estimates                               |
-| Ethnographic Analogies                         |
-| Carrying Capacity                              |
-| Economic modelling                             |
 | Extrapolation of buried individuals            |
@@
 From the Neolithic onwards, known settlement areas in Switzerland concentrate along its rivers and lakes [@christianlüthi2009].
-Thus, our working region offers excellent data for demographic estimation, but poses very specific problems for such an undertaking.
-We have high-resolution information on the temporal sequence of individual lakeside settlements by means of dendro data.
-In these cases, ^14^C data are not as abundant simply because they are inferior to dendro dating.
+Thus, our working region offers on the one hand excellent data for population estimation, but on the other hand poses very specific problems for such an undertaking.
+If we have high-resolution information on the temporal sequence of individual settlements at the lakeside settlements by means of dendro data, this also might cause a research problem with regard to the ^14^C data often used as a proxy.
 
@@
 The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements.
+This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.
 
-## Summed radiocarbon
+## Radiocarbon Based Settlement Count
 
 Summing up the probabilities of radiocarbon dates can undoubtedly be considered the standard approach in demographic reconstruction in archaeology today.
-The logic for this can be derived from the Poisson binomial distribution: such a distribution can be described as the sum of independent Bernoulli trials as it represents the probability estimate for the existence or use of a site per unit time.
+In our variant of the summation calibration, we sum up the individual data in each site so that each site receives a weight of 1.
+These individual probabilities are then summed up over all sites, resulting in a radiocarbon based settlement count (RBSC).
+In many respects, this follows the same logic as generally applied summation calibration, where binning achieves much the same outcome.
+
+The logic for RBSC can be derived from the Poisson binomial distribution: such a distribution can be described as the sum of independent Bernoulli trials as it represents the probability estimate for the existence or use of a site per unit time.
 The expected value per time unit for the number of simultaneously existing sites/settlements is the sum of the individual probabilities for the sites, i.e.
@@
 
-The dataset for the ^14^C sum calibration primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of Martínez-Grau et al.
+The dataset for the RBSC primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of Martínez-Grau et al.
 [-@martínez-grau2021].
@@
 We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites.
-For the creation of the sum calibration, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.
+For the creation of the RBSC, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.
 
@@
 ```
+
+Dendrodated lakeshore settlements and RBSC are two fairly direct proxies for settlement numbers over time.
+However, both are subject to certain constraints, especially on the Swiss Plateau, but also beyond.
+They require the discovery of corresponding settlement sites.
+This is subject to preservation and discovery filters, which distort the results of the derived estimates.
+These biases are not connected to past population trends, but result from research activity, preservation and publication status, as well as dating strategies.
+Therefore, in our model we supplement these two direct proxies with such proxies that are more independent of these biases.
 
@@
 
-We include relative dating information obtained from the heritage authorities of the Swiss cantons (Figure \@ref(fig:aoristmap)).
-These are primarily derived from scattered surface finds, often with a low dating accuracy (only in the range of archaeological periods), incorporated into our model as a typologically-dated, aorist time series.
-However, it is not dependent on radiocarbon dating and thus it avoids the methodological issues of sum calibration.
-Data from `r nrow(aorist)` sites were included in the aoristic sum.
+To add another archaeological indicator of occupation, we include relative dating information obtained from the Swiss cantonal archaeology/heritage management authorities (Figure \@ref(fig:aoristmap)).
+These are primarily derived from scattered surface finds, which often have a low dating accuracy.
+We incorporate this data into our model as a typologically dated aorist time series.
+The dating accuracy is only in the range of archaeological periods, but the advantage is that we are not bound to the conditions and problems of radiocarbon dating and thus methodological issues of sum calibration can be avoided (scatter, calibration curve effects).
+Furthermore, this data provides an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data.
+Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy offered by archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.
 
@@
 
-We assume that the higher the population density in an area, the greater the human influence on the natural environment [@zimmermann2004; @lechterbeck2014].
-Evidence of deforestation can therefore provide indications of population dynamics.
+The utilisation of this proxy is based on the assumption that the higher the population density in an area, the greater the human influence on the natural environment [@lechterbeck2014], and that the extensiveness of agricultural activity in an area is related to human population density [@zimmermann2004].
+Evidence of land clearance in pollen diagrams can therefore provide further indications of population dynamics where humans can be assumed to be the main driver of this process, which is the case in much of Europe.
 The full procedure for deriving this proxy from several different pollen diagrams is detailed in a previous publication [@heitz2021].
-Here, we use five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes.
+In this study, five pollen diagrams from sites mainly in the hinterland of the large Alpine lakes were used.
+The technical steps are also documented in the accompanying R compendium.
+The percentage pollen data, based on a pollen sum of all terrestrial taxa of the individual sites, was combined into one data set by means of a principal component analysis (Figure \@ref(fig:pollenproxy)).
+Only terrestrial pollen taxa with a frequency of more than one third and, if present, with an average frequency of at least 0.1% were selected to reduce potential disturbance by rare species.
+Cereal pollen was explicitly retained as an important anthropogenic indicator.
+As each sample is absolutely dated, the data on the x-axis can be plotted against the openness value on the y-axis to obtain a time series time series for land clearance.
 
@@
 
-Sum calibration, openness and the dendro-dated settlement data was smoothed by a moving average with a 50 years window, corresponding to the unified sampling interval for all proxies.
+RBSC, openness and the dendro-dated settlement data was smoothed by a moving average with a 50 years window, corresponding to the unified sampling interval for all proxies.
+The choice of interval reflects a trade-off between the data quality of the proxies and the complexity of the model.
 The aoristic sum was not smoothed, because it already has a very coarse temporal resolution.
 In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps.
-Population development is simulated in a 'process model' using a Poisson process.
+Population development is simulated in a 'process model' using a Negative Binomial process.
 
@@
 They follow two principles.
-First, a hidden or latent process is assumed, representing the state of the variable of interest $z$ through the entire time series.
+First, a hidden or latent process is assumed, representing the state of the (vector) variable of interest $z$ through the entire time series.
 Every state of variable z as $z_t$ at a specific time is bound by a Markov process to the state of variable $z$ at time before, at $t-1$.
@@
 $$
-z_t = \beta z_{t-1} + \epsilon_t
+z_t = \mathbf{\beta} z_{t-1} + \epsilon_t
 \tag{1}
@@
 $$
-y_t = \alpha z_t + \eta_t
+x_t = \mathbf{\alpha} z_t + \eta_t
 \tag{2}
@@
 The observations are a function of the states only and the parameter $\alpha$ allows the observation at time t to be a biased estimate of the state at time t." [@auger-methe_guide_2021, 4-5].
-Dabei sind $\epsilon$ die process variation, und $\eta$ der observation error, welche mit einer passenden Verteilung modelliert werden können.
+Here $\epsilon$ is the process variation, and $\eta$ the observation error, which can be modelled with a suitable distribution.
+In our current implementation we ignore $\alpha$ and thus assume that $\mathbf{x}_t$ is an unbiased observation.
+We compensate for this (see below) by defining $\beta$ as a simplex with sum 1.
+This leads to the intuition that each of the proxies is seen as more or less good compared to the others, but in their sum they represent the best possible description of $z_t$ (given the available data).
 
-Um die Struktur der Abhängigkeiten zu verdeutlichen, kann man das Modell auch in terms of probability distributions darstellen:
+To clarify the structure of the dependencies, the model can also be represented in terms of probability distributions:
 
 $$
-f(z_t | z_{t-1}, \theta_p), \quad t = 1, ... , T
+f(z_t | z_{t-1}, \mathbf{\theta}_p), \quad t = 1, ... , T
 \tag{3}
@@
 $$
-f(y_t | z_{t}, \theta_o), \quad t = 1, ... , T
+f(x_t | z_{t}, \mathbf{\theta}_o), \quad t = 1, ... , T
 \tag{4}
@@
 
-Dabei sind f and g are two normal probability density functions, and $\theta_p$ and $\theta_o$ are vectors of parameters associated with each equation.
+Here f and g are two probability density functions, and $\theta_p$ and $\theta_o$ are vectors of parameters associated with each equation.
 Eq.
@@
 
-This structure makes these models particularly suitable for demographic reconstruction using archaeological and other data.
-Population density itself is not directly measurable: all we have at our disposal are observations derived by unknown transfer functions.
+This structure makes these models particularly suitable for population reconstruction using archaeological and other data.
+Population density itself is not directly measurable: all we have at our disposal are observations derived by unknown linking functions.
 
 Our overall model is broken down into several hierarchically-connected individual elements.
-The process model represents the demographic development itself, without already being explicitly parameterised with data.
+The process model represents the population development itself, without already being explicitly parameterised with data.
 Here we assume that the latent variable 'number of sites' ($z$) is strongly autocorrelated across different time periods ($z_t$).
@@
 $$
-\lambda_t = z_{t-1} e^{\mathbf{\beta^T \delta \mathbf{y}_t}}
+\lambda_t = z_{t-1} e^{\mathbf{\beta^T \delta \mathbf{x}_t}}
 \tag{7}
@@
 
-$\mathbf{\beta}$ is a parameter vector of weightings that sums to $1$, a vector transpose is indicated by $^T$, and $y_t$ is a vector of proxy difference values.
+$\mathbf{\beta}$ is a parameter vector of weightings that sums to $1$, a vector transpose is indicated by $^T$, and $x_t$ is a vector of proxy difference values.
 
-Using the logarithm as a link function ensures that $\lambda$, which must always be positive, can also be described by variables that may also be negative.
+Using a link function ensures that $\lambda$, which must always be positive, can also be described by variables that may also be negative.
 $\beta$ serves as slope factor, as in a normal linear regression.
@@
 
-Since $\lambda$ and $z$ are essentially in the same range (e.g.
-if $lambda=1$, the expected value for $z$ would also be 1), $z_{t-1}$ must also be log-transformed for the congruence of both values.
+A Poisson distribution has the constraint that the variance is equal to the mean.
+This distribution may have been empirically very useful for modelling respective processes.
+However, we were advised in the peer review of the article that this limitation may not be justified.
+Therefore, we have rebuilt and run the model using a Negative Binomial Distribution, and present here the results of this process model, contrasting them with those of the Poisson model.
+
+A negative binomial distribution can be seen as a generalisation of a Poisson distribution.
+In this case, the expected value and variance are parameterised separately, which gives the model an additional degree of freedom.
+This affects the credibility intervals, which become correspondingly wider.
+A negative binomial distribution is described by a parameter of the probability of success $p$, and a parameter $r$, which indicates the rate.
+Therefore, the implementation of the process model in this variant looks like this:
+
+$$
+z_t \sim NB(r,p_t) \\
+p_t = \frac{r}{r+\lambda_t} \\
+r \sim halfCauchy(0,25)
+\tag{8}
+$$
+
+The negative binomial distribution has a variance $\frac{r (1-p)}{p^2}$, whereby the distribution becomes identical to the Poisson distribution in the limit $r \to \inf$.
+In practise, values above 50 are considered to produce a distribution that approaches a Poisson Distribution.
+In the model, $r$ is estimated from the data using a half-Cauchy distribution with a location parameter of 25 as prior.
+This represents a weakly informative prior according to the standard specifications for a scale hyperparameter [@gelman_prior_2006].
+
 Population size $z_t$ as well as population change $\lambda_t$ are time-dependent.
@@
 \end{aligned}
-\tag{8}
+\tag{9}
 $$
-A gamma distribution centres probability in the range $[0,1[$; adding 1 makes this range $[1-2[$.
+A gamma distribution is definied [0, inf).
+This parameterisation used here centres the probability mass in the range 0-1; adding 1 makes this range 1-2.
 This prevents the number of sites from explosively increasing between two time periods, which would lead to problems for the convergence of the model.
@@
 
-In this initial implementation, the observational model is essentially a Poisson regression, where the proxies are used to inform the change in the number of settlements between time steps.
+In this initial implementation, the observational model is essentially a Negative Binomial regression, where the proxies are used to inform the change in the number of settlements between time steps.
 The individual proxies were z-normalised and absolute differences between time steps were then computed.
@@
 \begin{aligned}
-y'_t &= \frac{y_t - \bar{y}} {\sigma^2_y}\ |\ \sigma^2_y := Standard\ Deviation  \\
-\delta y'_t &= y'_t - y'_{t-1}
+x'_t &= \frac{x_t - \bar{x}} {\sigma^2_x}\ |\ \sigma^2_x := Standard\ Deviation  \\
+\delta x'_t &= x'_t - x'_{t-1}
 \end{aligned}
-\tag{9}
+\tag{10}
 $$
@@
 There are two reasons for this: First, the proxies we use are obviously autocorrelated.
-This can be seen as empirical confirmation that such an autocorrelation can also be regarded as probable for the latent variable of population size behind it.
+This can be seen as empirical confirmation that such an autocorrelation can also be regarded as probable for the latent variable of number of sites behind it.
 On the other hand, a model that does not assume such an autocorrelation would mean that population development is primarily controlled by migration, since the population already present in an area would play no role in this.
-We think it more likely that it is primarily births and death events, together with migration, that are responsible for population numbers and their change.
-Therefore, we view the population at time t as a function of the population at time t-1 plus a change that can be estimated from the change in the proxies.
+We think it more likely that it is primarily births and death events, together with migration, that are responsible for number of sites and their change.
+Therefore, we view the population at time t (and therefore the number of sites) as a function of the population at time t-1 plus a change that can be estimated from the change in the proxies.
 In this way, we make all time steps interdependent, which constrains the model in a comprehensible and justifiable way, and thus narrows the probability range for the prediction by providing additional information in the model.
-Here, $\beta_i$ is a scaling factor that represents the influence of the respective proxy.
+Here, $\beta_i$ is a scaling factor that represents the influence of the changes in the individual proxies.
 It is a confidence value of the model for the respective proxy, so that the sum of all $\beta_i$ results in 1.
@@
 \sum_{i=1}^n \beta_i = 1
-\tag{10}
+\tag{11}
 $$
@@
 \end{aligned}
-\tag{11}
+\tag{12}
 $$
 
-Intuitively, we consider the sum of the proxies as determinant of the change in number of settlements.
+Intuitively, we consider the sum of the changes of the proxies as best possible representation of the change in number of settlements, given the data that we have.
 That is, the share of each individual proxy is variable and is estimated within the model.
-This share is recorded within the model as the parameter p.
+This is represented by $\beta$ in the mathematical model.
+In the computational model, we named this parameter p (for percentage influence)
 
-The error value is represented by the Poisson process in the process model.
+Another way of approaching this would be to take an end-to-end Bayesian approach [eg.
+@price_end__end_2021].
+Here we would model the relationships between the proxies and the population size directly, and let the model estimate the parameters of this process.
+This was our original approach.
+We had to abandon it for the time being due to the underdetermined nature of such a model with our data.
+However, we are convinced that such an approach would lead to a better model that propagates uncertainty more appropriately and avoids certain biases associated with the conversion to a fixed time series and the associated biases due to preprocessing.
+We plan to revert to this more principled approach when applying our model on a larger scale with a regionalisation and an extension of the proxies.
+
+The error value is represented by the Negative Binomial process in the process model.
 In this implementation, the model finds the best possible combination between the individual proxies to describe a settlement dynamic.
@@
 
+## Negative Binomial Model
+
 ```{r popdensplot, fig.cap="Estimate of population density predicted by the model. The four input proxies are also plotted (scaled) for comparison."}
-knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popdens_estimation.pdf"))
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popdens_estimation_neg_binomial.pdf"))
 ```
@@
 ```{r varcoeffplot, fig.cap="Variability of the model estimate of population density over time, with the estimate itself for reference."}
-knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popvar_plot.pdf"))
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popvar_plot_neg_binomial.pdf"))
 ```
@@
 contradictions in the proxies.
-The coefficient of variation is 0.13 for the beginning and 0.1 for the end of the estimate, with the greatest variability (0.47) seen around 2150 BCE.
+The coefficient of variation is 0.13 for the beginning and 0.1 for the end of the estimate, with the greatest variability (0.59) seen around 2150 BCE.
 This is not surprising as there are fewer archaeological contexts recorded from the earlier phase of the Early Bronze Age, c.
@@
 The beginning and end of the time series are relatively clearly determined, resulting from the *a priori* setting of final population density, but also from the uniformity of the proxies during these periods.
-Overall, the variability is relatively stable over the entire estimation and averages 33% of the respective mean.
+Overall, the variability is relatively stable over the entire estimation and averages 35% of the respective mean.
 
 ```{r pdistribution, fig.cap="Distribution of influence ratios of proxies on model's final estimation of number of sites."}
-knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "p_estimation.pdf"))
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "p_estimation_neg_binomial.pdf"))
 ```
@@
 The parameter $p$ reflects the relative weight given to the individual proxies.
-Its posterior distribution (Figure \@ref(fig:pdistribution)) shows that the model weights the openness indicator the highest, averaging slightly above 60%, followed by the sum calibration, with an average of about 20%.
+Its posterior distribution (Figure \@ref(fig:pdistribution)) shows that the model weights the openness indicator the highest, averaging slightly above 60%, followed by the RBSC, with an average of about 20%.
 The aoristic sum is slightly above 10%, whereas the importance of the dendro-dated settlements is below 10%.
@@
 The aoristic sum is flat for long periods, making it difficult to integrate with other proxies.
-The sum calibration shows very strong short-term fluctuations, at least partly due to the calibration curve, which suggests that it does not reliably represent a continuous population trend.
+The RBSC shows very strong short-term fluctuations, at least partly due to the calibration curve, which suggests that it does not reliably represent a continuous population trend.
 Its fluctuations have an impact on the model's estimate, albeit to a lesser extent than the general trend.
+
+## Poisson Model
+
+In the model, the parameter $r$ was also estimated for the negative binomial model, which indicates the convergence of the distribution to a Poisson distribution.
+If this value exceeds 50, then a Poisson model is essentially given.
+The posterior distribution of this parameter looks as follows:
+
+```{r rdistribution, fig.cap="Posterior distribution of the rate parameter r."}
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "r.png"))
+```
+
+Clearly, the central tendency of this distribution is above 50.
+The long tail is caused by the fact that above a certain threshold the exact parameter r no longer makes a difference, since the distribution approaches a Poisson distribution.
+We can therefore assume that the result essentially speaks for a support of a Poisson distribution from the data.
+If we take this as a basis, then the mean values of the estimate are essentially identical, but the credibility intervals are smaller (Fig.
+\@ref(fig:popdensplotpoisson))
+
+```{r popdensplotpoisson, fig.cap="Estimate of population density predicted by the model, here with the Poisson distribution in the Process model.
+The four input proxies are also plotted (scaled) for comparison."}
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "popdens_estimation.pdf"))
+```
+
+## Sensitivity Analysis
+
+The model has essentially two predefined parameters.
+These are the mean site size and the end value for the population.
+The mean site size should not have a significant influence on the analysis, as it represents a pure scaling factor.
+In contrast, the given start value defines the course of the reconstruction and should therefore have a dominant influence on the model estimates.
+
+In the sensitivity analysis (see supplementary material or GitHub repository), the model was run with 20 parameter settings each, with the one used in the model forming the middle of all cases.
+
+```{r sensanalysismeansitesize, fig.cap="Estimate of population density predicted by the model in a sensibility analysis sweep over the parameter mean site size."}
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "sens_analysis_mean_site_size.png"))
+```
+
+```{r sensanalysisnend, fig.cap="Estimate of population density predicted by the model in a sensibility analysis sweep over the parameter final population density."}
+knitr::include_graphics(file.path(rstudioapi::getActiveProject(), "figures", "sens_analysis_nend.png"))
+```
+
+The analyses carried out (Fig.
+\@ref(fig:sensanalysismeansitesize) resp.
+\@ref(fig:sensanalysisnend)) show a clear pattern and confirm the assumptions.
+The model is not sensitive to mean site size, but is determined by final population density.
 
@@
 Comparing the model's overall estimate with the individual proxies provides several insights into the quality of these records.
-The sum calibration, currently the most frequently used proxy for (relative) population change in prehistory, has its large fluctuations dampened when considered alongside other proxies.
+The sum calibration (here: RBSC), currently the most frequently used proxy for (relative) population change in prehistory, has its large fluctuations dampened when considered alongside other proxies.
 This is especially ture of the first fluctuation shortly after 4000 BCE.
-The expected increase in archaeological remains with the onset of Neolithisation is still clearly visible, but the overall curve is much flatter than the sum calibration itself.
+The expected increase in archaeological remains with the onset of Neolithisation is still clearly visible, but the overall curve is much flatter than the RBSC itself.
 The period between 3950 and 3700 BCE, contemporaneous with the first major settlement of the Three Lakes regions' lakeshores, coincides with a noticeable plateau in the calibration curve, producing an overestimation of the ^14^C density.
@@
 
-On average, the model weights the sum calibration at about 20%, significantly less than the 60% afforded to the openness indicator.
+On average, the model weights the RBSC at about 20%, significantly less than the 60% afforded to the openness indicator.
 After an initial increase, which is easily explained by spread of agriculture, the openness indicator tends to fluctuate less and thus has a dampening effect on the overall estimate.
-In general, this trend in the sum calibration is well reflected in land openness, while changes within the Neolithic and Bronze Age are more gradual.
+In general, this trend in the RBSC is well reflected in land openness, while changes within the Neolithic and Bronze Age are more gradual.
+However, it should be noted that temporal changes in land use strategies are not yet modelled in the current implementation.
+This is clearly a potential for improvement for future model generations.
 
@@
   file = normalizePath(
-            file.path(here(), "data","preprocessed_data", "popdens_summary.csv"
+            file.path(here(), "data","preprocessed_data", "popdens_summary_neg_poisson.csv"
                       )
@@
 
-The key advance in the model we present is the ability to estimate, in absolute terms, past population sizes and the uncertainty accompanying our present knowledge.
+The model we present is a step towards a fully integrated hierarchical Bayesian model for estimating absolute population sizes.
+With this model we can already combine different data (qualities) and evaluate the reliability of the estimates at different points in time.
+Currently, absolute estimation is still done using an arbitrary starting value.
 These estimates can be a basis for further studies where relative measures of population development are not helpful, such as long-term land use studies.
@@
 Jan Kolář was supported by a long-term research development project (RV67985939) and by a grant from the Czech Science Foundation (19-20970Y).
+Julian Laabs was also funded by the German Research Foundation (DFG), Collaborative Research Centre 1266 'Scales of Transformation' (project number 2901391021).
 We also thank the Institute of Archaeological Sciences of the University of Bern for its support and faith in the outcome of our modelling project.
-Finally, we thank (already) the unknown reviewers for their helpful comments, which will certainly improve this manuscript significantly.
+Finally, we thank Michael Holton Price and W.
+Christopher Carleton for their really thorough review, helpful comments and valuable suggestions, which will certainly improve this manuscript significantly.
 

### 2022-12-09 manuscript/paper_revision_1.Rmd
Corrections and comments
Comments in <!-- HTML comments -->. NB.: I haven't re-knitted anything.

@@
 ---
-title: "Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the North of the Swiss Alps"
+title: "Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from north of the Swiss Alps"
 author:
@@
 In this pilot study, we present an approach that we believe can be usefully transferred to other regions, as well as refined and extended to greatly advance our understanding of prehistoric demography.
-  Here, we present a Bayesian hierarchical model suite that uses Negative Binomial linear model and state-space representation to produce absolute estimates of past population size and density.
-At its core, the statistical model is as follows: if the proxies as a whole have a positive delta, a Negative Binomial draw for the new number of settlements is biased by a positive amount (and similarly for negative deltas).
-Using the  area North of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
+  <!-- JR: For me this is far too much methodological detail for an abstract; I prefer the submitted version.
+I think simply removing "regression" is sufficient to head-off potential for confusion with an explanatory model.
+-->Here, we present a Bayesian hierarchical model suite that uses negative binomial linear model and state-space representation to produce absolute estimates of past population size and density.
+At its core, the statistical model is as follows: if the proxies as a whole have a positive delta, a negative binomial draw for the new number of settlements is biased by a positive amount (and similarly for negative deltas).
+Using an area north of the main ridge of the Swiss Alps in prehistoric times (6000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone.
 The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research.
@@
   - Our initial model produces more robust, high-resolution estimates of past population dynamics than previous, single-proxy approaches
-  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (6000–1000 BCE) based on an initial value
+  - We provide absolute estimates of population size and density on an area north of the Swiss Alps in prehistoric times (6000–1000 BCE) based on an initial value
 ---
@@
 
+<!-- JR:  Okay I know I'm reopening an old issue here, but I really don't think "the north of the Swiss Alps" works as a geographic descriptor in English.
+If I encountered it for the first time I'd think it meant all of Europe north of the Alps.
+Removing the definite article helps a bit, but it's still confusing and awkward.
+I preferred "Swiss plateau" but, if that's unacceptable, can we just say "Northeast Switzerland" or something?
+-->
+
 Prehistorians have long recognised demography as a fundamental force in human cultural evolution [@childe_man_1936].
@@
 [-@kintigh_grand_2014] list human influence, dominance, population size, and population growth amongst their 'grand challenges' for archaeology in the 21st century.
-
 In particular, the 'dates as data' technique [@rick_dates_1987], using the frequency of radiocarbon dates as a proxy for population dynamics, has been significantly developed in the last decade [e.g.
@@
 This approach has contributed greatly to our understanding of prehistoric demography, but is not without its critics [@attenbrow_hiscock_2015; @carleton_groucutt_2020; @price_end__end_2021].
-While the methodology continues to evolve and address these critiques [@crema_review_2022], it remains subject to fundamental problems.
-We believe that these problems cannot be overcome by methodological refinements in this area alone.
-One of those problems is, that most analysis of this kind rely on a single proxy [@french_manifesto_2021; @schmidt_approaching_2021].
-Other data for demographic processes, such as age composition of grave groups [@bocquet_appel_explaining_2008; @kohler_long_2014] exist, as does the use of environmental data [@lechterbeck2014].
-The combination of a large number of indicators can compensate for the weaknesses of the individual proxies used and open up completely new possibilities for insights [@wolpert_past_2021, 10].
-A Bayesian approach offers a robust, quantitative methodology for inferring prehistoric population dynamics from multiple proxies, including summed radiocarbon dates.
-Problems such as statistical dispersion of sum calibration and effects of the calibration curve will be mitigated respectively detected by this combination with indicators that are not subject to these effects.
-Moreover, in our approach, the mutual evaluation of the proxies in terms of concordance penalises divergent behaviour of an indicator (due to intrinsic biases) by reducing its relevance for the resulting estimate.
-While this does not eliminate these problems entirely, it does reduce them in a numerically manageable way for large data sets.
+While the methodology continues to evolve to address these critiques, methodological refinements alone cannot overcome the fundamental issues that limit its use as a direct proxy for past population [@crema_review_2022].
+Similarly, while there are several other archaeological [e.g.
+@bocquet_appel_explaining_2008; @kohler_long_2014] and environmental [@lechterbeck2014] proxies for demographic processes, analyses that rely on any one of these proxies alone are inevitably limited by the nature of that proxy [@french_manifesto_2021; @schmidt_approaching_2021].
 
-In the following article we present our methodology and demonstrate its application with a case study in the area north of the Swiss Alps.
-We use a type of Bayesian Generalised Linear Model [@kruschke2015] to combine the different proxies.
+<!-- JR: I've replaced "indicator" with "proxy" in this section, but I think it's still mixed elsewhere. I don't mind which one we use, but I think we should stick to one term. -->
+
+In this article we present a new, Bayesian methodology that combines multiple proxies to infer prehistoric population dynamics.
+As a case study, we apply this method to a region north of the Swiss Alps, deriving an absolute estimate of regional population density between 6000 and 1000 BCE from a combination of radiocarbon, dendrochronology, aoristic and landscape openness data.
+This Bayesian approach offers a robust, quantitative methodology for demographic inference that not only compensates for the weaknesses of individual types of record, but can open up entirely new areas of insight [@wolpert_past_2021, 10].
+Issues inherent to single proxies (e.g.
+for radiocarbon, statistical dispersion from summing and distortions transferred from the calibration curve) are detected by cross-comparison with proxies that are not subject to these effects.
+The Bayesian algorithm penalises divergent behaviour in any one proxy (due to intrinsic biases) by reducing its relevance for the resulting estimate.
+While this does not eliminate these problems entirely, it does mitigate them in a numerically manageable way for large data sets.
+<!-- JR: Not convinced that the following belongs in the introduction. Don't we say all this in the methods section? --->
+Specifically, our method uses a type of Bayesian generalised linear model [@kruschke2015] to combine different proxies of past population density.
 Their assessment (weight and credibility intervals) results from the modelling and the combination of the data itself and is not defined a priori.
@@
 In this sense, it is not a deductive model for testing a hypothesis, but a reconstructive (abductive) one, such it is also used in environmental reconstructions [@inkpen_explaining_2009].
+<!-- JR: end -->
 In this respect, we hope to convince the readers that such an approach is far better suited to realising reliable reconstructions of past population developments than any of the previous approaches.
@@
 Our case study targets the period between 6000–1000 BCE.
-The lower limit of this time window was chosen to avoid the so-called 'Hallstatt plateau' in the Northern Hemisphere radiocarbon calibration curve, which causes difficulties for the using sum calibration and similar approaches.
+The lower limit of this time window was chosen to avoid the so-called 'Hallstatt plateau' in the northern hemisphere radiocarbon calibration curve, which causes difficulties for the using sum calibration and similar approaches.
 The upper limit coincides with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence.
@@
 
-The dataset we use for the number of dendro-dated wetland settlements in the Three Lakes region was collected by Julian Laabs for his PhD thesis [@laabs2019].
-The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements.
+We used a dataset containing the number of dendro-dated wetland settlements in the Three Lakes region [@laabs2019].<!-- JR: As Julian is a co-author, I don't think it's necessary to attribute him specially in the text?
+--> The time series used here runs from `r abs(min(dendro$date))` to `r abs(max(dendro$date))` BCE, and contains the number of chronologically registered fell phases at individual settlements.
 This results in a time series that reflects the settlement of the lakeshores in the Neolithic and Bronze Age periods.
 
-## Radiocarbon Based Settlement Count
+## Radiocarbon-based settlement count
 
-Summing up the probabilities of radiocarbon dates can undoubtedly be considered the standard approach in demographic reconstruction in archaeology today.
-In our variant of the summation calibration, we sum up the individual data in each site so that each site receives a weight of 1.
+We obtained radiocarbon dates for the study region from the XRONOS database (https://xronos.ch), supplemented by dates collected by Laabs [-@laabs2019] and Martínez-Grau et al.
+[-@martínez-grau2021].
+The dataset contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites (see Figure \@ref(fig:c14map)).
+The dates range from `r max(c14$bp)` to `r min(c14$bp)` uncal BP – corresponding to a calibrated range wider than our target of 6000–1000 BCE in order to minimise boundary effects.
+
+We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites.
+For the creation of the RBSC, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.
+
+<!-- JR: I'm not convinced about the remainder of this section or by the phrase "radiocarbon-based site settlement".
+Basically we are following the standard summed radiocarbon method, and calling it something different here is likely to confuse many.
+And while I agree with the intepretation in terms of probability theory, it seems out of place here.
+-->
+Following the standard approach to sum calibration, we summed up the individual data in each site so that each site receives a weight of 1.
 These individual probabilities are then summed up over all sites, resulting in a radiocarbon based settlement count (RBSC).
@@
 
-The dataset for the RBSC primarily consists of data from the XRONOS database (https://xronos.ch), supplemented by dates from the unpublished PhD thesis of Julian Laabs [@laabs2019] and the data collection of Martínez-Grau et al.
-[-@martínez-grau2021].
-It contains a total of `r nrow(c14)` single ^14^C data from `r length(unique(c14$site))` sites (see Figure \@ref(fig:c14map)).
-The dates in the dataset range in ^14^C years from `r max(c14$bp)` to `r min(c14$bp)` uncal BP.
-This time window extends beyond the study horizon in order to minimise boundary effects.
-
-We binned the data at site levels to obtain a temporally dispersed count and thus an expected value of contemporaneous ^14^C dated sites.
-For the creation of the RBSC, the corresponding functions of the R package rcarbon [@crema2021] were used with their default settings.
-
 ```{r c14map, fig.cap="The location of the ^14^C dated sites in the dataset."}
@@
 
+<!-- JR: I am not sure the following paragraph is necessary, considering we have already made the case for "single proxy bad, multi-proxy good" several times.
+If we do keep it, it would perhaps fit better at the top of materials section (i.e., explaining the choice of proxies before we explain what they are).
+-->
 Dendrodated lakeshore settlements and RBSC are two fairly direct proxies for settlement numbers over time.
@@
 
-To add another archaeological indicator of occupation, we include relative dating information obtained from the Swiss cantonal archaeology/heritage management authorities (Figure \@ref(fig:aoristmap)).
-These are primarily derived from scattered surface finds, which often have a low dating accuracy.
+We obtained relative dating information from the Swiss cantonal archaeology/heritage management authorities (Figure \@ref(fig:aoristmap)).
+This is primarily derived from scattered surface finds, which often have a low dating accuracy.
 We incorporate this data into our model as a typologically dated aorist time series.
@@
 Furthermore, this data provides an independent indicator with regard to the methodology of the ^14^C data, even if they are influenced by similar transmission filters and archaeological conditions as the evaluation of ^14^C data.
-Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy offered by archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.
+Data from `r nrow(aorist)` sites were included in the aoristic sum, which is a very rough indicator due to the low dating accuracy offered by archaeological phases, but which nevertheless has an important role in the normalisation of the data due to its independence from calibration effects.<!-- JR: Okay, really nitpicking here, but...
+are they independent of calibration effects?
+We generally assign absolute dates to periods by interpreting radiocarbon chronologies and if, say, the boundary between two periods falls on a calibration plateau, the date ranges for those periods are going to be correspondingly imprecise, no?
+--->
 
@@
 
+<!-- JR: So I feel like we've ended up with a lot more *technical* detail in the preceeding four sections, when what the reviewers asked for was a more thorough justification of *why* they're population proxies.
+I don't think this is necessarily an improvement...
+-->
+
 ```{r pollenproxy, fig.cap="Value on the first dimension of the PCA against dating of the samples for the individual pollen profiles and their combined average value as the openness indicator."}
@@
 The aoristic sum was not smoothed, because it already has a very coarse temporal resolution.
-In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements located in the north of the Swiss Alps.
+In the construction of our 'observational model', we considered all these proxies as informative of the number of settlements located north of the Swiss Alps.
 Population development is simulated in a 'process model' using a Negative Binomial process.
@@
 However, we were advised in the peer review of the article that this limitation may not be justified.
-Therefore, we have rebuilt and run the model using a Negative Binomial Distribution, and present here the results of this process model, contrasting them with those of the Poisson model.
+Therefore, we have rebuilt and run the model using a negative binomial Distribution, and present here the results of this process model, contrasting them with those of the Poisson model.
 
@@
 We also thank the Institute of Archaeological Sciences of the University of Bern for its support and faith in the outcome of our modelling project.
-Finally, we thank Michael Holton Price and W.
-Christopher Carleton for their really thorough review, helpful comments and valuable suggestions, which will certainly improve this manuscript significantly.
+Finally, we would like to thank reviewers Michael Holton Price and W.
+Christopher Carleton, as well as editor Enrico Crema, for an extremely thorough review process.
+Their helpful comments and valuable suggestions have improved both this paper and the model it describes signficiantly.
 

### 2023-04-24 analysis/analysis.Rmd

@@
 
-The data set for the 14C sum calibration is composed of the above-mentioned sources.
+The data set for the 14C sum calibration is composed of the above-mentioned sources. This data consists of 1135 dates from 246 sites.
 
@@
 
-Next, we calibrate the 14C data using rcarbon: <!--# uncal vs cal? -->
+Next, we calibrate the 14C data using rcarbon:
 
@@
 ```{r}
-plot(1950-sumcal$grid$calBP, sumcal$grid$PrDens,type="l", main = paste0("n=", length(c14.calibrated)))
-
+plot(1950-sumcal$grid$calBP, sumcal$grid$PrDens,type="l")
 ```
@@
 ```{r}
-all_proxies[,2:5] <- all_proxies[,2:5] %>% scale() %>% diff() %>% rbind(0,.)
+all_proxies_diff <- all_proxies
+all_proxies_diff[,2:5] <- all_proxies_diff[,2:5] %>% scale() %>% diff() %>% rbind(0,.)
+```
+
+We can also visualise these deltas of the proxies:
+
+```{r}
+all_proxies_diff %>% pivot_longer(2:5) %>% ggplot(aes(x = age, y = value)) + geom_line() + facet_wrap(. ~ name)
 ```
@@
 redux_factor <- 1
-redux_nrow <- round(nrow(all_proxies) * redux_factor)
-redux_seq <- round(seq(from=1,to=nrow(all_proxies), length.out = redux_nrow))
-
-model_data <- all_proxies[redux_seq,]
+redux_nrow <- round(nrow(all_proxies_diff) * redux_factor)
+redux_seq <- round(seq(from=1,to=nrow(all_proxies_diff), length.out = redux_nrow))
+
+model_data <- all_proxies_diff[redux_seq,]
 ```

### 2023-04-24 analysis/analysis.Rmd
Simplify file paths
Removing unnecessary calls to normalizePath and file.path()

@@
 ```{r}
-pollen_locations <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "pollen_locations.csv")))
+pollen_locations <- read.csv(here("data","raw_data", "pollen_locations.csv"))
 pollen_locations
@@
 ```{r}
-source(file = normalizePath(file.path(here(), "code", "neotoma_helpers.R")))
+source(here("code", "neotoma_helpers.R"))
 ```
@@
 ```{r}
-common_names <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "species.csv")), row.names = 1)
+common_names <- read.csv(here("data", "raw_data", "species.csv"), row.names = 1)
 common_names
@@
 ```{r}
-ggsave(normalizePath(file.path(here(), "figures","pollenproxy.pdf"), mustWork = F),
+ggsave(here("figures","pollenproxy.pdf"),
        plot = pollenproxy_plot, width = 21, height = 29.7/2, units = "cm")
@@
 ```{r}
-c14 <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "14c_swiss_plateau.csv")))
+c14 <- read.csv(here("data","raw_data", "14c_swiss_plateau.csv"))
 ```
@@
 ```{r}
-sites <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "site_data_fuzzy.csv")))
+sites <- read.csv(here("data","raw_data", "site_data_fuzzy.csv"))
 ```
@@
 
-chrono_rough <- read.csv(file = normalizePath(
-  file.path(here(), "data","raw_data", "chrono_rough.csv")
-  ),
+chrono_rough <- read.csv(here("data","raw_data", "chrono_rough.csv"),
                          row.names = 1)
 
-chrono_vrough <- read.csv(file = normalizePath(
-  file.path(here(), "data","raw_data", "chrono_very_rough.csv")
-  ),
+chrono_vrough <- read.csv(here("data","raw_data", "chrono_very_rough.csv"),
                           row.names = 1)
@@
 ```{r}
-lss <- read.csv(file = normalizePath(file.path(here(), "data","raw_data", "dendro_years.csv")))
+lss <- read.csv(here("data","raw_data", "dendro_years.csv"))
 ```
@@
 ```{r}
-write.csv(all_proxies, file = normalizePath(
-  file.path(here(), "data","preprocessed_data", "all_proxies.csv"
-            ),
-  mustWork = F)
-  )
+write.csv(all_proxies, here("data","preprocessed_data", "all_proxies.csv"))
 ```
@@
 ```{r}
-all_proxies <- read.csv(file = normalizePath(
-  file.path(here(), "data","preprocessed_data", "all_proxies.csv")
-  ),
+all_proxies <- read.csv(here("data","preprocessed_data", "all_proxies.csv"),
                         row.names = 1)
@@
 
-all_proxies_orig <- read.csv(file = normalizePath(file.path(here(), "data","preprocessed_data", "all_proxies.csv")), row.names = 1)
+all_proxies_orig <- read.csv(here("data","preprocessed_data", "all_proxies.csv"), row.names = 1)
 
@@
 write.csv(PopDens_summary, 
-          file = normalizePath(
-            file.path(here(), "data","preprocessed_data", "popdens_summary.csv"
-                      ),
-            mustWork = F)
-          )
+          here("data", "preprocessed_data", "popdens_summary.csv"))
 ```
@@
 ```{r}
-all_proxies_orig <- read.csv(file = normalizePath(file.path(here(), "data","preprocessed_data", "all_proxies.csv")), row.names = 1)
+all_proxies_orig <- read.csv(here("data","preprocessed_data", "all_proxies.csv"), 
+                             row.names = 1)
 
@@
 ```{r}
-ggsave(normalizePath(file.path(here(), "figures","popdens_estimation.pdf"), mustWork = F),
-       plot = popdens_plot_adjusted, width = 21, height = 29.7/2, units = "cm")
+ggsave(here("figures", "popdens_estimation.pdf"),
+       plot = popdens_plot_adjusted, 
+       width = 21, height = 29.7/2, units = "cm")
 ```
@@
 ```{r}
-ggsave(normalizePath(file.path(here(), "figures","beta_estimation.pdf"), mustWork = F),
+ggsave(here("figures","beta_estimation.pdf"),
        plot = beta_plot, width = 21, height = 29.7/2, units = "cm")
@@
 ```{r}
-ggsave(normalizePath(file.path(here(), "figures","popvar_plot.pdf"), mustWork = F),
+ggsave(here("figures","popvar_plot.pdf"),
        plot = popvar_plot, width = 21, height = 29.7/2, units = "cm")

### 2023-04-24 analysis/analysis.Rmd
Refactor maps in analysis
Removing external/online dependencies: webshot, magick, natural earth data (now in raw_data), stamen

@@
 library(sf)
-library(mapview)
+library(ggspatial)
 library(vegan)
@@
 
+```{r}
+source(here("code/map_helpers.R"))
+```
+
 ## Data collection
@@
 pollen_locations.sf <- st_as_sf(pollen_locations, 
-                                coords = c("lng", "lat"), # coordinate columns
-                                crs = 4326)  # coordinate references system in epsg))
-pollen_locations.sf.lv95 <- st_transform(pollen_locations.sf, crs = 2056)
-
-pollenmap <- tempfile(fileext = ".png")
-mapshot(
-  mapview(pollen_locations.sf),
-  file = pollenmap
-  )
-
-img <- magick::image_read(pollenmap)
-old.par <- par(mar = c(0, 0, 0, 0))
-plot(img)
-par(old.par)
+                                coords = c("lng", "lat"), crs = 4326)
+
+base_map() + 
+  geom_sf(data = pollen_locations.sf, inherit.aes = FALSE) +
+  base_coord()
 ```
@@
 ```{r}
-c14map <- tempfile(fileext = ".png")
-mapshot(
-  mapview(st_as_sf(c14,
-                   coords = c("lng", "lat"), 
-                   crs = 4326),layer.name = "14C Data"
-          ),
-  file = c14map
-  )
-
-img <- magick::image_read(c14map)
-old.par <- par(mar = c(0, 0, 0, 0))
-plot(img)
-par(old.par)
+c14.sf <- c14 |>
+  distinct(site, lat, lng) |>
+  st_as_sf(coords = c("lng", "lat"),  crs = 4326)
+
+base_map() +
+  layer_spatial(c14.sf) +
+  base_coord()
 ```
@@
 ```{r}
-sitesmap <- tempfile(fileext = ".png")
-mapshot(
-  mapview(st_as_sf(sites,
+sites.sf <- st_as_sf(sites,
                    coords = c("LV03_Rechtswert_fuzzy", "LV03_Hochwert_fuzzy"),
-                   crs = 21781),
-          layer.name = "Sites"
-  ),
-  file = sitesmap
-)
-img <- magick::image_read(sitesmap)
-old.par <- par(mar = c(0, 0, 0, 0))
-plot(img)
-par(old.par)
+                   crs = 21781)
+
+base_map() + 
+  layer_spatial(sites.sf) +
+  base_coord()
 ```

### 2023-05-03 analysis/alt_analysis.Rmd

@@
 It is now a Negative Binomial Distribution, where both the expected value and the variance can vary.
-The expected value is still derived from $\lambda_t$, but this is integrated into the model via a success parameter $prob_t$, which results as follows:
+The expected value is still derived from $\lambda_t$, but this is integrated into the model via a success parameter $p_t$, which results as follows:
 
 $$
-prob_t = \frac{r}{r+\lambda_t}
+p_t = \frac{r}{r+\lambda_t}
 $$
@@
   # Estimate the initial state vector of population abundances
-  nEnd ~ dnorm(nSites[nYears] * MeanSiteSize / AreaSwissPlateau, sd=0.5)
+  nEnd ~ dnorm(z[nYears] * MeanSiteSize / AreaSwissPlateau, sd=0.5)
 
@@
     # The actual number of sites per year
-    nSites[t] ~ dnegbin(prob[t],r)
-    prob[t] <- r/(r+lambda[t])
+    z[t] ~ dnegbin(p[t],r)
+    p[t] <- r/(r+lambda[t])
 
@@
 Finally, we have to define which parameters (apart from the top-level stochastic nodes of the model) are to be observed and stored.
-We also include the population density as a core parameter and the relative proportion of proxies p:
+We also include the population density as a core parameter and the relative proportion of proxies beta:
 
 ```{r}
-model_monitors <- c("PopDens", "p", "r")
+model_monitors <- c("PopDens", "beta", "r")
 ```
@@
 
-p_mat <- matrix(ncol = 4, nrow = n_prior)
+beta_mat <- matrix(ncol = 4, nrow = n_prior)
 for(i in 1:n_prior){
- p_mat[i,] <- rdirch(1, c(alpha[[1]][i], alpha[[2]][i], alpha[[3]][i], alpha[[4]][i]))
+ beta_mat[i,] <- rdirch(1, c(alpha[[1]][i], alpha[[2]][i], alpha[[3]][i], alpha[[4]][i]))
 }
-MCMCtrace(collector, params = "p", priors = p_mat, pdf = F, ind = T, Rhat = T)
+MCMCtrace(collector, params = "beta", priors = beta_mat, pdf = F, ind = T, Rhat = T)
 ```
 
-And then for lambda and nSites (respectively derived for PopDens):
+And then for lambda and z (respectively derived for PopDens):
 
 ```{r}
-nSites <- matrix(ncol = model_constants$nYears, nrow = n_prior)
+z <- matrix(ncol = model_constants$nYears, nrow = n_prior)
 MeanSiteSize <- rpois(n_prior, 50)
@@
 
-nSites[,model_constants$nYears] <- rpois(n_prior,nEnd / MeanSiteSize * model_constants$AreaSwissPlateau)
+z[,model_constants$nYears] <- rpois(n_prior,nEnd / MeanSiteSize * model_constants$AreaSwissPlateau)
 
 for(i in (model_constants$nYears-1):1) {
-  nSites[,i] <- rpois(n_prior, nSites[,i+1])
+  z[,i] <- rpois(n_prior, z[,i+1])
 }
 
-PopDens <- nSites * MeanSiteSize / model_constants$AreaSwissPlateau
+PopDens <- z * MeanSiteSize / model_constants$AreaSwissPlateau
 
@@
 ```{r}
-ggsave(normalizePath(file.path(here(), "figures","p_estimation_neg_binomial.pdf"), mustWork = F),
-       plot = p_plot, width = 21, height = 29.7/2, units = "cm")
+ggsave(normalizePath(file.path(here(), "figures","beta_estimation_neg_binomial.pdf"), mustWork = F),
+       plot = beta_plot, width = 21, height = 29.7/2, units = "cm")
 ```

### 2023-05-03 analysis/analysis.Rmd
Replace neotoma with neotoma2
Which is now on CRAN

@@
 ```{r message=FALSE}
+library(analogue)
+library(coda)
+library(doParallel)
+library(fmcmc)
+library(ggdist)
+library(ggspatial)
+library(grid)
+library(here)
+library(MCMCvis)
+library(neotoma2)
+library(nimble)
+library(rcarbon)
+library(RcppRoll)
+library(reshape2)
+library(sf)
 library(tidyverse)
-library(ggdist)
-library(reshape2)
-library(here)
-library(RcppRoll)
-library(sf)
-library(ggspatial)
 library(vegan)
-library(neotoma)
-library(analogue)
-library(rcarbon)
-library(doParallel)
-library(nimble)
-library(coda)
-library(MCMCvis)
-library(fmcmc)
-library(grid)
 ```
@@
 require(aoristAAR)
-```
-Also, neotoma is not longer available on cran, and it has been replaced by neotoma2, so we had to replace it. That is why we have to load it again:
-
-```{r}
-if(!require('neotoma2')) devtools::install_github('NeotomaDB/neotoma2')
-require(neotoma2)
 ```

### 2023-07-05 analysis/analysis.Rmd
Revise section and chunk titles

@@
 
-## Loading the necessary libraries
+## Dependencies
 
@@
 
-```{r message=FALSE}
+```{r dependencies, message=FALSE}
 library(analogue)
@@
 
-```{r message=FALSE}
+```{r dev-dependencies, message=FALSE}
 if(!require('devtools')) install.packages('devtools')
@@
 
-```{r}
-source(here("code/map_helpers.R"))
-```
-
-## Data collection
+```{r source-dependencies}
+source(here("code", "map_helpers.R"))
+source(here("code", "neotoma_helpers.R"))
+```
+
+## Proxy data
 
@@
 
-### Openness index
+### Pollen (openness index)
 
@@
 
-```{r}
-pollen_locations <- read.csv(here("data","raw_data", "pollen_locations.csv"))
-pollen_locations
+```{r data-pollen-locations}
+(pollen_locations <- read.csv(here("data","raw_data", "pollen_locations.csv")))
 ```
@@
 
-```{r}
+```{r map-pollen-locations}
 pollen_locations.sf <- st_as_sf(pollen_locations, 
@@
 
-We download the data from neotoma:
-
-```{r message=FALSE, warning=FALSE}
+We download data for these locations using the neotoma2 package and some helper functions defined in `code/neotoma_helpers.R`.
+We retain only records that include the NISP, which we convert into a percentage.
+
+```{r data-pollen, message=FALSE, warning=FALSE}
 pollen_data <- get_downloads(pollen_locations$neotoma_id, all_data=T)
-```
-
-Then we load a few help scripts which can be found under "code/neotoma_helpers.R".
-
-```{r}
-source(here("code", "neotoma_helpers.R"))
-```
-
-Now we filter the data sets so that we only retain the pollen data that record "NISP".
-
-```{r}
 pollen_data.samples <- samples(pollen_data)
-
 pollen_data.samples.filtered <- pollen_data.samples %>% subset(element == "pollen" & units == "NISP")
-```
-
-Then we convert the NISP into percentages:
-
-```{r}
 pollen_data_comp <- transform_to_percent(pollen_data.samples.filtered)
@@
 
-### ^14^C Summenkalibration
-
-The data set for the 14C sum calibration is composed of the above-mentioned sources. This data consists of 1135 dates from 246 sites.
-
-```{r}
+### Summed radiocarbon
+
+The data set for the ^14^C sum calibration is composed of the above-mentioned sources. This data consists of 1135 dates from 246 sites.
+
+```{r read-c14}
 c14 <- read.csv(here("data","raw_data", "14c_swiss_plateau.csv"))
@@
 
-```{r}
+```{r plot-c14}
 c14.sf <- c14 |>
@@
 
-```{r}
+```{r calibrate-c14}
 c14.calibrated <- calibrate(x = c14$bp, errors = c14$std, verbose = F)
@@
 
-```{r}
+```{r sum-c14}
 sumcal <- spd(c14.calibrated, timeRange = 1950-timeframe, bins = c14$site, runm = sw, verbose = F)
@@
 
-```{r}
+```{r plot-c14-sum}
 plot(1950-sumcal$grid$calBP, sumcal$grid$PrDens,type="l")
@@
 
-```{r}
+```{r resample-c14-sum}
 res.sumcal <- sumcal$grid[sumcal$grid$calBP %in%
@@
 
-### Aoristic Site Count
+### All settlements
 
@@
 
-```{r}
+```{r read-sites}
 sites <- read.csv(here("data","raw_data", "site_data_fuzzy.csv"))
@@
 
-```{r}
+```{r plot-sites}
 sites.sf <- st_as_sf(sites,
@@
 
-### Dendro Dated Settlements
+### Lakeshore settlements
 
@@
 
-```{r}
+```{r read-lss}
 lss <- read.csv(here("data","raw_data", "dendro_years.csv"))
@@
 
-```{r}
+```{r plot-lss}
 plot(res.lss$date, res.lss$moving_mean,type="l")
@@
 
-```{r}
+```{r resample-lss}
 res.lss <- res.lss[res.lss$date %in% seq(from=timeframe[1],
@@
 
-### Combining the proxies
+## Combined proxies
 
@@
 
-```{r}
+```{r plot-proxies}
 ggplot() +
@@
 
-```{r}
+```{r combine-proxies}
 all_proxies <- merge(res.sumcal, res.open_final, by.x = "calBP", by.y = "age") %>%
@@
 
-```{r}
+```{r cache-proxies}
 write.csv(all_proxies, here("data","preprocessed_data", "all_proxies.csv"))
@@
 
-## Transforming the Data
+### Transforming the Data
 
@@
 
-```{r}
+```{r read-cached-proxies}
 all_proxies <- read.csv(here("data","preprocessed_data", "all_proxies.csv"),
@@
 
-## The Model
+## Model definition
 
@@
 
-```{r}
+```{r define-model}
 model_code <- nimbleCode( {
@@
 
-```{r}
+```{r define-model-constants}
 model_constants <- list(
@@
 
-```{r}
-
+```{r define-model-inits}
 # Calculation of an exponential increase of the factor 10
@@
 
-```{r}
+```{r define-run-params}
 batches <- 100000 # run in batches of this length
@@
 
-```{r}
+```{r define-run-monitors}
 model_monitors <- c("PopDens", "beta")
@@
 
-```{r}
+```{r setup-parallel}
 
@@
 
-```{r}
+```{r run-model-inital}
 start_time <- Sys.time()
@@
 
-```{r}
+```{r run-model-convergence}
 # Initialize with non-convergence
@@
 
-```{r}
+```{r run-model-final}
 do_final_run = F

### 2023-09-27 manuscript/paper_revision_1.Rmd
slight changes in the course of a course

@@
       email: jan.kolar@ibot.cas.cz
-      institute: [IBOT, IAMFAMU]
+      institute: [IBOT, IAMFAMU, IOA]
       correspondence: false
@@
   - IAMFAMU: Institute of Archaeology and Museology, Faculty of Arts, Masaryk University
+  - IOA: Institute of Archaeology, University College London
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 # Code availability
-The computer code used to generate the Bayesian Population model is provided in full in the Supplementary Information, together with information about the program and version used.
+The code used to generate the Bayesian Population model is provided in full in the Supplementary Information, together with information about the program and version used.
 The R code and Data are available online at https://github.com/MartinHinz/bayesian.demographic.reconstruction.2022 and is archived at https://doi.org/10.5281/zenodo.6594498.
